### 2017-07-27 ModelTesting/ModelTestingSummary.Rmd
Added in the files for model testing

---
title: "Model Testing Summary"
author: "Topher Weiss-Lehman"
date: ""
output: pdf_document
---

\begin{center}
    \begin{tabular}{ | p{3.5cm} | p{8.9cm} | p{3.5cm} |}
    \hline
    \bf{Functionality tested} & \bf{Test details} & \bf{Test outcome} \\
    \hline
    SaveParams() & SaveParams.R tests that all input parameters are delisted and saved in an executable file that can be later read into R & `r readLines("SaveParams.txt")` \\
    \hline
    ChangeClimate() & ChangeClimate.R tests both the functionality and ability to return useful error messages of the ChangeClimate function & `r readLines("ChangeClimate.txt")` \\
    \hline
    PopMatColNames() & PopMatColNames.R checks the example functionality as well as the proper output for different numbers of fitness and dispersal alleles & `r readLines("PopMatColNames.txt")` \\
    \hline
    Initialize() & Initialize.R checks for correct dimensionality of generated population matrices and tests that the values generated for the matrix conform to given parameters such as specified distributions and sex ratios & `r readLines("Initialize.txt")` \\
    \hline
    GetEnvQual() \& CalcEnvMean() & EnvCalcFxns.R tests both of these functions by checking for symmetry and equivalence across spatial scales & `r readLines("EnvCalcFxns.txt")` \\
    \hline
    CalcTraits() & CalcTraits.R tests the dimenstionality of the output and the distributional accuracy of the function & `r readLines("CalcTraits.txt")` \\
    \hline
    Disperse() & Disperse.R tests the distinction among the three possible dispersal kernels as well as the effect of increasing landscape width on dispersal patterns & `r readLines("Disperse.txt")` \\
    \hline
    RelFit() & RelFits.R tests the range of values returned by RelFit() and the effect of increasing or decreasing the degree of local adaptation and the amount of trait variance present on the calculated fitnesses & `r readLines("RelFits.txt")` \\
    \hline
    Inheritence() & Inheritence.R tests for correct segregation of independent alleles as well as correct behavior in the mutation process & `r readLines("Inheritence.txt")` \\
    \hline
    MatFill() & MatFill.R tests for the correct dimensionality of the returned matrices and the correct elements in them based on input parents, locations, etc. & `r readLines("MatFill.txt")` \\
    \hline
    Reproduce() & Reproduce.R checks for helpful error messages, cases for which there should be zero growth on average, and detioriating growth with increased distance from an optimum due to local adaptation & `r readLines("Reproduce.txt")` \\
    \hline
    No dispersal & NoDisperse.R uses the full model to test the case of zero dispersal and ensure that populations do not leave their staring patch & `r readLines("NoDisperse.txt")` \\
    \hline
    Stationary range & StationaryRange.R uses the full model to check that populations with the ability to grow and disperse fill up a stationary range boundary, but do not move beyond it & `r readLines("StationaryRange.txt")` \\
    \hline
    Moving environment & MovingEnv.R uses the full model to check that populations with the ability to grow, but not disperse, react to a moving environment by eventually going extinct & `r readLines("MovingEnv.txt")` \\
    \hline
    \end{tabular}
\end{center}

### 2017-07-27 ModelTesting/CalcTraits.R

@@ -1,0 +1,60 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the CalcTraits function, (1) use the Initialize function to create
+#    an input matrix and ensure CalcTraits returns ouput of the proper
+#    dimensionality, (2) Plot histograms of all the trait values for all
+#    individuals. Since they are summing across normally distributed random
+#    variables, they should conform to a normal distribution with a mean of
+#    2*NumLoci*TraitMean and a standard deviation of sqrt(2*NumLoci*TraitDiv^2).
+#    For the dispersal trait, first take the log of the trat values, then
+#    compare to this distribution since the dispersal sum is exponentiated.
+
+NumFitLoci <- 12
+NumDispLoci <- 7
+TrialPopSize <- 1000
+FitMean <- 2
+FitDiv <- 3
+DispMean <- 0.2
+DispDiv <- 6
+
+ColNames <- PopMatColNames(nFit = NumFitLoci, nDisp = NumDispLoci, monoecious = TRUE)
+FitCols <- grep("^fit", ColNames)
+DispCols <- grep("^disp", ColNames)
+TrialPop <- Initialize(ColumnNames = ColNames, FitCols = FitCols, DispCols = DispCols, 
+                       PopSize = TrialPopSize, BetaInit = 0, width = 3, 
+                       SexRatio = 0.5, FitInit = FitMean, FitDiv = FitDiv, 
+                       DispInit = DispMean, DispDiv = DispDiv)
+
+TrialTraits <- CalcTraits(population = 1:nrow(TrialPop), PopMat = TrialPop, 
+                          FitColumns = FitCols, DispColumns = DispCols)
+
+if(nrow(TrialTraits) == TrialPopSize & ncol(TrialTraits) == 3){
+     Check1 <- TRUE
+} else{
+     Check1 <- FALSE
+}
+
+
+hist(TrialTraits[,"fit"], freq = FALSE)
+xseq <- seq(0, 100, length.out = 1000)
+lines(x = xseq, dnorm(x = xseq, mean = 2*NumFitLoci*FitMean, sd = sqrt(2*NumFitLoci*FitDiv^2)))
+
+hist(log(TrialTraits[,"disp"]), freq = FALSE)
[…]

### 2017-07-27 ModelTesting/ChangeClimate.R

@@ -1,0 +1,47 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the ChangeClimate function, first choose simple values of climate 
+#    speed and initial conditions so that the answer is obvious from manual
+#    calculations. Then supply a negative value of speed and ensure a helpful
+#    error message is returned, followed by a length of time of 0 to ensure
+#    that an empty vector is returned.
+
+# First try a straitforward climate change
+Beta1 <- 0
+ClimateSpeed <- 1
+ChangeDuration <- 10
+# These values should produce a vector of length 10 with values 1, 2, 3, etc.
+
+TestRun <- ChangeClimate(BetaInit = Beta1, LengthShift = ChangeDuration, 
+                         ClimSpeed = ClimateSpeed)
+
+if(sum(TestRun == 1:10) == 10){
+     FirstCheck <- TRUE
+} else{
+     FirstCheck <- FALSE
+}
+
+# Next try a negative speed
+TestNegative <- ChangeClimate(BetaInit = Beta1, LengthShift = ChangeDuration,
+                              ClimSpeed = -1)
+if(is.null(TestNegative)){
+     SecondCheck <- TRUE
+} else{
+     SecondCheck <- FALSE
+}
+
+# Finally try a shift length of 0
+Test0 <- ChangeClimate(BetaInit = Beta1, LengthShift = 0, ClimSpeed = ClimateSpeed)
+if(is.null(Test0)){
+     ThirdCheck <- TRUE
+} else{
+     ThirdCheck <- FALSE
+}
+
+if(FirstCheck & SecondCheck & ThirdCheck){
+     writeLines("Pass", con = "ModelTesting/ChangeClimate.txt")
+} else{
+     writeLines("Fail", con = "ModelTesting/ChangeClimate.txt")
+}

### 2017-07-27 ModelTesting/Disperse.R

@@ -1,0 +1,152 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the dispersal function, first use the previously vetted PopMatColNames,
+#    Initialize, and CalcTraits functions to generate the necessary input. Then,
+#    (1) set the landscape width to 5 with a large population size and use the
+#    dispersal function to generate post dispersal locations according to
+#    all three dispersal kernels. Examine them with a heat map and confirm
+#    that the kernels have some of the expected differences in patterns.
+#    (2) Repeat the above process with larger values of the dispersal trait in
+#    the Intialize function and confirm that it leads to greater dispersal in
+#    all three kernels.
+#    (3) For a single kernel, keep all other parameters the same, but set the
+#    width to 5, 10, and 20 with population sizes that will result to equivalent
+#    average abundances across the width of the landscape (e.g. 1,000, 10,000, and
+#    20,000).
+
+library(plot3D)
+
+AllCols <- PopMatColNames(nFit = 4, nDisp = 4, monoecious = TRUE)
+FitCols <- grep("^fit", AllCols)
+DispCols <- grep("^disp", AllCols)
+PreDispMat <- Initialize(ColumnNames = AllCols, FitCols = FitCols, DispCols = DispCols,
+                         PopSize = 1000, BetaInit = 0, width = 5, FitInit = 0,
+                         FitDiv = 1, DispInit = 0, DispDiv = 0.2)
+PreDispMat[,"x0"] <- PreDispMat[,"x1"]
+PreDispMat[,"y0"] <- PreDispMat[,"y1"]
+
+TraitMat <- CalcTraits(population = 1:1000, PopMat = PreDispMat, FitColumns = FitCols,
+                       DispColumns = DispCols)
+
+NormDisp <- Disperse(PopMat = PreDispMat, traits = TraitMat, width = 5, kern = "norm",
+                     PatchScale = 1)
[…]

### 2017-07-27 ModelTesting/Inheritence.R

@@ -1,0 +1,36 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the Inheritence function, I will use small, manually created input
+#    population matrices to test specific cases. (1) Test cases with low and
+#    high mutation rates and variance to ensure that the function behaves as
+#    expected. (2) Set mutation to 0 and make a small input matrix with all
+#    unique allele values to ensure that segregation is working correctly.
+
+AllCols <- PopMatColNames(nFit = 1, nDisp = 4, monoecious = TRUE)
+FitCols <- grep("^fit", AllCols)
+DispCols <- grep("^disp", AllCols)
+PopMat <- matrix(NA, nrow = 6, ncol = length(AllCols))
+colnames(PopMat) <- AllCols
+PopMat[,5:6] <- 1:12
+PopMat[,7:14] <- 1:48
+
+parents <- matrix(1:6, nrow = 3, ncol = 2)
+
+Inheritence(Cols = FitCols, parents = parents, PopMat = PopMat, U = c(0.5, 0.5),
+            Vm = c(1, 1))
+Inheritence(Cols = FitCols, parents = parents, PopMat = PopMat, U = c(0.5, 0.5),
+            Vm = c(0, 0))
+Inheritence(Cols = FitCols, parents = parents, PopMat = PopMat, U = c(0, 0),
+            Vm = c(1, 1))
+
+
+Inheritence(Cols = DispCols, parents = parents, PopMat = PopMat, U = c(0.5, 0.5),
+            Vm = c(1, 1))
+Inheritence(Cols = DispCols, parents = parents, PopMat = PopMat, U = c(0.5, 0.5),
+            Vm = c(0, 0))
+Inheritence(Cols = DispCols, parents = parents, PopMat = PopMat, U = c(0, 0),
+            Vm = c(1, 1))
+
+writeLines("Pass", con = "ModelTesting/Inheritence.txt")

### 2017-07-27 ModelTesting/PopMatColNames.R

@@ -1,0 +1,32 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the PopMatColNames function, first ensure that the example functionality
+#    works regardless of the other arguments. Then, check that it returns the
+#    expected output for different combinations of number of alleles and
+#    life history syndromes (monoecious vs. dioecious).
+
+Test1 <- PopMatColNames(example = TRUE)
+Test2 <- PopMatColNames(nFit = 10, nDisp, monoecious = FALSE, example = TRUE)
+if(sum(Test1 == Test2) == length(Test1)){
+     Check1 <- TRUE
+} else{
+     Check1 <- FALSE
+}
+
+Test3 <- PopMatColNames(nFit = 3, nDisp = 4, monoecious = TRUE)       # Should be 18
+Test4 <- PopMatColNames(nFit = 1, nDisp = 1, monoecious = FALSE)      # Should be 9
+Test5 <- PopMatColNames(nFit = 20, nDisp = 100, monoecious = TRUE)    # Should be 244
+
+if( (length(Test3) == 18) & (length(Test4) == 9) & (length(Test5) == 244) ){
+     Check2 <- TRUE
+} else{
+     Check2 <- FALSE
+}     
+     
+if( Check1 & Check2 ){     
+     writeLines("Pass", con = "ModelTesting/PopMatColNames.txt")
+} else{
+     writeLines("Fail", con = "ModelTesting/PopMatColNames.txt")
+}

### 2017-07-27 ModelTesting/Reproduce.R

@@ -1,0 +1,161 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# Check the negative R error message
+# Check appropriate lack of growth for very large population sizes
+# Use width of 1 to allow for more precise control of Nt
+# Check that setting R to 0 results in equal population size (on average)
+# Check that setting K to 0 results in no growth (might have to add in a check to the main function for this)
+# Check that moving the spatial location to more and more maladapted localities results in reduced growth on average
+
+AllCols <- PopMatColNames(nFit = 1, nDisp = 1, monoecious = TRUE)
+FitCols <- grep("^fit", AllCols)
+DispCols <- grep("^disp", AllCols)
+PopMat <- Initialize(ColumnNames = AllCols, FitCols = FitCols, DispCols = DispCols,
+                     PopSize = 50, BetaInit = 0, width = 1, FitInit = 0.25,
+                     FitDiv = 0, DispInit = 1, DispDiv = 1)
+TraitMat <- CalcTraits(population = 1:50, PopMat = PopMat, FitColumns = FitCols,
+                       DispColumns = DispCols)
+
+# Set necessary parameters
+LocalSel <- 2
+tau <- 100
+beta <- 0
+omega <- 0.25
+gamma <- 1
+R0 <- 2
+K0 <- 100
+U <- c(0.25, 0.25)
+Vm <- c(0.1, 0.1)
+names(U) <- c("fit", "disp")
+names(Vm) <- c("fit", "disp")
+PatchScale <- 1
+
+# Test that a useful error message is provided
+Reproduce(beta = beta, gamma = gamma, tau = tau, omega = omega, R0 = R0, K0 = K0,
+          U = U, Vm = Vm, LocalSel = LocalSel, traits = TraitMat, PopMat = PopMat,
+          EnvGradType = "k", ColumnNames = AllCols, FitColumns = FitCols, 
+          DispColumns = DispCols, PatchScale = PatchScale)
+
+Check1 <- TRUE
+
+# Check the negative R error message
+Reproduce(beta = beta, gamma = gamma, tau = tau, omega = omega, R0 = -2, K0 = K0,
+          U = U, Vm = Vm, LocalSel = LocalSel, traits = TraitMat, PopMat = PopMat,
[…]

### 2017-07-27 ModelTesting/SaveParams.R

@@ -1,0 +1,53 @@
+# Set the working directory and load in the simulation functions
+setwd("~/Desktop/RangeShifts/ShiftingSlopes/")
+source("SimulationFunctions.R")
+
+# To test the SaveParams() function of the model, create a list of parameter
+#    values to save, run the function, and check that it creates an executable
+#    file that will load in values identical to those created initially.
+
+OriginalParams <- list(beta = 1, gamma = 2, tau = 3, LocalSel = 4, omega = 5,
+                       U = c(6,7), Vm = c(8,9), nFit = 10, nDisp = 11, R0 = 12,
+                       K0 = 13, width = 14, kern = 15, EnvGradType = 16,
+                       monoecious = 17, BurnIn = 18, BurnOut = 19, LengthShift = 20,
+                       ClimSpeed = 21, InitPopSize = 22, BetaInit = 23, FitInit = 24,
+                       FitDiv = 25, DispInit = 26, DispDiv = 27)
+
+SaveParams(parameters = OriginalParams, FilePath = "ModelTesting")
+source("ModelTesting/parameters.R")
+
+if( (OriginalParams$beta == beta) &
+    (OriginalParams$gamma == gamma) &
+    (OriginalParams$tau == tau) &
+    (OriginalParams$LocalSel == LocalSel) &
+    (OriginalParams$omega == omega) &
+    (OriginalParams$U[1] == U[1]) &
+    (OriginalParams$U[2] == U[2]) &
+    (OriginalParams$Vm[1] == Vm[1]) &
+    (OriginalParams$Vm[2] == Vm[2]) &
+    (OriginalParams$nFit == nFit) &
+    (OriginalParams$nDisp == nDisp) &
+    (OriginalParams$R0 == R0) &
+    (OriginalParams$K0 == K0) &
+    (OriginalParams$width == width) &
+    (OriginalParams$kern == kern) &
+    (OriginalParams$EnvGradType == EnvGradType) &
+    (OriginalParams$monoecious == monoecious) &
+    (OriginalParams$BurnIn == BurnIn) &
+    (OriginalParams$BurnOut == BurnOut) &
+    (OriginalParams$LengthShift == LengthShift) &
+    (OriginalParams$ClimSpeed == ClimSpeed) &
+    (OriginalParams$InitPopSize == InitPopSize) &
+    (OriginalParams$BetaInit == BetaInit) &
+    (OriginalParams$FitInit == FitInit) &
[…]

### 2017-09-06 ModelTesting/ModelTestingSummary.Rmd
Did some more model testing

@@
     \hline
-    No dispersal & NoDisperse.R uses the full model to test the case of zero dispersal and ensure that populations do not leave their staring patch & `r readLines("NoDisperse.txt")` \\
+    No dispersal & NoDisperse.R uses the full model to test the case of zero dispersal and ensure that populations do not leave their staring patch & `r readLines("NoDisperse/NoDisperse.txt")` \\
     \hline
-    Stationary range & StationaryRange.R uses the full model to check that populations with the ability to grow and disperse fill up a stationary range boundary, but do not move beyond it & `r readLines("StationaryRange.txt")` \\
+    Stationary range & StationaryRange.R uses the full model to check that populations with the ability to grow and disperse fill up a stationary range boundary, but do not move beyond it & `r readLines("StationaryRange/StationaryRange.txt")` \\
     \hline
-    Moving environment & MovingEnv.R uses the full model to check that populations with the ability to grow, but not disperse, react to a moving environment by eventually going extinct & `r readLines("MovingEnv.txt")` \\
+    Moving environment & MovingEnv.R uses the full model to check that populations with the ability to grow, but not disperse, react to a moving environment by eventually going extinct & `r readLines("MovingEnv/MovingEnv.txt")` \\
     \hline

### 2017-09-06 ModelTesting/NoDisperse/NoDisperse.R

@@ -2,7 +2,7 @@
 #    absence of dispersal. 
 
 setwd("~/Desktop/RangeShifts/ShiftingSlopes/ModelTesting/NoDisperse/")
-source("~/Desktop/RangeShifts/ShiftingSlopes/SimulationFunctions.R")
+source("~/Desktop/RangeShifts/ShiftingSlopes/SimFunctions.R")
 
 # First set all parameters
 BetaInit <- 0
@@ -21,7 +21,7 @@ width <- 10
 kern <- "exp"
 EnvGradType <- "K"
 monoecious <- TRUE
-BurnIn <- 5
+BurnIn <- 100
 BurnOut <- 0
 LengthShift <- 0
 ClimSpeed <- 0
@@ -31,29 +31,50 @@ FitDiv <- 0.025
 DispInit <- -100
 DispDiv <- 0
 PatchScale <- 10
+NumRands <- 1000000
+SexRatio <- 0.5
+
 
 parameters <- list(BetaInit, gamma, tau, LocalSel, omega, U, Vm, nFit, nDisp, R0,
                    K0, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
                    LengthShift, ClimSpeed, InitPopSize, FitInit, FitDiv, DispInit,
-                   DispDiv, PatchScale)
+                   DispDiv, PatchScale, NumRands, SexRatio)
 names(parameters) <- c("BetaInit", "gamma", "tau", "LocalSel", "omega", "U", "Vm", 
                        "nFit", "nDisp", "R0", "K0", "width", "kern", "EnvGradType", 
                        "monoecious", "BurnIn", "BurnOut", "LengthShift", "ClimSpeed", 
                        "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
-                       "PatchScale")
+                       "PatchScale", "NumRands", "SexRatio")
 
 FullSim(parameters = parameters, parallel = FALSE)
 
 
-# Graph the resulting abundance patterns
-quartz(width = 7, height = 5)
-par(mfrow = c(2, 5), mar = c(0, 0, 0, 0), oma = c(5, 4, 4, 2) + 0.1)
+# Graph the resulting abundance patterns through time as a composite graph
+TimeVec <- 1:BurnIn
 load("Sim1/SummaryStats.Rdata")
-OccPatches <- rep(NA, width)
-for(i in 1:width){
-     PopData <- PatchAbund[,i,]
[…]

### 2017-09-06 ModelTesting/NoDisperse/Sim3/parameters.R

@@ -1,26 +1,0 @@
-BetaInit <- 0
-gamma <- 0.025
-tau <- 276.25
-LocalSel <- 1
-omega <- 3
-U <- c(0.02,0)
-Vm <- c(4e-04,0)
-names(U) <- names(Vm) <- c("fit", "disp")
-nFit <- 5
-nDisp <- 5
-R0 <- 1
-K0 <- 10
-width <- 10
-kern <- "exp"
-EnvGradType <- "K"
-monoecious <- TRUE
-BurnIn <- 5
-BurnOut <- 0
-LengthShift <- 0
-ClimSpeed <- 0
-InitPopSize <- 50
-FitInit <- 0.05
-FitDiv <- 0.025
-DispInit <- -100
-DispDiv <- 0
-PatchScale <- 10

### 2018-04-17 Manuscript/Main.tex
Added the manuscript .tex file

\documentclass[11pt, oneside]{article}
\usepackage{geometry}                		
\geometry{letterpaper}                   		
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{hyperref}
\usepackage{fancybox}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}			
\usepackage{natbib}				
\usepackage{fullpage}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{color}

%\usepackage{figcaps}
%\usepackage[tablesfirst,nolists]{endfloat}
\usepackage{authblk}

\DeclareRobustCommand{\firstsecond}[2]{#1}

\newcommand{\mb}{\mathbf}
\newcommand{\bs}{\boldsymbol}
\newcommand{\wt}{\widetilde}
\newcommand{\s}{^{(s)}}

\title{Local adaptation and dispersal evolution interact to drive population response to climate change}

\date{}

\author[1]{Christopher Weiss-Lehman}
\author[1]{Allison Shaw}

\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}

\begin{document}
\maketitle

\doublespacing
\linenumbers

\section{Introduction}
Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions (CITATION). These range shifts are generally expected to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise (CITATION). In fact, contemporary range shifts have already been observed in a wide range of taxa (CITATION). Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate (CITATION), the creation of novel species assemblages as not all species shift their ranges at the same rate or at all (CITATION), or both. It is therefore crucial to understand the dynamics of such climate induced range shifts to better inform current and future conservation work.

While contemporary range shifts are a relatively new phenomenon, important insights can be gained from the related but distinct process of range expansion. Range expansions have been studied for decades, leading to a robust understanding of both the ecological (CITATIONS) and evolutionary (CITATIONS) mechanisms responsible for shaping such expansions. For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability (CITATIONS), recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time (CITATIONS). As range shifts also involve a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions (CITATIONS). However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have important differences as well.

In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions. Range expansions typically begin from the successful establishment and subsequent spread of a small, founding population as typically occurs with the range expansions of invasive and reintroduced species (CITATIONS). Such founding populations represent samples from some larger source population and as such lack any initial spatial population structure. In contrast, range shifts involve entire populations with existing spatial structure within the initially stable range (CITATION). Such spatial structure can take the form of local adaptation within the range, degree of transition in population size from range core to edge, or some combination of the two (CITATION).

These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change (CITATIONS). For example, the nature of the gradient forming the range edge (e.g. declines in birth rates vs. increases in extinction risk) can directly change the probability of extinction a species faces during a climate driven range shift (CITATION). A population's risk of extinction during a range shift has also been related to local adaptation within the range. Specifically, a broad environmental niche (i.e. little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes (CITATION). Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal evolution in an asexual species (CITATION). While this result was independent of the structure of the gradient forming the range edge (CITATION), this is unlikely to hold true for sexually reproducing species. As the type of gradient forming the range edge can influence the distribution of dispersal phenotypes within the range (CITATION), it is likely that this would also influence the dynamics of local adaptation in sexually reproducing species in which different locally adapted genotypes are joined via gene flow. In fact, pollen dispersal in flowering plants has been shown to dramatically reduce the likelihood of a range shift in response to climate change (CITATION). Evolution during range shifts due to local adaptation and dispersal evolution are likely important drivers of range shift dynamics (CITATION) and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.

%Citation order for above paragraph: Henry et al. 2013; Atkins et al. 2010; Hargreaves et al. 2015; Hargreaves et al. 2015; Henry et al. 2013; Aguilee et al. 2016; Van Petegem et al. 2016

Here, we use a complex, individual-based model incorporating evolution in both fitness and dispersal distance to determine the interaction between local adaptation and the starkness of the gradient at the range edge in driving population dynamics during climate change. Using this model, we simultaneously vary the strength of local adaptation, the starkness of the range boundary, and the speed of climate change to ascertain the relationship among them and how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate. We additionally compare the dynamics of successful versus doomed populations to understand the factors contributing to population extinction and identify warning signs for populations in danger of extinction as a result of a range shift.

\section{Methods}
\subsection{Overview}
\subsubsection{Purpose}
This model tests an evolving population's ability to track a changing climate under a variety of conditions. Specifically, populations are simulated with (1) stark or gradual range boundaries, (2) strong or weak local selection pressures, and (3) varying speeds of climate change. In all simulations, individual dispersal and relative fitness are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time. All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change. Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat for the population (see the \textit{Submodels} section below). Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.

\subsubsection{State variables and scales}
The model simulates monoecious (i.e. no defined sex) individuals characterized by spatial coordinates for their location and diploid loci for both fitness and dispersal. Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis. To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same row of the landscape. Patches are defined by the location of the patch center in x and y coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (see the \textit{Submodels} section below).

The abiotic environment is defined by a parameter denoting the center of the range, the slope of the range boundary, and the width of the range. The range can then shift in space by altering the parameter defining the range center, which is how climate change is implemented in the model. Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries. The shape and severity of this gradient is defined by the same parameter used to set the slope of the range boundary along with one other parameter describing the strength of local selection along the gradient (greater values of this parameter will lead to steeper gradients in phenotypic optima within the range). The attached figures demonstrate the shape of the range and phenotypic optima functions and how they vary with different parameter values (see equations in \textit{Submodels}).

\subsubsection{Process overview and scheduling}
Time is also modeled in discrete intervals defining single generations of the population. Within each generation, individuals first reproduce within their natal patch. After reproduction, all adults die and the offspring disperse according to their dispersal kernels (\textit{Submodels}), resulting in discrete, non-overlapping generations. Reproduction occurs according to a stochastic implementation of the logistic growth equation taking into account the mean fitness of individuals within the patch (\textit{Submodels}). For each offspring, two parents are randomly sampled (with replacement and weighted according to their fitness) from the local population. While all individuals are monoecious, selfing is allowed only when there is a single individual within a patch. Thus, high fitness individuals may produce many offspring while individuals with relatively low fitness may not produce any. Individuals inherit one allele from each parent at each loci (except in the case of selfing when individuals inherit both copies of a single parent's alleles). Loci are assumed to segregate independently and inherited loci are subject to mutation (\textit{Submodels}). Individual dispersal distance for each offspring is drawn from a dispersal kernel defined by their dispersal trait and the direction of dispersal is drawn from a uniform distribution. The next generation then begins with all offspring in their post dispersal locations and ready to reproduce.

\subsection{Design concepts}
\subsubsection{Emergence}
Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both fitness and dispersal traits during climate change. These are all examined in the context of their impact on the population's ability to keep pace with the changing climate.

\subsubsection{Stochasticity}
All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci. Environmental parameters are fixed, however, and the process of climate change (i.e. the movement of environmentally suitable habitat through time) is deterministic. Thus, the model accounts for demographic stochasticity but ignores environmental stochasticity.

\subsubsection{Interactions}
Individuals in the model interact via mating and density-dependent competition within patches. Other important interactions are the relationship between dispersal evolution and local adaption, particularly in edge populations, and how this relationship impacts a population's ability to avoid extinction and track a changing climate. These interactions are examined in the context of different configurations of the initial, stable range and different speeds of climate change.

\subsubsection{Desired output}
After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits). If a population went extinct during the model run, the time of extinction is recorded. Throughout the simulation, certain aggregated values are calculated and recorded for each occupied patch including the population size and the mean and variance of allele values for each trait.

\subsection{Details}
\subsubsection{Initialization}
The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below. Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions. This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation.

\subsubsection{Submodels}
\paragraph{Environmentally suitable habitat}
Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$). The maximum achievable carrying capacity ($K_{max}$) occurs in the center of the species' range and declines with increasing distance from the center. Specifically, the carrying capacity at a location $x$ is defined as a combination of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows 
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
\end{cases}
\end{equation}
where $beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure 1). To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau}} & x > \beta_{t}
\end{cases}	
\end{equation}
yielding a derivative of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).

Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure 1). The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
\begin{equation}
K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
\end{equation}

By varying the parameters defining $f(x,t)$, we can change both the total achievable carrying capacity of the population throughout the range (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$). We are primarily interested in the effect of altering the slope at which $K_{x}$ declines at the range boundaries, however as doing so can also alter the total achievable carrying capacity of the population, we control for this confounding factor by fixing the total area under the curve $f(x,t)$. The indefinite integral of $f(x,t)$ can be shown to be
\begin{equation}
\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1}{\gamma}
\end{equation}
which can be solved for $\tau$. Thus, if the total area under the curve is fixed, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.

Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change. During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
\begin{equation}
\beta_{t}=\nu\eta(t-\hat{t})
\end{equation}
where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $\eta$ is the spatial scale of each patch, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.

\paragraph{Local adaptation}
To allow an arbitrary degree of local adaptation within the range, the local phenotypic optima ($z_{opt,x}$) is set as follows
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
where $\lambda$ defines the strength of local selection with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range. Individual fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype (CITATION). All loci are assumed to contribute additively to the phenotype with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values for the fitness trait.

\paragraph{Population dynamics}
Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model (CITATIONS). To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$). The expected number of new offspring in patch $x$ at time $t+1$ is then given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

Parentage of the offspring is then assigned by random sampling of the local male and female population. The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any. Each offspring inherits one allele per locus from each parent, assuming no linkage among loci. After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.

\paragraph{Mutation}
Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from tlheir parents. The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$). Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation is
\begin{equation}
\frac{U^{T}}{2L^{T}}
\end{equation}
Mutational effects are drawn from a normal distribution with mean $0$ and a standard deviation of
\begin{equation}
\sqrt{V_{m}^{T}U^{T}}
\end{equation}
By defining the mutation process in this manner rather than setting a probability of mutation and mutational effect directly, similar mutational dynamics can be imposed regardless of the number of loci used in the simulation.

\paragraph{Dispersal}
Finally, individuals disperse according to an exponential dispersal kernel defined by each individual's dispersal phenotype. An individual's dispersal phenotype is the expected dispersal distance and is given by
\begin{equation}
d_{i} = \frac{D\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
\end{equation}
where $D$ is the maximum expected dispersal distance in terms of discrete patches, $\eta$ is the spatial scale of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $D$, and the summation is the sum of all alleles contributing to dispersal. Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis. The expected dispersal distance, $d_{i}$ is then used to draw a realized distance from an exponential dispersal kernel. The direction of dispersal is drawn from a uniform distribution bounded by $0$ and $2\pi$. If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ axis, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ axis, thus wrapping the top and bottom edges of the landscape to avoid edge effects. Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure 1).

\subsection{Simulation descriptions}
To explore the interacting role of local adaptation and the slope of the decline in population abundance at the range edge on trait evolution and the dynamics of range shifts, we simulated $100$ populations each over a range of parameter values varying both the degree of local adaptation and the decline in population abundance at the range edge. Additionally, for each parameter combination we simulated three different speeds of climate change. Under

%\bibliographystyle{amnatnat}
%\bibliography{main_bib}

\end{document}

### 2018-04-17 Manuscript/Main.tex
Updated the manuscript files with figures and citations. There is now a complete draft of the introduction and methods

@@
 \section{Introduction}
-Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions (CITATION).
-These range shifts are generally expected to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise (CITATION).
-In fact, contemporary range shifts have already been observed in a wide range of taxa (CITATION).
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate (CITATION), the creation of novel species assemblages as not all species shift their ranges at the same rate or at all (CITATION), or both.
+Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global, penuelas2003global, hansen2001global, scholze2006climate}.
+These range shifts are generally expected to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
+In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, walther2002ecological, parmesan2003globally, parmesan2006ecological, parmesan1999poleward, perry2005climate}.
+Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{davis2001range, parmesan2006ecological, zhu2012failure, sekercioglu2008climate}, the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel, gilman2010framework, williams2007novel}, or both.
 It is therefore crucial to understand the dynamics of such climate induced range shifts to better inform current and future conservation work.
@@
 While contemporary range shifts are a relatively new phenomenon, important insights can be gained from the related but distinct process of range expansion.
-Range expansions have been studied for decades, leading to a robust understanding of both the ecological (CITATIONS) and evolutionary (CITATIONS) mechanisms responsible for shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability (CITATIONS), recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time (CITATIONS).
-As range shifts also involve a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions (CITATIONS).
-However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have important differences as well.
+Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{skellam1951random, hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, burton2010trade, excoffier2009genetic, kubisch2014and} mechanisms responsible for shaping such expansions.
+For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, phillips2015evolutionary}.
+As range shifts also involve a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}.
+However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.
 
 In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions.
-Range expansions typically begin from the successful establishment and subsequent spread of a small, founding population as typically occurs with the range expansions of invasive and reintroduced species (CITATIONS).
+Range expansions typically begin from the successful establishment and subsequent spread of a small, founding population as typically occurs with the range expansions of invasive and reintroduced species~\citep{hastings2005spatial}.
 Such founding populations represent samples from some larger source population and as such lack any initial spatial population structure.
-In contrast, range shifts involve entire populations with existing spatial structure within the initially stable range (CITATION).
-Such spatial structure can take the form of local adaptation within the range, degree of transition in population size from range core to edge, or some combination of the two (CITATION).
+In contrast, range shifts involve entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}.
+Such spatial structure can take the form of local adaptation within the range, degree of transition in population size from range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
 
-These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change (CITATIONS).
+These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change.
 For example, the nature of the gradient forming the range edge (e.g.
 declines in birth rates vs.
-increases in extinction risk) can directly change the probability of extinction a species faces during a climate driven range shift (CITATION).
+increases in extinction risk) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
 A population's risk of extinction during a range shift has also been related to local adaptation within the range.
 Specifically, a broad environmental niche (i.e.
-little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes (CITATION).
-Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal evolution in an asexual species (CITATION).
-While this result was independent of the structure of the gradient forming the range edge (CITATION), this is unlikely to hold true for sexually reproducing species.
-As the type of gradient forming the range edge can influence the distribution of dispersal phenotypes within the range (CITATION), it is likely that this would also influence the dynamics of local adaptation in sexually reproducing species in which different locally adapted genotypes are joined via gene flow.
-In fact, pollen dispersal in flowering plants has been shown to dramatically reduce the likelihood of a range shift in response to climate change (CITATION).
-Evolution during range shifts due to local adaptation and dispersal evolution are likely important drivers of range shift dynamics (CITATION) and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
+little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes~\citep{atkins2010local}.
+Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal evolution in an asexual species~\citep{hargreaves2015fitness}, though it is unclear if this result will hold for sexually reproducing species in which locally adapted genotypes interact via gene flow.
+As the gradient forming the range edge can influence the distribution of dispersal phenotypes within the range~\citep{henry2013eco}, it is likely that this would also result in an altered interaction between dispersal evolution and local adaptation in sexually reproducing species.
+In fact, pollen dispersal in flowering plants has been shown to dramatically reduce the likelihood of a range shift in response to climate change~\citep{aguilee2016pollen}.
+Evolution during range shifts due to local adaptation and dispersal evolution are both potentially important drivers of range shift dynamics~\citep{van2016spatial} and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
 
-%Citation order for above paragraph: Henry et al. 2013; Atkins et al. 2010; Hargreaves et al. 2015; Hargreaves et al. 2015; Henry et al. 2013; Aguilee et al. 2016; Van Petegem et al. 2016
-
-Here, we use a complex, individual-based model incorporating evolution in both fitness and dispersal distance to determine the interaction between local adaptation and the starkness of the gradient at the range edge in driving population dynamics during climate change.
-Using this model, we simultaneously vary the strength of local adaptation, the starkness of the range boundary, and the speed of climate change to ascertain the relationship among them and how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
-We additionally compare the dynamics of successful versus doomed populations to understand the factors contributing to population extinction and identify warning signs for populations in danger of extinction as a result of a range shift.
+Here, we use a complex, individual-based model incorporating evolution in both fitness and dispersal distance to determine the interaction between local adaptation and the starkness of the gradient at the range edge in driving population dynamics during climate change for a sexually reproducing species.
+Using this model, we simultaneously vary the strength of local adaptation, the starkness of the range boundary, and the speed of climate change to ascertain the relationship among them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
+We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction and identify warning signs for populations in danger of extinction as a result of a range shift.
 
@@
 All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change.
-Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat for the population (see the \textit{Submodels} section below).
+Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see the \textit{Submodels} section below).
 Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.
@@
 \subsubsection{State variables and scales}
-The model simulates monoecious (i.e.
-no defined sex) individuals characterized by spatial coordinates for their location and diploid loci for both fitness and dispersal.
+The model simulates a population of male and female individuals characterized by spatial coordinates for their location and diploid loci for both fitness and dispersal.
 Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis.
@@
 
-The abiotic environment is defined by a parameter denoting the center of the range, the slope of the range boundary, and the width of the range.
+The abiotic environment is defined by a parameter denoting the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}).
 The range can then shift in space by altering the parameter defining the range center, which is how climate change is implemented in the model.
 Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries.
-The shape and severity of this gradient is defined by the same parameter used to set the slope of the range boundary along with one other parameter describing the strength of local selection along the gradient (greater values of this parameter will lead to steeper gradients in phenotypic optima within the range).
-The attached figures demonstrate the shape of the range and phenotypic optima functions and how they vary with different parameter values (see equations in \textit{Submodels}).
+The severity of this gradient is defined by a separate parameter independent of the other environmental parameters (greater values of this parameter will lead to steeper gradients in phenotypic optima within the range).
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
+\vspace{-5mm}
+\caption[LoF entry]{Environmental quality, as defined by $f(x,t)$, over a spatial gradient.
+The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
+Additionally, the lattice of discrete patches in which population dynamics occur is shown beneath.
+This lattice of $\eta$ x $\eta$ patches are mapped to environmental conditions along the $x$ axis using $f(x,t)$ and are environmentally constant along the $y$ axis.
+Dispersal is unbounded in the $x$ direction and implemented with wrapping boundaries in the $y$ direction.}
+\label{fig:EnvFunction}
+\end{figure}
 
@@
 Within each generation, individuals first reproduce within their natal patch.
-After reproduction, all adults die and the offspring disperse according to their dispersal kernels (\textit{Submodels}), resulting in discrete, non-overlapping generations.
-Reproduction occurs according to a stochastic implementation of the logistic growth equation taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
-For each offspring, two parents are randomly sampled (with replacement and weighted according to their fitness) from the local population.
-While all individuals are monoecious, selfing is allowed only when there is a single individual within a patch.
-Thus, high fitness individuals may produce many offspring while individuals with relatively low fitness may not produce any.
-Individuals inherit one allele from each parent at each loci (except in the case of selfing when individuals inherit both copies of a single parent's alleles).
-Loci are assumed to segregate independently and inherited loci are subject to mutation (\textit{Submodels}).
-Individual dispersal distance for each offspring is drawn from a dispersal kernel defined by their dispersal trait and the direction of dispersal is drawn from a uniform distribution.
-The next generation then begins with all offspring in their post dispersal locations and ready to reproduce.
+After reproduction, all adults die and the offspring disperse according to their dispersal kernels (\textit{Submodels}), resulting in discrete, non-overlapping generations (See Figure~\ref{fig:LifeCycle}).
+Reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
+Parental pairs form via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with high fitness are likely to produce multiple offspring while individuals with no fitness may not produce any.
+Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process (\textit{Submodels}).
+After reproduction, all individuals in the parental generation perish, forming discrete non-overlapping generations.
+The offspring then disperse according to their dispersal traits to start the next generation in their post dispersal locations.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
+\vspace{-5mm}
+\caption[LoF entry]{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
+Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice.
+After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
+\label{fig:LifeCycle}
+\end{figure}
 
@@
 the movement of environmentally suitable habitat through time) is deterministic.
-Thus, the model accounts for demographic stochasticity but ignores environmental stochasticity.
+Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.
 
@@
 This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation.
+See Table 1 for a full list of all parameter values used in the simulations described here.
 
@@
 \end{equation}
-where $beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure 1).
+where $beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}).
 To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
@@
 
-Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure 1).
+Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure~\ref{fig:EnvFunction}).
 The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
@@
 \begin{equation}
-\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1}{\gamma}
+\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
 \end{equation}
@@
 \end{equation}
-where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype (CITATION).
+where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype~\citep{lande1976natural}.
 All loci are assumed to contribute additively to the phenotype with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values for the fitness trait.
@@
 \paragraph{Population dynamics}
-Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model (CITATIONS).
+Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
 To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$).
@@
 \subsection{Simulation descriptions}
-To explore the interacting role of local adaptation and the slope of the decline in population abundance at the range edge on trait evolution and the dynamics of range shifts, we simulated $100$ populations each over a range of parameter values varying both the degree of local adaptation and the decline in population abundance at the range edge.
+To explore the interacting role of local adaptation and the slope of the decline in population abundance at the range edge on trait evolution and the dynamics of range shifts, we simulated $100$ populations each over a range of parameter values varying both the degree of local adaptation and the decline in population abundance at the range edge (See Table 1).
 Additionally, for each parameter combination we simulated three different speeds of climate change.
-Under
+All simulations were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
 
-%\bibliographystyle{amnatnat}
-%\bibliography{main_bib}
+\bibliographystyle{amnatnat}
+\bibliography{main_bib}
 

### 2018-04-19 Manuscript/Main.tex
Updated the manuscript with parameter values and an additional citation in the introduction

@@
 Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal evolution in an asexual species~\citep{hargreaves2015fitness}, though it is unclear if this result will hold for sexually reproducing species in which locally adapted genotypes interact via gene flow.
-As the gradient forming the range edge can influence the distribution of dispersal phenotypes within the range~\citep{henry2013eco}, it is likely that this would also result in an altered interaction between dispersal evolution and local adaptation in sexually reproducing species.
+As the gradient forming the range edge can influence the distribution of dispersal phenotypes within the range~\citep{henry2013eco, hargreaves2014evolution}, it is likely that this would also result in an altered interaction between dispersal evolution and local adaptation in sexually reproducing species.
 In fact, pollen dispersal in flowering plants has been shown to dramatically reduce the likelihood of a range shift in response to climate change~\citep{aguilee2016pollen}.
@@
 \paragraph{Mutation}
-Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from tlheir parents.
+Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents.
 The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$).
@@
 If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ axis, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ axis, thus wrapping the top and bottom edges of the landscape to avoid edge effects.
-Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure 1).
+Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure~\ref{fig:EnvFunction}).
+
+\begin{table}
+\renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
+    \hline
+    Parameter & Description & Value \\ \hline \hline
+    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ \\
+    $\beta_{1}$ & Center of environmentally suitable habitat before climate change & $0$ \\
+    $\eta$ & Spatial dimensions of habitat patches in continuous space & $50$ \\
+    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ direction & $10$ \\
+    $\hat{t}$ & Last time point of stable climate conditions & $2000$ \\
+    $t_{\Delta}$ & Duration of climate change & $100$ \\
+    $t_{max}$ & Total number of time points in the simulation & $2150$ \\
+    $R$ & Intrinsic growth rate of the population & $2$ \\
+    $K_{max}$ & Maximum achievable carrying capacity in the range & $100$ \\ 
+    $\psi$ & Expected sex ratio in the population & $0.5$ \\
+    $D$ & Maximum achievable dispersal phenotype & $1000$ \\
+    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
+    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
+    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ for each trait \\
+    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ for each trait \\
+    $L^{T}$ & The number of diploid loci defining each trait & $5$ for each trait \\
+    $\mu_{1}^{f}$ & The initial mean allele value for the fitness trait & $0$ \\
+    $\mu_{1}^{d}$ & The initial mean allele value for the dispersal trait & $-1$ \\
+    $\sigma_{1}^{f}$ & The initial standard deviation of allele values for fitness & $0.025$ \\
+    $\sigma_{1}^{d}$ & The initial standard deviation of allele values for dispersal & $1$ \\
+    \hline
+  \end{tabular}
+\caption[LoF entry]{Simulation parameters held constant across all scenarios presented here.}
+\label{table:ConstPars}
+\end{table}
+
+\begin{table}
+\renewcommand{\arraystretch}{1.5}
+  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
+    \hline
+    Habitat gradient at range edge & Strength of local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
+     & None & $0.0025$ & $250$ & $0$ \\
+    Shallow & Moderate & $0.0025$ & $250$ & $0.004$ \\
+     & Strong & $0.0025$ & $250$ & $0.008$ \\ \hline
+     & None & $0.025$ & $421.479$ & $0$ \\
+    Moderate & Moderate & $0.025$ & $421.479$ & $0.004$ \\
+     & Strong & $0.025$ & $421.479$ & $0.008$ \\ \hline
+     & None & $0.25$ & $421.48$ & $0$ \\
+    Stark & Moderate & $0.25$ & $421.48$ & $0.004$ \\
+     & Strong & $0.25$ & $421.48$ & $0.008$ \\ 
+    \hline
+  \end{tabular}
+\caption[LoF entry]{Descriptions and parameter values for the $9$ different scenarios examined here. For each scenario, $\nu$ (the speed of climate change) was varied among $0.5$, $1$, and $2$.}
+\label{table:Scenarios}
+\end{table}
 
 \subsection{Simulation descriptions}
-To explore the interacting role of local adaptation and the slope of the decline in population abundance at the range edge on trait evolution and the dynamics of range shifts, we simulated $100$ populations each over a range of parameter values varying both the degree of local adaptation and the decline in population abundance at the range edge (See Table 1).
-Additionally, for each parameter combination we simulated three different speeds of climate change.
+Most of the parameters described above were held constant over all simulations we performed (Table ~\ref{table:ConstPars}).
+We varied certain parameter values to explore the interacting roles of local adaptation, the habitat gradient at the range edge, and the speed of climate change in the dynamics of range shifts.
+Specifically, we considered a factorial combination of three severities of habitat quality decline at the range edge, three degrees of local adaptation within the range, and three speeds of climate change (Table~\ref{table:Scenarios}) for a total of $27$ different model scenarios.
+For each scenario, we performed $100$ simulations to assess the interacting ecological and evolutionary dynamics of climate driven range shifts.
 All simulations were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.

### 2018-06-14 Manuscript/Main.tex
Updated the manuscript with Allison's comments

@@
 \author[1]{Christopher Weiss-Lehman}
-\author[1]{Allison Shaw}
+\author[1]{Allison K. Shaw}
 
@@
 Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global, penuelas2003global, hansen2001global, scholze2006climate}.
-These range shifts are generally expected to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
+These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
 In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, walther2002ecological, parmesan2003globally, parmesan2006ecological, parmesan1999poleward, perry2005climate}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{davis2001range, parmesan2006ecological, zhu2012failure, sekercioglu2008climate}, the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel, gilman2010framework, williams2007novel}, or both.
-It is therefore crucial to understand the dynamics of such climate induced range shifts to better inform current and future conservation work.
+Such range shifts present significant challenges to current and future conservation efforts as they can result in both the extinction of populations failing to track a changing climate~\citep{davis2001range, parmesan2006ecological, zhu2012failure, sekercioglu2008climate}, and the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel, gilman2010framework, williams2007novel}, or both.
+It is therefore crucial to understand the dynamics of such climate-induced range shifts to better inform current and future conservation work.
 
-While contemporary range shifts are a relatively new phenomenon, important insights can be gained from the related but distinct process of range expansion.
+While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion.
 Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{skellam1951random, hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, burton2010trade, excoffier2009genetic, kubisch2014and} mechanisms responsible for shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, phillips2015evolutionary}.
-As range shifts also involve a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}.
+For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, phillips2015evolutionary, shaw2015dispersal}.
+Since range shifts, like range expansions, have a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}.
 However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.
@@
 In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions.
-Range expansions typically begin from the successful establishment and subsequent spread of a small, founding population as typically occurs with the range expansions of invasive and reintroduced species~\citep{hastings2005spatial}.
-Such founding populations represent samples from some larger source population and as such lack any initial spatial population structure.
-In contrast, range shifts involve entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}.
-Such spatial structure can take the form of local adaptation within the range, degree of transition in population size from range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
+Range expansions, especially for invasive and reintroduced species, typically begin from the successful establishment and subsequent spread of a small, founding population~\citep{hastings2005spatial}.
+Such founding populations often represent samples from some larger source population and as such lack any initial spatial population structure.
+In contrast, range shifts start from entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}.
+Such spatial structure can take the form of local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
 
 These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change.
-For example, the nature of the gradient forming the range edge (e.g.
+For example, the mechanism responsible for the gradient in population size from core to edge (e.g.
 declines in birth rates vs.
@@
 little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes~\citep{atkins2010local}.
-Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal evolution in an asexual species~\citep{hargreaves2015fitness}, though it is unclear if this result will hold for sexually reproducing species in which locally adapted genotypes interact via gene flow.
-As the gradient forming the range edge can influence the distribution of dispersal phenotypes within the range~\citep{henry2013eco, hargreaves2014evolution}, it is likely that this would also result in an altered interaction between dispersal evolution and local adaptation in sexually reproducing species.
-In fact, pollen dispersal in flowering plants has been shown to dramatically reduce the likelihood of a range shift in response to climate change~\citep{aguilee2016pollen}.
+Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
+However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and fitness are directly linked via gene flow.
+For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
+In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
+Additionally, the distribution of dispersal phenotypes within a population can be influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts.
 Evolution during range shifts due to local adaptation and dispersal evolution are both potentially important drivers of range shift dynamics~\citep{van2016spatial} and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
 
-Here, we use a complex, individual-based model incorporating evolution in both fitness and dispersal distance to determine the interaction between local adaptation and the starkness of the gradient at the range edge in driving population dynamics during climate change for a sexually reproducing species.
-Using this model, we simultaneously vary the strength of local adaptation, the starkness of the range boundary, and the speed of climate change to ascertain the relationship among them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
-We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction and identify warning signs for populations in danger of extinction as a result of a range shift.
+Here, we develop a complex, individual-based model of a sexually reproducing species, which incorporates evolution in both fitness (via local adaptation) and dispersal distance to determine the interaction between the strength of local selection and the starkness of the gradient at the range edge in driving population dynamics during climate change.
+Using this model, we simultaneously vary the strength of local selection and the starkness of the range boundary to ascertain the relationship between them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
+We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction during climate-induced range shifts.
 
@@
 This model tests an evolving population's ability to track a changing climate under a variety of conditions.
-Specifically, populations are simulated with (1) stark or gradual range boundaries, (2) strong or weak local selection pressures, and (3) varying speeds of climate change.
+Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the strength of local selection.
 In all simulations, individual dispersal and relative fitness are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
@@
 \subsubsection{State variables and scales}
-The model simulates a population of male and female individuals characterized by spatial coordinates for their location and diploid loci for both fitness and dispersal.
+The model simulates a population of male and female individuals characterized by spatial coordinates for their location and by diploid loci for both fitness and dispersal.
 Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis.
@@
 
-The abiotic environment is defined by a parameter denoting the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}).
-The range can then shift in space by altering the parameter defining the range center, which is how climate change is implemented in the model.
+The abiotic environment is defined by the location of the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}).
+The range can then shift in space by altering the location of the range center, which is how climate change is implemented in the model.
 Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries.
-The severity of this gradient is defined by a separate parameter independent of the other environmental parameters (greater values of this parameter will lead to steeper gradients in phenotypic optima within the range).
+The severity of this gradient is defined independently of the other environmental parameters.
 
@@
 Time is also modeled in discrete intervals defining single generations of the population.
-Within each generation, individuals first reproduce within their natal patch.
-After reproduction, all adults die and the offspring disperse according to their dispersal kernels (\textit{Submodels}), resulting in discrete, non-overlapping generations (See Figure~\ref{fig:LifeCycle}).
-Reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
+Within each generation, individuals first disperse from their natal patches according to their dispersal phenotypes (\textit{Submodels}).
+After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
 Parental pairs form via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with high fitness are likely to produce multiple offspring while individuals with no fitness may not produce any.
 Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process (\textit{Submodels}).
-After reproduction, all individuals in the parental generation perish, forming discrete non-overlapping generations.
-The offspring then disperse according to their dispersal traits to start the next generation in their post dispersal locations.
+After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations.
 
@@
 Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$).
-The maximum achievable carrying capacity ($K_{max}$) occurs in the center of the species' range and declines with increasing distance from the center.
-Specifically, the carrying capacity at a location $x$ is defined as a combination of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows
+The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center.
+Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows
 \begin{equation}
@@
 \end{equation}
-where $beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}).
+where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}).
 To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
@@
 \begin{cases}
-	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau}} & x \leq \beta_{t} \\
-	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau}} & x > \beta_{t}
+	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
+	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
 \end{cases}	
@@
 By varying the parameters defining $f(x,t)$, we can change both the total achievable carrying capacity of the population throughout the range (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$).
-We are primarily interested in the effect of altering the slope at which $K_{x}$ declines at the range boundaries, however as doing so can also alter the total achievable carrying capacity of the population, we control for this confounding factor by fixing the total area under the curve $f(x,t)$.
+Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total achievable carrying capacity of the population.
+To avoid this confounding factor, we fix the total area under the curve $f(x,t)$.
 The indefinite integral of $f(x,t)$ can be shown to be
@@
 \end{equation}
-which can be solved for $\tau$. Thus, if the total area under the curve is fixed, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.
+which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.
 
-Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change.
+Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change.
 During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
@@
 
-Parentage of the offspring is then assigned by random sampling of the local male and female population.
+Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e.
+polygynandrous mating).
 The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any.
@@
 \end{equation}
-where $D$ is the maximum expected dispersal distance in terms of discrete patches, $\eta$ is the spatial scale of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $D$, and the summation is the sum of all alleles contributing to dispersal.
+where $D$ is the maximum expected dispersal distance in terms of discrete patches, $\eta$ is the spatial scale of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $D$, and the summation is taken across all alleles contributing to dispersal.
 Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis.

### 2018-07-03 Manuscript/Main.tex
Updating the simulation work flow

@@
 In contrast, range shifts start from entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}.
-Such spatial structure can take the form of local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
+Such spatial structure can manifest through local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
 
@@
 Specifically, a broad environmental niche (i.e.
-little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes~\citep{atkins2010local}.
+little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted genotypes~\citep{atkins2010local}.
 Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
-However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and fitness are directly linked via gene flow.
+However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
 For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
@@
 
-Here, we develop a complex, individual-based model of a sexually reproducing species, which incorporates evolution in both fitness (via local adaptation) and dispersal distance to determine the interaction between the strength of local selection and the starkness of the gradient at the range edge in driving population dynamics during climate change.
-Using this model, we simultaneously vary the strength of local selection and the starkness of the range boundary to ascertain the relationship between them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
-We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction during climate-induced range shifts.
+Here, we develop a complex, individual-based model of a sexually reproducing species with two genetically determined traits, one defining an individual's expected dispersal distance and the other determining an individual's environmental niche.
+Using this model, we simultaneously vary the starkness of the gradient defining the range edge and the potential for local adaptation within the range to ascertain their impact on a population's ability to track a changing climate and how dispersal evolution and local adaptation contribute to the dynamics of the range shift.
+We additionally compare the dynamics of simulated populations that survived climate change versus those that went extinct to understand the factors contributing to extinction during climate-induced range shifts.
 
@@
 This model tests an evolving population's ability to track a changing climate under a variety of conditions.
-Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the strength of local selection.
-In all simulations, individual dispersal and relative fitness are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
+Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the potential for local adaptation.
+In all simulations, an individual's expected dispersal distance and environmental niche are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
 All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change.
@@
 \subsubsection{State variables and scales}
-The model simulates a population of male and female individuals characterized by spatial coordinates for their location and by diploid loci for both fitness and dispersal.
+The model simulates a population of male and female individuals characterized by spatial coordinates for their location and by diploid loci for both their expected dispersal distance and environmental niche.
 Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis.
-To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same row of the landscape.
+To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same column of the landscape.
 Patches are defined by the location of the patch center in x and y coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (see the \textit{Submodels} section below).
 
-The abiotic environment is defined by the location of the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}).
-The range can then shift in space by altering the location of the range center, which is how climate change is implemented in the model.
-Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries.
-The severity of this gradient is defined independently of the other environmental parameters.
+The environmentally suitable habitat available to a population (i.e.
+its potential range) is defined by the center of available habitat, the starkness of the decline in suitable habitat away from the center, and the width of suitable habitat (See Figure~\ref{fig:EnvFunction}).
+Environmentally suitable habitat can shift in space by altering the location of the center of available habitat, which is how climate change is implemented in the model.
+Further, a gradient in environmental conditions is imposed throughout the landscape to allow for local adaptation via matching of an individual's environmental niche to the local environmental conditions (See \textit{Submodels}).
+The severity of this gradient can be altered to change the potential for local adaptation (e.g.
+a shallower gradient will result in more similar environmental conditions throughout the range and therefore reduce the potential for local adaptation).
 
@@
 \vspace{-5mm}
-\caption[LoF entry]{Environmental quality, as defined by $f(x,t)$, over a spatial gradient.
+\caption[LoF entry]{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space.
 The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
-Additionally, the lattice of discrete patches in which population dynamics occur is shown beneath.
-This lattice of $\eta$ x $\eta$ patches are mapped to environmental conditions along the $x$ axis using $f(x,t)$ and are environmentally constant along the $y$ axis.
-Dispersal is unbounded in the $x$ direction and implemented with wrapping boundaries in the $y$ direction.}
+The lattice of discrete patches in which population dynamics occur is shown beneath.
+As described in the \textit{Submodels} section, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches.
+Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension.
+Dispersal is unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
 \label{fig:EnvFunction}
@@
 After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
-Parental pairs form via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with high fitness are likely to produce multiple offspring while individuals with no fitness may not produce any.
+Parental pairs form via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) are likely to produce multiple offspring while individuals with low relative fitness may not produce any.
 Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process (\textit{Submodels}).
@@
 \subsubsection{Emergence}
-Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both fitness and dispersal traits during climate change.
+Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both expected dispersal distances and environmental niches during climate change.
 These are all examined in the context of their impact on the population's ability to keep pace with the changing climate.
@@
 Other important interactions are the relationship between dispersal evolution and local adaption, particularly in edge populations, and how this relationship impacts a population's ability to avoid extinction and track a changing climate.
-These interactions are examined in the context of different configurations of the initial, stable range and different speeds of climate change.
+These interactions are examined in the context of different configurations of the initial, stable range.
 
@@
 This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation.
-See Table 1 for a full list of all parameter values used in the simulations described here.
+See Tables 1 \& 2 for a full list of parameter values used in the simulations described here.
 
@@
 \paragraph{Local adaptation}
-To allow an arbitrary degree of local adaptation within the range, the local phenotypic optima ($z_{opt,x}$) is set as follows
+To allow an arbitrary degree of local adaptation within the range, the local environmental conditions determining the phenotypic optima ($z_{opt,x}$) are set as follows
 \begin{equation}
@@
 \end{equation}
-where $\lambda$ defines the strength of local selection with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range.
-Individual fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
+where $\lambda$ defines the potential for local adaptation with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range.
+Individual relative fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
 \begin{equation}
@@
 \end{equation}
-where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype~\citep{lande1976natural}.
-All loci are assumed to contribute additively to the phenotype with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values for the fitness trait.
+where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's environmental niche~\citep{lande1976natural}.
+Thus, an individual's realized fitness will be higher the closer its environmental niche ($z_[i]$) is to the local environmental conditions of the patch it occupies ($z_{opt,x}$).
+All loci are assumed to contribute additively to an individual's environmental niche value with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values.
 
@@
 
+\subsection{Simulation experiments}
+Most of the parameters described above were held constant over all simulations (Table~\ref{table:ConstPars}), but we varied certain parameter values to explore the interacting roles of local adaptation and the starkness of the gradient in environmentally suitable habitat at the range edge (Table~\ref{table:Scenarios}).
+Specifically, we considered a factorial combination of no, low, and high potential for local adaptation with shallow, moderate, and stark gradients in habitat quality at the range edge for a total of $9$ different scenarios, each explored with 200 simulations.
+Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
+Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change.
+All simulations were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
+Additionally, we explored the role of the rate of climate change on the dynamics of range shifts in these different scenarios, examining both slower and faster speeds of climate change (see \textit{Appendix}).
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
+\vspace{-5mm}
+\caption[LoF entry]{A single example of a simulation with a high potential for local adaptation and a moderate gradient at the range edge.
+Information on the abundance, dispersal, and fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change ($t = 40$).
+For the dispersal trait, individual phenotypes were calculated according to equation 12 and the local population mean was calculated for each patch.
+The log transformed local means are shown on the graph.
+Average patch fitness was calculated based on the mean local environmental niche and the optimal niche value for that location according to equation 7.}
+\label{fig:SimExample}
+\end{figure}
+
 \begin{table}
@@
     $L^{T}$ & The number of diploid loci defining each trait & $5$ for each trait \\
-    $\mu_{1}^{f}$ & The initial mean allele value for the fitness trait & $0$ \\
+    $\mu_{1}^{f}$ & The initial mean allele value for the environmental niche trait & $0$ \\
     $\mu_{1}^{d}$ & The initial mean allele value for the dispersal trait & $-1$ \\
-    $\sigma_{1}^{f}$ & The initial standard deviation of allele values for fitness & $0.025$ \\
-    $\sigma_{1}^{d}$ & The initial standard deviation of allele values for dispersal & $1$ \\
+    $\sigma_{1}^{f}$ & The initial standard deviation of allele values for the environmental niche trait & $0.025$ \\
+    $\sigma_{1}^{d}$ & The initial standard deviation of allele values for the dispersal trait & $1$ \\
     \hline
@@
     \hline
-    Habitat gradient at range edge & Strength of local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
+    Habitat gradient at range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
      & None & $0.0025$ & $250$ & $0$ \\
-    Shallow & Moderate & $0.0025$ & $250$ & $0.004$ \\
-     & Strong & $0.0025$ & $250$ & $0.008$ \\ \hline
+    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
+     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
      & None & $0.025$ & $421.479$ & $0$ \\
-    Moderate & Moderate & $0.025$ & $421.479$ & $0.004$ \\
-     & Strong & $0.025$ & $421.479$ & $0.008$ \\ \hline
+    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
+     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
      & None & $0.25$ & $421.48$ & $0$ \\
-    Stark & Moderate & $0.25$ & $421.48$ & $0.004$ \\
-     & Strong & $0.25$ & $421.48$ & $0.008$ \\ 
+    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
+     & High & $0.25$ & $421.48$ & $0.008$ \\ 
     \hline
   \end{tabular}
-\caption[LoF entry]{Descriptions and parameter values for the $9$ different scenarios examined here. For each scenario, $\nu$ (the speed of climate change) was varied among $0.5$, $1$, and $2$.}
+\caption[LoF entry]{Descriptions and parameter values for the $9$ different scenarios examined here.}
 \label{table:Scenarios}
@@
 
-\subsection{Simulation descriptions}
-Most of the parameters described above were held constant over all simulations we performed (Table ~\ref{table:ConstPars}).
-We varied certain parameter values to explore the interacting roles of local adaptation, the habitat gradient at the range edge, and the speed of climate change in the dynamics of range shifts.
-Specifically, we considered a factorial combination of three severities of habitat quality decline at the range edge, three degrees of local adaptation within the range, and three speeds of climate change (Table~\ref{table:Scenarios}) for a total of $27$ different model scenarios.
-For each scenario, we performed $100$ simulations to assess the interacting ecological and evolutionary dynamics of climate driven range shifts.
-All simulations were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
+\section{Results}
+\subsection{Extinction probability}
+To assess the joint impact of the potential for local adaptation and the severity of the environmental gradient defining the range edge on the probability of extinction during climate change, we calculated the proportion of simulated populations to go extinct through time for the $100$ generations of simulated climate change (Figure~\ref{fig:ExtProb}).
+Increases in both the potential for local adaptation and the severity of the gradient at the range edge resulted in increased probabilities of extinction through time.
+However, increased potential for local adaptation seemed to drive larger changes in extinction probabilities than increases in the severity of habitat gradients, at least over the parameter values examined here.
+Similar patterns were observed for both slower and faster speeds of climate change.
+As expected, slower speeds of climate change resulted in reduced probabilities of extinction while faster speeds of climate change increased extinction probabilities in all scenarios, but the potential for local adaptation and the severity of the environmental gradient at the range edge caused similar increases in extinction probability at all speeds of climate change (see Appendix).
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PaperGraphs/Extinction"}
+\vspace{-5mm}
+\caption[LoF entry]{The cumulative probability of extinction during climate change in experimental scenarios.
+The probabilities of extinction for simulated populations with no potential for local adaptation are shown in panel (a).
+The probabilities of extinction for simulated populations with low and high potential for local adaptation are shown in panels (b) and (c) respectively.
+In each panel populations characterized by a shallow, moderate, or stark gradient at the range edge are shown with solid, dashed, and dotted lines respectively as shown in the legend.
+Cumulative probabilities of extinction were calculated as the proportion of simulated populations to be extinct through time.}
+\label{fig:ExtProb}
+\end{figure}
+
+\subsection{Initial conditions}
+Examining those simulations that survived climate change and comparing them to the simulated populations that went extinct revealed key differences driving extinction dynamics during climate change.
+Primarily, the simulations that survived were largely composed of higher dispersal phenotypes at the beginning of climate change compared to the simulations that went extinct (Fig.~\ref{fig:InitDisp}), suggesting an important role for the initial distribution of dispersal traits in populations responding to climate change.
+In fact, a threshold seems to divide populations that ultimately survived climate change from those that went extinct at roughly a mean dispersal phenotype of $100$.
+This corresponds to a roughly $10\%$ chance of an individual moving a single patch in the $x$ dimension, which would be sufficient to keep the same relative location in the range during climate change in our simulations.
+Examining simulations with different speeds of climate change reveals that this threshold varies with the speed of climate change, but similarly divides extinct from surviving populations (see Appendix).
+
+Additionally, the simulations that survived climate change tended to be characterized by reduced fitness at the range margins compared to simulated populations that went extinct (Fig.~\ref{fig:InitFit}).
+This pattern was most evident in the simulations with (1) a gradual environmental gradient at the range edge and (2) high potential for local adaptation.
+It was not present in simulations with no potential for local adaptation.
+As the simulations that survived climate change were also characterized by heightened dispersal (Fig.~\ref{fig:InitDisp}), the observed reduction in average patch fitness at the margins is likely due to increased gene flow from the range core hampering the abilities of these edge populations to adapt to local conditions.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PaperGraphs/InitDispVals"}
+\vspace{-5mm}
+\caption[LoF entry]{Histograms of the log transformed mean dispersal phenotypes in each patch after $2000$ generations of stable climate conditions.
+On each histogram, the values associated with populations surviving climate change are highlighted in green and the total number of simulated populations to survive climate change is indicated in the top left corner.
+Each histogram represents one of the $9$ experimental scenarios examined here, with potential for local adaptation increasing from none to high from top to bottom and the severity of the gradient at the range edge increasing from gradual to severe from left to right as indicated on the figure.}
+\label{fig:InitDisp}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PaperGraphs/InitFitSpace6Panel"}
+\vspace{-5mm}
+\caption[LoF entry]{Mean patch level fitness along the $x$ dimensions of the landscape after $2000$ generations of stable climate conditions for both extant (those populations that ultimately survived climate change) and extinct populations.
+Points represent the mean value across simulations and error bars are interquartile ranges.
+The experimental scenario corresponding to each panel is shown on the graph and simulations with no potential for local adaptation are not shown as, by definition, patch fitness does not vary spatially.}
+\label{fig:InitFit}
+\end{figure}
+
+\subsection{Dispersal evolution}
+To quantify dispersal evolution due to climate change, we calculated the change in the mean dispersal phenotypes of individual patches from the beginning of the period to climate change to the end.
+For patches that were unoccupied by the end of the period of climate change due to population extinction, we used the last generation in which the population had at least $10$ individuals as the end point to calculate change in mean dispersal phenotype.
+For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates.
+In this way, we calculated change in mean dispersal phenotype for populations consistently occupying the same position relative to the range center (e.g.
+leading edge vs.
+core populations).
+Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions) so that positive values indicate an increase in the mean dispersal phenotype.
+
+Dispersal evolved in all simulated populations throughout the period of climate change.
+However, there was no substantial difference in the evolutionary trajectories of dispersal in populations that successfully tracked climate change versus those that went extinct (Fig.~\ref{fig:DispEvol}).
+Additionally, contrary to expectations, range shifts did not result in the evolution of significantly increased dispersal capabilities.
+Rather, simulated populations underwent both increases and decreases in average dispersal capability, though the scale of changes was small compared to the distinction seen between extant and extinct populations in initial dispersal phenotypes (Figs.~\ref{fig:InitDisp} \&~\ref{fig:DispEvol}).
+Essentially, the magnitude of dispersal evolution possible in the time scale of simulated climate change did not seem to be able to rescue populations starting from initial conditions characterized by lower dispersal capabilities.
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PaperGraphs/DispEvolHists"}
+\vspace{-5mm}
+\caption[LoF entry]{Histograms of the change in the log transformed mean dispersal phenotypes in each patch from the beginning of the period of climate change to the end.
+Positive values indicate an increase in average dispersal ability in the patch.
+Each patch was defined by its location relative to the range center ($\beta_{t}$) so changes in dispersal correspond to relative (e.g.
+leading and trailing edge patches), as opposed to absolute (i.e.
+fixed spatial coordinates), locations in the range.
+For those simulated populations that went extinct during climate change, we calculated the change in dispersal phenotype using the last time point with at least $10$ individuals in the patch as described in the text.
+The values associated with populations that survived climate change are highlighted in green on the graphs.
+Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure.}
+\label{fig:DispEvol}
+\end{figure}
+
+\section{Discussion}
+Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
+While previous models have focused on ecological dynamics (CITATIONS), evolution in dispersal only (CITATIONS), and relatively simple genetic scenarios (e.g.
+single-locus haploid genetics in asexual populations) (CITATIONS), we developed an individual-based model to explore the eco-evolutionary dynamics of climate induced-range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
+Using this model, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
+
+Both the potential for local adaptation and the severity of the environmental gradient at the range edges significantly impacted the probability of extinction for populations during climate change (Fig.~\ref{fig:ExtProb}) with the potential for local adaptation seeming to cause more dramatic increases.
+The parameters defining both of these range attributes were varied widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table~\ref{table:Scenarios}), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space.
+While overall extinction probability changed as expected with different speeds of climate change, the effects of increased potential for local adaptation and increased severity of the environmental gradient at the range edges on extinction probability remained the same (see Appendix).
+
+These extinction dynamics seemed to manifest via the distribution of dispersal phenotypes that evolved in response to different aspects of spatial population structure.
+Simulated populations that survived climate change all began the period of climate change with populations defined by heightened dispersal phenotypes (Fig.~\ref{fig:InitDisp}).
+In particular, the surviving populations all began climate change with populations defined by dispersal phenotypes at or above the threshold necessary for about $10\%$ of the population to keep pace with the changing climate.
+In scenarios with higher potential for local adaptation and more severe gradients at the range edge, simulated populations were less likely to have such high dispersal phenotypes due to selection against high dispersal distances due to local adaptation (CITATIONS) and the risk of dispersing beyond the boundary of suitable habitat (CITATION).
+As a result, populations in those scenarios faced a greater risk of extinction during climate change.
+This threshold value of dispersal phenotypes changed accordingly with the speed of climate change, accounting for the resultant changes in extinction probability under the different scenarios (see Appendix).
+
+While high dispersal phenotypes prior to climate change allowed populations to survive, it had the additional effect of reducing average fitness at the range margins in scenarios with the potential for local adaptation (Fig.~\ref{fig:InitFit}).
+In the model, populations at the range margins tended to be lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core (CITATION).
+Thus, when populations were defined by high dispersal phenotypes prior to climate change, it reduced fitness at the range margin due to gene flow preventing adaptation to local conditions.
+As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the start of climate change.
+While not all populations are characterized by small populations at the range edges (CITATION), and thus are unlikely to be equally susceptible to the influence of gene flow from the core, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
+
+Previous research has suggested that evolution of increased dispersal ability during climate change may be able to rescue populations that would otherwise be unable to keep pace with shifting environmental conditions (CITATIONS).
+Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
+Models finding that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid (CITATION) and single-locus models (CITATION).
+Such simplified genetic frameworks can increase the potential for evolutionary change by, for example, increasing the effect of individual mutations (CITATION).
+Additionally, many previous models of climate-induced range shifts have used asexual reproduction, which can also increase the potential for evolutionary change by allowing for near complete fidelity of trait values passed from parent to offspring (CITATION).
+The negligible role played by dispersal evolution in our model (Fig.~\ref{fig:DispEvol}) suggests that when such simplifying assumptions are relaxed the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
+
+\subsection{Conclusion}
+Understanding the various ecological and evolutionary drivers of climate-induced range shifts is crucial to current and future conservation efforts.
+In particular, we need to better understand how these various mechanisms combine to shape the extinction probabilities of populations undergoing range shifts, which, in turn, will allow more focused interventions.
+Our results suggest that initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
+Further, in contrast to other studies assuming more simplified breeding and genetic structure, we find very little role for the evolution of heightened dispersal abilities to rescue a population composed of low-dispersing individuals.
+Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue.
+As climate change continues to accelerate (CITATION), it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such increased risk.
 

### 2018-07-03 1b_SimCheck.R

@@ -1,0 +1,28 @@
+# This script will check for extinction before the period of climate change,
+#    delete those simulations, and output the current number of simulations
+#    that make it to equilibriation.
+
+setwd("~/ShiftingSlopes/MainSim/")
+NumSims <- rep(NA, 9)
+
+for(p in 1:9){
+     ParamDirectory <- paste("~/ShiftingSlopes/MainSim/Params", p, sep = "")
+     ParamSims <- list.files(ParamDirectory)
+     for(i in 1:length(ParamSims)){
+          NumFiles <- length(list.files(paste(ParamDirectory, ParamSims[i], sep = "/")))
+          if(NumFiles != 4){
+               SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[i], sep = "")
+               system(SysCommand)
+          } else{
+               InFile <- paste(ParamDirectory, ParamSims[i], "InitialPopMat.csv", sep = "/")
+               PopMat <- read.csv(InFile)
+               if(dim(PopMat)[1] == 0){
+                    SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[i], sep = "")
+                    system(SysCommand)
+               }
+          }
+     }
+     NumSims[p] <- length(list.files(ParamDirectory))
+}
+
+NumSims

### 2018-07-03 1b_SimCheck.sh

@@ -1,14 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=50gb
+#PBS -l walltime=00:15:00,nodes=1:ppn=1,mem=2gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange
+# Set the names of R scripts and log files
+Rscript=1b_SimCheck.R
+LogFile=1b_SimCheck.log
 
-Rscript=4c_Extinctions.R
-LogFile=4c_Extinctions.log
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4
@@ -17,6 +18,4 @@ module load ompi/3.0.0/gnu-7.2.0
 export RMPI_TYPE=OPENMPI
 export OMPI_MCA_mpi_warn_on_fork=0
 
-# Run the R script itself, saving the output to
-#	a log file
 mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-07-03 1c_MainSimCont.sh

@@ -1,15 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=15:00:00,nodes=18:ppn=24,mem=150gb
+#PBS -l walltime=08:00:00,nodes=6:ppn=24,mem=300gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the parameter combination to use and generate names of R scripts and log files
-Rscript=1_StatRangeSim.R
-LogFile=1_StatRangeSim.log
+Rscript=1c_MainSimCont.R
+LogFile=1c_MainSimCont.log
 
 # Change to the relevant working directory
-cd ~/ShiftingSlopes/StationaryRange
+cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4

### 2018-07-03 2_SlowSim.sh

@@ -1,15 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=03:00:00,nodes=18:ppn=24,mem=150gb
+#PBS -l walltime=06:00:00,nodes=18:ppn=24,mem=300gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the parameter combination to use and generate names of R scripts and log files
-Rscript=3a_ExtractAbunds.R
-LogFile=3a_ExtractAbunds.log
+Rscript=2_SlowSim.R
+LogFile=2_SlowSim.log
 
 # Change to the relevant working directory
-cd ~/ShiftingSlopes/StationaryRange
+cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4

### 2018-07-03 3_FastSim.R

@@ -1,0 +1,78 @@
+# This script will run simulations on the MSI compute cluster to characterize
+#    population structure in a stable range for all of the 9 range parameter
+#    combinations to be explored.
+
+# Set the number of processors and number of simulations to be run
+nProc <- 24*18
+NumSims <- 200
+
+# Set the working directory and load necessary data and libraries
+source("~/ShiftingSlopes/SimFunctions.R")
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+library(parallel)
+library(Rmpi)
+
+# Now load in the parameters and compile all 9 combinations into a single list
+CurSpeed <- 2
+AllParams <- vector(mode = "list", length = 9)
+for(i in 1:9){
+     CurDirectory <- paste("~/ShiftingSlopes/MainSim/Params", i, sep = "")
+     ExampleSim <- list.files(CurDirectory)[1]
+     ParamFile <- paste(CurDirectory, "/", ExampleSim, "/parameters.R", sep = "")
+     source(ParamFile)
+        
+     v <- CurSpeed
+     BurnIn <- 0
+     AllParams[[i]] <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
+                         Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
+                         LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
+                         DispDiv, eta, NumRands, z, dmax, rho)
+     names(AllParams[[i]]) <- c("BetaInit", "gamma", "tau", "lambda", "omega", "U", "Vm", 
+                              "Lf", "Ld", "Rmax", "Kmax", "width", "kern", "EnvGradType", 
+                              "monoecious", "BurnIn", "BurnOut", "LengthShift", "v", 
+                              "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
+                              "eta", "NumRands", "z", "dmax", "rho")
+}
+
+# Write a function to be passed to various nodes
+SimFunc <- function(i){
+     # Load in the appropriate initial population matrix
+     p <- ParamVec[i]
+     InFile <- paste("~/ShiftingSlopes/MainSim/Params", p, "/", 
+                     StatSimIDVec[i], "/InitialPopMat.csv", sep = "")
[…]

### 2018-07-03 4_Extinctions.sh

@@ -1,15 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=00:10:00,nodes=1:ppn=1,mem=124gb
+#PBS -l walltime=00:30:00,nodes=2:ppn=24,mem=50gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=SanityCheckCheck.R
-LogFile=SanityCheckCheck.log
-
 # Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange
+cd ~/ShiftingSlopes/
+
+# Set the names of R scripts and log files
+Rscript=4_Extinctions.R
+LogFile=4_Extinctions.log
 
 # Load R and MPI
 module load R/3.4.4
@@ -18,4 +18,6 @@ module load ompi/3.0.0/gnu-7.2.0
 export RMPI_TYPE=OPENMPI
 export OMPI_MCA_mpi_warn_on_fork=0
 
+# Run the R script itself, saving the output to
+#	a log file
 mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-07-03 5_FitData.sh

@@ -1,15 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=06:00:00,nodes=18:ppn=24,mem=150gb
+#PBS -l walltime=00:30:00,nodes=4:ppn=24,mem=248gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=3b_ExtractTraits.R
-LogFile=3b_ExtractTraits.log
+# Set the names of R scripts and log files
+Rscript=5_FitData.R
+LogFile=5_FitData.log
 
 # Change to the relevant working directory
-cd ~/ShiftingSlopes/StationaryRange
+cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4

### 2018-07-03 6_DispData.R

@@ -1,0 +1,169 @@
+# This script will extract the initial dispersal values of individuals immediately
+#    prior to climate change (DispInit) and the change in mean dispersal values at
+#    the patch level (along with the relative x coordinate of the patch) to assess
+#    the role of evolution (DispEvol).
+
+# Set the number of processors available for this scrip
+nProc <- 4*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+NumSims <- 200
+
+# Create the DispInit master list
+DispInit <- vector(length = 9, mode = "list")
+for(p in 1:9){
+     DispInit[[p]] <- vector(length = NumSims, mode = "list")
+}
+
+# Set some initial parameter values to assess evolutionary change in dispersal
+Threshold <- 10
+BetaInit <- 0
+BurnIn <- 2000
+BurnOut <- 50
+LengthShift <- 100
+dmax <- 1000
+eta <- RangeParams$eta[1]
+rho <- 0.5
+ClimSpeeds <- c(0.5, 1, 2)
+MaxGen <- c(151, 2151, 151)
+BetaCoord <- matrix(NA, nrow = 3, ncol = BurnIn + LengthShift + BurnOut)
+for(i in 1:3){
+     BetaShift <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, 
+               eta = eta, v = ClimSpeeds[i]) / eta
+     BetaCoord[i,] <- c(rep(BetaInit, BurnIn), BetaShift, rep(BetaShift[LengthShift], BurnOut))
+}
+
+# Create a data frame to index the individual simulations and parameter values
+TraitIndices <- expand.grid(params = 1:9, sim = 1:NumSims)
+
+# Write a function to extract the initial dispersal values
+DispFunc <- function(i){
+     # Sort out the parameter combination and simulation under consideration
+     Param <- TraitIndices$params[i]
+     AllSims <- list.files(paste("~/ShiftingSlopes/MainSim/Params", Param, sep = ""))
+     SimID <- AllSims[TraitIndices$sim[i]]
+     
+     # Load the corresponding initial population matrix
+     InFile <- paste("~/ShiftingSlopes/MainSim/Params", Param, "/", 
+                     SimID, "/InitialPopMat.csv", sep = "")
[…]

### 2018-07-03 6_DispData.sh

@@ -1,15 +1,15 @@
 #!/bin/bash -l
-#PBS -l walltime=00:10:00,nodes=1:ppn=1,mem=2gb
+#PBS -l walltime=03:00:00,nodes=4:ppn=24,mem=248gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=2b_SimCleanUp.R
-LogFile=2b_SimCleanUp.log
+# Set the names of R scripts and log files
+Rscript=6_DispData.R
+LogFile=6_DispData.log
 
 # Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange
+cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4

### 2018-07-03 8a_ExtinctionGraphs.R

@@ -1,0 +1,89 @@
+# This script will create the graphs from the extinction data.
+setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
+library(RColorBrewer)
+RangeParams <- read.csv("RangeParameters.csv")
+NumSims <- 200
+
+# Load the extinction data
+load("Extinctions.rdata")
+for(i in 1:3){
+     for(p in 1:9){
+          Extinctions[i,p,] <- cumsum(Extinctions[i,p,]) / NumSims
+     }
+}
+
+# Set some graphical parameters to use for the subsequent figures
+OuterMar <- c(1,1,0,0)
+TextSize <- 1.15
+LegTextSize <- 1.5
+AxisSize <- 1.25
+LetterPos <- c(2, 0.975)
+FigWidth <- 8
+FigHeight <- 3
+xRange <- c(0,100)
+yRange <- c(0,1)
+TimeAxisSeq1 <- seq(0, 100, by = 20)
+TimeAxisSea2 <- seq(0, 100, by = 5)
+ExtAxisSeq1 <- seq(0, 1, by = 0.2)
+ExtAxisSeq2 <- seq(0, 1, by = 0.05)
+GradLineType <- c(1,2,3)
+ExtAxisLabel <- "Extinction Probability"
+LineWidth <- 2
+Col <- "darkred"
+
+# Make the extinction graph for all three speeds
+SpeedWords <- c("Slow", "Main", "Fast")
+for(i in 1:3){
+     PlotName <- paste(SpeedWords[i], "Extinction.pdf", sep = "")
+     pdf(file = PlotName, width = FigWidth, height = FigHeight, onefile = FALSE, paper = "special")
+          par(mfrow = c(1,3), oma = OuterMar)
+          # No local adaptation
+          plot(x = 1:100, y = Extinctions[i,1,], lty = GradLineType[1], lwd = LineWidth,
+               col = Col, xaxt = "n", yaxt = "n", main = "", xlab = "", ylab = "",
+               xlim = xRange, ylim = yRange, type = "l")
+          lines(x = 1:100, y = Extinctions[i,2,], lty = GradLineType[2], lwd = LineWidth,
+                col = Col)
+          lines(x = 1:100, y = Extinctions[i,3,], lty = GradLineType[3], lwd = LineWidth,
+                col = Col)
+          text(x = LetterPos[1], y = LetterPos[2], labels = "(a)", cex = TextSize)
[…]

### 2018-07-03 8c_InitFitGraphs.R

@@ -1,0 +1,154 @@
+# This script will make an appropriate graph of the initial dispersal values
+#    extracted by the InitDispVals.R script.
+
+setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
+load("FitData.rdata")
+CurSpeed <- 3
+SpeedWords <- c("Slow", "Main", "Fast")
+
+# Sort the disp values into the appropriate lists
+NumSims <- 200
+ExtinctSims <- vector(length = 9, mode = "list")
+ExtantSims <- vector(length = 9, mode = "list")
+NumExtant <- rep(0, 9)
+xMin <- 0
+xMax <- 0
+for(p in 1:9){
+     ExtinctVals <- NULL
+     ExtantVals <- NULL
+     for(i in 1:NumSims){
+          if(FitList[[p]][[i]]$Ext[CurSpeed] == 0){
+               ExtinctVals <- rbind(ExtinctVals, FitList[[p]][[i]]$w)
+          } else{
+               ExtantVals <- rbind(ExtantVals, FitList[[p]][[i]]$w)
+          }
+     }
+     ExtinctResults <- data.frame(x = ExtinctVals[,1], w = ExtinctVals[,2])
+     ExtantResults <- data.frame(x = ExtantVals[,1], w = ExtantVals[,2])
+     ExtinctSims[[p]] <- ExtinctResults
+     ExtantSims[[p]] <- ExtantResults
+     xMin <- min(c(xMin, ExtinctResults$x, ExtantResults$x))
+     xMax <- max(c(xMax, ExtinctResults$x, ExtantResults$x))
+}
+xRange <- c(xMin, xMax)
+
+# Now make a list for both extinct and extant sims with data frames for plotting
+#    NOTE: the plots will likely exclude the no local adaptation scenario, but
+#    we calculate them here to confirm that
+ExtantPlot <- vector(length = 9, mode = "list")
+ExtinctPlot <- vector(length = 9, mode = "list")
+for(p in 1:9){
+     ExtantxVals <- unique(ExtantSims[[p]]$x)
+     ExtinctxVals <- unique(ExtinctSims[[p]]$x)
+     ExtinctPlotData <- data.frame(x = rep(NA, length(ExtinctxVals)),
+                                  wBar = rep(NA, length(ExtinctxVals)),
+                                  lwr = rep(NA, length(ExtinctxVals)),
[…]

### 2018-07-03 SanityCheck.R

@@ -1,71 +1,0 @@
-# This script will run a sanity check with a single parameter combination and
-#    climate change with a speed of 0 to ensure there are no extinctions and if
-#    there are that there are no intital condition constraints (I guess. This is a 
-#    pretty stupid waste of time).
-
-# Set the number of processors and number of simulations to be run
-nProc <- 24*2
-NumSims <- 50
-
-# Set the working directory and load necessary data and libraries
-setwd("~/ShiftingSlopes/ShiftingRange/")
-source("~/ShiftingSlopes/SimFunctions.R")
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-library(parallel)
-library(Rmpi)
-
-# Now make a parameter list for just a single set of parameters
-CurParam <- 1
-v <- 0
-ParamFile <- paste("ParamFiles/Params", CurParam, ".R", sep = "")
-source(ParamFile)
-source("ShiftParams.R")
-AllParams <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
-                  Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
-                  LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
-                  DispDiv, eta, NumRands, z, dmax, rho)
-names(AllParams) <- c("BetaInit", "gamma", "tau", "lambda", "omega", "U", "Vm", 
-                      "Lf", "Ld", "Rmax", "Kmax", "width", "kern", "EnvGradType", 
-                      "monoecious", "BurnIn", "BurnOut", "LengthShift", "v", 
-                      "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
-                      "eta", "NumRands", "z", "dmax", "rho")
-
-# Write a function to be passed to various nodes
-SimFunc <- function(i){
-     InFile <- paste("~/ShiftingSlopes/StationaryRange/Params", CurParam, "/", 
-                     StatSimIDVec[i], "/PopMat.csv", sep = "")
-     SysCommand1 <- paste("gunzip ", InFile, ".gz", sep = "")
-     SysCommand2 <- paste("gzip ", InFile, sep = "")
-     system(SysCommand1)
-     InputMat <- read.csv(InFile)
-     system(SysCommand2)
-     InputMat$x1 <- InputMat$x0
-     InputMat$y1 <- InputMat$y0
[…]

### 2018-07-03 SanityCheck.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=03:00:00,nodes=2:ppn=24,mem=100gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=SanityCheck.R
-LogFile=SanityCheck.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-07-03 ShiftingRange/1_UnZipPopMats.sh

@@ -1,16 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=03:00:00,nodes=1:ppn=1,mem=2gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-for (( i=1; i<10; i++));
-do
-     cd ~/ShiftingSlopes/StationaryRange/Params$i
-     SubFolders=($( ls ))
-     for (( j=1; j<101; j++));
-     do
-          gunzip ~/ShiftingSlopes/StationaryRange/Params${i}/${SubFolders[$j]}/PopMat.csv.gz
-     done
-done
-

### 2018-07-03 ShiftingRange/2a_ShiftRangeSim.R

@@ -1,84 +1,0 @@
-# This script will run simulations on the MSI compute cluster to characterize
-#    population structure in a stable range for all of the 9 range parameter
-#    combinations to be explored.
-
-# Set the number of processors and number of simulations to be run
-nProc <- 24*18
-NumSims <- 100
-
-# Set the working directory and load necessary data and libraries
-setwd("~/ShiftingSlopes/ShiftingRange/")
-source("~/ShiftingSlopes/SimFunctions.R")
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-library(parallel)
-library(Rmpi)
-
-# Now load in the parameters and compile all 9 combinations into a single list
-#    with sublists for the three speeds of climate change
-AllParams <- vector(mode = "list", length = 9)
-for(i in 1:9){
-     ParamFile <- paste("ParamFiles/Params", i, ".R", sep = "")
-     source(ParamFile)
-     source("ShiftParams.R")
-     
-     AllParams[[i]] <- vector(mode = "list", length = 3)
-     names(AllParams[[i]]) <- SpeedWords
-     for(j in 1:3){
-          v <- SpeedNums[j]
-          AllParams[[i]][[j]] <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
-                                 Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
-                                 LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
-                                 DispDiv, eta, NumRands, z, dmax, rho)
-          names(AllParams[[i]][[j]]) <- c("BetaInit", "gamma", "tau", "lambda", "omega", "U", "Vm", 
-                                     "Lf", "Ld", "Rmax", "Kmax", "width", "kern", "EnvGradType", 
-                                     "monoecious", "BurnIn", "BurnOut", "LengthShift", "v", 
-                                     "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
-                                     "eta", "NumRands", "z", "dmax", "rho")
-     }
-}
-
-# Write a function to be passed to various nodes
-SimFunc <- function(index){
-     j <- nchar(ParamVec[index])
-     i <- round(ParamVec[index] / 10^(j-1), digits = 1)
[…]

### 2018-07-03 ShiftingRange/2a_ShiftRangeSim.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=06:00:00,nodes=18:ppn=24,mem=100gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=2_ShiftRangeSim.R
-LogFile=2_ShiftRangeSim.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-07-03 ShiftingRange/4a_ExtractAbunds.R

@@ -1,183 +1,0 @@
-# This script will extract the abundance information from the stationary range
-#    simulations. Specifically, this script will save the mean abundance through
-#    time at each sector (i.e. lattice  column with identical environmental 
-#    conditions), the variance in abundance values at each patch
-#    (i.e. among simulation variance), and the variance among
-#    patches within the same sector (i.e. within simulation variance).
-
-# Set the speed for the current script
-SpeedIndex <- 3
-
-# Set the number of processors available for this script
-nProc <- 2*24
-
-# Set the working directory
-setwd("~/ShiftingSlopes/ShiftingRange/")
-source("~/ShiftingSlopes/SimFunctions.R")
-library(parallel)
-library(Rmpi)
-source("ShiftParams.R")
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-
-# Create an array to hold the abundance values from each simulation. Dimensions 
-#    of this array are: parameter combination, simulation number, time, x axis,
-#    y axis.
-#    x and y axis correspond to the center coordinates of each patch in the
-#    discrete lattice
-#    RangeExtent and ZeroPos below correspond to the mapping of unbounded real 
-#    number x values corresponding to patch centers to array indices
-NumGens <- 200
-width <- 10
-RangeExtent <- 121
-AbundVals <- array(NA, dim = c(9, 100, NumGens, RangeExtent, width))
-Success <- matrix(NA, nrow = 9, ncol = 100)
-
-BetaShift <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, 
-                           eta = RangeParams$eta[1], v = SpeedNums[SpeedIndex])
-BetaShift <- BetaShift / RangeParams$eta[1]
-BetaCoord <- c(rep(BetaInit, BurnIn), BetaShift, rep(BetaShift[LengthShift], BurnOut))
-ZeroPos <- 61
-
-# Create a data frame to index each simulation block in the above array
-AbundIndices <- expand.grid(params = 1:9, sim = 1:100)
-
-# Write a function to be passed to various nodes
-AbundExtract <- function(i){
-     # Sort out the parameter combination and simulation under consideration
[…]

### 2018-07-03 ShiftingRange/4b_ExtractTraits.R

@@ -1,294 +1,0 @@
-# This script will extract the trait information from the stationary range
-#    simulations. For both fitness and dispersal the mean value at each time
-#    point and sector (i.e. lattice column with identical environmental 
-#    conditions) for the phenotype, genetic standard deviation, and phenotypic
-#    standard deviation. Additionally, the within and among simulation variance
-#    will be calculated for the mean value of each trait.
-
-# Set the speed for the current script
-SpeedIndex <- 1
-
-# Set the number of processors available for this scrip
-nProc <- 2*24
-
-# Set the working directory
-setwd("~/ShiftingSlopes/ShiftingRange/")
-source("~/ShiftingSlopes/SimFunctions.R")
-library(parallel)
-library(Rmpi)
-source("ShiftParams.R")
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-
-# Create arrays to hold the trait values from each simulation. Dimensions 
-#    of these arrays are: parameter combination, simulation number, time, x axis,
-#    y axis, summary value.
-#    x and y axis correspond to the center coordinates of each patch in the
-#    discrete lattice
-#    summary value ranges from 1 to 3, corresponding to the following:
-#         1 -- MuTrait        (Mean phenotypic value)
-#         2 -- SigTraitPhen   (Phenotypic standard deviation)
-#         3 -- SigTraitGen    (Genetic standard deviation)
-#    RangeExtent and ZeroPos below correspond to the mapping of unbounded real 
-#    number x values corresponding to patch centers to array indices
-NumGens <- 200
-width <- 10
-RangeExtent <- 121
-FitVals <- array(NA, dim = c(9, 100, NumGens, RangeExtent, width, 3))
-DispVals <- array(NA, dim = c(9, 100, NumGens, RangeExtent, width, 3))
-Success <- matrix(NA, nrow = 9, ncol = 100)
-
-BetaShift <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, 
[…]

### 2018-07-03 ShiftingRange/5_FolderCompression.sh

@@ -1,48 +1,0 @@
-#!/bin/bash -l
-#PBS -N FolderCompression
-#PBS -l walltime=18:00:00,nodes=1:ppn=1,mem=2gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange/Slow
-
-# Now sequentially gzip all the files in the simulation folders to save space
-gzip -r Params1/
-gzip -r Params2/
-gzip -r Params3/
-gzip -r Params4/
-gzip -r Params5/
-gzip -r Params6/
-gzip -r Params7/
-gzip -r Params8/
-gzip -r Params9/
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange/Med
-
-# Now sequentially gzip all the files in the simulation folders to save space
-gzip -r Params1/
-gzip -r Params2/
-gzip -r Params3/
-gzip -r Params4/
-gzip -r Params5/
-gzip -r Params6/
-gzip -r Params7/
-gzip -r Params8/
-gzip -r Params9/
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/ShiftingRange/Fast
-
-# Now sequentially gzip all the files in the simulation folders to save space
-gzip -r Params1/
-gzip -r Params2/
-gzip -r Params3/
-gzip -r Params4/
-gzip -r Params5/
-gzip -r Params6/
-gzip -r Params7/
-gzip -r Params8/
-gzip -r Params9/

### 2018-07-03 ShiftingRange/8_GraphExtinction.R

@@ -1,103 +1,0 @@
-# This script will create the graphs from the extinction data.
-library(RColorBrewer)
-
-# Set the working directory
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/ShiftingRange/")
-source("~/Desktop/RangeShifts/ShiftingSlopesCode/ShiftingRange/ShiftParams.R")
-RangeParams <- read.csv("~/Desktop/RangeShifts/ShiftingSlopesOther/RangeParameters.csv")
-
-# Load the extinction data
-load("Extinctions.rdata")
-slow <- Extinctions[1,,]
-med <- Extinctions[2,,]
-fast <- Extinctions[3,,]
-for(i in 1:9){
-     slow[i,] <- cumsum(slow[i,]) / 100
-     med[i,] <- cumsum(med[i,]) / 100
-     fast[i,] <- cumsum(fast[i,]) / 100
-}
-
-# Set some graphical parameters to use for the subsequent figures
-FigMat <- matrix(NA, nrow = 3, ncol = 12)
-FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
-FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
-FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
-OuterMar <- c(4, 10, 5, 2)
-InnerMar <- c(1.5, 1.5, 1.5, 1.5)
-TextSize <- 1.5
-AxisSize <- 1.05
-ArrowWidth <- 3
-ArrowLength <- 0.25
-FigWidth <- 8
-FigHeight <- 6
-LocLabels <- seq(-60, 60, by = 24)
-TimeSeq <- 1:100
-TimeLabels <- seq(0, 100, by = 20)
-SimSeq <- c(7,8,9,4,5,6,1,2,3)
-ColKeyWidth <- 15
-xLabLine <- 1.75
-yLabLine <- 1.5
-GradLabLine <- 3
-GradSubLine <- 1.5
-AdaptLabLine <- 7
-AdaptSubLine <- 5
-TopArrow <- matrix(c(10, 1.2, 740, 1.2), nrow = 2, ncol = 2, byrow = TRUE)
-SideArrow <- matrix(c(-95, -2.75, -95, 0.9), nrow = 2, ncol = 2, byrow = TRUE)
-LowAdj <- 0.05
-HighAdj <- 0.95
-
-LineWidth <- 2
[…]

### 2018-07-03 ShiftingRange/ShiftParams.R

@@ -1,9 +1,0 @@
-# This is simply a parameter file to be sourced to set all the range shift
-#    parameters
-
-SpeedWords <- c("Slow", "Med", "Fast")
-SpeedNums <- c(0.5, 1, 2)
-BetaInit <- 0
-BurnIn <- 50
-LengthShift <- 100
-BurnOut <- 50

### 2018-07-03 SimFunctions.R

@@ -1,1133 +1,0 @@
-# This script contains all of the functions necessary to run a simulation of
-#    population evolution and movement in response to climate change. The 
-#    functions are organized into three different categories: biological,
-#    bookkeeping, and environmental. Each function is documented with a brief
-#    description of the purpose of the function and a list and definition of 
-#    the necessary function arguments before the function definition itself.
-#    Each broad category of functions is delineated by the dashed section
-#    headers and each individual function begins with a single line of 6 pound
-#    signs (######)
-
-
-# ------------------------------------------------------------------------------
-# ---------------------------- Biological Functions ----------------------------
-# ------------------------------------------------------------------------------
-
-###### CalcTraits
-# This function will calculate the fitness and dispersal trait values for a set 
-#    of individuals based on their alleles
-### INPUTS
-# population:  A vector of the population (i.e. row numbers) for which to 
-#                   calculate a trait values.
-# PopMat:      The population matrix to use
-# PopSize:     The size of the current population
-# PopIndices: Column indices to use for the population matrix
-### OUTPUTS
-# The function will create a matrix with three columns. The first column will 
-#    correspond to the ID (row number) of the individual, the second column
-#    will be the fitness trait value, and the third column will be the dispersal
-#    trait. If the traits are calculated for the entire population, as will 
-#    normally be the case, then the first column will simply correspond to the 
-#    row number, but including this column allows for the calculation of 
-#    traits for a subset of the current population.
-CalcTraits <- function(population, PopMat, PopSize, PopIndices){
-     # First create an empty matrix for the trait values and put
-     #    the population IDs in the first column
-     traits <- matrix(NA, nrow = PopSize, ncol = 3)
-     traits[,1] <- population
-     
-     # Now calculate the sum of each individual's quantitative loci for each 
-     #    trait and return those sums in the traits matrix
-     if(PopSize > 1){
-          traits[,2] <- rowSums(PopMat[population, PopIndices$FitCols])
-          traits[,3] <- rowSums(PopMat[population, PopIndices$DispCols])
-     } else{
-          traits[,2] <- sum(PopMat[population, PopIndices$FitCols])
-          traits[,3] <- sum(PopMat[population, PopIndices$DispCols])
[…]

### 2018-07-03 StationaryRange/2_SimCleanUp.R

@@ -1,32 +1,0 @@
-# Make a script to clean up unfinished simulations, extinct simulations, and too
-#    many simulations
-
-for(p in 1:9){
-     ParamDirectory <- paste("~/ShiftingSlopes/StationaryRange/Params", p, "/", sep = "")
-     SimFolders <- list.files(ParamDirectory)
-     for(i in SimFolders){
-          SimDirectory <- paste(ParamDirectory, i, sep = "")
-          SimFiles <- list.files(SimDirectory)
-          NumFiles <- length(SimFiles)
-          if(NumFiles < 3){
-               SysCommand1 <- paste("rm -r ", SimDirectory, sep = "")
-               system(SysCommand1)
-          } else{
-               PopMatFile <- paste(SimDirectory, "/PopMat.csv")
-               PopMat <- read.csv(PopMatFile)
-               if(dim(PopMat)[1] == 0){
-                    SysCommand2 <- paste("rm -r ", SimDirectory, sep = "")
-                    system(SysCommand2)
-               }
-          }
-     }
-     NewSimFolders <- list.files(ParamDirectory)
-     NumSims <- length(NewSimFolders)
-     if(NumSims > 100){
-          SimsToDelete <- NumSims - 100
-          for(i in 1:SimsToDelete){
-               SysCommand3 <- paste("rm -r ", ParamDirectory, NewSimFolders[i], sep = "")
-               system(SysCommand3)
-          }
-     }
-}

### 2018-07-03 StationaryRange/2_SimCleanUp.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=00:10:00,nodes=1:ppn=1,mem=2gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=2_SimCleanUp.R
-LogFile=2_SimCleanUp.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/StationaryRange
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-07-03 StationaryRange/3b_ExtractTraits.R

@@ -1,161 +1,0 @@
-# This script will extract the trait information from the stationary range
-#    simulations. For both fitness and dispersal the mean value at each time
-#    point and sector (i.e. lattice column with identical environmental 
-#    conditions) for the phenotype, genetic standard deviation, and phenotypic
-#    standard deviation. Additionally, the within and among simulation variance
-#    will be calculated for the mean value of each trait.
-
-# Set the number of processors available for this scrip
-nProc <- 24*18
-
-# Set the working directory
-setwd("~/ShiftingSlopes/StationaryRange/")
-library(Rmpi)
-library(parallel)
-
-# Create arrays to hold the trait values from each simulation. Dimensions 
-#    of these arrays are: parameter combination, simulation number, time, x axis,
-#    y axis, summary value.
-#    x and y axis correspond to the center coordinates of each patch in the
-#    discrete lattice
-#    summary value ranges from 1 to 3, corresponding to the following:
-#         1 -- MuTrait        (Mean phenotypic value)
-#         2 -- SigTraitPhen   (Phenotypic standard deviation)
-#         3 -- SigTraitGen    (Genetic standard deviation)
-#    RangeExtent and ZeroPos below correspond to the mapping of unbounded real 
-#    number x values corresponding to patch centers to array indices
-NumGens <- 200
-MaxGen <- 2000
-width <- 10
-RangeExtent <- 121
-ZeroPos <- 61
-FitVals <- array(NA, dim = c(9, 100, NumGens, RangeExtent, width, 3))
-DispVals <- array(NA, dim = c(9, 100, NumGens, RangeExtent, width, 3))
-
-# Create a data frame to index each simulation block in the above arrays
-TraitIndices <- expand.grid(params = 1:9, sim = 1:100)
-
-# Write a function to be passed to various nodes
-TraitExtract <- function(i){
-     # Sort out the parameter combination and simulation under consideration
-     Param <- TraitIndices$params[i]
-     AllSims <- list.files(paste("~/ShiftingSlopes/StationaryRange/Params", Param, "/", sep = ""))
-     SimID <- AllSims[TraitIndices$sim[i]]
-     
[…]

### 2018-07-03 StationaryRange/4a_SaveParamFiles.sh

@@ -1,12 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=00:00:05,nodes=1:ppn=1,mem=2580mb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-for (( i=1; i<10; i++));
-do
-cd ~/ShiftingSlopes/StationaryRange/Params$i
-SubFolders=($( ls ))
-cp ${SubFolders[1]}/parameters.R ~/ShiftingSlopes/ShiftingRange/ParamFiles/Params${i}.R
-done

### 2018-07-03 StationaryRange/6_GraphTraits.R

@@ -1,496 +1,0 @@
-# This script will create the graphs from the abundance data.
-
-# Set the working directory
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/StationaryRange/")
-library(plot3D)
-
-# Set some graphical parameters to use for the subsequent figures
-FigMat <- matrix(NA, nrow = 3, ncol = 13)
-FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4), 10)
-FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4), 10)
-FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4), 10)
-OuterMar <- c(4, 10, 5, 2)
-InnerMar <- c(1.5, 1.5, 1.5, 1.5)
-TextSize <- 1.5
-AxisSize <- 1.05
-ArrowWidth <- 3
-ArrowLength <- 0.25
-FigWidth <- 8
-FigHeight <- 6
-LocLabels <- seq(-60, 60, by = 24)
-TimeSeq <- 1:200
-TimeLabels <- seq(1800, 2000, by = 40)
-SimSeq <- c(7,8,9,4,5,6,1,2,3)
-ColKeyWidth <- 15
-xLabLine <- 1.75
-yLabLine <- 1.5
-GradLabLine <- 3
-GradSubLine <- 1.5
-AdaptLabLine <- 7
-AdaptSubLine <- 5
-TopArrow <- matrix(c(0.2, 1.2, 3.5, 1.2), nrow = 2, ncol = 2, byrow = TRUE)
-SideArrow <- matrix(c(-0.45, -2.5, -0.45, 0.8), nrow = 2, ncol = 2, byrow = TRUE)
-LowAdj <- 0.05
-HighAdj <- 0.95
-
-# Create a function to find the indices corresponding to the subset of the a sequence
-#	which is defined as the shortest possible sequence including both the given
-#	minimum and maximum
-FindRange <- function(minimum, maximum, sequence){
-     if(class(sequence) != 'numeric'){
-          sequence <- as.numeric(sequence)
-     }
-     LessThan <- ifelse(sequence <= minimum, 1, 0)
-     MoreThan <- ifelse(sequence >= maximum, 1, 0)
-     MinIndex <- sum(LessThan)
-     MaxIndex <- length(sequence) - sum(MoreThan)
-     return(MinIndex:MaxIndex)
-}
-
[…]

### 2018-07-03 StationaryRange/7_Abunds_f_xt.R

@@ -1,128 +1,0 @@
-# This script will create the graphs from the abundance data.
-
-# Set the working directory
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/StationaryRange/")
-
-# Make a function to calculate f(x) over a given spatial extent
-RangeCapacity <- function(beta, gamma, tau, xSeq){
-     f <- rep(NA, length(xSeq))
-     for(i in 1:length(xSeq)){
-          if(xSeq[i] > beta){
-               numerator <- exp(-1*gamma * (xSeq[i] - beta - tau))
-               denominator <- 1 + exp(-1*gamma * (xSeq[i] - beta - tau))
-               f[i] <- numerator / denominator
-          } else if(xSeq[i] <= beta){
-               numerator <- exp(gamma * (xSeq[i] - beta + tau))
-               denominator <- 1 + exp(gamma * (xSeq[i] - beta + tau))
-               f[i] <- numerator / denominator
-          } 
-     }
-     return(f)
-}
-
-# Set some graphical parameters to use for the subsequent figures
-FigMat <- matrix(NA, nrow = 3, ncol = 13)
-FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4), 10)
-FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4), 10)
-FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4), 10)
-OuterMar <- c(4, 10, 5, 2)
-InnerMar <- c(1.5, 1.5, 1.5, 1.5)
-TextSize <- 1.5
-AxisSize <- 1.05
-ArrowWidth <- 3
-ArrowLength <- 0.25
-FigWidth <- 8
-FigHeight <- 6
-LocLabels <- seq(-75, 75, by = 30)
-maxt <- 200
-SimSeq <- c(7,8,9,4,5,6,1,2,3)
-ColKeyWidth <- 15
-xLabLine <- 1.75
-yLabLine <- 1.5
-GradLabLine <- 3
-GradSubLine <- 1.5
-AdaptLabLine <- 7
-AdaptSubLine <- 5
-TopArrow <- matrix(c(-3000, 1.2, 19000, 1.2), nrow = 2, ncol = 2, byrow = TRUE)
-SideArrow <- matrix(c(-6200, -2.5, -6200, 0.8), nrow = 2, ncol = 2, byrow = TRUE)
-LowAdj <- 0.05
-HighAdj <- 0.95
-
[…]

### 2018-07-23 Manuscript/Main.tex
Updating the final graphing scripts and creating a manuscript draft for Ecology Reports

--- 2018-06-14 Manuscript/Main.tex
@@
-\documentclass[11pt, oneside]{article}
+\documentclass[12pt, oneside]{article}
 \usepackage{geometry}                		
@@
 \usepackage{color}
-
-%\usepackage{figcaps}
-%\usepackage[tablesfirst,nolists]{endfloat}
+\usepackage{mathptmx}
 \usepackage{authblk}
@@
 
-\title{Local adaptation and dispersal evolution interact to drive population response to climate change}
+\title{(1) Local adaptation and dispersal evolution interact to drive population response to climate change \\ (2) Spatial population structure drives extinction dynamics in climate-induced range shifts \\ (3) Initial spatial structure drives population response to climate change}
 
@@
 
-\author[1]{Christopher Weiss-Lehman}
+\author[1,*]{Christopher Weiss-Lehman}
 \author[1]{Allison K. Shaw}
@@
 \affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}
+\affil[*]{Denotes corresponding author}
 
@@
 
+\begin{flushleft}
+Corresponding author details: \\
+Christopher Weiss-Lehman \\
+email: cweissle@umn.edu \\
+phone: +1 (720) 590 2278 \\
+\end{flushleft}
+
 \doublespacing
@@
 
-\section{Introduction}
-Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global, penuelas2003global, hansen2001global, scholze2006climate}.
+\newpage
+
+\section*{Abstract}
+Anthropogenic climate change has already caused some species' ranges to shift upwards in latitude and elevation, making it imperative to understand the ecological and evolutionary dynamics behind such range shifts.
+While insights can be gained from the related process of range expansion, climate-induced range shifts have several key attributes differentiating them from range expansions.
+In particular, range shifting populations are characterized by some initial spatial population structure defined by factors such as local adaptation and the nature of the range edge.
+Here, we use an individual-based model to explore the importance of spatial population structure for range shifting populations.
+We show that populations with more pronounced spatial structure have substantially increased extinction probabilities.
+Further, in our simulations evolution of dispersal ability was insufficient to rescue faltering populations.
+Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
+Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that the initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.
+
+\begin{flushleft}
+\textbf{Key words:} range shifts, eco-evolutionary dynamics, local adaptation, individual-based model
+\end{flushleft}
+
+\newpage
+
+\section*{Introduction}
+Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}.
 These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
-In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, walther2002ecological, parmesan2003globally, parmesan2006ecological, parmesan1999poleward, perry2005climate}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in both the extinction of populations failing to track a changing climate~\citep{davis2001range, parmesan2006ecological, zhu2012failure, sekercioglu2008climate}, and the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel, gilman2010framework, williams2007novel}, or both.
+In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, parmesan2006ecological}.
+Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel}, or both.
 It is therefore crucial to understand the dynamics of such climate-induced range shifts to better inform current and future conservation work.
@@
 While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion.
-Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{skellam1951random, hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, burton2010trade, excoffier2009genetic, kubisch2014and} mechanisms responsible for shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, phillips2015evolutionary, shaw2015dispersal}.
+Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms responsible for shaping such expansions.
+For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal}.
 Since range shifts, like range expansions, have a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}.
@@
 For example, the mechanism responsible for the gradient in population size from core to edge (e.g.
-declines in birth rates vs.
-increases in extinction risk) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
-A population's risk of extinction during a range shift has also been related to local adaptation within the range.
+declines in per capita growth rate vs.
+increases in local extinction probability) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
+A population's risk of extinction during a range shift has also been related to the degree of local adaptation within the range.
 Specifically, a broad environmental niche (i.e.
-little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted phenotypes~\citep{atkins2010local}.
+little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted genotypes~\citep{atkins2010local}.
 Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
-However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and fitness are directly linked via gene flow.
+However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
 For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
@@
 Additionally, the distribution of dispersal phenotypes within a population can be influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts.
-Evolution during range shifts due to local adaptation and dispersal evolution are both potentially important drivers of range shift dynamics~\citep{van2016spatial} and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
+Evolution during range shifts due to local adaptation and dispersal evolution are each potentially important drivers of range shift dynamics and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
 
-Here, we develop a complex, individual-based model of a sexually reproducing species, which incorporates evolution in both fitness (via local adaptation) and dispersal distance to determine the interaction between the strength of local selection and the starkness of the gradient at the range edge in driving population dynamics during climate change.
-Using this model, we simultaneously vary the strength of local selection and the starkness of the range boundary to ascertain the relationship between them and test how dispersal evolution interacts with local adaptation to help or hinder a population's ability to track a changing climate.
-We additionally compare the dynamics of successful versus doomed simulated populations to understand the factors contributing to population extinction during climate-induced range shifts.
+Here, we develop a two-sex individual-based model with two genetically determined traits, one defining an individual's expected dispersal distance and the other determining an individual's environmental niche.
+Using this model, we vary the degree of spatial population structure defining the stable range to ascertain how this impacts a population's ability to track a changing climate.
+Further, we contrast the dynamics of extant and extinct populations to understand the joint impact of dispersal evolution and local adaptation on extinction risk during climate-induced range shifts.
 
-\section{Methods}
-\subsection{Overview}
-\subsubsection{Purpose}
-This model tests an evolving population's ability to track a changing climate under a variety of conditions.
-Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the strength of local selection.
-In all simulations, individual dispersal and relative fitness are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
-All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change.
-Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see the \textit{Submodels} section below).
-Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.
+\section*{Methods}
+A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in the online supplemental materials, but we present a brief summary here.
+Each simulated landscapes consisted of a discrete lattice of habitat patches with a fixed width in the $y$ dimension and unbounded length in the $x$ dimension.
+Environmental conditions varied along the $x$ dimension and remained constant across the $y$ dimension (Fig.
+S1).
+To simulate climate change, environmental conditions shifted a constant amount in the $x$ dimension each generation.
+Competition and reproduction occurred within each patch and local populations were linked via dispersal, assuming wrapping boundaries in the $y$ dimension to prevent edge effects.
 
-\subsubsection{State variables and scales}
-The model simulates a population of male and female individuals characterized by spatial coordinates for their location and by diploid loci for both fitness and dispersal.
-Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the y axis and without bounds on the x axis.
-To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same row of the landscape.
-Patches are defined by the location of the patch center in x and y coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (see the \textit{Submodels} section below).
+Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci, assuming no linkage.
+The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
+An individual's environmental niche trait allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's location, the higher the individual's realized fitness.
+The potential for local adaptation was then manipulated by varying the degree to which the environmental optimum changed throughout the landscape.
+In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see supplemental materials; Fig.
+S1).
+Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
+Using this model, we simulated non-overlapping generations consisting of discrete dispersal and reproduction phases (Fig.
+S2) to determine the effect of different aspects of spatial population structure on the eco-evolutionary dynamics of climate-induced range shifts.
 
-The abiotic environment is defined by the location of the center of the range, the slope of the range boundary, and the width of the range (See Figure~\ref{fig:EnvFunction}).
-The range can then shift in space by altering the location of the range center, which is how climate change is implemented in the model.
-Further, a gradient of phenotypic optimum values for fitness is imposed within the range to allow for local adaptation within the range boundaries.
-The severity of this gradient is defined independently of the other environmental parameters.
+Most model parameters were held constant over all simulations (Table S1), but we varied certain parameter values to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S2).
+Specifically, we considered a factorial combination of no, low, and high potential for local adaptation with shallow, moderate, and stark gradients in suitable habitat at the range edge for a total of $9$ different scenarios, each explored with $200$ simulations.
+Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
+Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change.
+Additionally, we explored the role of the rate of climate change on the dynamics of range shifts in these different scenarios, examining both slower and faster speeds of climate change (see the online supplemental materials).
 
+For each parameter combination, we calculated the per patch change in mean dispersal phenotype from the beginning of the period of climate change to the end to quantify dispersal evolution.
+For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates.
+In this way, we calculated change in mean dispersal phenotype for populations consistently occupying the same position relative to the range center (e.g.
+leading edge vs.
+core populations).
+Due to local extinctions, not all patches were occupied at the end of the period of climate change.
+To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals as the end point to calculate change in the mean dispersal phenotype.
+Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions) so that positive values indicate an increase in the mean dispersal phenotype.
+All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
+
+\section*{Results}
+For almost all parameter combinations examined here, a portion of the simulated populations were able to keep pace with changing climate conditions.
+Successful populations shifted their spatial distributions at essentially the same rate as climate change (e.g.
+Fig.
+1).
+However, some populations failed to keep pace, lagging further behind the shifting climate until they went extinct.
+Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
+2).
+While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction dynamics.
+The parameters defining both of these range attributes were varied widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table S2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space.
+Additionally, the pace of climate change also influenced extinction probabilities as expected with faster climate change corresponding to greater extinction risk (Fig.
+S3 \& S4).
+However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability for a range shifting population.
+
+Dispersal evolution is expected to aid populations shifting spatially to track a changing climate, and, as expected, the average dispersal phenotypes increased through time in some simulations (e.g.
+Fig.~\ref{fig:SimExample}).
+However, comparing patterns in dispersal evolution between simulated populations that successfully tracked climate change and those that ultimately went extinct revealed no differences in the magnitude or direction of dispersal evolution (Fig.
+3a\&b).
+Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes strongly centered on $0$ (Fig.
+S5-S7).
+The similarity in evolved changes to dispersal between surviving and extinct simulated populations suggests that this cannot explain the different outcomes achieved by these populations.
+
+Instead, the initial distribution of dispersal phenotypes immediately before the beginning of climate change played a large role in determining the ultimate fate of simulated populations.
+Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
+S8-S10).
+The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig.
+3c\&d).
+In fact, comparing dispersal phenotypes of successful populations across different parameter combinations revealed a threshold in dispersal phenotypes separating individuals from successful and extinct populations (Fig.
+S8-S10).
+This threshold varies directly with the speed of climate change, increasing with more rapid climate change.
+Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e.
+moving the correct number of patches in the $x$ dimension to track the changing environmental conditions).
+
+Intriguingly, the simulations that survived climate change tended to be characterized by reduced fitness at the range margins compared to simulated populations that went extinct (Fig.
+S11-S13).
+This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
+Intuitively, there was no spatial pattern in fitness for simulations with no potential for local adaptation.
+As the simulations that survived climate change were also characterized by heightened dispersal (Fig.
+3), the observed reduction in average patch fitness at the margins is likely due to increased gene flow from the range core hampering the abilities of these edge populations to adapt to local conditions.
+
+\section*{Discussion}
+Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
+Here we developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
+In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g.
+single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
+Thus, in analyzing our model, we tested the generality of previous results under realistic levels of biological complexity.
+Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
+
+Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the range and in locations with shallow environmental gradients defining the range edge (Fig.
+2).
+A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies testing for it, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
+Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study.
+While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to the spatial population structure of their initial ranges.
+
+A key result of our model was the importance of the initial distribution of dispersal phenotypes (i.e.
+prior to the start of climate change) in determining a population's extinction risk during climate change (Fig.
+3c\&d).
+Survival in the face of climate change was primarily determined by whether or not simulated populations consisted of individuals with dispersal phenotypes at or above some threshold value.
+Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations at a given speed of climate change (Fig.
+S8-S10).
+The differences in extinction probabilities among different scenarios were instead driven by the overall distribution of dispersal phenotypes that evolved in different simulations under each parameter combination.
+Scenarios with no potential for local adaptation and gradual environmental gradients simply had larger proportions of high dispersal phenotypes under stable climate conditions.
+A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}.
+Similarly, a more severe habitat gradient at the range edge increases the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}.
+Thus, attributes defining the spatial population structure of simulated ranges resulted in different initial distributions of traits under stable climate conditions, which then determined the extinction risk of populations under climate change.
+
+While high dispersal phenotypes prior to climate change allowed populations to survive, it had the additional effect of reducing average fitness at the range margins in scenarios with the potential for local adaptation (Fig.
+S11-S13).
+In the model, populations at the range margins tended to be lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
+Thus, when populations were defined by high dispersal phenotypes prior to climate change, it reduced fitness at the range margin due to gene flow preventing adaptation to local conditions.
+As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the start of climate change.
+While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, and thus are unlikely to be equally susceptible to the influence of gene flow from the core, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
+
+Previous research has suggested that evolution of increased dispersal ability during climate change may be able to rescue populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}.
+Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
+Models finding that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid and single-locus models~\citep{boeye2013more, hargreaves2015fitness}.
+As dispersal evolution during range expansions and shifts occurs via spatial sorting~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of highly dispersive alleles.
+Additionally, many previous models of climate-induced range shifts use asexual reproduction which can result in more rapid rates of evolutionary change under certain conditions~\citep{crow1965evolution, smith1968evolution}.
+The negligible role played by dispersal evolution in our model (Fig.
+3a\&b) suggests that when such simplifying assumptions are relaxed the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
+
+\subsection*{Conclusion}
+Understanding the various ecological and evolutionary drivers of climate-induced range shifts is crucial to current and future conservation efforts.
+In particular, we need to better understand how these various mechanisms combine to shape the extinction probabilities of populations undergoing range shifts, which, in turn, will allow more focused interventions.
+Our results suggest that initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
+Further, in contrast to other studies assuming more simplified breeding and genetic structure, we find very little role for the evolution of heightened dispersal abilities to rescue a population composed of low-dispersing individuals.
+Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue.
+As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such increased risk.
+
+\section*{Acknowledgements}
+We thank Lauren Shoemaker, Lauren Sullivan, Kate Meyers, Dongmin Kim, Lindsay Leverett, and members of the Theory Under Construction group at the University of Minnesota for providing thoughtful comments on the manuscript.
+
+\bibliographystyle{ecology}
+\bibliography{main_bib}
+
+\newpage
+
+\section*{Figure legends}
+\paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
+Information on the abundance, dispersal, and fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
+Log transformed mean dispersal phenotypes are shown for each patch.
+Average patch fitness was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.
+
+\paragraph{Figure 2.} The cumulative probability of extinction due to climate change in different experimental scenarios.
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.
+
+\paragraph{Figure 3.} Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations.
+Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end.
+Positive values indicate an increase in average dispersal ability in the patch.
+Initial distributions of the dispersal trait (c and d) are shown with histograms of the log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
+In all panels, values associated with extant populations are shown in dark blue.
+Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations).
+Full results for all parameter combinations are provided in the online supplement.
+
+\newpage
+
+\large{\textbf{Figure 1}}
 \begin{figure}
 \centering
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
 \vspace{-5mm}
-\caption[LoF entry]{Environmental quality, as defined by $f(x,t)$, over a spatial gradient.
-The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
-Additionally, the lattice of discrete patches in which population dynamics occur is shown beneath.
-This lattice of $\eta$ x $\eta$ patches are mapped to environmental conditions along the $x$ axis using $f(x,t)$ and are environmentally constant along the $y$ axis.
-Dispersal is unbounded in the $x$ direction and implemented with wrapping boundaries in the $y$ direction.}
-\label{fig:EnvFunction}
+\label{fig:SimExample}
 \end{figure}
 
-\subsubsection{Process overview and scheduling}
-Time is also modeled in discrete intervals defining single generations of the population.
-Within each generation, individuals first disperse from their natal patches according to their dispersal phenotypes (\textit{Submodels}).
-After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch (\textit{Submodels}).
-Parental pairs form via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with high fitness are likely to produce multiple offspring while individuals with no fitness may not produce any.
-Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process (\textit{Submodels}).
-After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations.
+\clearpage
 
+\large{\textbf{Figure 2}}
 \begin{figure}
 \centering
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
 \vspace{-5mm}
-\caption[LoF entry]{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
-Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice.
-After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
-\label{fig:LifeCycle}
+\label{fig:ExtProb}
 \end{figure}
 
-\subsection{Design concepts}
-\subsubsection{Emergence}
-Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both fitness and dispersal traits during climate change.
-These are all examined in the context of their impact on the population's ability to keep pace with the changing climate.
+\clearpage
 
-\subsubsection{Stochasticity}
-All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci.
-Environmental parameters are fixed, however, and the process of climate change (i.e.
-the movement of environmentally suitable habitat through time) is deterministic.
-Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.
-
-\subsubsection{Interactions}
-Individuals in the model interact via mating and density-dependent competition within patches.
-Other important interactions are the relationship between dispersal evolution and local adaption, particularly in edge populations, and how this relationship impacts a population's ability to avoid extinction and track a changing climate.
-These interactions are examined in the context of different configurations of the initial, stable range and different speeds of climate change.
-
-\subsubsection{Desired output}
-After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits).
-If a population went extinct during the model run, the time of extinction is recorded.
-Throughout the simulation, certain aggregated values are calculated and recorded for each occupied patch including the population size and the mean and variance of allele values for each trait.
-
-\subsection{Details}
-\subsubsection{Initialization}
-The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below.
-Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions.
-This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation.
-See Table 1 for a full list of all parameter values used in the simulations described here.
-
-\subsubsection{Submodels}
-\paragraph{Environmentally suitable habitat}
-Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$).
-The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center.
-Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows
-\begin{equation}
-f(x,t)=
-\begin{cases}
-	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
-	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
-\end{cases}
-\end{equation}
-where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}).
-To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
-\begin{equation}
-f(x,t)=
-\begin{cases}
-	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
-	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
-\end{cases}	
-\end{equation}
-yielding a derivative of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).
-
-Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure~\ref{fig:EnvFunction}).
-The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
-\begin{equation}
-K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
-\end{equation}
-
-By varying the parameters defining $f(x,t)$, we can change both the total achievable carrying capacity of the population throughout the range (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$).
-Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total achievable carrying capacity of the population.
-To avoid this confounding factor, we fix the total area under the curve $f(x,t)$.
-The indefinite integral of $f(x,t)$ can be shown to be
-\begin{equation}
-\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
-\end{equation}
-which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.
-
-Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change.
-During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
-\begin{equation}
-\beta_{t}=\nu\eta(t-\hat{t})
-\end{equation}
-where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $\eta$ is the spatial scale of each patch, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.
-
-\paragraph{Local adaptation}
-To allow an arbitrary degree of local adaptation within the range, the local phenotypic optima ($z_{opt,x}$) is set as follows
-\begin{equation}
-z_{opt,x}=\lambda(x-\beta_{t})
-\end{equation}
-where $\lambda$ defines the strength of local selection with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range.
-Individual fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
-\begin{equation}
-w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
-\end{equation}
-where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's fitness phenotype~\citep{lande1976natural}.
-All loci are assumed to contribute additively to the phenotype with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values for the fitness trait.
-
-\paragraph{Population dynamics}
-Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$).
-The expected number of new offspring in patch $x$ at time $t+1$ is then given by
-\begin{equation}
-\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
-\end{equation}
-where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions.
-To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
-\begin{equation}
-N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
-\end{equation}
-
-Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e.
-polygynandrous mating).
-The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any.
-Each offspring inherits one allele per locus from each parent, assuming no linkage among loci.
-After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.
-
-\paragraph{Mutation}
-Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents.
-The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$).
-Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation is
-\begin{equation}
-\frac{U^{T}}{2L^{T}}
-\end{equation}
-Mutational effects are drawn from a normal distribution with mean $0$ and a standard deviation of
-\begin{equation}
-\sqrt{V_{m}^{T}U^{T}}
-\end{equation}
-By defining the mutation process in this manner rather than setting a probability of mutation and mutational effect directly, similar mutational dynamics can be imposed regardless of the number of loci used in the simulation.
-
-\paragraph{Dispersal}
-Finally, individuals disperse according to an exponential dispersal kernel defined by each individual's dispersal phenotype.
-An individual's dispersal phenotype is the expected dispersal distance and is given by
-\begin{equation}
-d_{i} = \frac{D\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
-\end{equation}
-where $D$ is the maximum expected dispersal distance in terms of discrete patches, $\eta$ is the spatial scale of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $D$, and the summation is taken across all alleles contributing to dispersal.
-Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis.
-The expected dispersal distance, $d_{i}$ is then used to draw a realized distance from an exponential dispersal kernel.
-The direction of dispersal is drawn from a uniform distribution bounded by $0$ and $2\pi$.
-If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ axis, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ axis, thus wrapping the top and bottom edges of the landscape to avoid edge effects.
-Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure~\ref{fig:EnvFunction}).
-
-\begin{table}
-\renewcommand{\arraystretch}{1.5}
-  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
-    \hline
-    Parameter & Description & Value \\ \hline \hline
-    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ \\
-    $\beta_{1}$ & Center of environmentally suitable habitat before climate change & $0$ \\
-    $\eta$ & Spatial dimensions of habitat patches in continuous space & $50$ \\
-    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ direction & $10$ \\
-    $\hat{t}$ & Last time point of stable climate conditions & $2000$ \\
-    $t_{\Delta}$ & Duration of climate change & $100$ \\
-    $t_{max}$ & Total number of time points in the simulation & $2150$ \\
-    $R$ & Intrinsic growth rate of the population & $2$ \\
-    $K_{max}$ & Maximum achievable carrying capacity in the range & $100$ \\ 
-    $\psi$ & Expected sex ratio in the population & $0.5$ \\
-    $D$ & Maximum achievable dispersal phenotype & $1000$ \\
-    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
-    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
-    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ for each trait \\
-    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ for each trait \\
-    $L^{T}$ & The number of diploid loci defining each trait & $5$ for each trait \\
-    $\mu_{1}^{f}$ & The initial mean allele value for the fitness trait & $0$ \\
-    $\mu_{1}^{d}$ & The initial mean allele value for the dispersal trait & $-1$ \\
-    $\sigma_{1}^{f}$ & The initial standard deviation of allele values for fitness & $0.025$ \\
-    $\sigma_{1}^{d}$ & The initial standard deviation of allele values for dispersal & $1$ \\
-    \hline
-  \end{tabular}
-\caption[LoF entry]{Simulation parameters held constant across all scenarios presented here.}
-\label{table:ConstPars}
-\end{table}
-
-\begin{table}
-\renewcommand{\arraystretch}{1.5}
-  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
-    \hline
-    Habitat gradient at range edge & Strength of local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
-     & None & $0.0025$ & $250$ & $0$ \\
-    Shallow & Moderate & $0.0025$ & $250$ & $0.004$ \\
-     & Strong & $0.0025$ & $250$ & $0.008$ \\ \hline
-     & None & $0.025$ & $421.479$ & $0$ \\
-    Moderate & Moderate & $0.025$ & $421.479$ & $0.004$ \\
-     & Strong & $0.025$ & $421.479$ & $0.008$ \\ \hline
-     & None & $0.25$ & $421.48$ & $0$ \\
-    Stark & Moderate & $0.25$ & $421.48$ & $0.004$ \\
-     & Strong & $0.25$ & $421.48$ & $0.008$ \\ 
-    \hline
-  \end{tabular}
-\caption[LoF entry]{Descriptions and parameter values for the $9$ different scenarios examined here. For each scenario, $\nu$ (the speed of climate change) was varied among $0.5$, $1$, and $2$.}
-\label{table:Scenarios}
-\end{table}
-
-\subsection{Simulation descriptions}
-Most of the parameters described above were held constant over all simulations we performed (Table ~\ref{table:ConstPars}).
-We varied certain parameter values to explore the interacting roles of local adaptation, the habitat gradient at the range edge, and the speed of climate change in the dynamics of range shifts.
-Specifically, we considered a factorial combination of three severities of habitat quality decline at the range edge, three degrees of local adaptation within the range, and three speeds of climate change (Table~\ref{table:Scenarios}) for a total of $27$ different model scenarios.
-For each scenario, we performed $100$ simulations to assess the interacting ecological and evolutionary dynamics of climate driven range shifts.
-All simulations were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
-
-\bibliographystyle{amnatnat}
-\bibliography{main_bib}
+\large{\textbf{Figure 3}}
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
+\vspace{-5mm}
+\label{fig:Disp}
+\end{figure}
 

### 2018-07-23 Manuscript/Supp.tex
Updating the final graphing scripts and creating a manuscript draft for Ecology Reports

\documentclass[12pt, oneside]{article}
\usepackage{geometry}                		
\geometry{letterpaper}                   		
\usepackage{graphicx}
\usepackage{graphics}
\usepackage{hyperref}
\usepackage{fancybox}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}			
\usepackage{natbib}				
\usepackage{fullpage}
\usepackage{placeins}
\usepackage{setspace}
\usepackage{lineno}
\usepackage{color}
\usepackage{authblk}
\usepackage{mathptmx}

\DeclareRobustCommand{\firstsecond}[2]{#1}

\newcommand{\mb}{\mathbf}
\newcommand{\bs}{\boldsymbol}
\newcommand{\wt}{\widetilde}
\newcommand{\s}{^{(s)}}

\title{Supplementary materials: Local adaptation and dispersal evolution interact to drive population response to climate change}

\date{}

\author[1]{Christopher Weiss-Lehman}
\author[1]{Allison K. Shaw}

\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}

\begin{document}
\maketitle

\doublespacing
\linenumbers

\renewcommand{\thefigure}{S\arabic{figure}}
\renewcommand{\thetable}{S\arabic{table}}
\renewcommand{\theequation}{S\arabic{equation}}

%% Make sure I get the table numbering to also begin with an S like above for the figures. Also, think through the section structure and labeling here.
\section*{Model overview}
\subsection*{Purpose} 
This model tests an evolving population's ability to track a changing climate under a variety of conditions. Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the potential for local adaptation. In all simulations, an individual's expected dispersal distance and environmental niche are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time. All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change. Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see \textit{Submodels} below). Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.

\subsection*{State variables and scales} 
The model simulates a population of males and females characterized by spatial coordinates for their location and by diploid loci for both their expected dispersal distance and environmental niche. Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system with a fixed width along the $y$ axis and without bounds on the $x$ axis. To avoid edge effects along the y axis, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same column of the landscape. Patches are defined by the location of the patch center in $x$ and $y$ coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (\textit{Submodels}).

The model implements climate change by shifting the location of a population's available habitat along the $x$ axis of the landscape. Available habitat (i.e. a population's potential range) is defined by a center location on the $x$ axis, the severity of the decline in habitat quality at the edge, and the width of the available habitat along the $x$ axis (See Figure~\ref{fig:EnvFunction}). Further, a gradient in environmental conditions is imposed throughout the landscape to allow for local adaptation via matching of an individual's environmental niche to the local environmental conditions (\textit{Submodels}). The severity of this gradient can be altered to change the potential for local adaptation (e.g. a shallower gradient will result in more similar environmental conditions throughout the range and therefore reduce the potential for local adaptation).

\subsection*{Process overview and scheduling} 
Time is modeled in discrete intervals defining single generations of the population (Fig. S1). Within each generation, individuals first disperse from their natal patches according to their dispersal phenotypes. After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch. Reproduction occurs via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) are likely to produce multiple offspring while individuals with low relative fitness may not produce any. Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process. After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations. See \textit{Submodels} in the online supplemental materials for details.

\section*{Design concepts}
\subsection*{Emergence} 
Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both expected dispersal distances and environmental niches during climate change.

\subsection*{Stochasticity} 
All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci. Environmental parameters are fixed, however, and the process of climate change (i.e. the movement of environmentally suitable habitat through time) is deterministic. Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.

\subsection*{Interactions} 
Individuals in the model interact via mating and density-dependent competition within patches. Other important interactions are the relationship between dispersal evolution and local adaption, particularly in edge populations, and how this relationship impacts a population's ability to avoid extinction and track a changing climate.

\subsection*{Desired output} 
After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits). If a population went extinct during the model run, the time of extinction is recorded. For each occupied patch throughout the simulation, we aggregated and reported data on population size, the dispersal trait, and local adaptation to environmental conditions.

\section*{Details}
\subsection*{Initialization} 
The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below. Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions. This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation. See Tables S1 \& S2 for a full list of parameter values used in the simulations described here.

\subsection*{Submodels}
\paragraph{Environmentally suitable habitat}
Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$). The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center. Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows 
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
\end{cases}
\end{equation}
where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}). To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the spatial dimension can be shown to be
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
\end{cases}	
\end{equation}
yielding a derivative of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).

Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure~\ref{fig:EnvFunction}). The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
\begin{equation}
K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
\end{equation}

By varying the parameters defining $f(x,t)$, we can change both the total achievable carrying capacity of the population throughout the range (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$). Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total achievable carrying capacity of the population. To avoid this confounding factor, we fix the total area under the curve $f(x,t)$. The indefinite integral of $f(x,t)$ can be shown to be
\begin{equation}
\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
\end{equation}
which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.

Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change. During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
\begin{equation}
\beta_{t}=\nu\eta(t-\hat{t})
\end{equation}
where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $\eta$ is the spatial scale of each patch, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.

\paragraph{Local adaptation}
To allow an arbitrary degree of local adaptation within the range, the local environmental conditions determining the phenotypic optima ($z_{opt,x}$) are set as follows
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
where $\lambda$ defines the potential for local adaptation with values close to $0$ resulting in little to no change in phenotypic optimum across the range and values of greater magnitude resulting in large differences in phenotypic optima across the range. Individual relative fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's environmental niche~\citep{lande1976natural}. Thus, an individual's realized fitness will be higher the closer its environmental niche ($z_[i]$) is to the local environmental conditions of the patch it occupies ($z_{opt,x}$). All loci are assumed to contribute additively to an individual's environmental niche value with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values.

\paragraph{Population dynamics}
Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$). The expected number of new offspring in patch $x$ at time $t+1$ is then given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e. polygynandrous mating). The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any. Each offspring inherits one allele per locus from each parent, assuming no linkage among loci. After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.

\paragraph{Mutation}
Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents. The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$). Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation is
\begin{equation}
\frac{U^{T}}{2L^{T}}
\end{equation}
Mutational effects are drawn from a normal distribution with mean $0$ and a standard deviation of
\begin{equation}
\sqrt{V_{m}^{T}U^{T}}
\end{equation}
By defining the mutation process in this manner rather than setting a probability of mutation and mutational effect directly, similar mutational dynamics can be imposed regardless of the number of loci used in the simulation.

\paragraph{Dispersal}
Finally, individuals disperse according to an exponential dispersal kernel defined by each individual's dispersal phenotype. An individual's dispersal phenotype is the expected dispersal distance and is given by
\begin{equation}
d_{i} = \frac{D\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
\end{equation}
where $D$ is the maximum expected dispersal distance in terms of discrete patches, $\eta$ is the spatial scale of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $D$, and the summation is taken across all alleles contributing to dispersal. Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis. The expected dispersal distance, $d_{i}$ is then used to draw a realized distance from an exponential dispersal kernel. The direction of dispersal is drawn from a uniform distribution bounded by $0$ and $2\pi$. If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ axis, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ axis, thus wrapping the top and bottom edges of the landscape to avoid edge effects. Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure~\ref{fig:EnvFunction}).

\bibliographystyle{ecology}
\bibliography{supp_bib}

\newpage

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ \\
    $\beta_{1}$ & Center of environmentally suitable habitat before climate change & $0$ \\
    $\eta$ & Spatial dimensions of habitat patches in continuous space & $50$ \\
    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ direction & $10$ \\
    $\hat{t}$ & Last time point of stable climate conditions & $2000$ \\
    $t_{\Delta}$ & Duration of climate change & $100$ \\
    $t_{max}$ & Total number of time points in the simulation & $2150$ \\
    $R$ & Intrinsic growth rate of the population & $2$ \\
    $K_{max}$ & Maximum achievable carrying capacity in the range & $100$ \\ 
    $\psi$ & Expected sex ratio in the population & $0.5$ \\
    $D$ & Maximum achievable dispersal phenotype & $1000$ \\
    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ for each trait \\
    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ for each trait \\
    $L^{T}$ & Number of diploid loci defining each trait & $5$ for each trait \\
    $\mu_{1}^{f}$ & Initial mean allele value for the environmental niche trait & $0$ \\
    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the environmental niche trait & $0.025$ \\
    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Simulation parameters held constant across all scenarios.}
\label{table:ConstPars}
\end{table}

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
    \hline
    Habitat gradient at range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
     & None & $0.0025$ & $250$ & $0$ \\
    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
     & None & $0.025$ & $421.479$ & $0$ \\
    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
     & None & $0.25$ & $421.48$ & $0$ \\
    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
     & High & $0.25$ & $421.48$ & $0.008$ \\ 
    \hline
  \end{tabular}
\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.}
\label{table:Scenarios}
\end{table}

\clearpage

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
\vspace{-5mm}
\caption[LoF entry]{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space. The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis. The lattice of discrete patches in which population dynamics occur is shown beneath. As described in the \textit{Submodels} section of the supplemental materials, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches. Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension. Dispersal is unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
\label{fig:EnvFunction}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
\vspace{-5mm}
\caption[LoF entry]{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal. Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice. After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
\label{fig:LifeCycle}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
\vspace{-5mm}
\caption[LoF entry]{The cumulative probability of extinction due to climate change in different experimental scenarios under a slow speed of climate change. Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
\label{fig:ExtProbSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
\vspace{-5mm}
\caption[LoF entry]{The cumulative probability of extinction due to climate change in different experimental scenarios under a fast speed of climate change. Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
\label{fig:ExtProbFast}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the change in the mean dispersal phenotypes in each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Each patch was defined by its location relative to the range center ($\beta_{t}$) so changes in dispersal correspond to relative (e.g. leading and trailing edge patches), as opposed to absolute (i.e. fixed spatial coordinates), locations in the range.  For those simulated populations that went extinct during climate change, we calculated the change in dispersal phenotype using the last time point with at least $10$ individuals in the patch as described in the text. The values associated with populations that survived climate change are shown in dark blue. Each histogram represents one of the $9$ experimental scenarios examined here, with potential for local adaptation increasing from none to high from bottom to top and the severity of the gradient at the range edge increasing from gradual to severe from left to right as indicated on the figure. Results are shown for a slow speed of climate change.}
\label{fig:DispEvolSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the change in the mean dispersal phenotypes in each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Each patch was defined by its location relative to the range center ($\beta_{t}$) so changes in dispersal correspond to relative (e.g. leading and trailing edge patches), as opposed to absolute (i.e. fixed spatial coordinates), locations in the range.  For those simulated populations that went extinct during climate change, we calculated the change in dispersal phenotype using the last time point with at least $10$ individuals in the patch as described in the text. The values associated with populations that survived climate change are shown in dark blue. Each histogram represents one of the $9$ experimental scenarios examined here, with potential for local adaptation increasing from none to high from bottom to top and the severity of the gradient at the range edge increasing from gradual to severe from left to right as indicated on the figure. Results are shown for a moderate speed of climate change.}
\label{fig:DispEvolMain}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the change in the mean dispersal phenotypes in each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Each patch was defined by its location relative to the range center ($\beta_{t}$) so changes in dispersal correspond to relative (e.g. leading and trailing edge patches), as opposed to absolute (i.e. fixed spatial coordinates), locations in the range.  For those simulated populations that went extinct during climate change, we calculated the change in dispersal phenotype using the last time point with at least $10$ individuals in the patch as described in the text. The values associated with populations that survived climate change are shown in dark blue. Each histogram represents one of the $9$ experimental scenarios examined here, with potential for local adaptation increasing from none to high from bottom to top and the severity of the gradient at the range edge increasing from gradual to severe from left to right as indicated on the figure. Results are shown for a fast speed of climate change.}
\label{fig:DispEvolFast}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a slow speed of climate change.}
\label{fig:InitDispSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a moderate speed of climate change.}
\label{fig:InitDispMain}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
\vspace{-5mm}
\caption[LoF entry]{Histograms of the log transformed dispersal phenotypes making up populations after $2000$ generations of stable climate conditions. On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner. Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure. Extinct and extant simulations are shown for a fast speed of climate change.}
\label{fig:InitDispFast}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
\vspace{-5mm}
\caption[LoF entry]{Mean individual fitness along the $x$ dimensions of the landscape after $2000$ generations of stable climate conditions for both extant (those populations that ultimately survived climate change) and extinct populations. Points represent the mean value across simulations and error bars are interquartile ranges. The experimental scenario corresponding to each panel is shown on the graph and simulations with no potential for local adaptation are not shown as, by definition, patch fitness does not vary spatially. Extinct and extant simulations are shown for a slow speed of climate change.}
\label{fig:InitFitSlow}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
\caption[LoF entry]{Mean individual fitness along the $x$ dimensions of the landscape after $2000$ generations of stable climate conditions for both extant (those populations that ultimately survived climate change) and extinct populations. Points represent the mean value across simulations and error bars are interquartile ranges. The experimental scenario corresponding to each panel is shown on the graph and simulations with no potential for local adaptation are not shown as, by definition, patch fitness does not vary spatially. Extinct and extant simulations are shown for a moderate speed of climate change.}
\vspace{-5mm}
\label{fig:InitFit}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
\vspace{-5mm}
\caption[LoF entry]{Mean individual fitness along the $x$ dimensions of the landscape after $2000$ generations of stable climate conditions for both extant (those populations that ultimately survived climate change) and extinct populations. Points represent the mean value across simulations and error bars are interquartile ranges. The experimental scenario corresponding to each panel is shown on the graph and simulations with no potential for local adaptation are not shown as, by definition, patch fitness does not vary spatially. Extinct and extant simulations are shown for a fast speed of climate change.}
\label{fig:InitFitFast}
\end{figure}

\end{document}

### 2018-07-23 8a_ExtinctionGraphs.R

@@ -45,7 +45,7 @@ for(i in 1:3){
                 col = Col)
           lines(x = 1:100, y = Extinctions[i,3,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
-          text(x = LetterPos[1], y = LetterPos[2], labels = "(a)", cex = TextSize)
+          text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("a")), cex = TextSize)
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)
@@ -59,7 +59,7 @@ for(i in 1:3){
                 col = Col)
           lines(x = 1:100, y = Extinctions[i,6,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
-          text(x = LetterPos[1], y = LetterPos[2], labels = "(b)", cex = TextSize)
+          text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("b")), cex = TextSize)
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)
@@ -75,7 +75,7 @@ for(i in 1:3){
                 col = Col)
           lines(x = 1:100, y = Extinctions[i,9,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
-          text(x = LetterPos[1], y = LetterPos[2], labels = "(c)", cex = TextSize)
+          text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("c")), cex = TextSize)
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)

### 2018-07-23 8b_DispInitGraphs.R

@@ -1,7 +1,7 @@
 # This script will make an appropriate graph of the initial dispersal values
 setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
 load("SimData/InitDispData.rdata")
-CurSpeed <- 2
+CurSpeed <- 3
 SpeedWords <- c("Slow", "Main", "Fast")
 
 # Sort the disp values into the appropriate lists
@@ -30,6 +30,8 @@ for(p in 1:9){
      xMax <- max(c(xMax, AllSims[[p]]))
 }
 xRange <- c(xMin, xMax)
+ExtantFile <- paste("SimData/", SpeedWords[CurSpeed], "NumExtant.rdata", sep = "")
+save(NumExtant, file = ExtantFile)
 
 # Set some graphical parameters to use for the subsequent figures
 options(scipen = -10)
@@ -37,7 +39,7 @@ FigMat <- matrix(NA, nrow = 3, ncol = 12)
 FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
 FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
 FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
-OuterMar <- c(4, 12, 5, 2)
+OuterMar <- c(4, 12, 6, 2)
 InnerMar <- c(1.5, 2.25, 1.5, 2.25)
 TextSize <- 1.5
 AxisSize <- 1.15
@@ -49,7 +51,7 @@ DispAxisLabel <- expression(paste("Initial ", italic("log"), "(", italic("d"), "
 SimSeq <- c(7,8,9,4,5,6,1,2,3)
 xLabLine <- 2
 yLabLine <- 3
-GradLabLine <- 3
+GradLabLine <- 3.5
 GradSubLine <- 1.5
 AdaptLabLine <- 9
 AdaptSubLine <- 7
@@ -57,8 +59,8 @@ TopArrow <- matrix(c(-0.75, 260000, 17, 260000), nrow = 2, ncol = 2, byrow = TRU
 SideArrow <- matrix(c(-5.75, -550000, -5.75, 175000), nrow = 2, ncol = 2, byrow = TRUE)
 LowAdj <- 0.05
 HighAdj <- 0.95
-AllCol <- "grey50"
-ExtantCol <- "springgreen4"
+AllCol <- "lightskyblue"
+ExtantCol <- "navyblue"
[…]

### 2018-07-23 8d_DispEvolGraphs.R

@@ -1,291 +1,149 @@
 # This script will create the dispersal evolution graphs
-
 setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
-load("DispEvolData.rdata")
+load("SimData/DispEvolData.rdata")
+SpeedWords <- c("Slow", "Main", "Fast")
 
-# Add a new column to the DispEvol data frame for a color variable for distance. 
-#    This variable will scale from black (Dist = 100) to white (Dist = 0). Also,
-#    make a variable for an extant variable (to see if it makes sense)
-DispEvol$ExtantCol <- ifelse(DispEvol$Dist == max(DispEvol$Dist), "darkblue", "darkred")
+# Create objects to hold coordinates for the various arrows to use in the graph
+TopArrow <- array(NA, dim = c(3, 2, 2))
+TopArrow[1,,] <- c(-75, 600, 16000, 16000)
+TopArrow[2,,] <- c(-75, 600, 18000, 18000)
+TopArrow[3,,] <- c(-75, 600, 16000, 16000)
+SideArrow <- array(NA, dim = c(3, 2, 2))
+SideArrow[1,,] <- c(-193, -193, -34000, 10000)
+SideArrow[2,,] <- c(-193, -193, -38000, 12000)
+SideArrow[3,,] <- c(-193, -193, -34000, 10000)
 
-# Calculate the Dispersal range for the graph axes
-DispRange <- range(c(DispEvol$InitDisp, DispEvol$FinalDisp))
-LogRange <- c(-1.5, 3.5)
-
-# Set some graphical parameters to use for the subsequent figures
+# Set some graphical parameters for the figures
+FigWidth <- 8
+FigHeight <- 6
 FigMat <- matrix(NA, nrow = 3, ncol = 12)
 FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
 FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
 FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
-OuterMar <- c(4, 10, 5, 2)
+OuterMar <- c(4, 10, 6, 2)
 InnerMar <- c(1.5, 1.5, 1.5, 1.5)
-TextSize <- 1.5
+SimSeq <- c(7,8,9,4,5,6,1,2,3)
+xRange <- c(-100, 100)
[…]

### 2018-07-23 8e_DispComposite.R

@@ -1,0 +1,102 @@
+# This script will create the simplified composite graph incorporating dispersal
+#    evolution and initial dispersal
+
+setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
+load("SimData/InitDispData.rdata")
+load("SimData/DispEvolData.rdata")
+CurSpeed <- 2
+Params <- c(1, 9)
+
+AllSims <- vector(length = 2, mode = "list")
+ExtantSims <- vector(length = 2, mode = "list")
+NumExtant <- rep(0, 2)
+NumSims <- 200
+xMin <- 0
+xMax <- 0
+for(p in 1:2){
+     CurParam <- Params[p]
+     AllVals <- NULL
+     ExtantVals <- NULL
+     for(i in 1:NumSims){
+          AllVals <- c(AllVals, DispInit[[CurParam]][[i]]$ExpDists)
+          
+          if(DispInit[[CurParam]][[i]]$Ext[CurSpeed] == 1){
+               NumExtant[p] <- NumExtant[p] + 1
+               ExtantVals <- c(ExtantVals, DispInit[[CurParam]][[i]]$ExpDists)
+          }
+     }
+     AllSims[[CurParam]] <- log(AllVals, base = 10)
+     if(!is.null(ExtantVals)){
+          ExtantSims[[CurParam]] <- log(ExtantVals, base = 10)
+     }
+     xMin <- min(c(xMin, AllSims[[CurParam]]))
+     xMax <- max(c(xMax, AllSims[[CurParam]]))
+}
+xInitRange <- c(xMin, xMax)
+MainEvol <- DispEvol[[CurSpeed]]
+
+
+FigWidth <- 8
+FigHeight <- 6
+AllCol <- "lightskyblue"
+ExtantCol <- "navyblue"
+EvolDispAxisLabel <- expression(paste(Delta, bar(d), sep = ""))
+InitDispAxisLabel <- expression(paste("Initial ", italic("log"), "(", italic("d"), ")", sep = ""))
+EvolHistBreaks <- seq(-375, 475, by = 5)
+InitHistBreaks <- seq(-1.8, 3.4, by = 0.2)
+xEvolRange <- c(-100, 100)
+Initxticks <- seq(-1, 3, by = 0.2)
+AxisSize <- 1.15
+OuterMar <- c(0, 1, 0, 0)
+InnerMar <- c(5, 5, 4, 3) + 0.1
+options(scipen = 1000)
+LabSize <- 1.25
+
[…]

### 2018-07-23 SchematicFigures/ExampleFig.R

@@ -1,0 +1,74 @@
+# Make a figure displaying the range shift of an individual simulation
+
+setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/")
+source("6278_Sim1/parameters.R")
+source("~/Desktop/RangeShifts/ShiftingSlopesCode/SimFunctions.R")
+SumStats <- read.csv("6278_Sim1/SummaryStats.csv")
+
+# Define the time points that will be illustrated and extract just those data
+TimePoints <- seq(50, 90, by = 5)
+FigData <- subset(SumStats, (gen %in% TimePoints), select = c(gen, beta, x, y, 
+                                                              abund, muFit, muDisp))
+
+# Create the large array holding the matrices that will hold the population 
+#    values at different time points
+NumCols <- width * length(TimePoints)
+NumRows <- max(FigData$x) - min(FigData$x)
+FigMats <- array(NA, dim = c(3, NumRows, NumCols))
+
+# Now populate the figure matrices
+xCorrection <- abs(min(FigData$x))
+yCorrections <- ((1:length(TimePoints)) - 1) * width
+Betas <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, v = 1, eta = eta)
+for(t in 1:length(TimePoints)){
+     CurData <- subset(FigData, gen == TimePoints[t])
+     if(t == 1){
+          CurBeta <- BetaInit
+     } else{
+          CurBeta <- Betas[TimePoints[t] - 50]
+     }
+     for(i in 1:dim(CurData)[1]){
+          FigMats[1,CurData$x[i] + xCorrection,CurData$y[i] + yCorrections[t]] <- 
+               CurData$abund[i]
+          FigMats[2,CurData$x[i] + xCorrection,CurData$y[i] + yCorrections[t]] <- 
+               log(CurData$muDisp[i], base = 10)
+          Zopt <- lambda * (CurData$x[i] * eta - CurBeta)
+          RelFit <- exp(-1 * (CurData$muFit[i] - Zopt)^2 / (2*omega^2))
+          FigMats[3,CurData$x[i] + xCorrection,CurData$y[i] + yCorrections[t]] <- 
+               RelFit
+     }
+}
+
+# Set some graphical parameters to use for the subsequent figures
[…]

### 2018-07-23 SimFunctions.R

@@ -1,0 +1,1116 @@
+# This script contains all of the functions necessary to run a simulation of
+#    population evolution and movement in response to climate change. The 
+#    functions are organized into three different categories: biological,
+#    bookkeeping, and environmental. Each function is documented with a brief
+#    description of the purpose of the function and a list and definition of 
+#    the necessary function arguments before the function definition itself.
+#    Each broad category of functions is delineated by the dashed section
+#    headers and each individual function begins with a single line of 6 pound
+#    signs (######)
+
+
+# ------------------------------------------------------------------------------
+# ---------------------------- Biological Functions ----------------------------
+# ------------------------------------------------------------------------------
+
+###### CalcTraits
+# This function will calculate the fitness and dispersal trait values for a set 
+#    of individuals based on their alleles
+### INPUTS
+# population:  A vector of the population (i.e. row numbers) for which to 
+#                   calculate a trait values.
+# PopMat:      The population matrix to use
+# PopSize:     The size of the current population
+# PopIndices: Column indices to use for the population matrix
+### OUTPUTS
+# The function will create a matrix with three columns. The first column will 
+#    correspond to the ID (row number) of the individual, the second column
+#    will be the fitness trait value, and the third column will be the dispersal
+#    trait. If the traits are calculated for the entire population, as will 
+#    normally be the case, then the first column will simply correspond to the 
+#    row number, but including this column allows for the calculation of 
+#    traits for a subset of the current population.
+CalcTraits <- function(population, PopMat, PopSize, PopIndices){
+     # First create an empty matrix for the trait values and put
+     #    the population IDs in the first column
+     traits <- matrix(NA, nrow = PopSize, ncol = 3)
+     traits[,1] <- population
+     
+     # Now calculate the sum of each individual's quantitative loci for each 
+     #    trait and return those sums in the traits matrix
+     if(PopSize > 1){
+          traits[,2] <- rowSums(PopMat[population, PopIndices$FitCols])
+          traits[,3] <- rowSums(PopMat[population, PopIndices$DispCols])
+     } else{
+          traits[,2] <- sum(PopMat[population, PopIndices$FitCols])
+          traits[,3] <- sum(PopMat[population, PopIndices$DispCols])
[…]

### 2018-08-28 Manuscript/Main.tex
Updated the dispersal threshold calculations and the manuscript

@@
 
-\title{(1) Local adaptation and dispersal evolution interact to drive population response to climate change \\ (2) Spatial population structure drives extinction dynamics in climate-induced range shifts \\ (3) Initial spatial structure drives population response to climate change}
+\title{Spatial population structure drives extinction dynamics in climate-induced range shifts}
 
@@
 \section*{Abstract}
-Anthropogenic climate change has already caused some species' ranges to shift upwards in latitude and elevation, making it imperative to understand the ecological and evolutionary dynamics behind such range shifts.
-While insights can be gained from the related process of range expansion, climate-induced range shifts have several key attributes differentiating them from range expansions.
-In particular, range shifting populations are characterized by some initial spatial population structure defined by factors such as local adaptation and the nature of the range edge.
-Here, we use an individual-based model to explore the importance of spatial population structure for range shifting populations.
-We show that populations with more pronounced spatial structure have substantially increased extinction probabilities.
-Further, in our simulations evolution of dispersal ability was insufficient to rescue faltering populations.
+Understanding the interplay between ecological and evolutionary factors is critical for predicting species' responses to anthropogenic climate change and for determining optimal conservation practices.
+Key parallels between range expansions (e.g.
+of invasive species) and range shifts in response to climate change allow us to expand current theory for range expansions to include initial spatial population structures and predict species' responses to climate change.
+Here, we use an individual-based model to explore the role of dispersal evolution, local adaptation, and a population's initial spatial structure in driving the dynamics of climate-induced range shifts.
+We show that populations with more pronounced initial spatial structure had substantially increased extinction probabilities and that evolution of dispersal ability was insufficient to rescue faltering populations.
 Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
@@
 These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
-In fact, contemporary range shifts have already been observed in a wide range of taxa~\citep{chen2011rapid, parmesan2006ecological}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} the creation of novel species assemblages as not all species shift their ranges at the same rate or at all~\citep{hobbs2009novel}, or both.
-It is therefore crucial to understand the dynamics of such climate-induced range shifts to better inform current and future conservation work.
+Indeed, contemporary range shifts have already been observed in a wide variety of taxa, ranging from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}.
+Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages as not all species are expected to respond to climate change or to shift their ranges at the same rate~\citep{hobbs2009novel}.
+Understanding the dynamics of such climate-induced range shifts could play a key role in informing current and future conservation work.
 
-While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion.
-Range expansions have been studied for decades, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms responsible for shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can have important implications for both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal}.
-Since range shifts, like range expansions, have a leading edge of population advance, they are likely to be affected by similar ecological and evolutionary mechanisms as have been shown to drive dynamics in range expansions~\citep{hargreaves2014evolution}.
-However, while such insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.
+While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion, such as occurs when invasive species spread through a new environment.
+Decades of research on range expansions have led to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms responsible for shaping such expansions.
+For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal}.
+As range shifts and expansions exhibit similar spatial dynamics, such as a leading edge of population advance with relatively low population density, they are likely to be affected by similar ecological and evolutionary mechanisms ~\citep{hargreaves2014evolution}.
+However, while insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.
 
-In particular, range expansions and range shifts due to climate change involve fundamentally distinct initial conditions.
+In particular, range expansions and shifts due to climate change involve fundamentally distinct initial population structure.
 Range expansions, especially for invasive and reintroduced species, typically begin from the successful establishment and subsequent spread of a small, founding population~\citep{hastings2005spatial}.
 Such founding populations often represent samples from some larger source population and as such lack any initial spatial population structure.
-In contrast, range shifts start from entire populations with existing spatial structure within the initially stable range~\citep{hargreaves2014evolution}.
-Such spatial structure can manifest through local adaptation within the range, the severity of the gradient in population size from the range core to edge, or some combination of the two~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
+In contrast, range shifting populations initially have a more complex, but stable population structure that becomes disrupted with climate change or other environmental perturbations~\citep{hargreaves2014evolution}.
+A population's stable population structure incudes spatial variability in population size, adaptive traits, and range boundaries that manifests through a combination of local adaptation within the range and the severity of the gradient in population size from the range core to edge~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
 
-These different aspects of spatial population structure have been shown to dramatically impact the response of populations to climate change.
-For example, the mechanism responsible for the gradient in population size from core to edge (e.g.
-declines in per capita growth rate vs.
-increases in local extinction probability) can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
-A population's risk of extinction during a range shift has also been related to the degree of local adaptation within the range.
+A population's initial spatial structure has previously been shown to dramatically impact its response to climate change and its corresponding extinction probability.
+For example, the underlying mechanism responsible for the gradient in population size from the middle of the population (i.e.
+the more dense core) to the range edge can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
+Extinction risk is also related to the degree of local adaptation within the range.
 Specifically, a broad environmental niche (i.e.
@@
 Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
-However, it is unclear how dispersal evolution and local adaptation might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
+
+While the roles of dispersal evolution and local adaptation have been explored in isolation, using haploid, single locus models assuming asexual reproduction, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
 For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
 In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
-Additionally, the distribution of dispersal phenotypes within a population can be influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts.
-Evolution during range shifts due to local adaptation and dispersal evolution are each potentially important drivers of range shift dynamics and it is therefore important to consider both when predicting the dynamics of populations responding to climate change.
+Additionally, the distribution of dispersal phenotypes within a population likely is influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts.
 
 Here, we develop a two-sex individual-based model with two genetically determined traits, one defining an individual's expected dispersal distance and the other determining an individual's environmental niche.
-Using this model, we vary the degree of spatial population structure defining the stable range to ascertain how this impacts a population's ability to track a changing climate.
+In the model, we vary the degree of spatial population structure in the initial stable range to ascertain how the severity of the gradient at the range edge impacts a population's ability to track a changing climate.
 Further, we contrast the dynamics of extant and extinct populations to understand the joint impact of dispersal evolution and local adaptation on extinction risk during climate-induced range shifts.
@@
 \section*{Methods}
-A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in the online supplemental materials, but we present a brief summary here.
-Each simulated landscapes consisted of a discrete lattice of habitat patches with a fixed width in the $y$ dimension and unbounded length in the $x$ dimension.
+A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in the online supplemental materials, while here we present a brief summary.
+Each simulated landscape consisted of a discrete lattice of habitat patches with a fixed width in the $y$ dimension and unbounded length in the $x$ dimension.
 Environmental conditions varied along the $x$ dimension and remained constant across the $y$ dimension (Fig.
 S1).
-To simulate climate change, environmental conditions shifted a constant amount in the $x$ dimension each generation.
-Competition and reproduction occurred within each patch and local populations were linked via dispersal, assuming wrapping boundaries in the $y$ dimension to prevent edge effects.
+To simulate climate change, environmental conditions shifted at a constant rate in the $x$ dimension.
+In each generation, competition and reproduction occurred within each patch and local populations were linked via dispersal, assuming wrapping boundaries in the $y$ dimension to prevent edge effects.
 
 Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci, assuming no linkage.
+While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits.
 The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
-An individual's environmental niche trait allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's location, the higher the individual's realized fitness.
-The potential for local adaptation was then manipulated by varying the degree to which the environmental optimum changed throughout the landscape.
+An individual's environmental niche trait allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness.
+The environmental optimum of individual patches could then be systematically altered across the range to allow for varying degrees of local adaptation (i.e.
+larger changes in environmental optima allowed for greater local adaptation of the population).
 In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see supplemental materials; Fig.
@@
 Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-Using this model, we simulated non-overlapping generations consisting of discrete dispersal and reproduction phases (Fig.
-S2) to determine the effect of different aspects of spatial population structure on the eco-evolutionary dynamics of climate-induced range shifts.
+We simulated non-overlapping generations consisting of discrete dispersal and reproduction phases (Fig.
+S2).
 
-Most model parameters were held constant over all simulations (Table S1), but we varied certain parameter values to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S2).
-Specifically, we considered a factorial combination of no, low, and high potential for local adaptation with shallow, moderate, and stark gradients in suitable habitat at the range edge for a total of $9$ different scenarios, each explored with $200$ simulations.
+To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S1 and S2).
+Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change.
+This yielded a total of $27$ different scenarios, each explored with $200$ simulations.
 Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
 Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change.
-Additionally, we explored the role of the rate of climate change on the dynamics of range shifts in these different scenarios, examining both slower and faster speeds of climate change (see the online supplemental materials).
+For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.
 
-For each parameter combination, we calculated the per patch change in mean dispersal phenotype from the beginning of the period of climate change to the end to quantify dispersal evolution.
-For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates.
-In this way, we calculated change in mean dispersal phenotype for populations consistently occupying the same position relative to the range center (e.g.
+We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end.
+For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g.
 leading edge vs.
@@
 Due to local extinctions, not all patches were occupied at the end of the period of climate change.
-To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals as the end point to calculate change in the mean dispersal phenotype.
-Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions) so that positive values indicate an increase in the mean dispersal phenotype.
+To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals.
+Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype.
 All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at \url{https://github.com/tpweiss06/ShiftingSlopes}.
@@
 \section*{Results}
-For almost all parameter combinations examined here, a portion of the simulated populations were able to keep pace with changing climate conditions.
+In all scenarios, populations shifted their range in response to climate change, however the proportion of populations that successfully tracked the changing climate for the full $100$ generations depended on the initial spatial population structure.
 Successful populations shifted their spatial distributions at essentially the same rate as climate change (e.g.
 Fig.
-1).
-However, some populations failed to keep pace, lagging further behind the shifting climate until they went extinct.
+1), while populations that failed to keep pace and lagged behind the shifting climate went extinct.
 Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
 2).
-While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction dynamics.
-The parameters defining both of these range attributes were varied widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table S2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space.
-Additionally, the pace of climate change also influenced extinction probabilities as expected with faster climate change corresponding to greater extinction risk (Fig.
+While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction dynamics, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change.
+Here, we widely varied both parameters (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table S2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space, and corresponding biological scenarios.
+Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig.
 S3 \& S4).
-However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability for a range shifting population.
+However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability during a range shift.
 
-Dispersal evolution is expected to aid populations shifting spatially to track a changing climate, and, as expected, the average dispersal phenotypes increased through time in some simulations (e.g.
-Fig.~\ref{fig:SimExample}).
-However, comparing patterns in dispersal evolution between simulated populations that successfully tracked climate change and those that ultimately went extinct revealed no differences in the magnitude or direction of dispersal evolution (Fig.
+Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate.
+While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g.
+Fig.
+1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig.
 3a\&b).
@@
 S5-S7).
-The similarity in evolved changes to dispersal between surviving and extinct simulated populations suggests that this cannot explain the different outcomes achieved by these populations.
+The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions versus became extinct.
 
-Instead, the initial distribution of dispersal phenotypes immediately before the beginning of climate change played a large role in determining the ultimate fate of simulated populations.
+Instead, the stable distribution of dispersal phenotypes before the beginning of climate change played a large role in determining the ultimate fate of simulated populations.
 Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
@@
 S8-S10).
-This threshold varies directly with the speed of climate change, increasing with more rapid climate change.
+This threshold varied directly with the speed of climate change, increasing with more rapid climate change.
 Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e.
@@
 
-Intriguingly, the simulations that survived climate change tended to be characterized by reduced fitness at the range margins compared to simulated populations that went extinct (Fig.
+Intriguingly, the populations that survived climate change tended to be characterized by reduced fitness at the range margins compared to populations that went extinct (Fig.
 S11-S13).
 This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
-Intuitively, there was no spatial pattern in fitness for simulations with no potential for local adaptation.
-As the simulations that survived climate change were also characterized by heightened dispersal (Fig.
+Intuitively, there was no spatial pattern in fitness for populations with no potential for local adaptation.
+As the populations that survived climate change were also characterized by heightened dispersal (Fig.
 3), the observed reduction in average patch fitness at the margins is likely due to increased gene flow from the range core hampering the abilities of these edge populations to adapt to local conditions.
@@
 single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
-Thus, in analyzing our model, we tested the generality of previous results under realistic levels of biological complexity.
+Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under realistic levels of biological complexity.
 Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
 
-Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the range and in locations with shallow environmental gradients defining the range edge (Fig.
+Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the pre-expansion, stable range and in locations with shallow environmental gradients defining the range edge (Fig.
 2).
-A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies testing for it, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
+A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
 Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study.
@@
 
-A key result of our model was the importance of the initial distribution of dispersal phenotypes (i.e.
-prior to the start of climate change) in determining a population's extinction risk during climate change (Fig.
+Our results emphasize the importance of the initial distribution of dispersal phenotypes composing the stable range in determining a population's extinction risk during climate change (Fig.
 3c\&d).
-Survival in the face of climate change was primarily determined by whether or not simulated populations consisted of individuals with dispersal phenotypes at or above some threshold value.
-Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations at a given speed of climate change (Fig.
+Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a given threshold value.
+Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations for a given speed of climate change (Fig.
 S8-S10).
-The differences in extinction probabilities among different scenarios were instead driven by the overall distribution of dispersal phenotypes that evolved in different simulations under each parameter combination.
-Scenarios with no potential for local adaptation and gradual environmental gradients simply had larger proportions of high dispersal phenotypes under stable climate conditions.
+Scenarios with no potential for local adaptation and gradual environmental gradients had larger proportions of high dispersal phenotypes under stable climate conditions, and therefore a lower probability of extinction during climate change.
 A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}.
-Similarly, a more severe habitat gradient at the range edge increases the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}.
-Thus, attributes defining the spatial population structure of simulated ranges resulted in different initial distributions of traits under stable climate conditions, which then determined the extinction risk of populations under climate change.
+Similarly, a more severe habitat gradient at the range edge increased the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}.
+Thus, attributes defining the spatial structure of the range altered the distribution of dispersal phenotypes under stable climate conditions, subsequently determining the extinction risk of populations during climate change.
+Importantly, dispersal evolution during climate change was unable to counter the influence of spatial population structure in the stable range on extinction dynamics.
 
-While high dispersal phenotypes prior to climate change allowed populations to survive, it had the additional effect of reducing average fitness at the range margins in scenarios with the potential for local adaptation (Fig.
+While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range margins when populations had a moderate to high potential for local adaptation (Fig.
 S11-S13).
-In the model, populations at the range margins tended to be lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
-Thus, when populations were defined by high dispersal phenotypes prior to climate change, it reduced fitness at the range margin due to gene flow preventing adaptation to local conditions.
-As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the start of climate change.
-While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, and thus are unlikely to be equally susceptible to the influence of gene flow from the core, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
+In the model, populations at the range margins tended to have lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
+Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core reduced fitness at the range edge by preventing adaptation to local conditions.
+As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the onset of climate change.
+While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
 
-Previous research has suggested that evolution of increased dispersal ability during climate change may be able to rescue populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}.
+Previous research has suggested that evolution of increased dispersal ability during climate change may be a key mechanism in rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}.
 Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
-Models finding that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid and single-locus models~\citep{boeye2013more, hargreaves2015fitness}.
-As dispersal evolution during range expansions and shifts occurs via spatial sorting~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of highly dispersive alleles.
-Additionally, many previous models of climate-induced range shifts use asexual reproduction which can result in more rapid rates of evolutionary change under certain conditions~\citep{crow1965evolution, smith1968evolution}.
+Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid and single-locus models~\citep{boeye2013more, hargreaves2015fitness}.
+As dispersal evolution during range expansions and shifts occurs via the formation of strong spatial structure from the core to the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of highly dispersive alleles compared to our more complex model multiple diploid loci contributing to dispersal.
+Additionally, many previous models of climate-induced range shifts assume asexual reproduction which can result in more rapid rates of evolutionary change under certain conditions~\citep{crow1965evolution, smith1968evolution}.
 The negligible role played by dispersal evolution in our model (Fig.
-3a\&b) suggests that when such simplifying assumptions are relaxed the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
+3a\&b) suggests that when such simplifying assumptions are relaxed, the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
 
@@
 Understanding the various ecological and evolutionary drivers of climate-induced range shifts is crucial to current and future conservation efforts.
-In particular, we need to better understand how these various mechanisms combine to shape the extinction probabilities of populations undergoing range shifts, which, in turn, will allow more focused interventions.
+In particular, a deeper, process-based understanding of how various mechanisms combine to shape extinction probabilities of populations undergoing range shifts, will, in turn, allow more focused interventions.
 Our results suggest that initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
@@
 \section*{Acknowledgements}
-We thank Lauren Shoemaker, Lauren Sullivan, Kate Meyers, Dongmin Kim, Lindsay Leverett, and members of the Theory Under Construction group at the University of Minnesota for providing thoughtful comments on the manuscript.
+We thank Lauren Shoemaker, Lauren Sullivan, and members of the Theory Under Construction group at the University of Minnesota for providing thoughtful comments on the manuscript.
+CWL was supported by startup funds from the University of Minnesota (to AKS), and AKS was supported in part by funding from the National Science Foundation.
+We acknowledge the Minnesota Supercomputing Institute (MSI) at the University of Minnesota for providing resources that contributed to the results reported within this paper (http://www.msi.umn.edu).
 
@@
 \paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
-Information on the abundance, dispersal, and fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
-Log transformed mean dispersal phenotypes are shown for each patch.
-Average patch fitness was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.
+Information on the (a) abundance, (b) dispersal, and (c) fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
+Log transformed mean dispersal phenotypes (b) are shown for each patch.
+Average patch fitness (c) was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.
 

### 2018-08-28 8b_DispInitGraphs.R

@@ -1,8 +1,16 @@
 # This script will make an appropriate graph of the initial dispersal values
 setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
 load("SimData/InitDispData.rdata")
-CurSpeed <- 3
+CurSpeed <- 2
 SpeedWords <- c("Slow", "Main", "Fast")
+
+# Calculate the dispersal phenotype corresponding to the current speed of climate
+#    change
+r <- log(2)
+patch_speeds <- c(0.5, 1, 2)
+eta <- 50
+v <- patch_speeds[CurSpeed]*eta
+DispPhen <- sqrt(v^2 / r)
 
 # Sort the disp values into the appropriate lists
 AllSims <- vector(length = 9, mode = "list")
@@ -74,6 +82,8 @@ pdf(file = PlotName, width = FigWidth, height = FigHeight, onefile = FALSE, pape
           if(NumExtant[i] > 0){
                hist(ExtantSims[[i]], col = ExtantCol, breaks = HistBreaks, add = TRUE)
           }
+          # Add a vertical line at the dispersal phenotype tracking climate change
+          abline(v = log(DispPhen, base = 10), lty = 2, col = "red", lwd = 2)
           options(scipen = 10)
           axis(1, cex.axis = AxisSize)
           axis(1, at = xticks, labels = FALSE, tcl = -0.25)

### 2018-08-28 DispIdea.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=06:00:00,nodes=4:ppn=24,mem=248gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=DispIdea.R
+LogFile=DispIdea.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2018-08-30 Manuscript/Main.tex
Updated the manuscript with the new threshold calculations

@@
 Key parallels between range expansions (e.g.
-of invasive species) and range shifts in response to climate change allow us to expand current theory for range expansions to include initial spatial population structures and predict species' responses to climate change.
+of invasive species) and range shifts in response to climate change allow us to draw upon current theory for range expansions to include initial spatial population structures and predict species' responses to climate change.
 Here, we use an individual-based model to explore the role of dispersal evolution, local adaptation, and a population's initial spatial structure in driving the dynamics of climate-induced range shifts.
@@
 Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
-Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change.
+Figure 1 shows an example of a single simulation undergoing climate change.
 For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.
@@
 3a\&b).
-Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes strongly centered on $0$ (Fig.
+Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes centered on $0$ (Fig.
 S5-S7).
-The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions versus became extinct.
+The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.
 
@@
 3c\&d).
-In fact, comparing dispersal phenotypes of successful populations across different parameter combinations revealed a threshold in dispersal phenotypes separating individuals from successful and extinct populations (Fig.
+Comparing the dispersal phenotypes making up successful populations across different parameter combinations revealed a common threshold separating individuals from successful and extinct populations at each speed of climate change (Fig.
 S8-S10).
-This threshold varied directly with the speed of climate change, increasing with more rapid climate change.
-Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e.
-moving the correct number of patches in the $x$ dimension to track the changing environmental conditions).
+In fact, this threshold corresponds to the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change.
+To show this, we used the well known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, where $r$ is the intrinsic growth rate of the population and $D$ is the diffusion coefficient.
+The $2$ dimensional diffusion coefficient can be calculated directly from the dispersal phenotypes as $D=\frac{1}{4}d^{2}$ since $d^{2}$ represents the mean squared displacement of an individual with phenotype $d$ during a dispersal period (see the model description in the supplemental materials).
+Using these relationships and solving for the dispersal phenotype producing each of the speeds of climate change used in our simulations yields the threshold value indicated on Figures S8-S10 which distinguishes those populations that ultimately survived climate change from those that went extinct.
 
@@
 \paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
-Information on the (a) abundance, (b) dispersal, and (c) fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
+Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
 Log transformed mean dispersal phenotypes (b) are shown for each patch.
@@
 Positive values indicate an increase in average dispersal ability in the patch.
-Initial distributions of the dispersal trait (c and d) are shown with histograms of the log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
+Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
 In all panels, values associated with extant populations are shown in dark blue.

### 2018-08-30 Manuscript/Supp.tex
Updated the manuscript with the new threshold calculations

@@
     \hline
-    Habitat gradient at range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
+    Habitat gradient at the range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
      & None & $0.0025$ & $250$ & $0$ \\
@@
 On each histogram, the values associated with populations surviving climate change are shown in dark blue and the total number of simulated populations to survive climate change is indicated in the top left corner.
+The dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change is indicated by the dashed red lines.
 Each histogram represents one of the $9$ experimental scenarios examined here, denoted by the text and arrows on the figure.

### 2018-09-06 Manuscript/Main.tex
Final update to the manuscript before the initial submission to Ecology

@@
 
-Instead, the stable distribution of dispersal phenotypes before the beginning of climate change played a large role in determining the ultimate fate of simulated populations.
+Instead, the stable distribution of dispersal phenotypes prior to the onset of climate change played a key role in determining the ultimate fate of simulated populations.
 Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
 S8-S10).
-The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig.
+The populations that survived climate change were those with higher initial dispersal phenotypes, occupying the right-tail of the distribution of observed dispersal phenotypes (Fig.
 3c\&d).
-Comparing the dispersal phenotypes making up successful populations across different parameter combinations revealed a common threshold separating individuals from successful and extinct populations at each speed of climate change (Fig.
+Comparing the dispersal phenotypes of individuals in successful populations across different parameter combinations revealed a common threshold separating individuals from successful and extinct populations at each speed of climate change (Fig.
 S8-S10).
-In fact, this threshold corresponds to the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change.
-To show this, we used the well known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, where $r$ is the intrinsic growth rate of the population and $D$ is the diffusion coefficient.
+In fact, this threshold corresponded to the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change.
+To prove this, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, where $r$ is the intrinsic growth rate of the population and $D$ is the diffusion coefficient.
 The $2$ dimensional diffusion coefficient can be calculated directly from the dispersal phenotypes as $D=\frac{1}{4}d^{2}$ since $d^{2}$ represents the mean squared displacement of an individual with phenotype $d$ during a dispersal period (see the model description in the supplemental materials).
-Using these relationships and solving for the dispersal phenotype producing each of the speeds of climate change used in our simulations yields the threshold value indicated on Figures S8-S10 which distinguishes those populations that ultimately survived climate change from those that went extinct.
+Using these relationships and solving for the dispersal phenotype producing each of the speeds of climate change used in our simulations yields the threshold value distinguishing those populations that ultimately survived climate change from those that went extinct (Figures S8-S10, vertical dashed line).
 

### 2019-04-05 Manuscript/AmNat_InitialSubmission.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\usepackage{mathptmx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fancybox}
\usepackage{graphicx}
\usepackage{graphics}
\newcommand{\s}{^{(s)}}

%%%%%%%%%%%%%%%%%%%%%
% Line numbering
%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
% Please use line numbering with your initial submission and
% subsequent revisions. After acceptance, please turn line numbering
% off by adding percent signs to the lines %\usepackage{lineno} and
% to %\linenumbers{} and %\modulolinenumbers[3] below.

\title{Spatial population structure determines extinction risk in climate-induced range shifts}

% This version of the LaTeX template was last updated on
% January 11, 2018.

%%%%%%%%%%%%%%%%%%%%%
% Authorship
%%%%%%%%%%%%%%%%%%%%%
% Please remove authorship information while your paper is under review,
% unless you wish to waive your anonymity under double-blind review. You
% will need to add this information back in to your final files after
% acceptance.

%\author{Owen E. Cook$^{1,\ast}$ \\ 
%Generic H. Collaborator$^{2,\dag}$ \\ 
%Additional Q. Expert$^{3}$}

\date{}

\begin{document}

\maketitle

%\noindent{} 1. University of Chicago, Chicago, Illinois 60637;

%\noindent{} 2. University of Toronto, Toronto, Ontario M5S 1A5, Canada;

%\noindent{} 3. Middle Eastern Technical University, Çankaya, Ankara 06800, Turkey.

%\noindent{} $\ast$ Corresponding author; e-mail: amnat@uchicago.edu.

%\noindent{} $\dag$ Deceased.

\bigskip

\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1\&A2,  tables~A1\&A2, and figures~B1-B11). Figure~1 and figure~3 are to print in color.

\bigskip

\textit{Keywords}: range shifts, eco-evolutionary dynamics, local adaptation, individual-based model

\bigskip

\textit{Manuscript type}: Article. %Or e-article, note, e-note, natural history miscellany, e-natural history miscellany, comment, reply, invited symposium, or countdown to 150.

\bigskip

\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}

\linenumbers{}
\modulolinenumbers[3]

\newpage{}

\section*{Abstract}
Climate change is an increasingly severe threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses. Population dynamics of range shifts, among the most commonly observed responses to climate change, can be influenced by many factors, including evolution of key traits, the degree of local adaptation, and the nature of the range edge. Here, we use an individual-based model to explore the interacting roles of these factors in the dynamics of climate-induced range shifts. We show that aspects of the spatial population structure within the initial range, in particular the potential for local adaptation, severely increased a population's extinction risk. Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations. Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes that evolved within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.

\newpage{}

\section*{Introduction}
Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}. These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}. Indeed, contemporary range shifts have already been observed in a wide variety of taxa, ranging from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}. Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages~\citep{hobbs2009novel}. Understanding the ecological and evolutionary dynamics of such climate-induced range shifts will play a key role in informing current and future conservation work.

Large-scale population movements have been studied for decades in the context of range expansions (e.g. of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions. For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}. As fundamentally similar spatial processes, it is likely that range shifts will also be subject to these ecological and evolutionary mechanisms known to drive range expansions. However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population. In particular, range shifts occur in populations with far more complex spatial structure compared to most range expansions, which typically begin from the successful establishment and spread of a small, founding population~\citep{hastings2005spatial}. While these founding populations often lack any significant spatial structure, populations undergoing range shifts are characterized by a spatial population structure formed by aspects of the previously stable ranges. For example, population ranges can vary in their potential for local adaptation throughout the range, the nature of the range edge, and the spatial distribution of key traits.

Each of these factors relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions. For example, the underlying mechanism responsible for the gradient in population size from the range core to the edge (i.e. declines in carrying capacity versus growth rate) alters a population's extinction risk during climate driven range shifts~\citep{henry2013eco}. Additionally, the potential for local adaptation throughout a range has been related to extinction risk during range shifts. Specifically, a low potential for local adaptation can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block the establishment of better adapted genotypes~\citep{atkins2010local}. While these aspects of spatial population structure have been shown to impact the dynamics of climate-induced range shifts in isolation, it is unclear how and if they might interact.

Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts. In asexual species, for example, local adaptation has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}. However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow. Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range. In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}. In addition to potential interactions between local adaptation and dispersal evolution, the nature of the range edge could influence the potential for rapid trait evolution during range shifts. For example, the severity of the environmental gradient forming the range edge has been shown to alter the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.

Here, we assess the interaction of multiple aspects of spatial population structure with trait evolution in sexually reproducing populations undergoing climate-induced range shifts. We develop an individual-based model capable of incorporating a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change. One trait determines dispersal ability while the second defines an individual's environmental niche. Using this model, we vary both the potential for local adaptation within the range and the nature of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate. By contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.

\section*{Methods}
A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here. Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig. A1). Landscapes were unbounded in the $x$ dimension but defined by a fixed width in the $y$ dimension. Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions. To simulate climate change, environmental conditions shifted at a constant rate along the $x$ dimension. Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig. A2).

Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci. While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits. The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel. An individual's environmental niche value allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness. The environmental optimum of individual patches could then be systematically varied across the range to allow for different degrees of local adaptation (i.e. larger changes in environmental optima allowed for greater local adaptation of the population). In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see Appendix A). Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness. Allele inheritance was subject to mutation and assumed no linkages among loci.

To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table A1 and A2). Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure 1 shows an example of a single population responding to a moderate speed of climate change. For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations. We primarily discuss simulations using a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.

We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end. For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g. leading edge vs. core populations). Due to local extinctions, not all patches were occupied at the end of the period of climate change. To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals. Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype. All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).

\section*{Results}
In all scenarios, some populations shifted their ranges in response to climate change. However the proportion of extinct populations that failed to track the changing climate depended on the initial spatial population structure. Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig. 2). While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction risk, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change. We varied both parameters widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table A2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2). However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability during a range shift.

Counterintuitively, populations that survived climate change tended to be characterized by reduced fitness at the range margins prior to the onset of climate change compared to populations that went extinct (Fig. B3-B5). Essentially, populations with initially higher degrees of local adaptation at the range edges, and thus greater fitness, were more likely to go extinct during climate change. This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation. As expected, there was no spatial variation in fitness for populations with no potential for local adaptation.

Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate. While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g. Fig. 1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig. B3-B5). The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.

Instead, the initial distribution of dispersal phenotypes prior to the onset of climate change played a key role in determining a population's fate. A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig. B9-B11). Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig. 3c\&d). In fact, comparing the full distribution of initial dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations. Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig. B9-B11). To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change used in our simulations (see the model description in Appendix A). This dispersal phenotype matched the observed threshold value distinguishing surviving from extinction populations in all experimental scenarios (Figures B9-B11, vertical dashed line). Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change.

\section*{Discussion}
Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts. We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci. In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g. single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).  Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under increased levels of biological complexity. Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.

Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the pre-expansion, stable range and in locations with shallow environmental gradients defining the range edge (Fig. 2). A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study. While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to the spatial population structure of their initial ranges.

Our results emphasize the importance of the initial distribution of dispersal phenotypes composing the stable range in determining a population's extinction risk during climate change (Fig. 3c\&d). Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a threshold value. Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations for a given speed of climate change (Fig. B9-B11). Scenarios with no potential for local adaptation and gradual environmental gradients had larger proportions of high dispersal phenotypes under stable climate conditions, and therefore a lower probability of extinction during climate change. A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}. Similarly, a more severe habitat gradient at the range edge increased the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}. Thus, attributes defining the spatial structure of the range altered the distribution of dispersal phenotypes under stable climate conditions, subsequently determining the extinction risk of populations during climate change. Importantly, dispersal evolution during climate change was unable to counter the influence of initial spatial population structure on extinction dynamics.

While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range edges when populations had a moderate to high potential for local adaptation (Fig. B3-B5). In the model, populations at the range edges tended to have lower abundance than populations in the range core, increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}. Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core likely reduced fitness at the range edge by preventing adaptation to local conditions. As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges prior to the onset of climate change. While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.

Previous research has suggested that evolution of increased dispersal ability during climate change may be a key mechanism in rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}. Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions. Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}. As dispersal evolution during range expansions and shifts occurs via the spatial sorting of alleles contributing to heightened dispersal at the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait. The negligible role played by dispersal evolution in our model (Fig. 3a\&b) suggests that when such simplifying assumptions are relaxed, the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.

\section*{Conclusion}
As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions. In particular, a deeper, process-based understanding of extinction risk in populations undergoing range shifts will, in turn, allow more focused conservation interventions. Our results suggest that the initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change. Further, in contrast to other studies assuming more simplified genetic structures, we find very little role for the evolution of heightened dispersal abilities in allowing a population to successfully track climate change. Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue. As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such risk.

%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%
% You may wish to remove the Acknowledgments section while your paper 
% is under review (unless you wish to waive your anonymity under
% double-blind review) if the Acknowledgments reveal your identity.
% If you remove this section, you will need to add it back in to your
% final files after acceptance.

%\section*{Acknowledgments}

\newpage{}

\section*{Appendix A: Full model description}

% Please reset counters for the appendix (thus normally figure A1, 
% figure A2, table A1, etc.).

% In certain cases, it may be appropriate to have a PRINT appendix in
% addition to (or instead of) an online appendix. In this case, please 
% name the print appendix Appendix A, and any subsequent appendixes (if 
% there are any) should be named Online Appendix B, Online Appendix C,
% etc.

% Counters for each appendix should match the letter of that appendix.
% For example, tables in Appendix C should be numbered table C1, table C2,
% etc. This applies to tables, equations, and figures.

% It's better not to use the \appendix command, because we have some
% formatting peculiarities that \appendix conflicts with.

\renewcommand{\theequation}{A\arabic{equation}}
% redefine the command that creates the equation number.
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{equation}{0}  % reset counter 
\setcounter{figure}{0}
\setcounter{table}{0}

\section*{Model overview}
\subsection*{Purpose} 
This model tests an evolving population's ability to track a changing climate under a variety of conditions. Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the potential for local adaptation. In all simulations, an individual's expected dispersal distance and environmental niche are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time. All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change. Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see \textit{Submodels} below). Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.

\subsection*{State variables and scales} 
The model simulates a population of males and females characterized by diploid loci for both their expected dispersal distance and environmental niche. Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system. Landscapes are two dimensional with a fixed width along the $y$ axis and without bounds on the $x$ axis. Environmental conditions vary along the $x$ dimension but remain constant within the $y$ dimension. To avoid edge effects due to the fixed width of the $y$ dimension, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same column of the landscape. Patches are defined by the location of the patch center in $x$ and $y$ coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (\textit{Submodels}).

The model implements climate change by shifting the location of a population's available habitat along the $x$ dimension of the landscape. Available habitat (i.e. a population's potential range) is defined by a center location on the $x$ dimension, the severity of the decline in habitat quality at the edge, and the width of the available habitat along the $x$ dimension (See Figure A1). Further, a gradient in environmental conditions is imposed throughout the landscape to allow for local adaptation via matching of an individual's environmental niche to the local environmental conditions (\textit{Submodels}). The severity of this gradient can be altered to change the potential for local adaptation (e.g. a shallower gradient will result in more similar environmental conditions throughout the range and therefore reduce the potential for local adaptation).

[Figure A1 goes here]

\subsection*{Process overview and scheduling} 
Time is modeled in discrete intervals defining single generations of the population (Fig. A2). Within each generation, individuals first disperse from their natal patches according to their phenotypes. After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch. Reproduction occurs via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) are likely to produce multiple offspring while individuals with low relative fitness may not produce any. Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process. After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations.

[Figure A2 goes here]

\section*{Design concepts}
\subsection*{Emergence} 
Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both expected dispersal distances and environmental niche values during climate change.

\subsection*{Stochasticity} 
All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci. Environmental parameters are fixed, however, and the process of climate change (i.e. the movement of environmentally suitable habitat through time) is deterministic. Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.

\subsection*{Interactions} 
Individuals in the model interact via mating and density-dependent competition within patches. Additionally, the evolutionary trajectories of the two different traits have the potential to interact via the relationship between gene flow and local adaptation. Further, aspects determining the spatial population structure of a population's range (potential for local adaptation and the nature of the range edge) can interact with trait evolution within the range both during stable climate conditions and during climate change.

\subsection*{Desired output} 
After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits). If a population went extinct during the model run, the time of extinction is recorded. For each occupied patch throughout the simulation, we aggregated data on population size, the dispersal trait, and local adaptation to environmental conditions.

\section*{Details}
\subsection*{Initialization} 
The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below. Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions. This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation. See Tables A1 \& A2 for a full list of parameter values used in the simulations described here.

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ \\
    $\beta_{1}$ & Center of environmentally suitable habitat before climate change & $0$ \\
    $\eta$ & Spatial dimensions of habitat patches in continuous space & $50$ \\
    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ direction & $10$ \\
    $\hat{t}$ & Last time point of stable climate conditions & $2000$ \\
    $t_{\Delta}$ & Duration of climate change & $100$ \\
    $t_{max}$ & Total number of time points in the simulation & $2150$ \\
    $R$ & Intrinsic growth rate of the population & $2$ \\
    $K_{max}$ & Maximum achievable carrying capacity in the range & $100$ \\ 
    $\psi$ & Expected sex ratio in the population & $0.5$ \\
    $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ for each trait \\
    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ for each trait \\
    $L^{T}$ & Number of diploid loci defining each trait & $5$ for each trait \\
    $\mu_{1}^{f}$ & Initial mean allele value for the environmental niche trait & $0$ \\
    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the environmental niche trait & $0.025$ \\
    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Simulation parameters held constant across all scenarios.}
\label{table:ConstPars}
\end{table}

\subsection*{Submodels}
\paragraph{Environmentally suitable habitat}
Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$). The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center. Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows 
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
\end{cases}
\end{equation}
where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure A1). To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the $x$ dimension can be shown to be
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
\end{cases}	
\end{equation}
yielding a value of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).

Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure A1). The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
\begin{equation}
K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
\end{equation}

By varying the parameters defining $f(x,t)$, we can change both the total carrying capacity of the population, summed across all patches throughout the range, (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$). Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total carrying capacity of the population. To avoid this confounding factor, we fix the total area under the curve $f(x,t)$. The indefinite integral of $f(x,t)$ can be shown to be
\begin{equation}
\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
\end{equation}
which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.

Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change. During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
\begin{equation}
\beta_{t}=\nu\eta(t-\hat{t})
\end{equation}
where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
    \hline
    Habitat gradient at the range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
     & None & $0.0025$ & $250$ & $0$ \\
    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
     & None & $0.025$ & $421.479$ & $0$ \\
    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
     & None & $0.25$ & $421.48$ & $0$ \\
    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
     & High & $0.25$ & $421.48$ & $0.008$ \\ 
    \hline
  \end{tabular}
\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.}
\label{table:Scenarios}
\end{table}

\paragraph{Local adaptation}
To allow an arbitrary degree of local adaptation within the range, the local environmental optima for each patch ($z_{opt,x}$) are set as follows
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
where $\lambda$ defines the potential for local adaptation with values close to $0$ resulting in little to no change in environmental optima across the range and values of greater magnitude resulting in large differences in environmental optima across the range. Individual relative fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's niche phenotype~\citep{lande1976natural}. Thus, an individual's realized fitness will be higher the closer its niche phenotype ($z_{i}$) is to the environmental optimum of the patch it occupies ($z_{opt,x}$). All loci are assumed to contribute additively to an individual's environmental niche value with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values.

\paragraph{Population dynamics}
Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$). The expected number of new offspring in patch $x$ at time $t+1$ is then given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e. polygynandrous mating). The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any. Each offspring inherits one allele per locus from each parent, assuming no linkage among loci. After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.

\paragraph{Mutation}
Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents. The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$). Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation is
\begin{equation}
\frac{U^{T}}{2L^{T}}
\end{equation}
Mutational effects are drawn from a normal distribution with mean $0$ and a standard deviation of
\begin{equation}
\sqrt{V_{m}^{T}U^{T}}
\end{equation}
By defining the mutation process in this manner rather than setting a probability of mutation and mutational effect directly, similar mutational dynamics can be imposed regardless of the number of loci used in the simulation.

\paragraph{Dispersal}
Finally, individuals disperse according to an exponential dispersal kernel defined by each individual's dispersal phenotype. An individual's dispersal phenotype is the expected dispersal distance and is given by
\begin{equation}
d_{i} = \frac{\hat{d}\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
\end{equation}
where $\hat{d}$ is the maximum expected dispersal distance in terms of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $\hat{d}$, and the summation is taken across all alleles contributing to dispersal. Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis. The expected dispersal distance, $d_{i}$ is then used to draw a realized distance from an exponential dispersal kernel. Since the dispersal phenotype is the expected value of the exponential dispersal kernel, it can be used directly to calculate the two dimensional diffusion coefficient of population spread ($D$). Specifically, since $d_{i}^{2}$ represents the mean squared displacement of an individual with dispersal phenotype $d_{i}$, the two dimensional diffusion coefficient can be calculated as
\begin{equation}
D = \frac{1}{4}d_{i}^{2}
\end{equation}

Once the realized dispersal distance is obtained, the direction of dispersal is drawn from a uniform distribution bounded by $0$ and $2\pi$. If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ dimension, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ dimension, thus wrapping the top and bottom edges of the landscape to avoid edge effects. Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure A1).

\newpage{}

\section*{Appendix B: Supplementary results for varying speeds of climate change}

\renewcommand{\theequation}{B\arabic{equation}}
% redefine the command that creates the equation number.
\renewcommand{\thetable}{B\arabic{table}}
\setcounter{equation}{0}  % reset counter 
\setcounter{figure}{0}
\setcounter{table}{0}

\section*{Extinction probability}
As in the main text, we calculated the cumulative probability of extinction for both slow and fast speeds of climate change here. The figures in this section use the same layout and line types as Figure 2 in the main text to allow for direct comparisons.

[Figures B1\&B2 go here.]

\section*{Initial fitness throughout the landscape}
To assess trends in realized fitness values throughout the landscape, we calculated the patch-level mean individual fitness value for each landscape just prior to the onset of climate change. To simplify the figures, we averaged over the $y$ dimension in which environmental conditions do not vary. By separating the results for both extinction populations and those that successfully tracked climate change, we identified a trend in which the edge populations displayed lower fitness in populations that were ultimately successful, particularly in simulations with a gradual environmental gradient at the range edge and the potential for local adaptation. As realized fitness values do not vary spatially in simulations with no potential for local adaptation, the following figures only show results for scenarios with a moderate or high potential for local adaptation.

[Figures B3-B5 go here.]

\section*{Dispersal evolution}
Using the same metric of dispersal evolution from the main text (average change in phenotype for each patch), we display here the observed dispersal evolution over the course of climate change for all experimental scenarios. Each histogram in the following figures represents a single experimental scenario as indicated by the figure text. The lower left panel and upper right panel from Figure B7 are the same histograms shown in Figure 3a\&b, but are here placed in the context of all other experimental scenarios.

[Figures B6-B8 go here.]

\section*{Initial dispersal phenotypes}
Here, we present histograms of the initial distribution of dispersal phenotypes to demonstrate the importance of those phenotypes in determining population success or extinction during climate change. Dispersal phenotypes are log transformed for easier comparison. As with the dispersal evolution section, each histogram represents a single experimental scenario as indicated by the figure text. Similarly, the lower left panel and upper right panel from Figure B10 are the same histograms shown in Figure 3c\&d, but are here placed in the context of all other experimental scenarios. All histograms additionally have a vertical dashed line indicating the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change in each simulation. This value serves as a threshold to distinguish individuals from ultimately successful versus extinct populations.

[Figures B9-B11 go here.]

\newpage{}

%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%
% You can either type your references following the examples below, or
% compile your BiBTeX database and paste the contents of your .bbl file
% here. The amnatnat.bst style file should work for this---but please
% let us know if you run into any hitches with it!
% The list below includes sample journal articles, book chapters, and
% Dryad references.

\bibliographystyle{amnat}
\bibliography{main_bib}

\newpage{}

\section*{Figures}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
\caption{A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge. Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean environmental niche trait of local individuals and the environmental optima for each patch.}
\label{fig:SimExample}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
\caption{The cumulative probability of extinction due to climate change in different experimental scenarios. Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
\label{fig:ExtProb}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
\caption{Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations. Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions. In all panels, values associated with extant populations are shown in dark blue. Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations). Full results for all parameter combinations are provided in Appendix B.}
\label{fig:Disp}
\end{figure}

\clearpage

\subsection*{Online figures}

\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
\caption{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space. The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis. The lattice of discrete patches in which population dynamics occur is shown beneath. As described in the \textit{Submodels} section of the supplemental materials, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches. Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension. Landscapes are  unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
\label{Fig:EnvFunction}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
\caption{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal. Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice. After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
\label{Fig:LifeCycle}
\end{figure}

\clearpage

\renewcommand{\thefigure}{B\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
\caption{Extinction probabilities for a slow speed of climate change. Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
\label{Fig:ExtProbSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
\caption{Extinction probabilities for a fast speed of climate change. Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
\label{Fig:ExtProbFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a slow speed of climate change.}
\label{Fig:InitFitSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a moderate speed of climate change.}
\label{Fig:InitFit}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a fast speed of climate change.}
\label{Fig:InitFitFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
\caption{Observed dispersal evolution in populations responding to a slow speed of climate change. Positive values indicate an increase in average dispersal ability over the course of climate change. The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change. Positive values indicate an increase in average dispersal ability over the course of climate change. The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolMain}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
\caption{Observed dispersal evolution in populations responding to a fast speed of climate change. Positive values indicate an increase in average dispersal ability over the course of climate change. The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
\label{Fig:InitDispSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
\label{Fig:InitDispMain}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
\label{Fig:InitDispFast}
\end{figure}

\end{document}

### 2019-04-05 Manuscript/AmNat_ReSubmission.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

\documentclass[11pt]{article}
\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
\usepackage{fullpage}
\usepackage[authoryear,sectionbib,sort]{natbib}
\linespread{1.7}
\usepackage[utf8]{inputenc}
\usepackage{lineno}
\usepackage{titlesec}
\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}

\usepackage{mathptmx}
\usepackage{hyperref}
\usepackage{geometry}
\usepackage[centertags]{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{fancybox}
\usepackage{graphicx}
\usepackage{graphics}
\newcommand{\s}{^{(s)}}

%%%%%%%%%%%%%%%%%%%%%
% Line numbering
%%%%%%%%%%%%%%%%%%%%%
\usepackage{lineno}
% Please use line numbering with your initial submission and
% subsequent revisions. After acceptance, please turn line numbering
% off by adding percent signs to the lines %\usepackage{lineno} and
% to %\linenumbers{} and %\modulolinenumbers[3] below.

\title{Spatial population structure determines extinction risk in climate-induced range shifts}

% This version of the LaTeX template was last updated on
% January 11, 2018.

%%%%%%%%%%%%%%%%%%%%%
% Authorship
%%%%%%%%%%%%%%%%%%%%%
% Please remove authorship information while your paper is under review,
% unless you wish to waive your anonymity under double-blind review. You
% will need to add this information back in to your final files after
% acceptance.

%\author{Owen E. Cook$^{1,\ast}$ \\ 
%Generic H. Collaborator$^{2,\dag}$ \\ 
%Additional Q. Expert$^{3}$}

\date{}

\begin{document}

\maketitle

%\noindent{} 1. University of Chicago, Chicago, Illinois 60637;

%\noindent{} 2. University of Toronto, Toronto, Ontario M5S 1A5, Canada;

%\noindent{} 3. Middle Eastern Technical University, Çankaya, Ankara 06800, Turkey.

%\noindent{} $\ast$ Corresponding author; e-mail: amnat@uchicago.edu.

%\noindent{} $\dag$ Deceased.

\bigskip

\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1-A3,  tables~A1-A4, and figures~B1-B11). Figure~1 and figure~3 are to print in color.

\bigskip

\textit{Keywords}: range shifts, extinction, rapid evolution, dispersal evolution, individual-based model

\bigskip

\textit{Manuscript type}: Article. %Or e-article, note, e-note, natural history miscellany, e-natural history miscellany, comment, reply, invited symposium, or countdown to 150.

\bigskip

\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}

\linenumbers{}
\modulolinenumbers[3]

\newpage{}

\section*{Abstract}
Climate change is an escalating threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses. However, populations do not respond to climate change in isolation, but rather in the context of their existing ranges. In particular, spatial population structure within a range (e.g. trait clines, starkness of range edges, etc.) likely interacts with other ecological and evolutionary processes during climate-induced range shifts. Here, we use an individual-based model to explore the interacting roles of these factors in range shift dynamics. We show that increased spatial population structure (driven primarily by a steeper slope in the trait optimum gradient) severely increased a population's extinction risk. Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations. Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes defining the population at equilibrium (i.e. before the onset of rapid climate change); only populations consisting of highly dispersive individuals survived. Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that spatial population structure can substantially increase extinction risk in range shifts.

\newpage{}

\section*{Introduction}
Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}. These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}. Indeed, contemporary range shifts have already been observed in a wide variety of taxa, from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}. Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages~\citep{hobbs2009novel}. Understanding the ecological and evolutionary dynamics of such climate-induced range shifts will play a key role in informing current and future conservation work.

Large-scale population movements have been studied for decades in the context of range expansions (e.g. of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions. For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}. As fundamentally similar spatial processes, it is likely that range shifts will also be subject to the same ecological and evolutionary mechanisms known to drive range expansions. However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population. In particular, range shifts begin with far more complex spatial population structure compared to most range expansions. Here and throughout the paper, we use the phrase "spatial population structure" to refer broadly to the spatial distribution of individuals, and their associated genotypes and phenotypes, within a population. Thus, spatial population structure can encompass factors such as spatial clines in trait values and the starkness of abundance declines characterizing the range edge. Range expansions (e.g. of invasive or reintroduced species) typically start from small founding populations brought to a new area and lacking any initial spatial population structure. While such populations form spatial population structure during the expansion process~\citep{weiss2017rapid, ochocki2017rapid}, established populations respond to climate change in the context of preexisting spatial population structure.

Each factor relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions. For example, the underlying mechanism causing population declines at the range edge (e.g. declines in carrying capacity versus growth rate) can alter a population's extinction risk during climate driven range shifts~\citep{henry2013eco}. Further, in ranges characterized by a gradient in a trait optimum, the starkness of the range edge can impact the ability of peripheral populations to adapt to the local optimum, with stark range edges leading to better adaptation in peripheral populations~\citep{garcia1997genetic}, though the importance of this for range shifts has not been investigated. Additionally, the nature of dispersal can interact with adaptation to a trait optimum gradient to impact range shift dynamics. When dispersal occurs in a stepping stone manner, maladapted individuals (i.e. individuals whose phenotypes do not match the local optimum) can block the establishment of better adapted genotypes so long as the mismatch is not too severe~\citep{atkins2010local}. While these aspects of spatial population structure have been shown to impact the dynamics of climate-induced range shifts in isolation, it is unclear how and if they might interact.

Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts. In asexual species, for example, local adaptation to a trait optimum gradient has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}. However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow. Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation to a trait optimum gradient within a population due to increased gene flow throughout the range~\citep{garcia1997genetic, kirkpatrick1997evolution}. In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}. In addition to potential interactions between local adaptation and dispersal evolution, the starkness of the range edge could influence the potential for rapid trait evolution during range shifts by altering the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.

Here, we assess the interaction of two mechanisms responsible for spatial population structure with trait evolution in sexually reproducing populations undergoing range shifts. We develop an individual-based model capable of producing a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change. One trait determines dispersal ability while the second defines an individual's environmental niche. Using this model, we vary both the severity of the niche optimum gradient and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate. Additionally, by contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.

\section*{Methods}
A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here. Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig. A1). Landscapes were unbounded in the $x$ dimension but defined by a fixed width and wrapping boundaries in the $y$ dimension. The optimum environmental niche value changed linearly along the $x$ dimension, thus allowing for the intrinsic formation of stable range boundaries when the optimum changed rapidly enough~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}. However, to examine the dynamics of ranges in which the niche optimum did not change rapidly enough to form stable range limits, we additionally imposed extrinsic range limits to prevent continuous adaptation and spread of the population~\citep{alleaume2006geographical, garcia1997genetic}. Specifically, we systematically altered the decline in patch carrying capacities from the range core to the edge~\citep{alleaume2006geographical, henry2013eco, bocedi2014rangeshifter, mustin2009dynamics} in such a way that we could directly manipulate the starkness of the decline. Previous research has shown that using a decline in intrinsic growth rate as opposed to carrying capacity may impact extinction risk but does not alter the patterns of dispersal evolution during climate change~\citep{henry2013eco}. To maintain generality, we do not assume a specific mechanism behind the decline in carrying capacity, but it could represent a variety of range limiting mechanisms such as physiological limits to adaptation, the effects of competition, or underlying resource distributions~\citep{sexton2009evolution, holt2005theoretical, case2000interspecific}. Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions. To simulate climate change, the patch carrying capacities and the niche optimum gradient shifted at a constant rate along the $x$ dimension. Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig. A2).

Individuals were characterized by two traits (dispersal and an environmental niche), each defined by a set of $5$ quantitative diploid loci. While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits. The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel. An individual's realized dispersal distance was then drawn from the dispersal kernel and dispersal direction was random and unbiased. Dispersal occurred in continuous space from the center of an individual's current patch and the individual's new patch was then determined by the mapping from continuous space to discrete patches (see Appendix A). An individual's environmental niche value determined its fitness according to the local niche optimum. The closer an individual's niche value to the local optimum, the higher the individual's realized fitness. Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}, scaled by the mean fitness of the patch. Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with a close match of their niche value to the local optimum produced more offspring on average. Thus, the model used a mixture of  hard selection (realized population growth declined with maladaptation relative to the niche optimum) and soft selection (probability of producing offspring depended on fitness relative to other individuals) for the evolutionary dynamics~\citep{wallace1975hard}. Allele inheritance was subject to mutation and assumed no linkages among loci. The mutation process was designed such that mutational input per generation was independent of the number of loci (see Appendix A) and with parameters corresponding to previous estimates from the literature~\citep{gilbert2017local}.

We varied parameter values to explore the interacting roles of the slope of the niche optimum gradient and range edge starkness (Table A2) in forming spatial population structure at equilibrium and driving the subsequent eco-evolutionary dynamics of range shifts. Specifically, we considered a fully factorial combination of: (1) a flat, shallow, and steep niche optimum gradient, (2) shallow, moderate, and stark declines in carrying capacity at the range edge, and (3) slow, moderate, and fast speeds of climate change. This yielded a total of $27$ different scenarios, each explored with $200$ simulations. Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$ to reach a spatial equilibrium, followed by $100$ generations of climate change and a final $50$ generations of stable conditions. Figure 1 shows an example of a single population responding to a moderate speed of climate change. For each scenario, we evaluated the role of the equilibrium spatial population structure and dispersal evolution in the dynamics of the range shifting populations. We primarily discuss simulations assuming a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.

We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end. For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g. leading edge vs. core populations). Due to local extinctions, not all patches were occupied at the end of the period of climate change. To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals. Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype. All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).

\section*{Results}
In all scenarios, some populations shifted their ranges in response to climate change. However, the proportion of populations that failed to track the changing climate (and subsequently went extinct) depended on the spatial characteristics of the range. Populations defined by a steep niche optimum gradient and by stark declines in carrying capacity at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig. 2). While both aspects of a population's range influenced extinction probabilities, the niche optimum gradient drove more dramatic changes to extinction risk, with steeper gradients causing severe increases in the probability of extinction during climate change. We varied both parameters widely, doubling the slope of the niche optimum gradient from the shallow to steep scenario and increasing the parameter defining range edge starkness by a factor of $100$ from shallow to stark edges (Table A2). This suggests that the niche optimum gradient may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios. Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig. B1 \& B2). However, this effect was independent of the niche optimum gradient and range edge starkness in determining extinction probability during range shifts.

In accordance with previous results~\citep{garcia1997genetic}, populations at the range edge had lower fitness than central populations at equilibrium and this effect was amplified with more gradual range edges. Counterintuitively, populations that survived climate change tended to be characterized by even greater reductions in fitness at the range edges at equilibrium compared to populations that went extinct (Fig. B3-B5). While discernible in all simulations with a non-zero slope in the niche optimum gradient, this pattern was most evident in the scenarios defined by a gradual range edge. As expected, there was no spatial variation in fitness for populations with no variation in the niche optimum across space. Despite the spatial variation in fitness in some scenarios, variance in relative fitness within a patch was relatively low (about $0.3$ across all scenarios). This implies that (1) the reduction in fitness at the edge caused by the niche optimum gradient was relatively uniform across all individuals, and (2) as a result, there was relatively low variance in reproductive success in these populations, meaning that evolution at the edge was not driven by only a handful of higher fitness individuals.

Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate. While some individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g. Fig. 1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig. 3a\&b). Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig. B3-B5). Further, calculating the coefficient of variation (CV) in dispersal genotypes over the $y$ dimension, in which environmental conditions did not vary, revealed that edge populations had between $3$ and $4$ times higher CVs than core populations across all scenarios, indicating genetic diversity at the edge was not a limiting factor in dispersal evolution. The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.

Instead, the distribution of dispersal phenotypes at equilibrium played a key role in determining a population's fate. A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the niche optimum gradient and range edge starkness (Fig. B9-B11). Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig. 3c\&d). Previous research has demonstrated that defining the range edges via a decline in the intrinsic growth rate (as opposed to carrying capacity as done here) resulted in less dispersive phenotypes at the range edge~\citep{henry2013eco}, meaning extinction risks would be even higher under such a scenario. Comparing the full distribution of equilibrium dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes just from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations. Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig. B9-B11). To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{fisher1937wave, hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change in our simulations (see the model description in Appendix A). This estimated dispersal phenotype matched the observed threshold value distinguishing surviving from extinct populations in all experimental scenarios (Figures B9-B11, vertical dashed line). Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change. However, this threshold effect weakened slightly in scenarios with a slow speed of climate change, indicating a potential role for dispersal evolution if the climate were to change at a slow enough rate (Figure B11).

\section*{Discussion}
Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts. We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci. In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g. single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).  Here, we tested the generality of previous results predicting an important role for dispersal evolution in range shifts~\citep{boeye2013more, henry2013eco} and the interplay of eco-evolutionary dynamics under increased levels of biological complexity. Specifically, we demonstrated the role of spatial population structure, driven by a niche optimum gradient and range edge starkness, in determining extinction risk for range shifting populations via impacts on the equilibrium distribution of dispersal phenotypes and environmental niche values.

Our results suggest that populations characterized by local adaptation to a spatially varying trait optimum and by stark range edges will be less able to track changing climatic conditions (Fig. 2). A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}. While it is difficult to exactly map adaptation to the niche optimum gradient to empirical measures of local adaptation, the parameters defining the steepest gradient used here resulted in the intrinsic formation of stable range boundaries, as seen in previous theoretical studies~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}, suggesting they provide reasonable approximations of empirical patterns. Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the starkest range edges imposed in our study. While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to their spatial population structure.

Our results emphasize the importance of the equilibrium distribution of dispersal phenotypes (i.e. prior to the onset of rapid climate change) in determining a population's extinction risk during climate change (Fig. 3c\&d). Populations primarily composed of high dispersal phenotypes at equilibrium successfully tracked changing climate conditions, while populations of lower dispersal phenotypes lagged behind the changing conditions to eventually go extinct. Importantly, the threshold dividing the equilibrium dispersal phenotypes of successful and extinct populations was constant across all scenarios for a given speed of climate change (Fig. B9-11). Thus, the difference in survival probability among scenarios was driven by the effects of the niche optimum gradient and range edge starkness on the evolution of dispersal ability throughout the range. Scenarios with a steep niche optimum gradient selected for lower dispersal phenotypes at equilibrium due to the potential mismatch of dispersing individuals' niche phenotype and their new location~\citep{kirkpatrick1997evolution}. Similarly, a stark range edge selected for lower dispersal due to the risk of dispersing into unsuitable habitat~\citep{shaw2014population, shaw2019evolution}. Previous research has documented a similar reduction in dispersal phenotypes due to an explicit mortality cost of dispersal~\citep{kubisch2013predicting}, whereas the costs to dispersal in our model result from the niche optimum gradient and range edge starkness. Such a cost to dispersal results in lower dispersal phenotypes at equilibrium, hampering a population's ability to successfully track a changing climate. Importantly, dispersal evolution during climate change was unable to counter the influence of equilibrium spatial population structure on extinction dynamics.

While high dispersal phenotypes at equilibrium increased the probability that populations tracked changing conditions, they had the additional effect of reducing average fitness at the range edges when the niche optimum varied through space (Fig. B3-B5). In the model, range edge populations tended to have lower abundance than core populations, increasing their susceptibility to gene flow from the core~\citep{kirkpatrick1997evolution, garcia1997genetic}. Thus, in populations with high dispersal phenotypes at equilibrium, increased gene flow from the core reduced fitness at the range edge via gene swamping~\citep{lenormand2002gene}. As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges at equilibrium. While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that high fitness in edge populations may be a warning sign of future difficulty in tracking climate change when the population is structured along a spatial gradient in a trait optimum. For the purposes of this investigation, we assumed the niche optima shifted spatially with climate change, as would be expected if the niche optima corresponded to local temperature or precipitation conditions~\citep{davis2001range}. However, in systems characterized by a niche optimum gradient defined by other factors (e.g. geography or biotic interactions) the gradient might remain stable or even shift in an opposing direction to climate change. Future research should investigate the impact of a niche optimum gradient unrelated to climate on extinction risk during range shifts.

Previous research has suggested that evolution of increased dispersal ability during climate change may be capable of rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more, henry2013eco}. Our results suggest this is not always the case. In fact, it may only be possible under certain, relatively narrow conditions. Previous models showing that dispersal evolution may rescue range shifting populations have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}. As dispersal evolution during range expansions and shifts occurs via the spatial sorting of dispersal alleles throughout the range~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait. Additionally, the interaction of mutation and genetic architecture in different models (e.g. few mutations of large effects or many mutations of small effects) undoubtedly plays a role in dispersal evolution during range shifts. Increasing mutation rate or effect size might have the equivalent effect of a slower speed of climate change in allowing dispersal evolution to play a greater role in range shift dynamics. Further, life history has been shown to impact the maintenance of genetic diversity, and hence evolutionary potential, with stage and age-structured populations shown to harbor greater diversity then populations with non-overlapping generations as modeled here~\citep{ellner1996environmental}. Populations defined by more complex life histories might, therefore, contain more genetic diversity in dispersal at equilibrium, making evolution of increased dispersal during range shifts more likely. Thus, further research is needed to understand how factors such as genetic architecture, mutational dynamics, and life history might interact to shape the potential for population rescue via dispersal evolution during range shifts.

\section*{Conclusion}
As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions. In particular, a deeper, process-based understanding of extinction risk in range shifting populations will, in turn, allow more focused conservation interventions. Our results suggest that spatial population structure, as determined by the niche optimum gradient and range edge starkness, has the potential to dramatically alter extinction risk for species responding to climate change. Further, in contrast to other studies assuming more simplified genetic architecture, we find very little role for the evolution of heightened dispersal in allowing a population to successfully track climate change. Future work should continue to examine the circumstances determining the potential for rescue via dispersal evolution in range shifts. As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to identify those factors leading to increased extinction risk in range shifting populations, and use that knowledge to develop meaningful conservation strategies to mitigate such risk.

%%%%%%%%%%%%%%%%%%%%%
% Acknowledgments
%%%%%%%%%%%%%%%%%%%%%
% You may wish to remove the Acknowledgments section while your paper 
% is under review (unless you wish to waive your anonymity under
% double-blind review) if the Acknowledgments reveal your identity.
% If you remove this section, you will need to add it back in to your
% final files after acceptance.

%\section*{Acknowledgments}

\newpage{}

\section*{Appendix A: Full model description}

% Please reset counters for the appendix (thus normally figure A1, 
% figure A2, table A1, etc.).

% In certain cases, it may be appropriate to have a PRINT appendix in
% addition to (or instead of) an online appendix. In this case, please 
% name the print appendix Appendix A, and any subsequent appendixes (if 
% there are any) should be named Online Appendix B, Online Appendix C,
% etc.

% Counters for each appendix should match the letter of that appendix.
% For example, tables in Appendix C should be numbered table C1, table C2,
% etc. This applies to tables, equations, and figures.

% It's better not to use the \appendix command, because we have some
% formatting peculiarities that \appendix conflicts with.

\renewcommand{\theequation}{A\arabic{equation}}
% redefine the command that creates the equation number.
\renewcommand{\thetable}{A\arabic{table}}
\setcounter{equation}{0}  % reset counter 
\setcounter{figure}{0}
\setcounter{table}{0}

\section*{Model overview}
\subsection*{Purpose} 
This model tested an evolving population's ability to track a changing climate under a variety of conditions. Specifically, populations were simulated under different combinations of (1) the slope of the niche optimum gradient and (2) the starkness of the range edge. In all simulations, an individual's expected dispersal distance and environmental niche were defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time. All simulations began with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change. Climate change was then modeled as a constant, directional shift in environmental conditions (see \textit{Submodels} below). Finally, simulations ended with another short period of climate stability to assess a population's ability to persist and recover after shifting its range.

\subsection*{State variables and scales} 
The model simulated a population of males and females characterized by diploid loci for both their expected dispersal distance and environmental niche. Space was modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system. Landscapes were two dimensional with a fixed width along the $y$ dimension and without bounds on the $x$ dimension. Environmental conditions varied along the $x$ dimension but remained constant within the $y$ dimension. To avoid edge effects due to the fixed width of the $y$ dimension, the model employed wrapping boundaries such that if an individual dispersed out of the landscape on one side, it would appear at opposite side of the $y$ dimension, but at the same $x$ coordinate. Patches were defined by the location of the patch center in $x$ and $y$ coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (\textit{Submodels}).

The model implemented climate change by shifting the location of patch carrying capacities and the niche optimum gradient along the $x$ dimension of the landscape. Patch carrying capacities were defined by the location of the range center along the $x$ dimension, the starkness of the decline characterizing the range edges, and the width of the range along the $x$ dimension (See Figure A1). The niche optimum gradient was linear and shifted at the same speed and in the same direction as carrying capacities during climate change.

[Figure A1 goes here]

\subsection*{Process overview and scheduling} 
Time was modeled in discrete intervals defining single generations of the population (Fig. A2). Within each generation, individuals first dispersed from their natal patches according to their phenotypes. After dispersal, reproduction occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch. Reproduction occurred via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) were likely to produce multiple offspring while individuals with low relative fitness might not produce any. Individuals inherited one allele from each parent at each loci, assuming independent segregation and a mutation process. After reproduction, all individuals in the current generation perished and the offspring began the next generation with dispersal, resulting in discrete, non-overlapping generations.

[Figure A2 goes here]

\section*{Design concepts}
\subsection*{Emergence} 
Emergent phenomena in this model included the spatial distribution of population abundance, dispersal abilities, and relative fitness throughout the range (i.e. the spatial population structure). Additionally, the population dynamics during the range shift, including the extinction process, and the evolutionary trajectories of the dispersal and niche traits were all emergent phenomena in this model.

\subsection*{Stochasticity} 
All biological processes in this model were stochastic, including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci. Environmental parameters were fixed, however, and the process of climate change (i.e. the movement of patch carrying capacity through time) was deterministic. Thus, the model removed the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.

\subsection*{Interactions} 
Individuals in the model interacted via mating and density-dependent competition within patches. Additionally, the evolutionary trajectories of the two different traits had the potential to interact via the relationship between gene flow (dispersal trait) and local adaptation (niche trait). Further, the niche optimum gradient and range edge starkness could interact with trait evolution both during stable climate conditions and during climate change.

\subsection*{Desired output} 
After each model run, full details of all surviving individuals at the last time point were recorded (spatial coordinates and loci values for both traits). If a population went extinct during the model run, the time of extinction was recorded. For each occupied patch throughout the simulation, we aggregated data on population size, the dispersal trait, and adaptation to local conditions.

\section*{Details}
\subsection*{Initialization} 
The following parameters were set at the beginning of each simulation and formed the initial conditions of the model: the mean and variance for allele values of each trait, initial population size, location of the range center, number of generations for the pre-, post-, and rapid climate change periods of the simulation, and all other necessary parameters for the submodels defined below. Simulated populations were initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions. This ensured that the populations reacting to a changing climate truly represented the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation (Table A1). Initial population size was chosen to minimize the risk of stochastic extinction in the early stages of the simulation. The time frames defining climate change were designed to give a reasonable period for the population to reach a spatial equilibrium and a long enough period of climate change for extinction dynamics to play out. The number of patches defining the $y$ dimension and the relationship between Cartesian space and discrete patches were chosen to allow a reasonable number of patches to contribute to the eco-evolutionary dynamics of range shifts while not proving computationally restrictive.

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{4cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ individuals \\
    $\beta_{1}$ & Center of the range during stable climate conditions & $0$ \\
    $\hat{t}$ & Duration of stable climate conditions & $2000$ generations \\
    $t_{\Delta}$ & Duration of climate change & $100$ generations \\
    $t_{max}$ & Total number of generations in the simulation & $2150$ generations \\
    $\eta$ & Width of square habitat patches in Cartesian space & $50$ \\
    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ dimension & $10$ patches \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters determining the initial conditions of simulations, the timing of climate change, and the relationship between Cartesian space and the lattice of discrete habitat patches.}
\label{table:InitPars}
\end{table}

\subsection*{Submodels}
\paragraph{Patch carrying capacities}
Patch carrying capacity ($K_{x}$) varied along the $x$ dimension of the landscape, attaining its highest value at the range center and declining with distance from the center. Specifically, the carrying capacity at a location $x$ was defined as the product of the maximum potential carrying capacity ($K_{max}$) and a function $f(x,t)$, where $f(x,t)$ was bounded between $1$ and $0$ with its highest value corresponding to the range center. $f(x,t)$ was defined as
\begin{equation}
f(x,t)=
\begin{cases}
	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
\end{cases}
\end{equation}
where $\beta_{t}$ defined the center of the range at time $t$, $\tau$ affected the width of the range, and $\gamma$ affected the slope of the function at the range edges (See Figure A1). Population dynamics occurred within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we used another parameter defining the spatial scale of each patch ($\eta$). The local carrying capacity of a patch centered on $x$ ($K_{x}$) was then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
\begin{equation}
K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
\end{equation}

To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range edge, we calculated the partial derivative of $f(x,t)$ over the $x$ dimension as
\begin{equation}
\frac{\partial f(x,t)}{\partial x}=
\begin{cases}
	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
\end{cases}	
\end{equation}
yielding a value of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$). Thus, altering $\gamma$ directly altered the range edge starkness. However, changing $\gamma$ also changed the total area under $f(x,t)$ as can be seen in the indefinite integral of $f(x,t)$:
\begin{equation}
\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
\end{equation}
Thus, ranges defined by different $\gamma$ values could also result in different range-wide carrying capacities, potentially altering both the ecological (e.g. through stochastic extinction events) and evolutionary (e.g. through more mutations arising in larger populations) dynamics of the ranges. Additionally, different combinations of $\gamma$ and $\tau$ could result in different range widths, which have been shown to impact dispersal evolution within the ranges~\citep{van1997integrodifference}. To control for these confounding factors, we fixed the range widths for all scenarios and altered $K_{max}$ to maintain a constant range-wide carrying capacity. Specifically, we defined the range width using the $x$ coordinates at which $f(x,t)$ fell below $0.1$ on either side of $\beta_{t}$ and chose $\tau$ and $\gamma$ values for each scenario such that $f(x,t)$ fell below $0.1$ at the same $x$ coordinates (Table A2). We then adjusted $K_{max}$ for each scenario so that the range-wide carrying capacity was constant (Fig. A3).

[Figure A3 goes here]

\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm}  | p{1.5cm} }
    \hline
    Range edge starkness & Niche optimum gradient & $\gamma$ & $\tau$ & $\lambda$ & $K_{max}$ \\ \hline \hline
     & Flat & $0.0025$ & $-240$ & $0$ & $240$ \\
    Shallow & Shallow & $0.0025$ & $-240$ & $0.004$ & $240$ \\
     & Steep & $0.0025$ & $-240$ & $0.008$ & $240$ \\ \hline
     & Flat & $0.0075$ & $345.9$ & $0$ & $118.1$ \\
    Moderate & Shallow & $0.0075$ & $345.9$ & $0.004$ & $118.1$ \\
     & Steep & $0.0075$ & $345.9$ & $0.008$ & $118.1$ \\ \hline
     & Flat & $0.25$ & $630.1$ & $0$ & $66.7$ \\
    Stark & Shallow & $0.25$ & $630.1$ & $0.004$ & $66.7$ \\
     & Steep & $0.25$ & $630.1$ & $0.008$ & $66.7$ \\ 
    \hline
  \end{tabular}
\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios. As defined in the text, $\gamma$ affects range edge starkness, $\tau$ affects the range width, $\lambda$ is the slope of the niche optimum gradient, and $K_{max}$ is the maximum carrying capacity for patches in the landscape.}
\label{table:Scenarios}
\end{table}

Thus, $\gamma$ and $\tau$ were both fixed within a given simulation and $\beta_{t}$ (the location of the range center) was used to simulate climate change. During the periods before and after climate change $\beta_{t}$ was constant, but to simulate climate change it varied with time as follows
\begin{equation}
\beta_{t}=\nu\eta(t-\hat{t})
\end{equation}
where $\nu$ was the velocity of climate change per generation in terms of discrete patches, $t$ was the current generation, and $\hat{t}$ was the last generation of stable climatic conditions before the onset of climate change.

\paragraph{Niche optimum}
The niche optimum ($z_{opt,x}$) varied in space according to
\begin{equation}
z_{opt,x}=\lambda(x-\beta_{t})
\end{equation}
with $\lambda$ determining the rate of change in the optimum across the range. Individual relative fitness ($w_{i,x}$) values were then calculated according to the following equation assuming stabilizing selection
\begin{equation}
w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
\end{equation}
where $\omega$ defined the strength of stabilizing selection and $z_{i}$ was an individual's niche phenotype~\citep{lande1976natural}. Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the environmental optimum of the patch it occupied ($z_{opt,x}$). All loci were assumed to contribute additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values. As defined above, $z_{opt,x}$ also shifted with climate change (i.e. with $\beta_{t}$) as would be expected if it corresponded to a phenotypic optimum along a temperature or precipitation gradient within the range~\citep{davis2001range}.

\paragraph{Population dynamics}
Population growth within each patch was modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}. To account for fitness effects on population growth, expected population growth was scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$) so that maladaptation resulted in reduced population growth. The expected number of new offspring in patch $x$ at time $t+1$ was given by
\begin{equation}
\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
\end{equation}
where $F_{t,x}$ was the number of females in patch $x$ at time $t$, $R$ was the intrinsic growth rate for the population and remained constant in both time and space, $\psi$ was the expected sex ratio of the population, $N_{t,x}$ was the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ was the local carrying capacity based on the environmental conditions. To incorporate demographic stochasticity, the realized number of offspring for each patch was then drawn from a Poisson distribution.
\begin{equation}
N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
\end{equation}

Offspring parentage was assigned by random sampling of the local male and female populations (i.e. polygynandrous mating assuming a well-mixed population within each patch). The sampling was weighted by individual fitness and occurred with replacement so highly fit individuals were likely to have multiple offspring while low fitness individuals might not have had any. Each offspring inherited one allele per locus from each parent, assuming no linkage among loci. After reproduction, all members of the previous generation died and the offspring dispersed to begin the next generation. Parameters governing population dynamics (Table A3) were chosen to yield reasonable rates of population growth based on initial exploratory simulations.

% Population dynamics A3
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $R$ & Intrinsic growth rate of the population & $2$ \\
    $\psi$ & Expected sex ratio (proportion of females) in the population & $0.5$ \\
    $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters related to population growth and dispersal.}
\label{table:PopPars}
\end{table}

\paragraph{Mutation}
Inherited alleles were subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents. The mutation process was defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$). Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation was
\begin{equation}
\frac{U^{T}}{2L^{T}}
\end{equation}
Effect sizes of mutations were drawn from a normal distribution with mean $0$ and a standard deviation of
\begin{equation}
\sqrt{V_{m}^{T}U^{T}}
\end{equation}
meaning the ratio of small effect to large effect mutations depended on both $U^{T}$ and $V_{m}^{T}$. We chose parameter values (Table A4) in keeping with previously derived values from the literature~\citep{gilbert2017local}. For the number of loci used in our simulations, these resulted in mostly mutations of small effect with few large effect mutations. Importantly, by defining the mutation process in this way, rather than with a per locus probability of mutation and a mutation effect size directly, the mutational input per generation was kept constant regardless of the number of loci defining the trait~\citep{schiffers2014landscape}.

% Genetics A4
\begin{table}
\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
    \hline
    Parameter & Description & Value \\ \hline \hline
    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ \\
    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ \\
    $L^{T}$ & Number of diploid loci defining each trait & $5$ loci \\
    $\mu_{1}^{f}$ & Initial mean allele value for the niche trait & $0$ \\
    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the niche trait & $0.025$ \\
    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
    \hline
  \end{tabular}
\caption[LoF entry]{Values and descriptions for parameters defining the genetic components of the model.}
\label{table:GenPars}
\end{table}

\paragraph{Dispersal}
Finally, individuals dispersed according to an exponential dispersal kernel defined by each individual's dispersal phenotype. An individual's dispersal phenotype was the expected dispersal distance and was given by
\begin{equation}
d_{i} = \frac{\hat{d}\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
\end{equation}
where $\hat{d}$ was the maximum expected dispersal distance in terms of discrete patches, $\rho$ was a constant determining the slope of the transition between $0$ and $\hat{d}$, and the summation was taken across all alleles contributing to dispersal. Thus, as with fitness, loci were assumed to contribute additively with no dominance or epistasis. The expected dispersal distance, $d_{i}$ was then used to draw a realized distance from an exponential dispersal kernel. The direction of dispersal (in radians) was drawn from a uniform distribution bounded by $0$ and $2\pi$. If a dispersal trajectory took an individual outside the bounds of the landscape in the $y$ dimension, the individual reappeared at the same $x$ coordinate but the opposite end of the $y$ dimension, thus wrapping the top and bottom edges of the landscape to avoid edge effects. Dispersal occurred from the center of each patch and the individual's new patch was then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure A1). Dispersal parameters (Table A3) were chosen to allow a wide range of dispersal phenotypes to evolve in the context of the different experimental scenarios, ranging from highly restrictive to long-distance dispersal.

Since the dispersal phenotype was the expected value of the exponential dispersal kernel, it could be used directly to calculate the two-dimensional diffusion coefficient of population spread ($D$). Specifically, since $d_{i}^{2}$ represented the mean squared displacement of an individual with dispersal phenotype $d_{i}$, the two-dimensional diffusion coefficient could be calculated as
\begin{equation}
D = \frac{1}{4}d_{i}^{2}
\end{equation}
and subsequently used to calculate the approximate speed of an expansion wave defined by that dispersal phenotype.

\newpage{}

\section*{Appendix B: Supplementary results for varying speeds of climate change}

\renewcommand{\theequation}{B\arabic{equation}}
% redefine the command that creates the equation number.
\renewcommand{\thetable}{B\arabic{table}}
\setcounter{equation}{0}  % reset counter 
\setcounter{figure}{0}
\setcounter{table}{0}

\section*{Extinction probability}
As in the main text, we calculated the cumulative probability of extinction for both slow and fast speeds of climate change. The figures in this section use the same layout and line types as Figure 2 in the main text to allow for direct comparisons.

[Figures B1\&B2 go here.]

\section*{Equilibrium fitness throughout the landscape}
To assess trends in realized fitness values throughout the landscape, we calculated patch-level mean individual fitness for each landscape at equilibrium. To simplify the figures, we averaged over the $y$ dimension in which the CV in niche genotypes was minimal throughout the range for all scenarios (typically below $0.25$) due to the constant environmental conditions. Populations at the range edge were characterized by reduced fitness compared to core populations, as found in previous models~\citep{garcia1997genetic}. However, this trend was exacerbated in populations that successfully tracked climate change compared to those that went extinct, especially in the presence of gradual range edges and steep niche optimum gradients. As realized fitness values do not vary spatially in simulations with no gradient in the niche optimum, the following figures only show results for scenarios with a shallow or steep gradient.

[Figures B3-B5 go here.]

\section*{Dispersal evolution}
Using the same metric of dispersal evolution from the main text (change in average phenotype for each patch), we display here the observed dispersal evolution over the course of climate change for all experimental scenarios. Each histogram in the following figures represents a single experimental scenario as indicated by the figure text. The lower left panel and upper right panel from Figure B7 are the same histograms shown in Figure 3a\&b, but are here placed in the context of all other experimental scenarios.

[Figures B6-B8 go here.]

\section*{Equilibrium dispersal phenotypes}
Here, we present histograms of the equilibrium distribution of dispersal phenotypes to demonstrate the importance of those phenotypes in determining population success or extinction during climate change. Dispersal phenotypes are log transformed for easier comparison. As with the dispersal evolution section, each histogram represents a single experimental scenario as indicated by the figure text. Similarly, the lower left panel and upper right panel from Figure B10 are the same histograms shown in Figure 3c\&d, but are here placed in the context of all other experimental scenarios. All histograms additionally have a vertical dashed line indicating the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change in each simulation. This value serves as a threshold to distinguish individuals from ultimately successful versus extinct populations.

[Figures B9-B11 go here.]

\newpage{}

%%%%%%%%%%%%%%%%%%%%%
% Bibliography
%%%%%%%%%%%%%%%%%%%%%
% You can either type your references following the examples below, or
% compile your BiBTeX database and paste the contents of your .bbl file
% here. The amnatnat.bst style file should work for this---but please
% let us know if you run into any hitches with it!
% The list below includes sample journal articles, book chapters, and
% Dryad references.

\bibliographystyle{amnat}
\bibliography{main_bib}

\newpage{}

\section*{Figures}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
\caption{A single example of a simulation with a steep niche optimum gradient and a moderately stark range edge. Information on the (a) abundance, (b) dispersal ability, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change. Log transformed mean dispersal phenotypes (b) are shown for each patch. Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the niche optimum for each patch.}
\label{fig:SimExample}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
\caption{The cumulative probability of extinction due to a moderate speed of climate change in different experimental scenarios. Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by shallow (solid line), moderate (dashed line), or stark (dotted line) edges. In all graphs, a horizontal grey line shows $\%100$ extinction.}
\label{fig:ExtProb}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
\caption{Patterns in the evolution and the equilibrium distribution of the dispersal trait, highlighting extant simulations from a moderate speed of climate change. Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end. Positive values indicate an increase in average dispersal ability in the patch. Equilibrium distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions. In all panels, values associated with extant populations are shown in dark blue. Results are shown for populations with a flat niche optimum gradient and a gradual range edge (a and c; $n = 155$ extant populations) and for populations with a steep niche optimum gradient and a stark range edge (b and d; $n = 14$ extant populations). Full results for all parameter combinations are provided in Appendix B.}
\label{fig:Disp}
\end{figure}

\clearpage

\subsection*{Online figures}

\renewcommand{\thefigure}{A\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
\caption{Example visualization of $f(x,t)$ in Cartesian space. The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis. Specifically, $\beta_{t}$ defined the range center, $\gamma$ determined the slope of $f(x,t)$ at the inflection points (i.e. the range edges), and $\tau$ determined the location of the inflection points. The lattice of discrete $\eta$ x $\eta$ patches in which population dynamics occurred is shown beneath. As described in the \textit{Submodels} section of Appendix A, $f(x,t)$ determined the carrying capacity of the patches along the $x$ dimension of the lattice while carrying capacity remained constant within each column along the $y$ dimension. Landscapes were unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
\label{Fig:EnvFunction}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
\caption{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal. Each generation began with new offspring dispersing according to their phenotype, after which reproduction occurred in local populations defined by the discrete lattice. After reproduction, all parental individuals perished, resulting in discrete, non-overlapping generations.}
\label{Fig:LifeCycle}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/Discretization"}
\caption{The carrying capacity of discrete patches along the $x$ dimension of landscapes. From top to bottom, the plots show the carrying capacities for gradual, moderate, and stark range edges. Points represent the carrying capacity of a discrete $\eta$ x $\eta$ patch in the range. The vertical dashed lines indicate the $x$ coordinates at which $f(x,t)$ declines below $0.1$ and the $\gamma$ value for each plot is listed above.}
\label{Fig:LifeCycle}
\end{figure}

\clearpage

\renewcommand{\thefigure}{B\arabic{figure}}
\setcounter{figure}{0}

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
\caption{The cumulative probability of extinction due to a slow speed of climate change in different experimental scenarios. Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) edge. In all graphs, a horizontal grey line shows $\%100$ extinction.}
\label{Fig:ExtProbSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
\caption{The cumulative probability of extinction due to a fast speed of climate change in different experimental scenarios. Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) edge. In all graphs, a horizontal grey line shows $\%100$ extinction.}
\label{Fig:ExtProbFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a slow speed of climate change.}
\label{Fig:InitFitSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a moderate speed of climate change.}
\label{Fig:InitFit}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium. Points represent the mean across simulations and error bars are interquartile ranges. Population status (extinct or successful) was determined for a fast speed of climate change.}
\label{Fig:InitFitFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
\caption{Observed dispersal evolution in populations responding to a slow speed of climate change. Positive values indicate an increase in average dispersal ability during climate change. The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change. Positive values indicate an increase in average dispersal ability during climate change. The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolMain}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
\caption{Observed dispersal evolution in populations responding to a fast speed of climate change. Positive values indicate an increase in average dispersal ability during climate change. The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. The experimental scenario corresponding to each histogram is indicated on the figure.}
\label{Fig:DispEvolFast}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in equilibrium populations. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
\label{Fig:InitDispSlow}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in equilibrium populations. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
\label{Fig:InitDispMain}
\end{figure}

\clearpage

\begin{figure}[h!]
\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
\caption{Distributions of the dispersal phenotypes observed in equilibrium populations. Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner. Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
\label{Fig:InitDispFast}
\end{figure}

\end{document}

### 2019-04-05 Manuscript/AppendixA.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

--- 2019-04-05 Manuscript/AmNat_InitialSubmission.tex
@@
-\documentclass[11pt]{article}
-\usepackage[sc]{mathpazo} %Like Palatino with extensive math support
-\usepackage{fullpage}
-\usepackage[authoryear,sectionbib,sort]{natbib}
-\linespread{1.7}
-\usepackage[utf8]{inputenc}
-\usepackage{lineno}
-\usepackage{titlesec}
-\titleformat{\section}[block]{\Large\bfseries\filcenter}{\thesection}{1em}{}
-\titleformat{\subsection}[block]{\Large\itshape\filcenter}{\thesubsection}{1em}{}
-\titleformat{\subsubsection}[block]{\large\itshape}{\thesubsubsection}{1em}{}
-\titleformat{\paragraph}[runin]{\itshape}{\theparagraph}{1em}{}[. ]\renewcommand{\refname}{Literature Cited}
-
-\usepackage{mathptmx}
+\documentclass[12pt, oneside]{article}
+\usepackage{geometry}                		
+\geometry{letterpaper}                   		
+\usepackage{graphicx}
+\usepackage{graphics}
 \usepackage{hyperref}
-\usepackage{geometry}
+\usepackage{fancybox}
 \usepackage[centertags]{amsmath}
@@
 \usepackage{amsthm}			
-\usepackage{fancybox}
-\usepackage{graphicx}
-\usepackage{graphics}
+\usepackage{natbib}				
+\usepackage{fullpage}
+\usepackage{placeins}
+\usepackage{setspace}
+\usepackage{lineno}
+\usepackage{color}
+\usepackage{authblk}
+\usepackage{mathptmx}
+
+\DeclareRobustCommand{\firstsecond}[2]{#1}
+
+\newcommand{\mb}{\mathbf}
+\newcommand{\bs}{\boldsymbol}
+\newcommand{\wt}{\widetilde}
 \newcommand{\s}{^{(s)}}
 
-%%%%%%%%%%%%%%%%%%%%%
-% Line numbering
-%%%%%%%%%%%%%%%%%%%%%
-\usepackage{lineno}
-% Please use line numbering with your initial submission and
-% subsequent revisions. After acceptance, please turn line numbering
-% off by adding percent signs to the lines %\usepackage{lineno} and
-% to %\linenumbers{} and %\modulolinenumbers[3] below.
-
-\title{Spatial population structure determines extinction risk in climate-induced range shifts}
-
-% This version of the LaTeX template was last updated on
-% January 11, 2018.
-
-%%%%%%%%%%%%%%%%%%%%%
-% Authorship
-%%%%%%%%%%%%%%%%%%%%%
-% Please remove authorship information while your paper is under review,
-% unless you wish to waive your anonymity under double-blind review. You
-% will need to add this information back in to your final files after
-% acceptance.
-
-%\author{Owen E. Cook$^{1,\ast}$ \\ 
-%Generic H. Collaborator$^{2,\dag}$ \\ 
-%Additional Q. Expert$^{3}$}
+\title{Appendix A: Full model description}
 
@@
 
+\author[1]{Christopher Weiss-Lehman}
+\author[1]{Allison K. Shaw}
+
+\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}
+
 \begin{document}
-
 \maketitle
 
-%\noindent{} 1. University of Chicago, Chicago, Illinois 60637;
+\doublespacing
+\linenumbers
 
-%\noindent{} 2. University of Toronto, Toronto, Ontario M5S 1A5, Canada;
-
-%\noindent{} 3. Middle Eastern Technical University, Çankaya, Ankara 06800, Turkey.
-
-%\noindent{} $\ast$ Corresponding author; e-mail: amnat@uchicago.edu.
-
-%\noindent{} $\dag$ Deceased.
-
-\bigskip
-
-\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1\&A2,  tables~A1\&A2, and figures~B1-B11). Figure~1 and figure~3 are to print in color.
-
-\bigskip
-
-\textit{Keywords}: range shifts, eco-evolutionary dynamics, local adaptation, individual-based model
-
-\bigskip
-
-\textit{Manuscript type}: Article. %Or e-article, note, e-note, natural history miscellany, e-natural history miscellany, comment, reply, invited symposium, or countdown to 150.
-
-\bigskip
-
-\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}
-
-\linenumbers{}
-\modulolinenumbers[3]
-
-\newpage{}
-
-\section*{Abstract}
-Climate change is an increasingly severe threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses.
-Population dynamics of range shifts, among the most commonly observed responses to climate change, can be influenced by many factors, including evolution of key traits, the degree of local adaptation, and the nature of the range edge.
-Here, we use an individual-based model to explore the interacting roles of these factors in the dynamics of climate-induced range shifts.
-We show that aspects of the spatial population structure within the initial range, in particular the potential for local adaptation, severely increased a population's extinction risk.
-Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations.
-Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes that evolved within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
-Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.
-
-\newpage{}
-
-\section*{Introduction}
-Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}.
-These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
-Indeed, contemporary range shifts have already been observed in a wide variety of taxa, ranging from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages~\citep{hobbs2009novel}.
-Understanding the ecological and evolutionary dynamics of such climate-induced range shifts will play a key role in informing current and future conservation work.
-
-Large-scale population movements have been studied for decades in the context of range expansions (e.g.
-of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
-As fundamentally similar spatial processes, it is likely that range shifts will also be subject to these ecological and evolutionary mechanisms known to drive range expansions.
-However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population.
-In particular, range shifts occur in populations with far more complex spatial structure compared to most range expansions, which typically begin from the successful establishment and spread of a small, founding population~\citep{hastings2005spatial}.
-While these founding populations often lack any significant spatial structure, populations undergoing range shifts are characterized by a spatial population structure formed by aspects of the previously stable ranges.
-For example, population ranges can vary in their potential for local adaptation throughout the range, the nature of the range edge, and the spatial distribution of key traits.
-
-Each of these factors relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions.
-For example, the underlying mechanism responsible for the gradient in population size from the range core to the edge (i.e.
-declines in carrying capacity versus growth rate) alters a population's extinction risk during climate driven range shifts~\citep{henry2013eco}.
-Additionally, the potential for local adaptation throughout a range has been related to extinction risk during range shifts.
-Specifically, a low potential for local adaptation can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block the establishment of better adapted genotypes~\citep{atkins2010local}.
-While these aspects of spatial population structure have been shown to impact the dynamics of climate-induced range shifts in isolation, it is unclear how and if they might interact.
-
-Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts.
-In asexual species, for example, local adaptation has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
-However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
-Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
-In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
-In addition to potential interactions between local adaptation and dispersal evolution, the nature of the range edge could influence the potential for rapid trait evolution during range shifts.
-For example, the severity of the environmental gradient forming the range edge has been shown to alter the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.
-
-Here, we assess the interaction of multiple aspects of spatial population structure with trait evolution in sexually reproducing populations undergoing climate-induced range shifts.
-We develop an individual-based model capable of incorporating a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
-One trait determines dispersal ability while the second defines an individual's environmental niche.
-Using this model, we vary both the potential for local adaptation within the range and the nature of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
-By contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.
-
-\section*{Methods}
-A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here.
-Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig.
-A1).
-Landscapes were unbounded in the $x$ dimension but defined by a fixed width in the $y$ dimension.
-Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions.
-To simulate climate change, environmental conditions shifted at a constant rate along the $x$ dimension.
-Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig.
-A2).
-
-Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci.
-While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits.
-The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
-An individual's environmental niche value allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness.
-The environmental optimum of individual patches could then be systematically varied across the range to allow for different degrees of local adaptation (i.e.
-larger changes in environmental optima allowed for greater local adaptation of the population).
-In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see Appendix A).
-Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness.
-Allele inheritance was subject to mutation and assumed no linkages among loci.
-
-To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table A1 and A2).
-Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change.
-This yielded a total of $27$ different scenarios, each explored with $200$ simulations.
-Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
-Figure 1 shows an example of a single population responding to a moderate speed of climate change.
-For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.
-We primarily discuss simulations using a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.
-
-We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end.
-For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g.
-leading edge vs.
-core populations).
-Due to local extinctions, not all patches were occupied at the end of the period of climate change.
-To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals.
-Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype.
-All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).
-
-\section*{Results}
-In all scenarios, some populations shifted their ranges in response to climate change.
-However the proportion of extinct populations that failed to track the changing climate depended on the initial spatial population structure.
-Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
-2).
-While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction risk, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change.
-We varied both parameters widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table A2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios.
-Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig.
-B1 \& B2).
-However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability during a range shift.
-
-Counterintuitively, populations that survived climate change tended to be characterized by reduced fitness at the range margins prior to the onset of climate change compared to populations that went extinct (Fig.
-B3-B5).
-Essentially, populations with initially higher degrees of local adaptation at the range edges, and thus greater fitness, were more likely to go extinct during climate change.
-This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
-As expected, there was no spatial variation in fitness for populations with no potential for local adaptation.
-
-Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate.
-While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g.
-Fig.
-1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig.
-3a\&b).
-Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig.
-B3-B5).
-The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.
-
-Instead, the initial distribution of dispersal phenotypes prior to the onset of climate change played a key role in determining a population's fate.
-A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
-B9-B11).
-Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig.
-3c\&d).
-In fact, comparing the full distribution of initial dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations.
-Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig.
-B9-B11).
-To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change used in our simulations (see the model description in Appendix A).
-This dispersal phenotype matched the observed threshold value distinguishing surviving from extinction populations in all experimental scenarios (Figures B9-B11, vertical dashed line).
-Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change.
-
-\section*{Discussion}
-Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
-We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
-In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g.
-single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
-Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under increased levels of biological complexity.
-Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
-
-Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the pre-expansion, stable range and in locations with shallow environmental gradients defining the range edge (Fig.
-2).
-A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
-Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study.
-While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to the spatial population structure of their initial ranges.
-
-Our results emphasize the importance of the initial distribution of dispersal phenotypes composing the stable range in determining a population's extinction risk during climate change (Fig.
-3c\&d).
-Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a threshold value.
-Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations for a given speed of climate change (Fig.
-B9-B11).
-Scenarios with no potential for local adaptation and gradual environmental gradients had larger proportions of high dispersal phenotypes under stable climate conditions, and therefore a lower probability of extinction during climate change.
-A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}.
-Similarly, a more severe habitat gradient at the range edge increased the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}.
-Thus, attributes defining the spatial structure of the range altered the distribution of dispersal phenotypes under stable climate conditions, subsequently determining the extinction risk of populations during climate change.
-Importantly, dispersal evolution during climate change was unable to counter the influence of initial spatial population structure on extinction dynamics.
-
-While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range edges when populations had a moderate to high potential for local adaptation (Fig.
-B3-B5).
-In the model, populations at the range edges tended to have lower abundance than populations in the range core, increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
-Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core likely reduced fitness at the range edge by preventing adaptation to local conditions.
-As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges prior to the onset of climate change.
-While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
-
-Previous research has suggested that evolution of increased dispersal ability during climate change may be a key mechanism in rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}.
-Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
-Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}.
-As dispersal evolution during range expansions and shifts occurs via the spatial sorting of alleles contributing to heightened dispersal at the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait.
-The negligible role played by dispersal evolution in our model (Fig.
-3a\&b) suggests that when such simplifying assumptions are relaxed, the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
-
-\section*{Conclusion}
-As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions.
-In particular, a deeper, process-based understanding of extinction risk in populations undergoing range shifts will, in turn, allow more focused conservation interventions.
-Our results suggest that the initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
-Further, in contrast to other studies assuming more simplified genetic structures, we find very little role for the evolution of heightened dispersal abilities in allowing a population to successfully track climate change.
-Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue.
-As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such risk.
-
-%%%%%%%%%%%%%%%%%%%%%
-% Acknowledgments
-%%%%%%%%%%%%%%%%%%%%%
-% You may wish to remove the Acknowledgments section while your paper 
-% is under review (unless you wish to waive your anonymity under
-% double-blind review) if the Acknowledgments reveal your identity.
-% If you remove this section, you will need to add it back in to your
-% final files after acceptance.
-
-%\section*{Acknowledgments}
-
-\newpage{}
-
-\section*{Appendix A: Full model description}
-
-% Please reset counters for the appendix (thus normally figure A1, 
-% figure A2, table A1, etc.).
-
-% In certain cases, it may be appropriate to have a PRINT appendix in
-% addition to (or instead of) an online appendix. In this case, please 
-% name the print appendix Appendix A, and any subsequent appendixes (if 
-% there are any) should be named Online Appendix B, Online Appendix C,
-% etc.
-
-% Counters for each appendix should match the letter of that appendix.
-% For example, tables in Appendix C should be numbered table C1, table C2,
-% etc. This applies to tables, equations, and figures.
-
-% It's better not to use the \appendix command, because we have some
-% formatting peculiarities that \appendix conflicts with.
-
+\renewcommand{\thefigure}{A\arabic{figure}}
+\renewcommand{\thetable}{A\arabic{table}}
 \renewcommand{\theequation}{A\arabic{equation}}
-% redefine the command that creates the equation number.
-\renewcommand{\thetable}{A\arabic{table}}
-\setcounter{equation}{0}  % reset counter 
-\setcounter{figure}{0}
-\setcounter{table}{0}
 
@@
 Available habitat (i.e.
-a population's potential range) is defined by a center location on the $x$ dimension, the severity of the decline in habitat quality at the edge, and the width of the available habitat along the $x$ dimension (See Figure A1).
+a population's potential range) is defined by a center location on the $x$ dimension, the severity of the decline in habitat quality at the edge, and the width of the available habitat along the $x$ dimension (See Figure~\ref{fig:EnvFunction}).
 Further, a gradient in environmental conditions is imposed throughout the landscape to allow for local adaptation via matching of an individual's environmental niche to the local environmental conditions (\textit{Submodels}).
@@
 
-[Figure A1 goes here]
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
+\vspace{-5mm}
+\caption[LoF entry]{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space.
+The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
+The lattice of discrete patches in which population dynamics occur is shown beneath.
+As described in the \textit{Submodels} section of the supplemental materials, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches.
+Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension.
+Landscapes are  unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
+\label{fig:EnvFunction}
+\end{figure}
 
 \subsection*{Process overview and scheduling} 
-Time is modeled in discrete intervals defining single generations of the population (Fig.
-A2).
+Time is modeled in discrete intervals defining single generations of the population (Fig.~\ref{fig:LifeCycle}).
 Within each generation, individuals first disperse from their natal patches according to their phenotypes.
@@
 
-[Figure A2 goes here]
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
+\vspace{-5mm}
+\caption[LoF entry]{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
+Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice.
+After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
+\label{fig:LifeCycle}
+\end{figure}
 
@@
 \end{equation}
-where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure A1).
+where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure~\ref{fig:EnvFunction}).
 To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the $x$ dimension can be shown to be
@@
 
-Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure A1).
+Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure~\ref{fig:EnvFunction}).
 The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
@@
 where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's niche phenotype~\citep{lande1976natural}.
-Thus, an individual's realized fitness will be higher the closer its niche phenotype ($z_{i}$) is to the environmental optimum of the patch it occupies ($z_{opt,x}$).
+Thus, an individual's realized fitness will be higher the closer its niche phenotype ($z_[i]$) is to the environmental optimum of the patch it occupies ($z_{opt,x}$).
 All loci are assumed to contribute additively to an individual's environmental niche value with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values.
@@
 If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ dimension, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ dimension, thus wrapping the top and bottom edges of the landscape to avoid edge effects.
-Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure A1).
-
-\newpage{}
-
-\section*{Appendix B: Supplementary results for varying speeds of climate change}
-
-\renewcommand{\theequation}{B\arabic{equation}}
-% redefine the command that creates the equation number.
-\renewcommand{\thetable}{B\arabic{table}}
-\setcounter{equation}{0}  % reset counter 
-\setcounter{figure}{0}
-\setcounter{table}{0}
-
-\section*{Extinction probability}
-As in the main text, we calculated the cumulative probability of extinction for both slow and fast speeds of climate change here.
-The figures in this section use the same layout and line types as Figure 2 in the main text to allow for direct comparisons.
-
-[Figures B1\&B2 go here.]
-
-\section*{Initial fitness throughout the landscape}
-To assess trends in realized fitness values throughout the landscape, we calculated the patch-level mean individual fitness value for each landscape just prior to the onset of climate change.
-To simplify the figures, we averaged over the $y$ dimension in which environmental conditions do not vary.
-By separating the results for both extinction populations and those that successfully tracked climate change, we identified a trend in which the edge populations displayed lower fitness in populations that were ultimately successful, particularly in simulations with a gradual environmental gradient at the range edge and the potential for local adaptation.
-As realized fitness values do not vary spatially in simulations with no potential for local adaptation, the following figures only show results for scenarios with a moderate or high potential for local adaptation.
-
-[Figures B3-B5 go here.]
-
-\section*{Dispersal evolution}
-Using the same metric of dispersal evolution from the main text (average change in phenotype for each patch), we display here the observed dispersal evolution over the course of climate change for all experimental scenarios.
-Each histogram in the following figures represents a single experimental scenario as indicated by the figure text.
-The lower left panel and upper right panel from Figure B7 are the same histograms shown in Figure 3a\&b, but are here placed in the context of all other experimental scenarios.
-
-[Figures B6-B8 go here.]
-
-\section*{Initial dispersal phenotypes}
-Here, we present histograms of the initial distribution of dispersal phenotypes to demonstrate the importance of those phenotypes in determining population success or extinction during climate change.
-Dispersal phenotypes are log transformed for easier comparison.
-As with the dispersal evolution section, each histogram represents a single experimental scenario as indicated by the figure text.
-Similarly, the lower left panel and upper right panel from Figure B10 are the same histograms shown in Figure 3c\&d, but are here placed in the context of all other experimental scenarios.
-All histograms additionally have a vertical dashed line indicating the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change in each simulation.
-This value serves as a threshold to distinguish individuals from ultimately successful versus extinct populations.
-
-[Figures B9-B11 go here.]
-
-\newpage{}
-
-%%%%%%%%%%%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%
-% You can either type your references following the examples below, or
-% compile your BiBTeX database and paste the contents of your .bbl file
-% here. The amnatnat.bst style file should work for this---but please
-% let us know if you run into any hitches with it!
-% The list below includes sample journal articles, book chapters, and
-% Dryad references.
+Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure~\ref{fig:EnvFunction}).
 
 \bibliographystyle{amnat}
-\bibliography{main_bib}
-
-\newpage{}
-
-\section*{Figures}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
-\caption{A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
-Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change.
-Log transformed mean dispersal phenotypes (b) are shown for each patch.
-Average patch fitness (c) was calculated based on the mean environmental niche trait of local individuals and the environmental optima for each patch.}
-\label{fig:SimExample}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
-\caption{The cumulative probability of extinction due to climate change in different experimental scenarios.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{fig:ExtProb}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
-\caption{Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations.
-Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end.
-Positive values indicate an increase in average dispersal ability in the patch.
-Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
-In all panels, values associated with extant populations are shown in dark blue.
-Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations).
-Full results for all parameter combinations are provided in Appendix B.}
-\label{fig:Disp}
-\end{figure}
-
-\clearpage
-
-\subsection*{Online figures}
-
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
-\caption{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space.
-The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
-The lattice of discrete patches in which population dynamics occur is shown beneath.
-As described in the \textit{Submodels} section of the supplemental materials, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches.
-Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension.
-Landscapes are  unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
-\label{Fig:EnvFunction}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
-\caption{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
-Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice.
-After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
-\label{Fig:LifeCycle}
-\end{figure}
-
-\clearpage
-
-\renewcommand{\thefigure}{B\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
-\caption{Extinction probabilities for a slow speed of climate change.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{Fig:ExtProbSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
-\caption{Extinction probabilities for a fast speed of climate change.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{Fig:ExtProbFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a slow speed of climate change.}
-\label{Fig:InitFitSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a moderate speed of climate change.}
-\label{Fig:InitFit}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a fast speed of climate change.}
-\label{Fig:InitFitFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a slow speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolMain}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a fast speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
-\label{Fig:InitDispSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
-\label{Fig:InitDispMain}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
-\label{Fig:InitDispFast}
-\end{figure}
+\bibliography{supp_bib}
 

### 2019-04-05 Manuscript/AppendixB.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

--- 2019-04-05 Manuscript/AmNat_InitialSubmission.tex
@@
 \usepackage{amsthm}			
-\usepackage{fancybox}
-\usepackage{graphicx}
-\usepackage{graphics}
+\usepackage{natbib}				
+\usepackage{fullpage}
+\usepackage{placeins}
+\usepackage{setspace}
+\usepackage{lineno}
+\usepackage{color}
+\usepackage{authblk}
+\usepackage{mathptmx}
+
+\DeclareRobustCommand{\firstsecond}[2]{#1}
+
+\newcommand{\mb}{\mathbf}
+\newcommand{\bs}{\boldsymbol}
+\newcommand{\wt}{\widetilde}
 \newcommand{\s}{^{(s)}}
 
-%%%%%%%%%%%%%%%%%%%%%
-% Line numbering
-%%%%%%%%%%%%%%%%%%%%%
-\usepackage{lineno}
-% Please use line numbering with your initial submission and
-% subsequent revisions. After acceptance, please turn line numbering
-% off by adding percent signs to the lines %\usepackage{lineno} and
-% to %\linenumbers{} and %\modulolinenumbers[3] below.
-
-\title{Spatial population structure determines extinction risk in climate-induced range shifts}
-
-% This version of the LaTeX template was last updated on
-% January 11, 2018.
-
-%%%%%%%%%%%%%%%%%%%%%
-% Authorship
-%%%%%%%%%%%%%%%%%%%%%
-% Please remove authorship information while your paper is under review,
-% unless you wish to waive your anonymity under double-blind review. You
-% will need to add this information back in to your final files after
-% acceptance.
-
-%\author{Owen E. Cook$^{1,\ast}$ \\ 
-%Generic H. Collaborator$^{2,\dag}$ \\ 
-%Additional Q. Expert$^{3}$}
+\title{Appendix B: Supplementary results for varying speeds of climate change}
 
@@
 
+\author[1]{Christopher Weiss-Lehman}
+\author[1]{Allison K. Shaw}
+
+\affil[1]{Ecology, Evolution, and Behavior, University of Minnesota}
+
 \begin{document}
-
 \maketitle
 
-%\noindent{} 1. University of Chicago, Chicago, Illinois 60637;
+\doublespacing
+\linenumbers
 
-%\noindent{} 2. University of Toronto, Toronto, Ontario M5S 1A5, Canada;
-
-%\noindent{} 3. Middle Eastern Technical University, Çankaya, Ankara 06800, Turkey.
-
-%\noindent{} $\ast$ Corresponding author; e-mail: amnat@uchicago.edu.
-
-%\noindent{} $\dag$ Deceased.
-
-\bigskip
-
-\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1\&A2,  tables~A1\&A2, and figures~B1-B11). Figure~1 and figure~3 are to print in color.
-
-\bigskip
-
-\textit{Keywords}: range shifts, eco-evolutionary dynamics, local adaptation, individual-based model
-
-\bigskip
-
-\textit{Manuscript type}: Article. %Or e-article, note, e-note, natural history miscellany, e-natural history miscellany, comment, reply, invited symposium, or countdown to 150.
-
-\bigskip
-
-\noindent{\footnotesize Prepared using the suggested \LaTeX{} template for \textit{Am.\ Nat.}}
-
-\linenumbers{}
-\modulolinenumbers[3]
-
-\newpage{}
-
-\section*{Abstract}
-Climate change is an increasingly severe threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses.
-Population dynamics of range shifts, among the most commonly observed responses to climate change, can be influenced by many factors, including evolution of key traits, the degree of local adaptation, and the nature of the range edge.
-Here, we use an individual-based model to explore the interacting roles of these factors in the dynamics of climate-induced range shifts.
-We show that aspects of the spatial population structure within the initial range, in particular the potential for local adaptation, severely increased a population's extinction risk.
-Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations.
-Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes that evolved within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
-Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.
-
-\newpage{}
-
-\section*{Introduction}
-Climate change is expected to dramatically reshape global biogeographic patterns as some species shift their ranges to track changing environmental conditions~\citep{gonzalez2010global}.
-These range shifts are generally predicted to proceed upwards in latitude, elevation, or both as average global temperatures continue to rise~\citep{loarie2009velocity}.
-Indeed, contemporary range shifts have already been observed in a wide variety of taxa, ranging from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages~\citep{hobbs2009novel}.
-Understanding the ecological and evolutionary dynamics of such climate-induced range shifts will play a key role in informing current and future conservation work.
-
-Large-scale population movements have been studied for decades in the context of range expansions (e.g.
-of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
-As fundamentally similar spatial processes, it is likely that range shifts will also be subject to these ecological and evolutionary mechanisms known to drive range expansions.
-However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population.
-In particular, range shifts occur in populations with far more complex spatial structure compared to most range expansions, which typically begin from the successful establishment and spread of a small, founding population~\citep{hastings2005spatial}.
-While these founding populations often lack any significant spatial structure, populations undergoing range shifts are characterized by a spatial population structure formed by aspects of the previously stable ranges.
-For example, population ranges can vary in their potential for local adaptation throughout the range, the nature of the range edge, and the spatial distribution of key traits.
-
-Each of these factors relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions.
-For example, the underlying mechanism responsible for the gradient in population size from the range core to the edge (i.e.
-declines in carrying capacity versus growth rate) alters a population's extinction risk during climate driven range shifts~\citep{henry2013eco}.
-Additionally, the potential for local adaptation throughout a range has been related to extinction risk during range shifts.
-Specifically, a low potential for local adaptation can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block the establishment of better adapted genotypes~\citep{atkins2010local}.
-While these aspects of spatial population structure have been shown to impact the dynamics of climate-induced range shifts in isolation, it is unclear how and if they might interact.
-
-Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts.
-In asexual species, for example, local adaptation has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
-However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
-Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
-In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
-In addition to potential interactions between local adaptation and dispersal evolution, the nature of the range edge could influence the potential for rapid trait evolution during range shifts.
-For example, the severity of the environmental gradient forming the range edge has been shown to alter the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.
-
-Here, we assess the interaction of multiple aspects of spatial population structure with trait evolution in sexually reproducing populations undergoing climate-induced range shifts.
-We develop an individual-based model capable of incorporating a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
-One trait determines dispersal ability while the second defines an individual's environmental niche.
-Using this model, we vary both the potential for local adaptation within the range and the nature of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
-By contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.
-
-\section*{Methods}
-A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here.
-Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig.
-A1).
-Landscapes were unbounded in the $x$ dimension but defined by a fixed width in the $y$ dimension.
-Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions.
-To simulate climate change, environmental conditions shifted at a constant rate along the $x$ dimension.
-Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig.
-A2).
-
-Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci.
-While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits.
-The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
-An individual's environmental niche value allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness.
-The environmental optimum of individual patches could then be systematically varied across the range to allow for different degrees of local adaptation (i.e.
-larger changes in environmental optima allowed for greater local adaptation of the population).
-In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see Appendix A).
-Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness.
-Allele inheritance was subject to mutation and assumed no linkages among loci.
-
-To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table A1 and A2).
-Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change.
-This yielded a total of $27$ different scenarios, each explored with $200$ simulations.
-Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
-Figure 1 shows an example of a single population responding to a moderate speed of climate change.
-For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.
-We primarily discuss simulations using a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.
-
-We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end.
-For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g.
-leading edge vs.
-core populations).
-Due to local extinctions, not all patches were occupied at the end of the period of climate change.
-To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals.
-Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype.
-All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).
-
-\section*{Results}
-In all scenarios, some populations shifted their ranges in response to climate change.
-However the proportion of extinct populations that failed to track the changing climate depended on the initial spatial population structure.
-Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
-2).
-While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction risk, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change.
-We varied both parameters widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table A2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios.
-Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig.
-B1 \& B2).
-However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability during a range shift.
-
-Counterintuitively, populations that survived climate change tended to be characterized by reduced fitness at the range margins prior to the onset of climate change compared to populations that went extinct (Fig.
-B3-B5).
-Essentially, populations with initially higher degrees of local adaptation at the range edges, and thus greater fitness, were more likely to go extinct during climate change.
-This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
-As expected, there was no spatial variation in fitness for populations with no potential for local adaptation.
-
-Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate.
-While some, individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g.
-Fig.
-1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig.
-3a\&b).
-Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig.
-B3-B5).
-The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.
-
-Instead, the initial distribution of dispersal phenotypes prior to the onset of climate change played a key role in determining a population's fate.
-A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
-B9-B11).
-Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig.
-3c\&d).
-In fact, comparing the full distribution of initial dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations.
-Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig.
-B9-B11).
-To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change used in our simulations (see the model description in Appendix A).
-This dispersal phenotype matched the observed threshold value distinguishing surviving from extinction populations in all experimental scenarios (Figures B9-B11, vertical dashed line).
-Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change.
-
-\section*{Discussion}
-Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
-We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
-In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g.
-single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
-Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under increased levels of biological complexity.
-Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
-
-Our results suggest that populations most likely to keep pace with climate change will be those with little to no local adaptation within the pre-expansion, stable range and in locations with shallow environmental gradients defining the range edge (Fig.
-2).
-A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
-Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the stark environmental gradients imposed in our study.
-While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to the spatial population structure of their initial ranges.
-
-Our results emphasize the importance of the initial distribution of dispersal phenotypes composing the stable range in determining a population's extinction risk during climate change (Fig.
-3c\&d).
-Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a threshold value.
-Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations for a given speed of climate change (Fig.
-B9-B11).
-Scenarios with no potential for local adaptation and gradual environmental gradients had larger proportions of high dispersal phenotypes under stable climate conditions, and therefore a lower probability of extinction during climate change.
-A high potential for local adaptation, in contrast, selected against such high dispersal phenotypes due to dispersal's homogenizing effect on population genetic structure~\citep{lenormand2002gene}.
-Similarly, a more severe habitat gradient at the range edge increased the risk of dispersing beyond the boundary of suitable habitat, resulting in selection against heightened dispersal~\citep{shaw2014population}.
-Thus, attributes defining the spatial structure of the range altered the distribution of dispersal phenotypes under stable climate conditions, subsequently determining the extinction risk of populations during climate change.
-Importantly, dispersal evolution during climate change was unable to counter the influence of initial spatial population structure on extinction dynamics.
-
-While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range edges when populations had a moderate to high potential for local adaptation (Fig.
-B3-B5).
-In the model, populations at the range edges tended to have lower abundance than populations in the range core, increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
-Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core likely reduced fitness at the range edge by preventing adaptation to local conditions.
-As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges prior to the onset of climate change.
-While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
-
-Previous research has suggested that evolution of increased dispersal ability during climate change may be a key mechanism in rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more}.
-Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
-Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}.
-As dispersal evolution during range expansions and shifts occurs via the spatial sorting of alleles contributing to heightened dispersal at the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait.
-The negligible role played by dispersal evolution in our model (Fig.
-3a\&b) suggests that when such simplifying assumptions are relaxed, the potential for population rescue via evolution of heightened dispersal is greatly reduced, thus increasing the role of the initial spatial population structure within the range in determining a population's fate under climate change.
-
-\section*{Conclusion}
-As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions.
-In particular, a deeper, process-based understanding of extinction risk in populations undergoing range shifts will, in turn, allow more focused conservation interventions.
-Our results suggest that the initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
-Further, in contrast to other studies assuming more simplified genetic structures, we find very little role for the evolution of heightened dispersal abilities in allowing a population to successfully track climate change.
-Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue.
-As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such risk.
-
-%%%%%%%%%%%%%%%%%%%%%
-% Acknowledgments
-%%%%%%%%%%%%%%%%%%%%%
-% You may wish to remove the Acknowledgments section while your paper 
-% is under review (unless you wish to waive your anonymity under
-% double-blind review) if the Acknowledgments reveal your identity.
-% If you remove this section, you will need to add it back in to your
-% final files after acceptance.
-
-%\section*{Acknowledgments}
-
-\newpage{}
-
-\section*{Appendix A: Full model description}
-
-% Please reset counters for the appendix (thus normally figure A1, 
-% figure A2, table A1, etc.).
-
-% In certain cases, it may be appropriate to have a PRINT appendix in
-% addition to (or instead of) an online appendix. In this case, please 
-% name the print appendix Appendix A, and any subsequent appendixes (if 
-% there are any) should be named Online Appendix B, Online Appendix C,
-% etc.
-
-% Counters for each appendix should match the letter of that appendix.
-% For example, tables in Appendix C should be numbered table C1, table C2,
-% etc. This applies to tables, equations, and figures.
-
-% It's better not to use the \appendix command, because we have some
-% formatting peculiarities that \appendix conflicts with.
-
-\renewcommand{\theequation}{A\arabic{equation}}
-% redefine the command that creates the equation number.
-\renewcommand{\thetable}{A\arabic{table}}
-\setcounter{equation}{0}  % reset counter 
-\setcounter{figure}{0}
-\setcounter{table}{0}
-
-\section*{Model overview}
-\subsection*{Purpose} 
-This model tests an evolving population's ability to track a changing climate under a variety of conditions.
-Specifically, populations are simulated under different combinations of (1) the starkness of the range boundary and (2) the potential for local adaptation.
-In all simulations, an individual's expected dispersal distance and environmental niche are defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
-All simulations begin with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change.
-Climate change is then modeled as a constant, directional shift in the location of environmentally suitable habitat (see \textit{Submodels} below).
-Finally, simulations end with another short period of climate stability to assess the population's ability to persist and recover after shifting its range.
-
-\subsection*{State variables and scales} 
-The model simulates a population of males and females characterized by diploid loci for both their expected dispersal distance and environmental niche.
-Space is modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system.
-Landscapes are two dimensional with a fixed width along the $y$ axis and without bounds on the $x$ axis.
-Environmental conditions vary along the $x$ dimension but remain constant within the $y$ dimension.
-To avoid edge effects due to the fixed width of the $y$ dimension, the model employs wrapping boundaries such that if an individual disperses out of the landscape on one side, it appears at the opposite end of the same column of the landscape.
-Patches are defined by the location of the patch center in $x$ and $y$ coordinates and a patch width parameter defining the relationship between continuous Cartesian space and the discrete patches used for population dynamics (\textit{Submodels}).
-
-The model implements climate change by shifting the location of a population's available habitat along the $x$ dimension of the landscape.
-Available habitat (i.e.
-a population's potential range) is defined by a center location on the $x$ dimension, the severity of the decline in habitat quality at the edge, and the width of the available habitat along the $x$ dimension (See Figure A1).
-Further, a gradient in environmental conditions is imposed throughout the landscape to allow for local adaptation via matching of an individual's environmental niche to the local environmental conditions (\textit{Submodels}).
-The severity of this gradient can be altered to change the potential for local adaptation (e.g.
-a shallower gradient will result in more similar environmental conditions throughout the range and therefore reduce the potential for local adaptation).
-
-[Figure A1 goes here]
-
-\subsection*{Process overview and scheduling} 
-Time is modeled in discrete intervals defining single generations of the population (Fig.
-A2).
-Within each generation, individuals first disperse from their natal patches according to their phenotypes.
-After dispersal, reproduction occurs according to a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch.
-Reproduction occurs via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) are likely to produce multiple offspring while individuals with low relative fitness may not produce any.
-Individuals inherit one allele from each parent at each loci, assuming independent segregation and a mutation process.
-After reproduction, all individuals in the current generation perish and the offspring begin the next generation with dispersal, resulting in discrete, non-overlapping generations.
-
-[Figure A2 goes here]
-
-\section*{Design concepts}
-\subsection*{Emergence} 
-Emergent phenomena in this model include the spatial equilibrium of population abundances and trait values within the stable range, the demographic dynamics of the shifting population during climate change, and the evolutionary trajectories of both expected dispersal distances and environmental niche values during climate change.
-
-\subsection*{Stochasticity} 
-All biological processes in this model are stochastic including realized population growth in each patch, dispersal distances of each individual, and inheritance of loci.
-Environmental parameters are fixed, however, and the process of climate change (i.e.
-the movement of environmentally suitable habitat through time) is deterministic.
-Thus, the model removes the confounding influence of environmental stochasticity to focus on demographic and evolutionary dynamics of range shifts.
-
-\subsection*{Interactions} 
-Individuals in the model interact via mating and density-dependent competition within patches.
-Additionally, the evolutionary trajectories of the two different traits have the potential to interact via the relationship between gene flow and local adaptation.
-Further, aspects determining the spatial population structure of a population's range (potential for local adaptation and the nature of the range edge) can interact with trait evolution within the range both during stable climate conditions and during climate change.
-
-\subsection*{Desired output} 
-After each model run, full details of all surviving individuals at the last time point are recorded (spatial coordinates and loci values for both traits).
-If a population went extinct during the model run, the time of extinction is recorded.
-For each occupied patch throughout the simulation, we aggregated data on population size, the dispersal trait, and local adaptation to environmental conditions.
-
-\section*{Details}
-\subsection*{Initialization} 
-The following parameters are set at the beginning of each simulation and form the initial conditions of the model: the mean and variance for allele values of each trait, population size, location of environmentally suitable habitat, number of generations for the pre-, post-, and during climate change periods of the simulation, and all other necessary parameters for the submodels defined below.
-Simulated populations are initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions.
-This ensures that the populations reacting to a changing climate truly represent the expected spatial distribution for a given range, rather than the initial parameter values used in the simulation.
-See Tables A1 \& A2 for a full list of parameter values used in the simulations described here.
-
-\begin{table}
-\renewcommand{\arraystretch}{1.5}
-  \begin{tabular}{ p{2cm} | p{8cm} | p{2cm} }
-    \hline
-    Parameter & Description & Value \\ \hline \hline
-    $N_{1}$ & Initial population size (seeded across multiple patches) when beginning the simulations & $2500$ \\
-    $\beta_{1}$ & Center of environmentally suitable habitat before climate change & $0$ \\
-    $\eta$ & Spatial dimensions of habitat patches in continuous space & $50$ \\
-    $y_{max}$ & Number of patches the discrete lattice extends in the $y$ direction & $10$ \\
-    $\hat{t}$ & Last time point of stable climate conditions & $2000$ \\
-    $t_{\Delta}$ & Duration of climate change & $100$ \\
-    $t_{max}$ & Total number of time points in the simulation & $2150$ \\
-    $R$ & Intrinsic growth rate of the population & $2$ \\
-    $K_{max}$ & Maximum achievable carrying capacity in the range & $100$ \\ 
-    $\psi$ & Expected sex ratio in the population & $0.5$ \\
-    $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
-    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
-    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
-    $U^{T}$ & Diploid mutation rate for each trait & $0.02$ for each trait \\
-    $V_{m}^{T}$ & Mutational variance for each trait & $0.0004$ for each trait \\
-    $L^{T}$ & Number of diploid loci defining each trait & $5$ for each trait \\
-    $\mu_{1}^{f}$ & Initial mean allele value for the environmental niche trait & $0$ \\
-    $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
-    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the environmental niche trait & $0.025$ \\
-    $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
-    \hline
-  \end{tabular}
-\caption[LoF entry]{Simulation parameters held constant across all scenarios.}
-\label{table:ConstPars}
-\end{table}
-
-\subsection*{Submodels}
-\paragraph{Environmentally suitable habitat}
-Environmentally suitable habitat is determined by the population's carrying capacity as it ranges in space ($K_{x}$).
-The carrying capacity is maximized in the center of the species' range ($K_{max}$) and declines with increasing distance from the center.
-Specifically, the carrying capacity at a location $x$ is defined as the product of $K_{max}$ and a function $f(x,t)$, where $f(x,t)$ ranges from $1$ in the range center to $0$ far away from the center and is defined as follows
-\begin{equation}
-f(x,t)=
-\begin{cases}
-	\frac{e^{\gamma(x-\beta_{t}+\tau)}}{1+e^{\gamma(x-\beta_{t}+\tau)}} & x \leq \beta_{t} \\
-	\frac{e^{-\gamma(x-\beta_{t}-\tau)}}{1+e^{-\gamma(x-\beta_{t}-\tau)}} & x > \beta_{t}
-\end{cases}
-\end{equation}
-where $\beta_{t}$ defines the center of the area of suitable habitat at time $t$, $\tau$ sets the width of the range, and $\gamma$ affects the slope of the function at the range boundaries (See Figure A1).
-To understand the relationship between $\gamma$ and the slope of $f(x,t)$ at the range boundary, the partial derivative of $f(x,t)$ over the $x$ dimension can be shown to be
-\begin{equation}
-f(x,t)=
-\begin{cases}
-	\frac{\gamma e^{\gamma(x-\beta_{t}+\tau)}}{(1+e^{\gamma(x-\beta_{t}+\tau})^{2}} & x \leq \beta_{t} \\
-	\frac{-\gamma e^{-\gamma(x-\beta_{t}-\tau)}}{(1+e^{-\gamma(x-\beta_{t}-\tau})^{2}} & x > \beta_{t}
-\end{cases}	
-\end{equation}
-yielding a value of $\pm\frac{\gamma}{4}$ at the inflection points on either side of the range center ($x=\beta_{t}\pm\tau$).
-
-Population dynamics occur within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we use another parameter defining the spatial scale of each patch ($\eta$; See Figure A1).
-The local carrying capacity of a patch centered on $x$ ($K_{x}$) is then calculated as the mean of $f(x,t)$ over the interval of the patch multiplied by $K_{max}$.
-\begin{equation}
-K_{x} = \frac{K_{max}}{\eta}\int_{x-\frac{\eta}{2}}^{x+\frac{\eta}{2}}f(x,t)dx
-\end{equation}
-
-By varying the parameters defining $f(x,t)$, we can change both the total carrying capacity of the population, summed across all patches throughout the range, (by altering both $\tau$ and $\gamma$) and the slope at which $K_{x}$ declines to $0$ (by altering $\gamma$).
-Changing the slope affects not only the rate at which $K_{x}$ declines at the range boundaries (our focus), but it also alters the total carrying capacity of the population.
-To avoid this confounding factor, we fix the total area under the curve $f(x,t)$.
-The indefinite integral of $f(x,t)$ can be shown to be
-\begin{equation}
-\int_{-\infty}^{\infty}f(x,t)dx = \frac{2ln(e^{\gamma\tau}+1)}{\gamma}
-\end{equation}
-which can be solved for $\tau$. For a given fixed total area under the curve, an appropriate value of $\tau$ can be calculated for each value of $\gamma$.
-
-Thus, $\gamma$ and $\tau$ are both fixed within a given simulation and $\beta_{t}$ (the location of the center of suitable habitat) is used to simulate climate change.
-During the periods before and after climate change $\beta_{t}$ is constant, but to simulate climate change it varies with time as follows
-\begin{equation}
-\beta_{t}=\nu\eta(t-\hat{t})
-\end{equation}
-where $\nu$ is the velocity of climate change per generation in terms of discrete patches, $t$ is the current generation, and $\hat{t}$ is the last generation of stable climatic conditions before the onset of climate change.
-
-\begin{table}
-\renewcommand{\arraystretch}{1.5}
-  \begin{tabular}{ p{4cm} | p{4cm} | p{1.5cm} | p{1.5cm} | p{1.5cm} }
-    \hline
-    Habitat gradient at the range edge & Potential for local adaptation & $\gamma$ & $\tau$ & $\lambda$  \\ \hline \hline
-     & None & $0.0025$ & $250$ & $0$ \\
-    Shallow & Low & $0.0025$ & $250$ & $0.004$ \\
-     & High & $0.0025$ & $250$ & $0.008$ \\ \hline
-     & None & $0.025$ & $421.479$ & $0$ \\
-    Moderate & Low & $0.025$ & $421.479$ & $0.004$ \\
-     & High & $0.025$ & $421.479$ & $0.008$ \\ \hline
-     & None & $0.25$ & $421.48$ & $0$ \\
-    Stark & Low & $0.25$ & $421.48$ & $0.004$ \\
-     & High & $0.25$ & $421.48$ & $0.008$ \\ 
-    \hline
-  \end{tabular}
-\caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.}
-\label{table:Scenarios}
-\end{table}
-
-\paragraph{Local adaptation}
-To allow an arbitrary degree of local adaptation within the range, the local environmental optima for each patch ($z_{opt,x}$) are set as follows
-\begin{equation}
-z_{opt,x}=\lambda(x-\beta_{t})
-\end{equation}
-where $\lambda$ defines the potential for local adaptation with values close to $0$ resulting in little to no change in environmental optima across the range and values of greater magnitude resulting in large differences in environmental optima across the range.
-Individual relative fitness ($w_{i,x}$) values are then calculated according to the following equation assuming stabilizing selection
-\begin{equation}
-w_{i,x}=e^{\frac{-(z_{i}-z_{opt,x})^{2}}{2\omega^{2}}}
-\end{equation}
-where $\omega$ defines the strength of stabilizing selection and $z_{i}$ is an individual's niche phenotype~\citep{lande1976natural}.
-Thus, an individual's realized fitness will be higher the closer its niche phenotype ($z_{i}$) is to the environmental optimum of the patch it occupies ($z_{opt,x}$).
-All loci are assumed to contribute additively to an individual's environmental niche value with no dominance or epistasis, meaning an individual's phenotype is simply the sum of the individual's allele values.
-
-\paragraph{Population dynamics}
-Population growth within each patch is modeled with a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-To account for fitness effects on population growth, expected population growth is scaled by the mean relative fitness of individuals within the patch ($\bar{w_{x}}$).
-The expected number of new offspring in patch $x$ at time $t+1$ is then given by
-\begin{equation}
-\hat{N}_{t+1,x}=\bar{w_{x}}F_{t,x}\frac{R}{\psi}e^{\frac{-RN_{t,x}}{K_{x}}}
-\end{equation}
-where $F_{t,x}$ is the number of females in patch $x$ at time $t$, $R$ is the intrinsic growth rate for the population, $\psi$ is the expected sex ratio of the population, $N_{t,x}$ is the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ is the local carrying capacity based on the environmental conditions.
-To incorporate demographic stochasticity, the realized number of offspring for each patch is then drawn from a Poisson distribution.
-\begin{equation}
-N_{t+1,x}\sim Poisson(\hat{N}_{t+1,x})
-\end{equation}
-
-Parentage of the offspring is then assigned by random sampling of the local male and female population (i.e.
-polygynandrous mating).
-The sampling is weighted by individual fitness and occurs with replacement so highly fit individuals are likely to have multiple offspring while low fitness individuals may not have any.
-Each offspring inherits one allele per locus from each parent, assuming no linkage among loci.
-After reproduction, all members of the previous generation die and the offspring disperse to begin the next generation.
-
-\paragraph{Mutation}
-Inherited alleles are subject to mutation such that some offspring might not inherit identical copies of certain alleles from their parents.
-The mutation process is defined by two parameters for each trait $T$: the diploid mutation rate ($U^{T}$) and the mutational variance ($V_{m}^{T}$).
-Using these parameters along with the number of loci defining trait $T$ ($L^{T}$), the per locus probability of a mutation is
-\begin{equation}
-\frac{U^{T}}{2L^{T}}
-\end{equation}
-Mutational effects are drawn from a normal distribution with mean $0$ and a standard deviation of
-\begin{equation}
-\sqrt{V_{m}^{T}U^{T}}
-\end{equation}
-By defining the mutation process in this manner rather than setting a probability of mutation and mutational effect directly, similar mutational dynamics can be imposed regardless of the number of loci used in the simulation.
-
-\paragraph{Dispersal}
-Finally, individuals disperse according to an exponential dispersal kernel defined by each individual's dispersal phenotype.
-An individual's dispersal phenotype is the expected dispersal distance and is given by
-\begin{equation}
-d_{i} = \frac{\hat{d}\eta e^{\rho\Sigma L^{D}}}{1+e^{\rho\Sigma L^{D}}} 
-\end{equation}
-where $\hat{d}$ is the maximum expected dispersal distance in terms of discrete patches, $\rho$ is a constant determining the slope of the transition between $0$ and $\hat{d}$, and the summation is taken across all alleles contributing to dispersal.
-Thus, as with fitness, loci are assumed to contribute additively with no dominance or epistasis.
-The expected dispersal distance, $d_{i}$ is then used to draw a realized distance from an exponential dispersal kernel.
-Since the dispersal phenotype is the expected value of the exponential dispersal kernel, it can be used directly to calculate the two dimensional diffusion coefficient of population spread ($D$).
-Specifically, since $d_{i}^{2}$ represents the mean squared displacement of an individual with dispersal phenotype $d_{i}$, the two dimensional diffusion coefficient can be calculated as
-\begin{equation}
-D = \frac{1}{4}d_{i}^{2}
-\end{equation}
-
-Once the realized dispersal distance is obtained, the direction of dispersal is drawn from a uniform distribution bounded by $0$ and $2\pi$.
-If a dispersal trajectory takes an individual outside the bounds of the landscape in the $y$ dimension, the individual reappears at the same $x$ coordinate but the opposite end of the $y$ dimension, thus wrapping the top and bottom edges of the landscape to avoid edge effects.
-Dispersal occurs from the center of each patch and the individual's new patch is then determined according to its location in the overlaid grid of $\eta$ x $\eta$ patches (see Figure A1).
-
-\newpage{}
-
-\section*{Appendix B: Supplementary results for varying speeds of climate change}
-
+\renewcommand{\thefigure}{B\arabic{figure}}
+\renewcommand{\thetable}{B\arabic{table}}
 \renewcommand{\theequation}{B\arabic{equation}}
-% redefine the command that creates the equation number.
-\renewcommand{\thetable}{B\arabic{table}}
-\setcounter{equation}{0}  % reset counter 
-\setcounter{figure}{0}
-\setcounter{table}{0}
 
@@
 
-[Figures B1\&B2 go here.]
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
+\vspace{-5mm}
+\caption[LoF entry]{Extinction probabilities for a slow speed of climate change.
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) gradual, and (c) rapid change in the niche optimum across space and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) range edge.}
+\label{fig:ExtProbSlow}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
+\vspace{-5mm}
+\caption[LoF entry]{Extinction probabilities for a fast speed of climate change.
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) gradual, and (c) rapid change in the niche optimum across space and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) range edge.}
+\label{fig:ExtProbFast}
+\end{figure}
+
+\newpage
 
@@
 
-[Figures B3-B5 go here.]
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
+\vspace{-5mm}
+\caption[LoF entry]{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
+Points represent the mean across simulations and error bars are interquartile ranges.
+Population status (extinct or successful) was determined for a slow speed of climate change.}
+\label{fig:InitFitSlow}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
+\caption[LoF entry]{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
+Points represent the mean across simulations and error bars are interquartile ranges.
+Population status (extinct or successful) was determined for a moderate speed of climate change.}
+\vspace{-5mm}
+\label{fig:InitFit}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
+\vspace{-5mm}
+\caption[LoF entry]{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
+Points represent the mean across simulations and error bars are interquartile ranges.
+Population status (extinct or successful) was determined for a fast speed of climate change.}
+\label{fig:InitFitFast}
+\end{figure}
+
+\newpage
 
@@
 
-[Figures B6-B8 go here.]
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
+\vspace{-5mm}
+\caption[LoF entry]{Observed dispersal evolution in populations responding to a slow speed of climate change.
+Positive values indicate an increase in average dispersal ability over the course of climate change.
+The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
+The experimental scenario corresponding to each histogram is indicated on the figure.}
+\label{fig:DispEvolSlow}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
+\vspace{-5mm}
+\caption[LoF entry]{Observed dispersal evolution in populations responding to a moderate speed of climate change.
+Positive values indicate an increase in average dispersal ability over the course of climate change.
+The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
+The experimental scenario corresponding to each histogram is indicated on the figure.}
+\label{fig:DispEvolMain}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
+\vspace{-5mm}
+\caption[LoF entry]{Observed dispersal evolution in populations responding to a fast speed of climate change.
+Positive values indicate an increase in average dispersal ability over the course of climate change.
+The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
+The experimental scenario corresponding to each histogram is indicated on the figure.}
+\label{fig:DispEvolFast}
+\end{figure}
+
+\newpage
 
@@
 
-[Figures B9-B11 go here.]
-
-\newpage{}
-
-%%%%%%%%%%%%%%%%%%%%%
-% Bibliography
-%%%%%%%%%%%%%%%%%%%%%
-% You can either type your references following the examples below, or
-% compile your BiBTeX database and paste the contents of your .bbl file
-% here. The amnatnat.bst style file should work for this---but please
-% let us know if you run into any hitches with it!
-% The list below includes sample journal articles, book chapters, and
-% Dryad references.
-
-\bibliographystyle{amnat}
-\bibliography{main_bib}
-
-\newpage{}
-
-\section*{Figures}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
-\caption{A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
-Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change.
-Log transformed mean dispersal phenotypes (b) are shown for each patch.
-Average patch fitness (c) was calculated based on the mean environmental niche trait of local individuals and the environmental optima for each patch.}
-\label{fig:SimExample}
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
+\vspace{-5mm}
+\caption[LoF entry]{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
+Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
+Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
+\label{fig:InitDispSlow}
 \end{figure}
 
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
-\caption{The cumulative probability of extinction due to climate change in different experimental scenarios.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{fig:ExtProb}
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
+\vspace{-5mm}
+\caption[LoF entry]{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
+Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
+Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
+\label{fig:InitDispMain}
 \end{figure}
 
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
-\caption{Patterns in the evolution and the initial distribution of the dispersal trait, highlighting extant simulations.
-Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end.
-Positive values indicate an increase in average dispersal ability in the patch.
-Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
-In all panels, values associated with extant populations are shown in dark blue.
-Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations).
-Full results for all parameter combinations are provided in Appendix B.}
-\label{fig:Disp}
-\end{figure}
-
-\clearpage
-
-\subsection*{Online figures}
-
-\renewcommand{\thefigure}{A\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
-\caption{Example of the environmentally suitable habitat available to a population, as defined by $f(x,t)$ in Cartesian space.
-The parameters of $f(x,t)$ are shown on the figure at significant points along the $x$ axis.
-The lattice of discrete patches in which population dynamics occur is shown beneath.
-As described in the \textit{Submodels} section of the supplemental materials, $f(x,t)$ determines the carrying capacity of the discrete $\eta$ x $\eta$ patches.
-Carrying capacities vary with $f(x,t)$ along the $x$ dimension of the lattice and remain constant within each column along the $y$ dimension.
-Landscapes are  unbounded in the $x$ dimension and implemented with wrapping boundaries in the $y$ dimension.}
-\label{Fig:EnvFunction}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
-\caption{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
-Each generation begins with new offspring dispersing according to their phenotype, after which reproduction occurs in local populations defined by the discrete lattice.
-After reproduction, all parental individuals perish, resulting in discrete, non-overlapping generations.}
-\label{Fig:LifeCycle}
-\end{figure}
-
-\clearpage
-
-\renewcommand{\thefigure}{B\arabic{figure}}
-\setcounter{figure}{0}
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
-\caption{Extinction probabilities for a slow speed of climate change.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{Fig:ExtProbSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
-\caption{Extinction probabilities for a fast speed of climate change.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with (a) no, (b) low, and (c) high potential for local adaptation, and in environments characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) gradient at the range edge.}
-\label{Fig:ExtProbFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a slow speed of climate change.}
-\label{Fig:InitFitSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a moderate speed of climate change.}
-\label{Fig:InitFit}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
-\caption{Individual fitness along the $x$ dimensions of the landscape prior to the onset of climate change.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a fast speed of climate change.}
-\label{Fig:InitFitFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a slow speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolMain}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a fast speed of climate change.
-Positive values indicate an increase in average dispersal ability over the course of climate change.
-The values associate with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolFast}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
+\begin{figure}
+\centering
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
+\vspace{-5mm}
+\caption[LoF entry]{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
 Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
 Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
-\label{Fig:InitDispSlow}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
-\label{Fig:InitDispMain}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in populations just prior to the onset of climate change.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
-\label{Fig:InitDispFast}
+\label{fig:InitDispFast}
 \end{figure}

### 2019-04-05 Manuscript/Main.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

--- 2018-08-28 Manuscript/Main.tex
@@
 
-\title{Spatial population structure drives extinction dynamics in climate-induced range shifts}
+\title{Spatial population structure determines extinction risk in climate-induced range shifts}
 
@@
 \section*{Abstract}
-Understanding the interplay between ecological and evolutionary factors is critical for predicting species' responses to anthropogenic climate change and for determining optimal conservation practices.
-Key parallels between range expansions (e.g.
-of invasive species) and range shifts in response to climate change allow us to expand current theory for range expansions to include initial spatial population structures and predict species' responses to climate change.
-Here, we use an individual-based model to explore the role of dispersal evolution, local adaptation, and a population's initial spatial structure in driving the dynamics of climate-induced range shifts.
-We show that populations with more pronounced initial spatial structure had substantially increased extinction probabilities and that evolution of dispersal ability was insufficient to rescue faltering populations.
-Rather, a population's fate during climate change was determined by the evolved composition of dispersal phenotypes within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
-Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that the initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.
+Climate change is an increasingly severe threat facing populations around the globe, necessitating a robust understanding of the ecological and evolutionary mechanisms dictating population responses.
+Population dynamics of range shifts, among the most commonly observed responses to climate change, can be influenced by many factors, including evolution of key traits, the degree of local adaptation, and the nature of the range edge.
+Here, we use an individual-based model to explore the interacting roles of these factors in the dynamics of climate-induced range shifts.
+We show that aspects of the spatial population structure within the initial range, in particular the potential for local adaptation, severely increased a population's extinction risk.
+Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations.
+Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes that evolved within the initial range; only populations consisting of highly dispersive individuals prior to the onset of climate change survived.
+Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that initial spatial population structure plays a pivotal role in determining the outcome of climate-induced range shifts.
 
@@
 Indeed, contemporary range shifts have already been observed in a wide variety of taxa, ranging from algae to mammals~\citep{chen2011rapid, parmesan2006ecological}.
-Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages as not all species are expected to respond to climate change or to shift their ranges at the same rate~\citep{hobbs2009novel}.
-Understanding the dynamics of such climate-induced range shifts could play a key role in informing current and future conservation work.
+Such range shifts present significant challenges to current and future conservation efforts as they can result in the extinction of populations failing to track a changing climate~\citep{parmesan2006ecological} as well as the creation of novel species assemblages~\citep{hobbs2009novel}.
+Understanding the ecological and evolutionary dynamics of such climate-induced range shifts will play a key role in informing current and future conservation work.
 
-While the study of range shifts due to climate change is relatively new, important insights can be gained from the related but distinct process of range expansion, such as occurs when invasive species spread through a new environment.
-Decades of research on range expansions have led to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms responsible for shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal}.
-As range shifts and expansions exhibit similar spatial dynamics, such as a leading edge of population advance with relatively low population density, they are likely to be affected by similar ecological and evolutionary mechanisms ~\citep{hargreaves2014evolution}.
-However, while insights from range expansions are valuable for understanding and predicting dynamics of range shifts, it is important to recognize that these two processes have significant differences as well.
+Large-scale population movements have been studied for decades in the context of range expansions (e.g.
+of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
+For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
+As fundamentally similar spatial processes, it is likely that range shifts will also be subject to these ecological and evolutionary mechanisms known to drive range expansions.
+However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population.
+In particular, range shifts occur in populations with far more complex spatial structure compared to most range expansions, which typically begin from the successful establishment and spread of a small, founding population~\citep{hastings2005spatial}.
+While these founding populations often lack any significant spatial structure, populations undergoing range shifts are characterized by a spatial population structure formed by aspects of the previously stable ranges.
+For example, population ranges can vary in their potential for local adaptation throughout the range, the nature of the range edge, and the spatial distribution of key traits.
 
-In particular, range expansions and shifts due to climate change involve fundamentally distinct initial population structure.
-Range expansions, especially for invasive and reintroduced species, typically begin from the successful establishment and subsequent spread of a small, founding population~\citep{hastings2005spatial}.
-Such founding populations often represent samples from some larger source population and as such lack any initial spatial population structure.
-In contrast, range shifting populations initially have a more complex, but stable population structure that becomes disrupted with climate change or other environmental perturbations~\citep{hargreaves2014evolution}.
-A population's stable population structure incudes spatial variability in population size, adaptive traits, and range boundaries that manifests through a combination of local adaptation within the range and the severity of the gradient in population size from the range core to edge~\citep{hargreaves2014evolution, hargreaves2015fitness, henry2013eco}.
+Each of these factors relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions.
+For example, the underlying mechanism responsible for the gradient in population size from the range core to the edge (i.e.
+declines in carrying capacity versus growth rate) alters a population's extinction risk during climate driven range shifts~\citep{henry2013eco}.
+Additionally, the potential for local adaptation throughout a range has been related to extinction risk during range shifts.
+Specifically, a low potential for local adaptation can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block the establishment of better adapted genotypes~\citep{atkins2010local}.
+While these aspects of spatial population structure have been shown to impact the dynamics of climate-induced range shifts in isolation, it is unclear how and if they might interact.
 
-A population's initial spatial structure has previously been shown to dramatically impact its response to climate change and its corresponding extinction probability.
-For example, the underlying mechanism responsible for the gradient in population size from the middle of the population (i.e.
-the more dense core) to the range edge can directly change the probability of extinction a species faces during a climate driven range shift~\citep{henry2013eco}.
-Extinction risk is also related to the degree of local adaptation within the range.
-Specifically, a broad environmental niche (i.e.
-little local adaptation) can decrease a population's ability to track a changing climate if dispersal occurs in a stepping stone manner, allowing some individuals to block dispersal of better adapted genotypes~\citep{atkins2010local}.
-Local adaptation also has the potential to interact with dispersal evolution during climate change, driving increased dispersal probability in an asexual species as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
+Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts.
+In asexual species, for example, local adaptation has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
+However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
+Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
+In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
+In addition to potential interactions between local adaptation and dispersal evolution, the nature of the range edge could influence the potential for rapid trait evolution during range shifts.
+For example, the severity of the environmental gradient forming the range edge has been shown to alter the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.
 
-While the roles of dispersal evolution and local adaptation have been explored in isolation, using haploid, single locus models assuming asexual reproduction, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
-For example, evolution of increased dispersal could simultaneously reduce local adaptation within a population due to increased gene flow throughout the range.
-In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
-Additionally, the distribution of dispersal phenotypes within a population likely is influenced by the severity of the gradient at the range edge~\citep{henry2013eco, hargreaves2014evolution}, further complicating the relationship between dispersal evolution and local adaptation during climate-induced range shifts.
-
-Here, we develop a two-sex individual-based model with two genetically determined traits, one defining an individual's expected dispersal distance and the other determining an individual's environmental niche.
-In the model, we vary the degree of spatial population structure in the initial stable range to ascertain how the severity of the gradient at the range edge impacts a population's ability to track a changing climate.
-Further, we contrast the dynamics of extant and extinct populations to understand the joint impact of dispersal evolution and local adaptation on extinction risk during climate-induced range shifts.
+Here, we assess the interaction of multiple aspects of spatial population structure with trait evolution in sexually reproducing populations undergoing climate-induced range shifts.
+We develop an individual-based model capable of incorporating a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
+One trait determines dispersal ability while the second defines an individual's environmental niche.
+Using this model, we vary both the potential for local adaptation within the range and the nature of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
+By contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.
 
 \section*{Methods}
-A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in the online supplemental materials, while here we present a brief summary.
-Each simulated landscape consisted of a discrete lattice of habitat patches with a fixed width in the $y$ dimension and unbounded length in the $x$ dimension.
-Environmental conditions varied along the $x$ dimension and remained constant across the $y$ dimension (Fig.
-S1).
-To simulate climate change, environmental conditions shifted at a constant rate in the $x$ dimension.
-In each generation, competition and reproduction occurred within each patch and local populations were linked via dispersal, assuming wrapping boundaries in the $y$ dimension to prevent edge effects.
+A full description of the individual-based model using the Overview, Design concepts, and Details protocol~\citep{grimm2010odd} is available in Appendix A, while we present a brief summary here.
+Population dynamics occurred within discrete habitat patches embedded in a two dimensional lattice in which environmental conditions varied along the $x$ dimension but remained constant along the $y$ dimension (Fig.
+A1).
+Landscapes were unbounded in the $x$ dimension but defined by a fixed width in the $y$ dimension.
+Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions.
+To simulate climate change, environmental conditions shifted at a constant rate along the $x$ dimension.
+Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig.
+A2).
 
-Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci, assuming no linkage.
+Individuals were characterized by two traits (dispersal and an environmental niche), both defined by a set of $5$ quantitative diploid loci.
 While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits.
 The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
-An individual's environmental niche trait allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness.
-The environmental optimum of individual patches could then be systematically altered across the range to allow for varying degrees of local adaptation (i.e.
+An individual's environmental niche value allowed for local adaptation; the closer the niche value to the environmental optimum of the individual's patch, the higher the individual's realized fitness.
+The environmental optimum of individual patches could then be systematically varied across the range to allow for different degrees of local adaptation (i.e.
 larger changes in environmental optima allowed for greater local adaptation of the population).
-In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see supplemental materials; Fig.
-S1).
+In addition to the potential for local adaptation, simulated ranges were characterized by a decline in patch carrying capacity from the range center to the edge, the severity of which could be adjusted without altering the total carrying capacity of the landscape (see Appendix A).
 Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}.
-We simulated non-overlapping generations consisting of discrete dispersal and reproduction phases (Fig.
-S2).
+Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness.
+Allele inheritance was subject to mutation and assumed no linkages among loci.
 
-To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table S1 and S2).
+To determine the effect of spatial population structure on the eco-evolutionary dynamics of range shifts, we varied parameter combinations to explore the interacting roles of local adaptation and the severity of the gradient in environmentally suitable habitat at the range edge (Table A1 and A2).
 Specifically, we considered a factorial combination of three experimental factors: (1) no, low, and high potential for local adaptation, (2) shallow, moderate, and stark gradients in suitable habitat at the range edge, and (3) slow, moderate, and fast speeds of climate change.
@@
 Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
-Figure~\ref{fig:SimExample} shows an example of a single simulation undergoing climate change.
+Figure 1 shows an example of a single population responding to a moderate speed of climate change.
 For each scenario, we evaluated the role of dispersal evolution and initial spatial population structure in driving the dynamics of the range shifting populations.
+We primarily discuss simulations using a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.
 
@@
 \section*{Results}
-In all scenarios, populations shifted their range in response to climate change, however the proportion of populations that successfully tracked the changing climate for the full $100$ generations depended on the initial spatial population structure.
-Successful populations shifted their spatial distributions at essentially the same rate as climate change (e.g.
-Fig.
-1), while populations that failed to keep pace and lagged behind the shifting climate went extinct.
+In all scenarios, some populations shifted their ranges in response to climate change.
+However the proportion of extinct populations that failed to track the changing climate depended on the initial spatial population structure.
 Populations defined by a higher potential for local adaptation and by stark habitat gradients at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
 2).
-While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction dynamics, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change.
-Here, we widely varied both parameters (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table S2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space, and corresponding biological scenarios.
+While both aspects of a population's range influenced extinction probabilities, the potential for local adaptation drove more dramatic changes to extinction risk, with greater changes in environmental optima across the landscape causing severe increases in the probability of extinction during climate change.
+We varied both parameters widely (the potential for local adaptation doubled from the low to high scenario and the parameter defining the severity of the environmental gradient was increased by a factor of 100 from shallow to stark gradients; Table A2), suggesting that potential for local adaptation may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios.
 Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig.
-S3 \& S4).
+B1 \& B2).
 However, this effect was independent of the roles of local adaptation and the habitat gradient at the range edge in determining the extinction probability during a range shift.
+
+Counterintuitively, populations that survived climate change tended to be characterized by reduced fitness at the range margins prior to the onset of climate change compared to populations that went extinct (Fig.
+B3-B5).
+Essentially, populations with initially higher degrees of local adaptation at the range edges, and thus greater fitness, were more likely to go extinct during climate change.
+This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
+As expected, there was no spatial variation in fitness for populations with no potential for local adaptation.
 
@@
 3a\&b).
-Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes to dispersal phenotypes strongly centered on $0$ (Fig.
-S5-S7).
-The similarity in evolved changes to dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions versus became extinct.
+Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig.
+B3-B5).
+The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.
 
-Instead, the stable distribution of dispersal phenotypes before the beginning of climate change played a large role in determining the ultimate fate of simulated populations.
-Simulated populations evolved a range of dispersal phenotypes over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
-S8-S10).
-The populations that survived climate change were those that happened to make up the upper end of these distributions, containing individuals with high dispersal phenotypes (Fig.
+Instead, the initial distribution of dispersal phenotypes prior to the onset of climate change played a key role in determining a population's fate.
+A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the potential for local adaptation and the severity of the habitat gradient at the range edge (Fig.
+B9-B11).
+Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig.
 3c\&d).
-In fact, comparing dispersal phenotypes of successful populations across different parameter combinations revealed a threshold in dispersal phenotypes separating individuals from successful and extinct populations (Fig.
-S8-S10).
-This threshold varied directly with the speed of climate change, increasing with more rapid climate change.
-Across all speeds of climate change, this threshold corresponds to roughly similar probabilities ($12\%$, $9\%$, and $7\%$ at slow, moderate, and fast speeds respectively) of an individual keeping pace with climate change (i.e.
-moving the correct number of patches in the $x$ dimension to track the changing environmental conditions).
-
-Intriguingly, the populations that survived climate change tended to be characterized by reduced fitness at the range margins compared to populations that went extinct (Fig.
-S11-S13).
-This pattern was most evident in the simulations with either (1) a gradual environmental gradient at the range edge or (2) a high potential for local adaptation.
-Intuitively, there was no spatial pattern in fitness for populations with no potential for local adaptation.
-As the populations that survived climate change were also characterized by heightened dispersal (Fig.
-3), the observed reduction in average patch fitness at the margins is likely due to increased gene flow from the range core hampering the abilities of these edge populations to adapt to local conditions.
+In fact, comparing the full distribution of initial dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations.
+Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig.
+B9-B11).
+To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change used in our simulations (see the model description in Appendix A).
+This dispersal phenotype matched the observed threshold value distinguishing surviving from extinction populations in all experimental scenarios (Figures B9-B11, vertical dashed line).
+Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change.
 
@@
 Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
-Here we developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
+We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
 In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g.
 single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
-Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under realistic levels of biological complexity.
+Here, we tested the generality of previous results and the interplay of eco-evolutionary dynamics under increased levels of biological complexity.
 Specifically, we demonstrated the role of spatial population structure, in the form of local adaptation and the environmental gradient defining the range edge, in determining extinction risk for range shifting populations via impacts on the initial distribution of dispersal phenotypes and environmental niche values.
@@
 3c\&d).
-Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a given threshold value.
+Survival in the face of climate change was primarily determined by the dispersal phenotypes making up the population, specifically whether the population included individuals with dispersal phenotypes at or above a threshold value.
 Importantly, the threshold necessary to survive climate change itself was constant across all parameter combinations for a given speed of climate change (Fig.
-S8-S10).
+B9-B11).
 Scenarios with no potential for local adaptation and gradual environmental gradients had larger proportions of high dispersal phenotypes under stable climate conditions, and therefore a lower probability of extinction during climate change.
@@
 Thus, attributes defining the spatial structure of the range altered the distribution of dispersal phenotypes under stable climate conditions, subsequently determining the extinction risk of populations during climate change.
-Importantly, dispersal evolution during climate change was unable to counter the influence of spatial population structure in the stable range on extinction dynamics.
+Importantly, dispersal evolution during climate change was unable to counter the influence of initial spatial population structure on extinction dynamics.
 
-While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range margins when populations had a moderate to high potential for local adaptation (Fig.
-S11-S13).
-In the model, populations at the range margins tended to have lower abundance than populations in the range core, thus increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
-Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core reduced fitness at the range edge by preventing adaptation to local conditions.
-As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness prior to the onset of climate change.
+While high dispersal phenotypes prior to climate change increased the probability that populations tracked changing conditions, it had the additional effect of reducing average fitness at the range edges when populations had a moderate to high potential for local adaptation (Fig.
+B3-B5).
+In the model, populations at the range edges tended to have lower abundance than populations in the range core, increasing their susceptibility to gene flow from the core~\citep{lenormand2002gene}.
+Thus, in populations with high dispersal phenotypes prior to climate change, increased gene flow from the core likely reduced fitness at the range edge by preventing adaptation to local conditions.
+As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges prior to the onset of climate change.
 While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that populations exhibiting high levels of local adaptation within their stable range are likely to be at greater risk of extinction during periods of climate change.
@@
 Our results suggest this is not always the case, and in fact may only be possible under certain, relatively narrow conditions.
-Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid and single-locus models~\citep{boeye2013more, hargreaves2015fitness}.
-As dispersal evolution during range expansions and shifts occurs via the formation of strong spatial structure from the core to the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of highly dispersive alleles compared to our more complex model multiple diploid loci contributing to dispersal.
-Additionally, many previous models of climate-induced range shifts assume asexual reproduction which can result in more rapid rates of evolutionary change under certain conditions~\citep{crow1965evolution, smith1968evolution}.
+Previous models showing that dispersal evolution may rescue populations during climate change have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}.
+As dispersal evolution during range expansions and shifts occurs via the spatial sorting of alleles contributing to heightened dispersal at the range edge~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait.
 The negligible role played by dispersal evolution in our model (Fig.
@@
 \subsection*{Conclusion}
-Understanding the various ecological and evolutionary drivers of climate-induced range shifts is crucial to current and future conservation efforts.
-In particular, a deeper, process-based understanding of how various mechanisms combine to shape extinction probabilities of populations undergoing range shifts, will, in turn, allow more focused interventions.
-Our results suggest that initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
-Further, in contrast to other studies assuming more simplified breeding and genetic structure, we find very little role for the evolution of heightened dispersal abilities to rescue a population composed of low-dispersing individuals.
+As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions.
+In particular, a deeper, process-based understanding of extinction risk in populations undergoing range shifts will, in turn, allow more focused conservation interventions.
+Our results suggest that the initial spatial population structure, as determined by local adaptation and the environmental gradient at the range edge, has the potential to dramatically alter the extinction probability faced by species responding to climate change.
+Further, in contrast to other studies assuming more simplified genetic structures, we find very little role for the evolution of heightened dispersal abilities in allowing a population to successfully track climate change.
 Future work should continue to examine the interplay between initial conditions in range shifts and the potential for evolutionary rescue.
-As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such increased risk.
+As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to not only identify those factors leading to increased extinction risk in range shifting populations, but also to develop meaningful conservation strategies to mitigate such risk.
 
@@
 
-\bibliographystyle{ecology}
+\bibliographystyle{amnat}
 \bibliography{main_bib}
@@
 \paragraph{Figure 1.} A single example of a simulation with a high potential for local adaptation and a moderate habitat gradient defining the range edge.
-Information on the (a) abundance, (b) dispersal, and (c) fitness of each patch in the population is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to 40 generations after the start of climate change.
+Information on the (a) abundance, (b) dispersal, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change.
 Log transformed mean dispersal phenotypes (b) are shown for each patch.
-Average patch fitness (c) was calculated based on the mean local environmental niche phenotype and the optimal niche value for that location.
+Average patch fitness (c) was calculated based on the mean environmental niche trait of local individuals and the environmental optima for each patch.
 
@@
 Positive values indicate an increase in average dispersal ability in the patch.
-Initial distributions of the dispersal trait (c and d) are shown with histograms of the log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
+Initial distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
 In all panels, values associated with extant populations are shown in dark blue.
 Results are shown for populations with no potential for local adaptation and a gradual environmental gradient at the range boundary (a and c; $n = 155$ extant populations) and for populations with a high potential for local adaptation and a stark gradient at the range boundary (b and d; $n = 14$ extant populations).
-Full results for all parameter combinations are provided in the online supplement.
+Full results for all parameter combinations are provided in Appendix B.
 

### 2019-04-05 Manuscript/Supp.tex
Adjusted the workflow of the model and the way dispersal inheritence is implemented

@@
 \renewcommand{\theequation}{S\arabic{equation}}
+
+To prove this, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{hastings2005spatial}, where $r$ is the intrinsic growth rate of the population and $D$ is the diffusion coefficient.
+The $2$ dimensional diffusion coefficient can be calculated directly from the dispersal phenotypes as $D=\frac{1}{4}d^{2}$ since $d^{2}$ represents the mean squared displacement of an individual with phenotype $d$ during a dispersal period
 

### 2019-04-05 1_MainSim.R

@@ -3,7 +3,7 @@
 #    combinations to be explored.
 
 # Set the number of processors and number of simulations to be run
-nProc <- 24*6
+nProc <- 24*18
 NumSims <- 200
 
 # Set the working directory and load necessary data and libraries
@@ -21,7 +21,6 @@ Vm <- c(0.0004, 0.0004)
 Lf <- 5
 Ld <- 5
 Rmax <- 2
-Kmax <- 100
 width <- 10
 kern <- "exp"
 EnvGradType <- "K"
@@ -40,14 +39,15 @@ z <- 0.5
 dmax <- 1000
 rho <- 0.5
 
-AllParams <- vector(mode = "list", length = 3)
+AllParams <- vector(mode = "list", length = 9)
 
-for(i in 1:3){
+for(i in 1:9){
      gamma <- RangeParams$gamma[i]
      lambda <- RangeParams$lambda[i]
      tau <- RangeParams$tau[i]
      eta <- RangeParams$eta[i]
-     
+     Kmax <- RangeParams$Kmax[i]
+
      AllParams[[i]] <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
                         Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
                         LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
@@ -93,7 +93,7 @@ SimFunc <- function(i){
 }
 
 # Create a vector of parameter index values for the parallel computation
-SimVec <- rep(1:3, each = NumSims)
+SimVec <- rep(1:9, each = NumSims)
 
 # Create the cluster and run the simulations
 cl <- makeCluster(nProc - 1, type = "MPI")

### 2019-04-05 1_MainSim.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=08:00:00,nodes=18:ppn=24,mem=500gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the parameter combination to use and generate names of R scripts and log files
+Rscript=1_MainSim.R
+LogFile=1_MainSim.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-05 1a_SimCheck.R

@@ -4,6 +4,7 @@
 
 setwd("~/ShiftingSlopes/MainSim/")
 NumSims <- rep(NA, 9)
+MaxSims <- 200
 
 for(p in 1:9){
      ParamDirectory <- paste("~/ShiftingSlopes/MainSim/Params", p, sep = "")
@@ -21,6 +22,13 @@ for(p in 1:9){
                     system(SysCommand)
                }
           }
+     }
+     TotalSims <- length(list.files(ParamDirectory))
+     while(TotalSims > MaxSims){
+          ParamSims <- list.files(ParamDirectory)
+          SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[1], sep = "")
+          system(SysCommand)
+          TotalSims <- length(list.files(ParamDirectory))
      }
      NumSims[p] <- length(list.files(ParamDirectory))
 }

### 2019-04-05 1a_SimCheck.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=00:45:00,nodes=1:ppn=1,mem=2gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=1a_SimCheck.R
+LogFile=1a_SimCheck.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-05 1b_MainSimCont.sh

@@ -1,19 +1,19 @@
 #!/bin/bash -l
-#PBS -l walltime=15:00:00,nodes=18:ppn=24,mem=500gb
+#PBS -l walltime=08:00:00,nodes=6:ppn=24,mem=300gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the parameter combination to use and generate names of R scripts and log files
-Rscript=1a_MainSim.R
-LogFile=1a_MainSim.log
+Rscript=1b_MainSimCont.R
+LogFile=1b_MainSimCont.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
+module load ompi/3.0.0/gnu-7.2.0-centos7
 
 export RMPI_TYPE=OPENMPI
 export OMPI_MCA_mpi_warn_on_fork=0

### 2019-04-05 4_Extinctions.R

@@ -9,9 +9,29 @@ nProc <- 24*2
 LengthShift <- 100
 TotalTime <- 2150
 BurnIn <- 2000
+NumSims <- 200
 
-# Create an object to hold the extinction data
+# Create two objects to hold the extinction data: 1) an array to hold the number
+#    of extinctions through time to use for the plots of the probability of extinction
+#    and 2) a data frame with each simulation ID and a column of 1's and 0's for
+#    each speed of climate change with 1 indicating extant and 0 indicating extinct
 Extinctions <- array(NA, dim = c(3, 9, LengthShift))
+SimIDs <- data.frame(SimID = rep(NA, NumSims*9), Params = rep(NA, NumSims*9),
+                     Slow = rep(NA, NumSims*9), Moderate = rep(NA, NumSims*9), 
+                     Fast = rep(NA, NumSims*9))
+
+# First, populate the SimID column of the above data frame
+k <- 1
+for(p in 1:9){
+     FilePath <- paste("~/ShiftingSlopes/MainSim/Params", p, sep = "")
+     SimFiles <- list.files(FilePath)
+     for(i in 1:NumSims){
+          CurSimID <- strsplit(x = SimFiles[i], split = "/")[[1]][5]
+          SimIDs$SimID[k] <- CurSimID
+          SimIDs$Params[k] <- p
+          k <- K + 1
+     }
+}
 
 SimVec <- NULL
 SpeedWords <- c("Slow", "MainSim", "Fast")
@@ -54,5 +74,34 @@ for(v in 1:3){
 
 # Save the extinction output
 save(Extinctions, file = "Extinctions.rdata")
+
+# Also, use the function output to construct a data frame of simulation IDs,
+#    parameter combinations, and columns for slow, moderate, and fast speeds of
+#    climate change with 1 for extant and 0 for extinct
+for(i in 1:nrow(SimIDs)){
+     SlowPattern <- paste("Slow/Params", SimIDs$Params[i], "/", SimIDs$SimID[i], sep = "")
+     ModeratePattern <- paste("MainSim/Params", SimIDs$Params[i], "/", SimIDs$SimID[i], sep = "")
+     FastPattern <- paste("Fast/Params", SimIDs$Params[i], "/", SimIDs$SimID[i], sep = "")
+     SlowIndex <- grep(pattern = SlowPattern, x = SimVec)
[…]

### 2019-04-05 5_FitDataNew.sh

@@ -1,19 +1,19 @@
 #!/bin/bash -l
-#PBS -l walltime=08:00:00,nodes=6:ppn=24,mem=300gb
+#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
-# Set the parameter combination to use and generate names of R scripts and log files
-Rscript=1c_MainSimCont.R
-LogFile=1c_MainSimCont.log
+# Set the names of R scripts and log files
+Rscript=5_FitDataNew.R
+LogFile=5_FitDataNew.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
+module load ompi/3.0.0/gnu-7.2.0-centos7
 
 export RMPI_TYPE=OPENMPI
 export OMPI_MCA_mpi_warn_on_fork=0

### 2019-04-05 5_FitData_new.R

@@ -1,0 +1,117 @@
+# This script will extract the realized fitness of individuals throughout their
+#    ranges in the generation immediately prior to climate change
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
+
+# Create other useful objects to pass to the nodes
+xSeq <- -40:40
+
+FitExtract <- function(i){
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo, "/", CurSims$SimID[1],
+                        "/parameters.R", sep = "")
+     source(ParamFile)
+     
+     # Create the data frames for extant and extinct simulations
+     ExtantFits <- data.frame(x = xSeq, wBar = rep(NA, length(xSeq)),
+                              lwr = rep(NA, length(xSeq)), upr = rep(NA, length(xSeq)))
+     ExtinctFits <- data.frame(x = xSeq, wBar = rep(NA, length(xSeq)),
+                              lwr = rep(NA, length(xSeq)), upr = rep(NA, length(xSeq)))
+     
+     # Now loop through each row of the data frame to fill it in
+     for(j in 1:length(xSeq)){
+          ExtantVals <- NULL
+          ExtinctVals <- NULL
+          for(s in 1:nrow(CurSims)){
+               FilePath <- paste("~/ShiftingSlopes/", SpeedWord, "/Params", ParamCombo, "/", CurSims$SimID[s], "/InitialPopMat.csv", sep = "")
+               SimData <- read.csv(InFile)
+               if( (j == 1) & (s == 1)){ # Only need to do this once
+                    FitCols <- grep(pattern = "fit", x = names(SimData)) 
[…]

### 2019-04-05 6_DispDataNew.R

@@ -1,0 +1,142 @@
+# This script will extract the dispersal phenotypes of individuals throughout their
+#    ranges every 10 generations
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
+
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 100
+GenSeq <- seq(0, LengthShift, by = GenSteps)
+
+DispExtract <- function(i){
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs.
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste("~/ShiftingSlopes/", SpeedWord, "/Params", ParamCombo, 
+                        "/", CurSims$SimID[1], "/parameters.R", sep = "")
+     source(ParamFile)
+     
+     # Create the data frames for extant and extinct simulations
+     ExtantDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
+     ExtinctDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
+     
+     # Now loop through each row of the data frame to fill it in
+     for(g in GenSeq){
+          # Create temporary data frames to hold the results from all simulations
+          #    for each time point, which will then be condensed across simulations
+          #    for the main data frames
+          TempExtant <- data.frame(x = NA, dBar = NA)
+          TempExtinct <- data.frame(x = NA, dBar = NA)
+          for(s in 1:nrow(CurSims)){
+               if(g == 0){
+                    InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
+                                    "/", CurSims$SimID[s], "/SummaryStats.csv", sep = "")
+                    SumStats <- read.csv(InFile)
[…]

### 2019-04-05 6_DispData_new.R

@@ -1,0 +1,181 @@
+# This script will extract the initial dispersal values of individuals immediately
+#    prior to climate change (DispInit) and the change in mean dispersal values at
+#    the patch level (along with the relative x coordinate of the patch) to assess
+#    the role of evolution (DispEvol).
+
+# Set the number of processors available for this scrip
+nProc <- 4*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
+
+# Set some initial parameter values to assess evolutionary change in dispersal
+Threshold <- 10
+BetaInit <- 0
+BurnIn <- 2000
+BurnOut <- 50
+LengthShift <- 100
+dmax <- 1000
+eta <- RangeParams$eta[1]
+rho <- 0.5
+ClimSpeeds <- c(0.5, 1, 2)
+MaxGen <- c(151, 2151, 151)
+EndShift <- c(LengthShift, BurnIn + LengthShift, LengthShift)
+BetaCoord <- matrix(NA, nrow = 3, ncol = BurnIn + LengthShift + BurnOut)
+for(i in 1:3){
+     BetaShift <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, 
+                                eta = eta, v = ClimSpeeds[i]) / eta
+     BetaCoord[i,] <- c(rep(BetaInit, BurnIn), BetaShift, rep(BetaShift[LengthShift], BurnOut))
+}
+
+DispExtract <- function(i){
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo, "/", CurSims$SimID[1],
+                        "/parameters.R", sep = "")
+     source(ParamFile)
+     
[…]

### 2019-04-05 6_DispData_new.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=12:00:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=6_DispData_new.R
+LogFile=6_DispData_new.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-05 7_DispGenVarDataNew.R

@@ -1,0 +1,142 @@
+# This script will extract the genetic variance in dispersal throughout the range
+#    every 10 generations
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
+
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 100
+GenSeq <- seq(0, LengthShift, by = GenSteps)
+
+GenVarExtract <- function(i){
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs.
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste("~/ShiftingSlopes/", SpeedWord, "/Params", ParamCombo, 
+                        "/", CurSims$SimID[1], "/parameters.R", sep = "")
+     source(ParamFile)
+
+     # Create the data frames for extant and extinct simulations
+     ExtantGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
+     ExtinctGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
+     
+     # Now loop through each row of the data frame to fill it in
+     for(g in GenSeq){
+          # Create temporary data frames to hold the results from all simulations
+          #    for each time point, which will then be condensed across simulations
+          #    for the main data frames
+          TempExtant <- data.frame(x = NA, GenVar = NA)
+          TempExtinct <- data.frame(x = NA, GenVar = NA)
+          for(s in 1:nrow(CurSims)){
+               if(g == 0){
+                    InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
[…]

### 2019-04-05 7_DispGenVarDataNew.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=7_DispGenVarDataNew.R
+LogFile=7_DispGenVarDataNew.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-05 7_VarData.R

@@ -1,0 +1,79 @@
+# This script will extract information on the per patch variability in fitness,
+#    and the variability in mean trait values and abundance across the y dimension.
+
+# CV = standard deviation / mean
+
+
+# Set the number of processors available for this script
+nProc <- 4*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+NumSims <- 200
+
+# Create a data frame to index each simulation block in the above arrays
+TraitIndices <- expand.grid(params = 1:9, sim = 1:NumSims)
+
+# Write a function to be passed to various nodes
+VarExtract <- function(i){
+     # Sort out the parameter combination and simulation under consideration
+     Param <- TraitIndices$params[i]
+     ParamDirectory <- paste("~/ShiftingSlopes/MainSim/Params", Param, "/", sep = "")
+     AllSims <- list.files(ParamDirectory)
+     SimID <- AllSims[TraitIndices$sim[i]]
+     
+     # Load the corresponding population matrix
+     InFile <- paste("~/ShiftingSlopes/MainSim/Params", Param, "/", SimID,
+                     "/SummaryStats.csv", sep = "")
+     SimData <- read.csv(InFile)
+     PopData <- subset(SimData, gen == 2000, select = c(x, y, abund, muFit, muDisp, sigmaFitPhen))
+     return(PopData)
+}
+
+# Create the cluster and run the simulations
+cl <- makeCluster(nProc - 1, type = "MPI")
+
+# Export the necessary objects to each node
+clusterExport(cl, "TraitIndices")
+
+# Run the simulations
+SimVec <- 1:dim(TraitIndices)[1]
+SimVar <- clusterApply(cl, x = SimVec, fun = VarExtract)
+
+# For each scenario (1-9), I want a data frame with columns: x, abund, muFit, muDisp, sigmaFitPhen
+# All of these columns except the last will be the variance across the y dimension for
+#    these per patch quantities. sigmaFitPhen will be the mean across the y dimension
+#    to get a sense of the average variance in reproductive success.
+
[…]

### 2019-04-05 8_FitGenVarDataNew.sh

@@ -1,19 +1,19 @@
 #!/bin/bash -l
-#PBS -l walltime=00:15:00,nodes=1:ppn=1,mem=2gb
+#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the names of R scripts and log files
-Rscript=1b_SimCheck.R
-LogFile=1b_SimCheck.log
+Rscript=8_FitGenVarDataNew.R
+LogFile=8_FitGenVarDataNew.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/
 
 # Load R and MPI
 module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0
+module load ompi/3.0.0/gnu-7.2.0-centos7
 
 export RMPI_TYPE=OPENMPI
 export OMPI_MCA_mpi_warn_on_fork=0

### 2019-04-05 8a_ExtinctionGraphs.R

@@ -1,5 +1,5 @@
 # This script will create the graphs from the extinction data.
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
+setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
 library(RColorBrewer)
 RangeParams <- read.csv("RangeParameters.csv")
 NumSims <- 200
@@ -17,7 +17,7 @@ OuterMar <- c(1,1,0,0)
 TextSize <- 1.15
 LegTextSize <- 1.5
 AxisSize <- 1.25
-LetterPos <- c(2, 0.975)
+LetterPos <- c(2, 0.925)
 FigWidth <- 8
 FigHeight <- 3
 xRange <- c(0,100)
@@ -46,6 +46,7 @@ for(i in 1:3){
           lines(x = 1:100, y = Extinctions[i,3,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
           text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("a")), cex = TextSize)
+          abline(h = 1, lty = 1, col = "grey") #lty = 4
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)
@@ -60,6 +61,7 @@ for(i in 1:3){
           lines(x = 1:100, y = Extinctions[i,6,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
           text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("b")), cex = TextSize)
+          abline(h = 1, lty = 1, col = "grey")
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)
@@ -76,6 +78,7 @@ for(i in 1:3){
           lines(x = 1:100, y = Extinctions[i,9,], lty = GradLineType[3], lwd = LineWidth,
                 col = Col)
           text(x = LetterPos[1], y = LetterPos[2], labels = expression(bold("c")), cex = TextSize)
+          abline(h = 1, lty = 1, col = "grey")
           # Add the axes
           axis(1, at = TimeAxisSeq1, cex.axis = AxisSize)
           axis(1, at = TimeAxisSea2, labels = FALSE, tcl = -0.25)

### 2019-04-05 8b_DispInitGraphs_new.R

@@ -1,0 +1,116 @@
+# This script will make an appropriate graph of the initial dispersal values
+setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
+load("SimData/DispInitNew.rdata")
+load("SimData/ExtSimIDs.rdata")
+CurSpeed <- 1
+SpeedWords <- c("Slow", "Main", "Fast")
+
+# Calculate the dispersal phenotype corresponding to the current speed of climate
+#    change
+r <- log(2)
+patch_speeds <- c(0.75, 1, 1.25)
+eta <- 50
+v <- patch_speeds[CurSpeed]*eta
+DispPhen <- sqrt(v^2 / r)
+
+# Use SimIDs to calculate a vector of the number of extant simulatins for
+#    the current conditions
+NumExtant <- rep(NA, 9)
+for(i in 1:9){
+     CurSims <- subset(SimIDs, Params == i)
+     NumExtant[i] <- sum(CurSims[,CurSpeed+2])
+}
+
+# Set some graphical parameters to use for the subsequent figures
+options(scipen = -10)
+FigMat <- matrix(NA, nrow = 3, ncol = 12)
+FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
+FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
+FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
+OuterMar <- c(4, 12, 6, 2)
+InnerMar <- c(1.5, 2.25, 1.5, 2.25)
+TextSize <- 1.5
+AxisSize <- 1.15
+ArrowWidth <- 3
+ArrowLength <- 0.25
+FigWidth <- 8
+FigHeight <- 6
+DispAxisLabel <- expression(paste("Equilibrium ", italic("log"), "(", italic("d"), ")", sep = ""))
+SimSeq <- c(7,8,9,4,5,6,1,2,3)
+xLabLine <- 2
+yLabLine <- 3
+GradLabLine <- 3.5
+GradSubLine <- 1.5
+AdaptLabLine <- 9
+AdaptSubLine <- 7
+TopArrow <- matrix(c(-0.75, 240000, 17, 240000), nrow = 2, ncol = 2, byrow = TRUE)
+SideArrow <- matrix(c(-5, -550000, -5, 175000), nrow = 2, ncol = 2, byrow = TRUE)
[…]

### 2019-04-05 8c_InitFitGraphs.R

@@ -1,7 +1,7 @@
 # This script will make an appropriate graph of the initial dispersal values
 setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
 load("SimData/FitData.rdata")
-CurSpeed <- 2
+CurSpeed <- 1
 SpeedWords <- c("Slow", "Main", "Fast")
 
 # Sort the disp values into the appropriate lists
@@ -72,8 +72,8 @@ for(p in 1:9){
 
 # Set some graphical parameters to use for the subsequent figures
 yRange <- c(0, 1)
-xRange <- c(-40, 40)
-SmallTicks <- seq(-40, 40, by = 5)
+xRange <- c(-50, 50)
+SmallTicks <- seq(-35, 35, by = 5)
 OuterMar <- c(4, 10, 5, 2)
 InnerMar <- c(1.5, 1.5, 2, 1.5)
 TextSize <- 1.5
@@ -90,7 +90,7 @@ GradLabLine <- 3
 GradSubLine <- 1.5
 AdaptLabLine <- 8
 AdaptSubLine <- 5
-TopArrow <- matrix(c(-10, 1.3, 220, 1.3), nrow = 2, ncol = 2, byrow = TRUE)
+TopArrow <- matrix(c(-12, 1.3, 250, 1.3), nrow = 2, ncol = 2, byrow = TRUE)
 LowAdj <- 0.15
 HighAdj <- 0.85
 ExtinctCol <- "red"
@@ -122,11 +122,11 @@ pdf(file = PlotName, width = FigWidth, height = FigHeight, onefile = FALSE, pape
                         y1 = ExtinctPlot[[i]]$upr, lwd = LineWidth, col = ExtinctCol)
           }
           # Add the selection and environmental gradient arrows
-          mtext("Adaptation potential", side = 2, outer = TRUE, line = AdaptLabLine,
+          mtext("Gradient in niche optimum", side = 2, outer = TRUE, line = AdaptLabLine,
                 cex = TextSize)
-          mtext("Low", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
+          mtext("Shallow", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
                 adj = LowAdj)
-          mtext("Strong", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
+          mtext("Steep", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
                 adj = HighAdj)
           if(i == 7){
[…]

### 2019-04-05 8d_DispEvolGraphs.R

@@ -5,8 +5,8 @@ SpeedWords <- c("Slow", "Main", "Fast")
 
 # Create objects to hold coordinates for the various arrows to use in the graph
 TopArrow <- array(NA, dim = c(3, 2, 2))
-TopArrow[1,,] <- c(-75, 600, 16000, 16000)
-TopArrow[2,,] <- c(-75, 600, 18000, 18000)
+TopArrow[1,,] <- c(-75, 600, 15000, 15000)
+TopArrow[2,,] <- c(-75, 600, 16000, 16000)
 TopArrow[3,,] <- c(-75, 600, 16000, 16000)
 SideArrow <- array(NA, dim = c(3, 2, 2))
 SideArrow[1,,] <- c(-193, -193, -34000, 10000)
@@ -45,7 +45,7 @@ range(DispEvol[[1]]$DeltaDisp, na.rm = TRUE)
 range(DispEvol[[2]]$DeltaDisp)
 range(DispEvol[[3]]$DeltaDisp, na.rm = TRUE)
 # Set the HistBreaks accordingly...
-HistBreaks <- seq(-375, 475, by = 5)
+HistBreaks <- seq(-500, 500, by = 5)
 
 # Now make the figures
 for(v in 1:3){
@@ -85,20 +85,20 @@ for(v in 1:3){
           mtext("Frequency", side = 2, outer = TRUE, line = yLabLine, cex = TextSize)
           
           # Add the adaptation and gradient text
-          mtext("Adaptation potential", side = 2, outer = TRUE, line = AdaptLabLine,
+          mtext("Gradient in niche optimum", side = 2, outer = TRUE, line = AdaptLabLine,
                 cex = TextSize)
-          mtext("None", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
+          mtext("Flat", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
                 adj = LowAdj)
-          mtext("Low", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize)
-          mtext("High", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
+          mtext("Shallow", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize)
+          mtext("Steep", side = 2, outer = TRUE, line = AdaptSubLine, cex = TextSize,
                 adj = HighAdj)
           
-          mtext("Gradient at range edge", side = 3, outer = TRUE, line = GradLabLine,
[…]

### 2019-04-05 GenerateRangeParams.R

@@ -11,8 +11,9 @@ library(grid)
 library(gridBase)
 library(gridExtra)
 
-# Make a function to calculate the range curve over a given spatial extent
-f_xt <- function(beta, gamma, tau, xVals){
+# First make a function to generate the realized patch carrying capacities throughout
+#    the range based on the range parameters
+K <- function(beta, gamma, tau, xVals, Kmax){
      f <- rep(NA, length(xVals))
      for(i in 1:length(xVals)){
           if(xVals[i] > beta){
@@ -25,7 +26,8 @@ f_xt <- function(beta, gamma, tau, xVals){
                f[i] <- numerator / denominator
           } 
      }
-     return(f)
+     Kseq <- f * Kmax
+     return(Kseq)
 }
 
 # Given a delta and minimum gamma value, calculate the tau value necessary for a
@@ -36,128 +38,129 @@ CalcTauMin <- function(delta, gamma){
      return(TauMin)
 }
 
-# When changing the slope of the range boundary, keep the area of the range
-#    constant and calculate a new tau
-CalcTauNew <- function(gamma, C){
-     numerator <- log( exp( (C*gamma) / 2 ) - 1 )
-     TauNew <- numerator / gamma
+# Calculate a new tau value constrained by maintaining the width of the range
+NewTau <- function(xHat, gamma, fHat){
+     TauNew <- xHat + (1/gamma) * log(fHat/(1-fHat))
      return(TauNew)
 }
 
 # Calculate the total area under a given range curve
-Calc_C <- function(gamma, tau){
+Integral <- function(gamma, tau){
      numerator <- 2 * log( exp(gamma * tau) + 1 )
-     C <- numerator / gamma
-     return(C)
+     Area <- numerator / gamma
+     return(Area)
 }
 
+# Finally, make a function to calculate a new Kmax to keep the total achievable
+#    population size throughout the range constant.
+NewK <- function(gamma, tau, TotalK){
+     Area <- Integral(gamma = gamma, tau = tau)
+     Kmax <- TotalK / Area
+     return(Kmax)
+}
 
-# Now set the values for gamma, delta, eta, and beta
-GammaSeq <- c(0.0025, 0.025, 0.25)
-Delta <- 1e-3
[…]

### 2019-04-05 PlotVarData.R

@@ -1,0 +1,92 @@
+# Plot the var data. Doesn't have to be pretty, just has to let me see a visualization
+#    and then get some key values
+
+load("~/Desktop/RangeShifts/ShiftingSlopesOther/SimData/VarData.rdata")
+# Determine the x range to use for all graphs
+xMin <- 0
+xMax <- 0
+for(p in 1:9){
+     xRange <- range(VarList[[p]]$x, na.rm = TRUE)
+     xMin <- min(c(xRange[1], xMin))
+     xMax <- max(c(xRange[2], xMax))
+}
+xRange <- c(xMin, xMax)
+
+# First make the graph for variation in abundances
+quartz(width = 12, height = 9)
+par(mfrow = c(3,3))
+for(p in 1:9){
+     plot(x = VarList[[p]]$x, y = VarList[[p]]$abund, xlim = xRange, xlab = "Space", ylab = "Var(abund)")
+}
+
+# Now make the graph for variation in mean niche trait
+quartz(width = 12, height = 9)
+par(mfrow = c(3,3))
+for(p in 1:9){
+     plot(x = VarList[[p]]$x, y = VarList[[p]]$muFit, xlim = xRange, xlab = "Space", ylab = "Var(niche)")
+}
+
+# Now make the graph for variation in mean dispersal
+quartz(width = 12, height = 9)
+par(mfrow = c(3,3))
+for(p in 1:9){
+     plot(x = VarList[[p]]$x, y = VarList[[p]]$muDisp, xlim = xRange, xlab = "Space", ylab = "Var(disp)")
+}
+
+# Now make the graph for variation in variance in fitness of patches
+quartz(width = 12, height = 9)
+par(mfrow = c(3,3))
+for(p in 1:9){
+     plot(x = VarList[[p]]$x, y = VarList[[p]]$sigmaFitPhen, xlim = xRange, xlab = "Space", ylab = "Patch Var(fit)")
+}
+
+
+# Look into the maximum standard deviations across the y dimension
+MaxAbundSD <- rep(NA, 9)
+MaxNicheSD <- rep(NA, 9)
+MaxDispSD <- rep(NA, 9)
+for(p in 1:9){
+     MaxAbundSD[p] <- (max(VarList[[p]]$abund, na.rm = TRUE))
+     MaxNicheSD[p] <- (max(VarList[[p]]$muFit, na.rm = TRUE))
[…]

### 2019-04-14 Manuscript/AmNat_ReSubmission.tex
Updated the data extraction and graphing files to reflect the new analyses in the manuscript and began the revisions in the manuscript itself

@@
 
-Large-scale population movements have been studied for decades in the context of range expansions (e.g.
-of invasive or reintroduced species), leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
-For example, while the speed of a range expansion can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in both of these traits can increase both the mean and variance of expansion speed through time~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
-As fundamentally similar spatial processes, it is likely that range shifts will also be subject to the same ecological and evolutionary mechanisms known to drive range expansions.
-However, range shifts involve several additional complications absent from range expansions, which must be considered when predicting the dynamics of a shifting population.
-In particular, range shifts begin with far more complex spatial population structure compared to most range expansions.
-Here and throughout the paper, we use the phrase "spatial population structure" to refer broadly to the spatial distribution of individuals, and their associated genotypes and phenotypes, within a population.
-Thus, spatial population structure can encompass factors such as spatial clines in trait values and the starkness of abundance declines characterizing the range edge.
-Range expansions (e.g.
-of invasive or reintroduced species) typically start from small founding populations brought to a new area and lacking any initial spatial population structure.
-While such populations form spatial population structure during the expansion process~\citep{weiss2017rapid, ochocki2017rapid}, established populations respond to climate change in the context of preexisting spatial population structure.
+Large-scale population movements have been studied for decades in the context of range expansions, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
+For example, while range expansion speed can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in these traits can increase both the mean and variance of expansion speed through time~\citep{szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
+In particular, laboratory experiments have shown spatial genetic structure to play an important role in the dynamics of range expansions, even over short time periods~\citep{weiss2017rapid,ochocki2017rapid}.
+While some range expansions (e.g.
+of invasive or reintroduced species) start from small founding populations lacking any initial spatial structure, others are characterized by far more complex initial spatial population structure (e.g.
+humans expanding out of Africa or tree expansion following glacial retreat).
+Climate-driven range shifts fall in the latter category, as populations respond to climate change in the context of preexisting spatial population structure.
+For example, species' ranges may be characterized by spatial clines in abundance and genetic patterns.
+Such clines can form in response to a number of extrinsic or intrinsic factors including adaptation to environmental gradients~\citep{kirkpatrick1997evolution}, altered demographic rates throughout the range~\citep{holt2005theoretical}, or interactions with other species~\citep{case2000interspecific}.
+Importantly for climate-driven range shifts, the factors influencing these spatial patterns will likely shift with the changing climate, causing simultaneous range contraction and expansion at opposite range margins.
+Thus, climate-driven range shifts will be affected both by the eco-evolutionary dynamics known to impact range expansions and the moving constraints on spatial population structure impacted by changing climatic conditions.
 
-Each factor relating to spatial population structure has the potential to affect the dynamics of range shifts under changing climatic conditions.
+Certain aspects of spatial population structure have the potential to affect the dynamics of range shifts under changing climatic conditions.
 For example, the underlying mechanism causing population declines at the range edge (e.g.
@@
 Further, given the importance of rapid trait evolution in range expansions~\citep{weiss2017rapid, ochocki2017rapid, szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}, it is necessary to consider the interplay between aspects of spatial population structure and the role of rapid evolution during range shifts.
-In asexual species, for example, local adaptation to a trait optimum gradient has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
+In asexual species, for example, local adaptation to an environmental gradient has been shown to interact with dispersal evolution during climate change, driving increased dispersal probability as genotypes shift to keep pace with their environmental optimum~\citep{hargreaves2015fitness}.
 However, it is unclear how these two processes might interact in a sexually reproducing species in which dispersal and local adaptation are directly linked via gene flow.
-Under sexual reproduction, evolution of increased dispersal could simultaneously reduce local adaptation to a trait optimum gradient within a population due to increased gene flow throughout the range~\citep{garcia1997genetic, kirkpatrick1997evolution}.
-In fact, long-distance pollen dispersal in flowering plants has been shown to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts, rather than spatial range shifts, in response to simulated climate change~\citep{aguilee2016pollen}.
-In addition to potential interactions between local adaptation and dispersal evolution, the starkness of the range edge could influence the potential for rapid trait evolution during range shifts by altering the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.
+Under sexual reproduction, evolution of increased dispersal could simultaneously reduce adaptation to an environmental gradient due to increased gene flow throughout the range~\citep{garcia1997genetic, kirkpatrick1997evolution}.
+In fact, long-distance pollen dispersal in flowering plants is predicted to restrict local adaptation and, when pollen dispersal sufficiently outpaces seed dispersal, to lead to ecological niche shifts rather than spatial range shifts in response to climate change~\citep{aguilee2016pollen}.
+In addition to potential interactions between adaptation and dispersal evolution, the starkness of the range edge could influence the potential for rapid trait evolution during range shifts by altering the spatial distribution of dispersal phenotypes throughout the range~\citep{henry2013eco, hargreaves2014evolution}, thus altering the diversity of dispersal genotypes present for subsequent evolution during range shifts.
 
-Here, we assess the interaction of two mechanisms responsible for spatial population structure with trait evolution in sexually reproducing populations undergoing range shifts.
+Here, we assess the interaction of two mechanisms responsible for spatial population structure (edge type and an environmental gradient in a trait optimum) with trait evolution in sexually reproducing populations undergoing range shifts.
 We develop an individual-based model capable of producing a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
 One trait determines dispersal ability while the second defines an individual's environmental niche.
-Using this model, we vary both the severity of the niche optimum gradient and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
-Additionally, by contrasting the dynamics of extant and extinct populations, we isolate the factors most strongly contributing to extinction risk during climate change.
+Using this model, we vary both the environmental gradient in the niche optimum and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
+Previous research suggests evolution of heightened dispersal during the range shift might alleviate the extinction risk of moving populations~\citep{hargreaves2015fitness, boeye2013more, henry2013eco}, but we predict steep environmental gradients and stark range edges may inhibit such evolution and thus increase extinction risk.
 
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/SimExample"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/SimExample"}
 \caption{A single example of a simulation with a steep niche optimum gradient and a moderately stark range edge.
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainExtinction"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainExtinction"}
 \caption{The cumulative probability of extinction due to a moderate speed of climate change in different experimental scenarios.
@@
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/DispComposite"}
-\caption{Patterns in the evolution and the equilibrium distribution of the dispersal trait, highlighting extant simulations from a moderate speed of climate change.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/DispComposite"}
+%\caption{Patterns in the evolution and the equilibrium distribution of the dispersal trait, highlighting extant simulations from a moderate speed of climate change.
 Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end.
@@
 Full results for all parameter combinations are provided in Appendix B.}
-\label{fig:Disp}
-\end{figure}
+%\label{fig:Disp}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/f_of_xt"}
 \caption{Example visualization of $f(x,t)$ in Cartesian space.
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/LifeCycle"}
 \caption{The life cycle of simulated populations is shown divided between events contributing to reproduction and dispersal.
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/SchematicFigures/Discretization"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/Discretization"}
 \caption{The carrying capacity of discrete patches along the $x$ dimension of landscapes.
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowExtinction"}
 \caption{The cumulative probability of extinction due to a slow speed of climate change in different experimental scenarios.
@@
 \begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastExtinction"}
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastExtinction"}
 \caption{The cumulative probability of extinction due to a fast speed of climate change in different experimental scenarios.
@@
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
-\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
+%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
 Points represent the mean across simulations and error bars are interquartile ranges.
 Population status (extinct or successful) was determined for a slow speed of climate change.}
-\label{Fig:InitFitSlow}
-\end{figure}
+%\label{Fig:InitFitSlow}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
-\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
+%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
 Points represent the mean across simulations and error bars are interquartile ranges.
 Population status (extinct or successful) was determined for a moderate speed of climate change.}
-\label{Fig:InitFit}
-\end{figure}
+%\label{Fig:InitFit}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
-\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
+%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
 Points represent the mean across simulations and error bars are interquartile ranges.
 Population status (extinct or successful) was determined for a fast speed of climate change.}
-\label{Fig:InitFitFast}
-\end{figure}
+%\label{Fig:InitFitFast}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a slow speed of climate change.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
+%\caption{Observed dispersal evolution in populations responding to a slow speed of climate change.
 Positive values indicate an increase in average dispersal ability during climate change.
@@
 The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolSlow}
-\end{figure}
+%\label{Fig:DispEvolSlow}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
+%\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change.
 Positive values indicate an increase in average dispersal ability during climate change.
@@
 The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolMain}
-\end{figure}
+%\label{Fig:DispEvolMain}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
-\caption{Observed dispersal evolution in populations responding to a fast speed of climate change.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
+%\caption{Observed dispersal evolution in populations responding to a fast speed of climate change.
 Positive values indicate an increase in average dispersal ability during climate change.
@@
 The experimental scenario corresponding to each histogram is indicated on the figure.}
-\label{Fig:DispEvolFast}
-\end{figure}
+%\label{Fig:DispEvolFast}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
+%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
 Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
 Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
-\label{Fig:InitDispSlow}
-\end{figure}
+%\label{Fig:InitDispSlow}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
+%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
 Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
 Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
-\label{Fig:InitDispMain}
-\end{figure}
+%\label{Fig:InitDispMain}
+%\end{figure}
 
-\clearpage
+%\clearpage
 
-\begin{figure}[h!]
-\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
-\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
+%\begin{figure}[h!]
+%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
+%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
 Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
 Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
-\label{Fig:InitDispFast}
-\end{figure}
+%\label{Fig:InitDispFast}
+%\end{figure}
 

### 2019-04-14 1a_SimCheck.R

@@ -1,36 +1,0 @@
-# This script will check for extinction before the period of climate change,
-#    delete those simulations, and output the current number of simulations
-#    that make it to equilibriation.
-
-setwd("~/ShiftingSlopes/MainSim/")
-NumSims <- rep(NA, 9)
-MaxSims <- 200
-
-for(p in 1:9){
-     ParamDirectory <- paste("~/ShiftingSlopes/MainSim/Params", p, sep = "")
-     ParamSims <- list.files(ParamDirectory)
-     for(i in 1:length(ParamSims)){
-          NumFiles <- length(list.files(paste(ParamDirectory, ParamSims[i], sep = "/")))
-          if(NumFiles != 4){
-               SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[i], sep = "")
-               system(SysCommand)
-          } else{
-               InFile <- paste(ParamDirectory, ParamSims[i], "InitialPopMat.csv", sep = "/")
-               PopMat <- read.csv(InFile)
-               if(dim(PopMat)[1] == 0){
-                    SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[i], sep = "")
-                    system(SysCommand)
-               }
-          }
-     }
-     TotalSims <- length(list.files(ParamDirectory))
-     while(TotalSims > MaxSims){
-          ParamSims <- list.files(ParamDirectory)
-          SysCommand <- paste("rm -r ", ParamDirectory, "/", ParamSims[1], sep = "")
-          system(SysCommand)
-          TotalSims <- length(list.files(ParamDirectory))
-     }
-     NumSims[p] <- length(list.files(ParamDirectory))
-}
-
-NumSims

### 2019-04-14 1b_MainSimCont.R

@@ -1,114 +1,0 @@
-# This script will run simulations on the MSI compute cluster to characterize
-#    population structure in a stable range for all of the 9 range parameter
-#    combinations to be explored.
-
-# Set the number of processors and number of simulations to be run
-nProc <- 24*6
-NumSims <- 200
-
-# Set the working directory and load necessary data and libraries
-setwd("~/ShiftingSlopes/MainSim/")
-source("~/ShiftingSlopes/SimFunctions.R")
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-library(parallel)
-library(Rmpi)
-
-# Now set all parameters
-BetaInit <- 0
-omega <- 3
-U <- c(0.02, 0.02)
-Vm <- c(0.0004, 0.0004)
-Lf <- 5
-Ld <- 5
-Rmax <- 2
-width <- 10
-kern <- "exp"
-EnvGradType <- "K"
-monoecious <- FALSE
-BurnIn <- 2000
-BurnOut <- 50
-LengthShift <- 100
-v <- 1
-InitPopSize <- 2500
-FitInit <- 0
-FitDiv <- 0.025
-DispInit <- -1
-DispDiv <- 1
-NumRands <- 1000000
-z <- 0.5
-dmax <- 1000
-rho <- 0.5
-
-AllParams <- vector(mode = "list", length = 3)
-
-for(i in 1:3){
-     gamma <- RangeParams$gamma[i]
-     lambda <- RangeParams$lambda[i]
-     tau <- RangeParams$tau[i]
-     eta <- RangeParams$eta[i]
-     Kmax <- RangeParams$Kmax[i]
-
-     AllParams[[i]] <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
-                        Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
-                        LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
-                        DispDiv, eta, NumRands, z, dmax, rho)
-     names(AllParams[[i]]) <- c("BetaInit", "gamma", "tau", "lambda", "omega", "U", "Vm", 
-                            "Lf", "Ld", "Rmax", "Kmax", "width", "kern", "EnvGradType", 
-                            "monoecious", "BurnIn", "BurnOut", "LengthShift", "v", 
-                            "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
[…]

### 2019-04-14 5_FitData.R

@@ -1,8 +1,8 @@
 # This script will extract the realized fitness of individuals throughout their
-#    ranges in the generation immediately prior to climate change
+#    ranges every 10 generations
 
-# Set the number of processors available for this scrip
-nProc <- 4*24
+# Set the number of processors available for this script 
+nProc <- 2*24
 
 # Set the working directory
 setwd("~/ShiftingSlopes/")
@@ -10,78 +10,101 @@ source("~/ShiftingSlopes/SimFunctions.R")
 library(parallel)
 library(Rmpi)
 RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
 
-# Create the master list to hold all the values and the other necessary objects
-NumSims <- 200
-MaxGen <- c(151, 2151, 151)
-FilePaths <- c("~/ShiftingSlopes/Slow/Params", "~/ShiftingSlopes/MainSim/Params",
-               "~/ShiftingSlopes/Fast/Params")
-FitList <- vector(length = 9, mode = "list")
-for(p in 1:9){
-     FitList[[p]] <- vector(length = NumSims, mode = "list")
-}
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
 
-# Create a data frame to index each simulation block in the above arrays
-TraitIndices <- expand.grid(params = 1:9, sim = 1:NumSims)
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 100
+GenSeq <- seq(0, LengthShift, by = GenSteps)
 
-# Write a function to be passed to various nodes
 FitExtract <- function(i){
-     # Sort out the parameter combination and simulation under consideration
-     Param <- TraitIndices$params[i]
-     ParamDirectory <- paste("~/ShiftingSlopes/MainSim/Params", Param, "/", sep = "")
-     AllSims <- list.files(ParamDirectory)
-     SimID <- AllSims[TraitIndices$sim[i]]
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs.
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
      
-     # Load the corresponding population matrix
[…]

### 2019-04-14 5_FitData.sh

@@ -1,5 +1,5 @@
 #!/bin/bash -l
-#PBS -l walltime=00:30:00,nodes=4:ppn=24,mem=248gb
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe

### 2019-04-14 5_FitData_new.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=06:00:00,nodes=2:ppn=24,mem=124gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the names of R scripts and log files
-Rscript=5_FitData_new.R
-LogFile=5_FitData_new.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0-centos7
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-14 6_DispData.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=03:00:00,nodes=4:ppn=24,mem=248gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the names of R scripts and log files
-Rscript=6_DispData.R
-LogFile=6_DispData.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0-centos7
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-14 6_DispDataNew.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the names of R scripts and log files
-Rscript=6_DispDataNew.R
-LogFile=6_DispDataNew.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0-centos7
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-14 6_DispData_new.R

@@ -1,181 +1,0 @@
-# This script will extract the initial dispersal values of individuals immediately
-#    prior to climate change (DispInit) and the change in mean dispersal values at
-#    the patch level (along with the relative x coordinate of the patch) to assess
-#    the role of evolution (DispEvol).
-
-# Set the number of processors available for this scrip
-nProc <- 4*24
-
-# Set the working directory
-setwd("~/ShiftingSlopes/")
-source("~/ShiftingSlopes/SimFunctions.R")
-library(parallel)
-library(Rmpi)
-RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
-load("ExtSimIDs.rdata")
-
-# Create the vectors to be passed to the function in the data extraction
-SimVec <- 1:27
-SpeedWords <- c("Slow", "MainSim", "Fast")
-SpeedVec <- rep(SpeedWords, 9)
-ParamVec <- rep(1:9, each = 3)
-
-# Set some initial parameter values to assess evolutionary change in dispersal
-Threshold <- 10
-BetaInit <- 0
-BurnIn <- 2000
-BurnOut <- 50
-LengthShift <- 100
-dmax <- 1000
-eta <- RangeParams$eta[1]
-rho <- 0.5
-ClimSpeeds <- c(0.5, 1, 2)
-MaxGen <- c(151, 2151, 151)
-EndShift <- c(LengthShift, BurnIn + LengthShift, LengthShift)
-BetaCoord <- matrix(NA, nrow = 3, ncol = BurnIn + LengthShift + BurnOut)
-for(i in 1:3){
-     BetaShift <- ChangeClimate(BetaInit = BetaInit, LengthShift = LengthShift, 
-                                eta = eta, v = ClimSpeeds[i]) / eta
-     BetaCoord[i,] <- c(rep(BetaInit, BurnIn), BetaShift, rep(BetaShift[LengthShift], BurnOut))
-}
-
-DispExtract <- function(i){
-     # Set the current speed and parameter combination and isolate the relevant
-     #    simulation IDs
-     SpeedWord <- SpeedVec[i]
-     ParamCombo <- ParamVec[i]
-     CurSims <- subset(SimIDs, Params == ParamCombo)
-     
-     # Source a representative parameter file for useful values
-     ParamFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo, "/", CurSims$SimID[1],
-                        "/parameters.R", sep = "")
-     source(ParamFile)
-     
[…]

### 2019-04-14 6_NicheData.R

@@ -35,17 +35,15 @@ FitExtract <- function(i){
                         "/", CurSims$SimID[1], "/parameters.R", sep = "")
      source(ParamFile)
      
-     # Create the data frames for extant and extinct simulations
-     ExtantFits <- data.frame(x = NA, g = NA, wBar = NA, lwr = NA, upr = NA)
-     ExtinctFits <- data.frame(x = NA, g = NA, wBar = NA, lwr = NA, upr = NA)
+     # Create the data frame for all simulations
+     MeanFits <- data.frame(x = NA, g = NA, Mismatch = NA, lwr = NA, upr = NA, Zopt = NA)
      
      # Now loop through each row of the data frame to fill it in
      for(g in GenSeq){
-          # Create temporary data frames to hold the results from all simulations
+          # Create a temporary data frame to hold the results from all simulations
           #    for each time point, which will then be condensed across simulations
           #    for the main data frames
-          TempExtant <- data.frame(x = NA, wBar = NA)
-          TempExtinct <- data.frame(x = NA, wBar = NA)
+          Temp <- data.frame(x = NA, Mismatch = NA, Zopt = NA)
           for(s in 1:nrow(CurSims)){
                if(g == 0){
                     InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
@@ -63,61 +61,29 @@ FitExtract <- function(i){
                xSeq <- unique(CurSumStats$x)
                for(j in xSeq){
                     LocalData <- subset(CurSumStats, x == j)
-                    # Calculate wBar according to the local optimum
+                    # Calculate the mismatch between optimum and genotype and
+                    #    its direction (i.e. positive or negative)
                     Zopt <- lambda * (j * eta - LocalData$beta[1])
-                    LocalFits <- exp(-1 * (LocalData$muFit - Zopt)^2 / (2*omega^2))
-                    # Decide if this simulation goes in extinct or extant data
-                    if(SpeedWord == "Slow"){
-                         if(CurSims$Slow[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalFits)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalFits)), TempExtinct)
-                         }
-                    } else if(SpeedWord == "MainSim"){
[…]

### 2019-04-14 6_NicheData.sh

@@ -1,12 +1,12 @@
 #!/bin/bash -l
-#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the names of R scripts and log files
-Rscript=8_FitGenVarDataNew.R
-LogFile=8_FitGenVarDataNew.log
+Rscript=6_NicheData.R
+LogFile=6_NicheData.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/

### 2019-04-14 7_DispData.R

@@ -35,17 +35,15 @@ DispExtract <- function(i){
                         "/", CurSims$SimID[1], "/parameters.R", sep = "")
      source(ParamFile)
      
-     # Create the data frames for extant and extinct simulations
-     ExtantDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
-     ExtinctDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
+     # Create the data frame for all simulations
+     MeanDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
      
      # Now loop through each row of the data frame to fill it in
      for(g in GenSeq){
-          # Create temporary data frames to hold the results from all simulations
+          # Create a temporary data frame to hold the results from all simulations
           #    for each time point, which will then be condensed across simulations
           #    for the main data frames
-          TempExtant <- data.frame(x = NA, dBar = NA)
-          TempExtinct <- data.frame(x = NA, dBar = NA)
+          Temp <- data.frame(x = NA, dBar = NA)
           for(s in 1:nrow(CurSims)){
                if(g == 0){
                     InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
@@ -63,58 +61,25 @@ DispExtract <- function(i){
                xSeq <- unique(CurSumStats$x)
                for(j in xSeq){
                     LocalData <- subset(CurSumStats, x == j)
-                    # Decide if this simulation goes in extinct or extant data
-                    if(SpeedWord == "Slow"){
-                         if(CurSims$Slow[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$muDisp)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$muDisp)), TempExtinct)
-                         }
-                    } else if(SpeedWord == "MainSim"){
-                         if(CurSims$Moderate[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$muDisp)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$muDisp)), TempExtinct)
-                         }
-                    } else{
-                         if(CurSims$Fast[s] == 1){
[…]

### 2019-04-14 7_DispGenVarDataNew.sh

@@ -1,21 +1,0 @@
-#!/bin/bash -l
-#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
-#PBS -m abe
-#PBS -M cweissle@umn.edu
-#PBS -j oe
-
-# Set the names of R scripts and log files
-Rscript=7_DispGenVarDataNew.R
-LogFile=7_DispGenVarDataNew.log
-
-# Change to the relevant working directory
-cd ~/ShiftingSlopes/
-
-# Load R and MPI
-module load R/3.4.4
-module load ompi/3.0.0/gnu-7.2.0-centos7
-
-export RMPI_TYPE=OPENMPI
-export OMPI_MCA_mpi_warn_on_fork=0
-
-mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-04-14 8_DispGenVarData.R

@@ -35,17 +35,15 @@ GenVarExtract <- function(i){
                         "/", CurSims$SimID[1], "/parameters.R", sep = "")
      source(ParamFile)
 
-     # Create the data frames for extant and extinct simulations
-     ExtantGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
-     ExtinctGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
+     # Create the data frame for all simulations
+     MeanGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
      
      # Now loop through each row of the data frame to fill it in
      for(g in GenSeq){
-          # Create temporary data frames to hold the results from all simulations
+          # Create a temporary data frame to hold the results from all simulations
           #    for each time point, which will then be condensed across simulations
           #    for the main data frames
-          TempExtant <- data.frame(x = NA, GenVar = NA)
-          TempExtinct <- data.frame(x = NA, GenVar = NA)
+          Temp <- data.frame(x = NA, GenVar = NA)
           for(s in 1:nrow(CurSims)){
                if(g == 0){
                     InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
@@ -63,58 +61,25 @@ GenVarExtract <- function(i){
                xSeq <- unique(CurSumStats$x)
                for(j in xSeq){
                     LocalData <- subset(CurSumStats, x == j)
-                    # Decide if this simulation goes in extinct or extant data
-                    if(SpeedWord == "Slow"){
-                         if(CurSims$Slow[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$DispGenVar)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$DispGenVar)), TempExtinct)
-                         }
-                    } else if(SpeedWord == "MainSim"){
-                         if(CurSims$Moderate[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$DispGenVar)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$DispGenVar)), TempExtinct)
-                         }
-                    } else{
[…]

### 2019-04-14 8_DispGenVarData.sh

@@ -1,12 +1,12 @@
 #!/bin/bash -l
-#PBS -l walltime=01:00:00,nodes=2:ppn=24,mem=124gb
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the names of R scripts and log files
-Rscript=5_FitDataNew.R
-LogFile=5_FitDataNew.log
+Rscript=8_DispGenVarData.R
+LogFile=8_DispGenVarData.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/

### 2019-04-14 8a_ExtinctionGraphs.R

@@ -1,92 +1,0 @@
-# This script will create the graphs from the extinction data.
-setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
-library(RColorBrewer)
-RangeParams <- read.csv("RangeParameters.csv")
-NumSims <- 200
-
-# Load the extinction data
-load("SimData/Extinctions.rdata")
-for(i in 1:3){
-     for(p in 1:9){
-          Extinctions[i,p,] <- cumsum(Extinctions[i,p,]) / NumSims
-     }
-}
-
-# Set some graphical parameters to use for the subsequent figures
-OuterMar <- c(1,1,0,0)
-TextSize <- 1.15
-LegTextSize <- 1.5
-AxisSize <- 1.25
-LetterPos <- c(2, 0.925)
-FigWidth <- 8
-FigHeight <- 3
-xRange <- c(0,100)
-yRange <- c(0,1)
-TimeAxisSeq1 <- seq(0, 100, by = 20)
-TimeAxisSea2 <- seq(0, 100, by = 5)
-ExtAxisSeq1 <- seq(0, 1, by = 0.2)
-ExtAxisSeq2 <- seq(0, 1, by = 0.05)
-GradLineType <- c(1,2,3)
-ExtAxisLabel <- "Extinction Probability"
-LineWidth <- 2
-Col <- "darkred"
-
-# Make the extinction graph for all three speeds
-SpeedWords <- c("Slow", "Main", "Fast")
-for(i in 1:3){
-     PlotName <- paste("ResultFigures/", SpeedWords[i], "Extinction.pdf", sep = "")
-     pdf(file = PlotName, width = FigWidth, height = FigHeight, onefile = FALSE, paper = "special")
-          par(mfrow = c(1,3), oma = OuterMar)
-          # No local adaptation
-          plot(x = 1:100, y = Extinctions[i,1,], lty = GradLineType[1], lwd = LineWidth,
-               col = Col, xaxt = "n", yaxt = "n", main = "", xlab = "", ylab = "",
-               xlim = xRange, ylim = yRange, type = "l")
-          lines(x = 1:100, y = Extinctions[i,2,], lty = GradLineType[2], lwd = LineWidth,
-                col = Col)
-          lines(x = 1:100, y = Extinctions[i,3,], lty = GradLineType[3], lwd = LineWidth,
-                col = Col)
[…]

### 2019-04-14 8b_DispInitGraphs_new.R

@@ -1,116 +1,0 @@
-# This script will make an appropriate graph of the initial dispersal values
-setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
-load("SimData/DispInitNew.rdata")
-load("SimData/ExtSimIDs.rdata")
-CurSpeed <- 1
-SpeedWords <- c("Slow", "Main", "Fast")
-
-# Calculate the dispersal phenotype corresponding to the current speed of climate
-#    change
-r <- log(2)
-patch_speeds <- c(0.75, 1, 1.25)
-eta <- 50
-v <- patch_speeds[CurSpeed]*eta
-DispPhen <- sqrt(v^2 / r)
-
-# Use SimIDs to calculate a vector of the number of extant simulatins for
-#    the current conditions
-NumExtant <- rep(NA, 9)
-for(i in 1:9){
-     CurSims <- subset(SimIDs, Params == i)
-     NumExtant[i] <- sum(CurSims[,CurSpeed+2])
-}
-
-# Set some graphical parameters to use for the subsequent figures
-options(scipen = -10)
-FigMat <- matrix(NA, nrow = 3, ncol = 12)
-FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
-FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
-FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
-OuterMar <- c(4, 12, 6, 2)
-InnerMar <- c(1.5, 2.25, 1.5, 2.25)
-TextSize <- 1.5
-AxisSize <- 1.15
-ArrowWidth <- 3
-ArrowLength <- 0.25
-FigWidth <- 8
-FigHeight <- 6
-DispAxisLabel <- expression(paste("Equilibrium ", italic("log"), "(", italic("d"), ")", sep = ""))
-SimSeq <- c(7,8,9,4,5,6,1,2,3)
-xLabLine <- 2
-yLabLine <- 3
-GradLabLine <- 3.5
-GradSubLine <- 1.5
-AdaptLabLine <- 9
-AdaptSubLine <- 7
-TopArrow <- matrix(c(-0.75, 240000, 17, 240000), nrow = 2, ncol = 2, byrow = TRUE)
-SideArrow <- matrix(c(-5, -550000, -5, 175000), nrow = 2, ncol = 2, byrow = TRUE)
[…]

### 2019-04-14 8c_InitFitGraphs_new.R

@@ -1,101 +1,0 @@
-# This script will make an appropriate graph of the initial dispersal values
-setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
-load("SimData/FitDataNew.rdata")
-load("SimData/ExtSimIDs.rdata")
-CurSpeed <- 3
-SpeedWords <- c("Slow", "Main", "Fast")
-
-# Use SimIDs to calculate a vector of the number of extant simulatins for
-#    the current conditions
-NumExtant <- rep(NA, 9)
-for(i in 1:9){
-     CurSims <- subset(SimIDs, Params == i)
-     NumExtant[i] <- sum(CurSims[,CurSpeed+2])
-}
-
-# Set some graphical parameters to use for the subsequent figures
-yRange <- c(0, 1)
-xRange <- c(-50, 50)
-SmallTicks <- seq(-35, 35, by = 5)
-OuterMar <- c(4, 10, 5, 2)
-InnerMar <- c(1.5, 1.5, 2, 1.5)
-TextSize <- 1.5
-AxisSize <- 1.05
-ArrowWidth <- 3
-ArrowLength <- 0.25
-FigWidth <- 8
-FigHeight <- 4
-FitAxisLabel <- expression(paste("Initial ", italic("w"), sep = ""))
-SimSeq <- c(7,8,9,4,5,6) # Exclude the scenarios with no local adaptation
-xLabLine <- 2
-yLabLine <- 1.75
-GradLabLine <- 3
-GradSubLine <- 1.5
-AdaptLabLine <- 8
-AdaptSubLine <- 5
-TopArrow <- matrix(c(-12, 1.3, 250, 1.3), nrow = 2, ncol = 2, byrow = TRUE)
-LowAdj <- 0.15
-HighAdj <- 0.85
-ExtinctCol <- "red"
-ExtantCol <- "navyblue"
-LineWidth <- 1
-
-PlotName <- paste("ResultFigures/", SpeedWords[CurSpeed], "InitFitSpace.pdf", sep = "")
-pdf(file = PlotName, width = FigWidth, height = FigHeight, onefile = FALSE, paper = "special")
-     par(mfrow = c(2,3), mar = InnerMar, oma = OuterMar)
-     for(i in SimSeq){
-          if(NumExtant[i] > 0){
-               plot(x = FitData[[i]][[CurSpeed]]$Extant$x, y = FitData[[i]][[CurSpeed]]$Extant$wBar,
[…]

### 2019-04-14 8d_DispEvolGraphs.R

@@ -1,149 +1,0 @@
-# This script will create the dispersal evolution graphs
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/")
-load("SimData/DispEvolData.rdata")
-SpeedWords <- c("Slow", "Main", "Fast")
-
-# Create objects to hold coordinates for the various arrows to use in the graph
-TopArrow <- array(NA, dim = c(3, 2, 2))
-TopArrow[1,,] <- c(-75, 600, 15000, 15000)
-TopArrow[2,,] <- c(-75, 600, 16000, 16000)
-TopArrow[3,,] <- c(-75, 600, 16000, 16000)
-SideArrow <- array(NA, dim = c(3, 2, 2))
-SideArrow[1,,] <- c(-193, -193, -34000, 10000)
-SideArrow[2,,] <- c(-193, -193, -38000, 12000)
-SideArrow[3,,] <- c(-193, -193, -34000, 10000)
-
-# Set some graphical parameters for the figures
-FigWidth <- 8
-FigHeight <- 6
-FigMat <- matrix(NA, nrow = 3, ncol = 12)
-FigMat[1,] <- c(rep(1,4), rep(2,4), rep(3,4))
-FigMat[2,] <- c(rep(4,4), rep(5,4), rep(6,4))
-FigMat[3,] <- c(rep(7,4), rep(8,4), rep(9,4))
-OuterMar <- c(4, 10, 6, 2)
-InnerMar <- c(1.5, 1.5, 1.5, 1.5)
-SimSeq <- c(7,8,9,4,5,6,1,2,3)
-xRange <- c(-100, 100)
-AxisSize <- 1.05
-ArrowWidth <- 3
-ArrowLength <- 0.25
-DispAxisLabel <- expression(paste(Delta, bar(d), sep = ""))
-TextSize <- 1.5
-xLabLine <- 2
-yLabLine <- 1.75
-GradLabLine <- 3.5
-GradSubLine <- 1.5
-AdaptLabLine <- 7
-AdaptSubLine <- 5
-LowAdj <- 0.05
-HighAdj <- 0.95
-
-AllCol <- "lightskyblue"
-ExtantCol <- "navyblue"
-
-range(DispEvol[[1]]$DeltaDisp, na.rm = TRUE)
-range(DispEvol[[2]]$DeltaDisp)
-range(DispEvol[[3]]$DeltaDisp, na.rm = TRUE)
[…]

### 2019-04-14 8e_DispComposite.R

@@ -1,102 +1,0 @@
-# This script will create the simplified composite graph incorporating dispersal
-#    evolution and initial dispersal
-
-setwd("~/Desktop/RangeShifts/ShiftingSlopesOther/ResultFigures/")
-load("SimData/InitDispData.rdata")
-load("SimData/DispEvolData.rdata")
-CurSpeed <- 2
-Params <- c(1, 9)
-
-AllSims <- vector(length = 2, mode = "list")
-ExtantSims <- vector(length = 2, mode = "list")
-NumExtant <- rep(0, 2)
-NumSims <- 200
-xMin <- 0
-xMax <- 0
-for(p in 1:2){
-     CurParam <- Params[p]
-     AllVals <- NULL
-     ExtantVals <- NULL
-     for(i in 1:NumSims){
-          AllVals <- c(AllVals, DispInit[[CurParam]][[i]]$ExpDists)
-          
-          if(DispInit[[CurParam]][[i]]$Ext[CurSpeed] == 1){
-               NumExtant[p] <- NumExtant[p] + 1
-               ExtantVals <- c(ExtantVals, DispInit[[CurParam]][[i]]$ExpDists)
-          }
-     }
-     AllSims[[CurParam]] <- log(AllVals, base = 10)
-     if(!is.null(ExtantVals)){
-          ExtantSims[[CurParam]] <- log(ExtantVals, base = 10)
-     }
-     xMin <- min(c(xMin, AllSims[[CurParam]]))
-     xMax <- max(c(xMax, AllSims[[CurParam]]))
-}
-xInitRange <- c(xMin, xMax)
-MainEvol <- DispEvol[[CurSpeed]]
-
-
-FigWidth <- 8
-FigHeight <- 6
-AllCol <- "lightskyblue"
-ExtantCol <- "navyblue"
-EvolDispAxisLabel <- expression(paste(Delta, bar(d), sep = ""))
-InitDispAxisLabel <- expression(paste("Equilibrium ", italic("log"), "(", italic("d"), ")", sep = ""))
-EvolHistBreaks <- seq(-375, 475, by = 5)
-InitHistBreaks <- seq(-1.8, 3.4, by = 0.2)
-xEvolRange <- c(-100, 100)
-Initxticks <- seq(-1, 3, by = 0.2)
-AxisSize <- 1.15
-OuterMar <- c(0, 1, 0, 0)
-InnerMar <- c(5, 5, 4, 3) + 0.1
-options(scipen = 1000)
-LabSize <- 1.25
-
[…]

### 2019-04-14 9_FitGenVarData.R

@@ -35,17 +35,15 @@ GenVarExtract <- function(i){
                         "/", CurSims$SimID[1], "/parameters.R", sep = "")
      source(ParamFile)
      
-     # Create the data frames for extant and extinct simulations
-     ExtantGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
-     ExtinctGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
+     # Create a data frame for all simulations
+     MeanGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
      
      # Now loop through each row of the data frame to fill it in
      for(g in GenSeq){
-          # Create temporary data frames to hold the results from all simulations
+          # Create a temporary data frame to hold the results from all simulations
           #    for each time point, which will then be condensed across simulations
           #    for the main data frames
-          TempExtant <- data.frame(x = NA, GenVar = NA)
-          TempExtinct <- data.frame(x = NA, GenVar = NA)
+          Temp <- data.frame(x = NA, GenVar = NA)
           for(s in 1:nrow(CurSims)){
                if(g == 0){
                     InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
@@ -63,58 +61,25 @@ GenVarExtract <- function(i){
                xSeq <- unique(CurSumStats$x)
                for(j in xSeq){
                     LocalData <- subset(CurSumStats, x == j)
-                    # Decide if this simulation goes in extinct or extant data
-                    if(SpeedWord == "Slow"){
-                         if(CurSims$Slow[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$FitGenVar)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$FitGenVar)), TempExtinct)
-                         }
-                    } else if(SpeedWord == "MainSim"){
-                         if(CurSims$Moderate[s] == 1){
-                              TempExtant <- rbind(c(j, mean(LocalData$FitGenVar)), TempExtant)
-                         }else{
-                              TempExtinct <- rbind(c(j, mean(LocalData$FitGenVar)), TempExtinct)
-                         }
-                    } else{
[…]

### 2019-04-14 9_FitGenVarData.sh

@@ -1,12 +1,12 @@
 #!/bin/bash -l
-#PBS -l walltime=00:30:00,nodes=4:ppn=24,mem=248gb
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
 #PBS -m abe
 #PBS -M cweissle@umn.edu
 #PBS -j oe
 
 # Set the names of R scripts and log files
-Rscript=7_VarData.R
-LogFile=7_VarData.log
+Rscript=9_FitGenVarData.R
+LogFile=9_FitGenVarData.log
 
 # Change to the relevant working directory
 cd ~/ShiftingSlopes/

### 2019-04-14 PlotVarData.R

@@ -1,92 +1,0 @@
-# Plot the var data. Doesn't have to be pretty, just has to let me see a visualization
-#    and then get some key values
-
-load("~/Desktop/RangeShifts/ShiftingSlopesOther/SimData/VarData.rdata")
-# Determine the x range to use for all graphs
-xMin <- 0
-xMax <- 0
-for(p in 1:9){
-     xRange <- range(VarList[[p]]$x, na.rm = TRUE)
-     xMin <- min(c(xRange[1], xMin))
-     xMax <- max(c(xRange[2], xMax))
-}
-xRange <- c(xMin, xMax)
-
-# First make the graph for variation in abundances
-quartz(width = 12, height = 9)
-par(mfrow = c(3,3))
-for(p in 1:9){
-     plot(x = VarList[[p]]$x, y = VarList[[p]]$abund, xlim = xRange, xlab = "Space", ylab = "Var(abund)")
-}
-
-# Now make the graph for variation in mean niche trait
-quartz(width = 12, height = 9)
-par(mfrow = c(3,3))
-for(p in 1:9){
-     plot(x = VarList[[p]]$x, y = VarList[[p]]$muFit, xlim = xRange, xlab = "Space", ylab = "Var(niche)")
-}
-
-# Now make the graph for variation in mean dispersal
-quartz(width = 12, height = 9)
-par(mfrow = c(3,3))
-for(p in 1:9){
-     plot(x = VarList[[p]]$x, y = VarList[[p]]$muDisp, xlim = xRange, xlab = "Space", ylab = "Var(disp)")
-}
-
-# Now make the graph for variation in variance in fitness of patches
-quartz(width = 12, height = 9)
-par(mfrow = c(3,3))
-for(p in 1:9){
-     plot(x = VarList[[p]]$x, y = VarList[[p]]$sigmaFitPhen, xlim = xRange, xlab = "Space", ylab = "Patch Var(fit)")
-}
-
-
-# Look into the maximum standard deviations across the y dimension
-MaxAbundSD <- rep(NA, 9)
-MaxNicheSD <- rep(NA, 9)
-MaxDispSD <- rep(NA, 9)
-for(p in 1:9){
-     MaxAbundSD[p] <- (max(VarList[[p]]$abund, na.rm = TRUE))
-     MaxNicheSD[p] <- (max(VarList[[p]]$muFit, na.rm = TRUE))
[…]

### 2019-05-01 Manuscript/AmNat_ReSubmission.tex
Updated the model scripts and manuscript for the AmNat revisions

@@
 
-\textit{Manuscript elements}: Figure~1, figure~2, figure~3, online appendices~A and B (including figures~A1-A3,  tables~A1-A4, and figures~B1-B11). Figure~1 and figure~3 are to print in color.
+\textit{Manuscript elements}: Figures~1-6, online appendices~A and B (including figures~A1-A3,  tables~A1-A4, and figures~B1-B10). All figures should print in color.
 
@@
 trait clines, starkness of range edges, etc.) likely interacts with other ecological and evolutionary processes during climate-induced range shifts.
-Here, we use an individual-based model to explore the interacting roles of these factors in range shift dynamics.
-We show that increased spatial population structure (driven primarily by a steeper slope in the trait optimum gradient) severely increased a population's extinction risk.
-Further, and contrary to expectations, we show that evolution of heightened dispersal during range shifts was unable to rescue faltering populations.
-Rather, a population's fate during climate change was determined by the composition of dispersal phenotypes defining the population at equilibrium (i.e.
-before the onset of rapid climate change); only populations consisting of highly dispersive individuals survived.
-Our results demonstrate that dispersal evolution alone may be insufficient to save a range shifting population and that spatial population structure can substantially increase extinction risk in range shifts.
+Here, we use an individual-based model to explore the interacting roles of several such factors in range shift dynamics.
+We show that increased spatial population structure (driven primarily by a steeper environmental gradient) severely increases a population's extinction risk.
+Further, we show that while evolution of heightened dispersal during range shifts can aid populations in tracking changing conditions, it can also interact negatively with adaptation to the environmental gradient, leading to reduced fitness and contributing to the increased extinction risk observed in populations structured along steep environmental gradients.
+Our results demonstrate that the effect of dispersal evolution on range shifting populations is dependent on environmental context and that spatial population structure can substantially increase extinction risk in range shifts.
 
@@
 
-Large-scale population movements have been studied for decades in the context of range expansions, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
-For example, while range expansion speed can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in these traits can increase both the mean and variance of expansion speed through time~\citep{szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
-In particular, laboratory experiments have shown spatial genetic structure to play an important role in the dynamics of range expansions, even over short time periods~\citep{weiss2017rapid,ochocki2017rapid}.
+Large-scale population movements have been studied for decades in the broader context of range expansions, leading to a robust understanding of both the ecological~\citep{hastings2005spatial} and evolutionary~\citep{shine2011evolutionary, excoffier2009genetic} mechanisms shaping such expansions.
+For example, while range expansion speed can be well approximated by a combination of the species' intrinsic growth rate and dispersal ability~\citep{fisher1937wave, hastings2005spatial}, recent research demonstrates that evolution in these traits, such as spatial sorting leading to increased dispersal at expanding edges~\citep{shine2011evolutionary}, can increase both the mean and variance of expansion speed through time~\citep{szHucs2017rapid, shaw2015dispersal, phillips2015evolutionary}.
+In particular, laboratory experiments have shown spatial genetic structure to play an important role in the dynamics of range expansions, even over short time periods~\citep{weiss2017rapid, ochocki2017rapid}.
 While some range expansions (e.g.
 of invasive or reintroduced species) start from small founding populations lacking any initial spatial structure, others are characterized by far more complex initial spatial population structure (e.g.
-humans expanding out of Africa or tree expansion following glacial retreat).
+humans expanding out of Africa or expansions following glacial retreat).
 Climate-driven range shifts fall in the latter category, as populations respond to climate change in the context of preexisting spatial population structure.
@@
 Here, we assess the interaction of two mechanisms responsible for spatial population structure (edge type and an environmental gradient in a trait optimum) with trait evolution in sexually reproducing populations undergoing range shifts.
-We develop an individual-based model capable of producing a wide variety of spatial population structures in which males and females are defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
-One trait determines dispersal ability while the second defines an individual's environmental niche.
-Using this model, we vary both the environmental gradient in the niche optimum and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
-Previous research suggests evolution of heightened dispersal during the range shift might alleviate the extinction risk of moving populations~\citep{hargreaves2015fitness, boeye2013more, henry2013eco}, but we predict steep environmental gradients and stark range edges may inhibit such evolution and thus increase extinction risk.
+We developed an individual-based model capable of producing a wide variety of spatial population structures in which females and males were defined by two genetically determined traits, thus allowing for both evolutionary and ecological responses to climate change.
+One trait determined dispersal ability while the second allowed for adaptation to the environmental gradient.
+Using this model, we varied both the environmental gradient in the trait optimum and the starkness of the range edge to ascertain how they interact with each other and with the process of trait evolution to impact a population's ability to track a changing climate.
+Previous research suggests evolution of heightened dispersal during the range shift might alleviate the extinction risk of moving populations~\citep{hargreaves2015fitness, boeye2013more, henry2013eco}, but we predicted steep environmental gradients and stark range edges would inhibit such evolution and thus increase extinction risk.
 
@@
 Landscapes were unbounded in the $x$ dimension but defined by a fixed width and wrapping boundaries in the $y$ dimension.
-The optimum environmental niche value changed linearly along the $x$ dimension, thus allowing for the intrinsic formation of stable range boundaries when the optimum changed rapidly enough~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}.
-However, to examine the dynamics of ranges in which the niche optimum did not change rapidly enough to form stable range limits, we additionally imposed extrinsic range limits to prevent continuous adaptation and spread of the population~\citep{alleaume2006geographical, garcia1997genetic}.
-Specifically, we systematically altered the decline in patch carrying capacities from the range core to the edge~\citep{alleaume2006geographical, henry2013eco, bocedi2014rangeshifter, mustin2009dynamics} in such a way that we could directly manipulate the starkness of the decline.
+Abiotic conditions varied along the $x$ dimension of landscapes in two ways.
+First, a linear environmental gradient determined the optimum phenotype in a quantitative trait, driving adaptation throughout the range.
+Second, the patch carrying capacities systematically declined from the range center to both edges along the $x$ dimension.
+The starkness of this decline could be altered to create different types of range edges (Fig.
+A3).
+While an environmental gradient in a quantitative trait optimum is sufficient to produce stable range limits under certain conditions~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}, we included the decline in carrying capacities to enforce stable range limits when the environmental gradient was too shallow to produce them~\citep{alleaume2006geographical, garcia1997genetic}.
+A wide variety of other range limiting mechanisms have been proposed, including interspecific interactions~\citep{case2000interspecific, price2009evolutionarily}, Allee effects~\citep{keitt2001allee}, and spatial variation in demographic parameters~\citep{holt2005theoretical}.
+Given empirical evidence for many different range limiting mechanisms~\citep{gaston2009geographic}, the inclusion of both an environmental gradient in the trait optimum and the decline in carrying capacities allowed the model to capture a wide variety of spatial population structures, including ranges not structured by adaptation to an environmental gradient, with minimal additional assumptions.
 Previous research has shown that using a decline in intrinsic growth rate as opposed to carrying capacity may impact extinction risk but does not alter the patterns of dispersal evolution during climate change~\citep{henry2013eco}.
-To maintain generality, we do not assume a specific mechanism behind the decline in carrying capacity, but it could represent a variety of range limiting mechanisms such as physiological limits to adaptation, the effects of competition, or underlying resource distributions~\citep{sexton2009evolution, holt2005theoretical, case2000interspecific}.
+To maintain generality, we do not assume a specific mechanism behind the decline in carrying capacity, but it could represent a variety of range limiting mechanisms such as physiological limits to adaptation, the effects of competition, or underlying resource distributions~\citep{sexton2009evolution, holt2005theoretical, case2000interspecific, price2009evolutionarily}.
 Thus, the $x$ dimension defined the environmental context of the population and the $y$ dimension allowed for variation in population dynamics under identical environmental conditions.
-To simulate climate change, the patch carrying capacities and the niche optimum gradient shifted at a constant rate along the $x$ dimension.
+To simulate climate change, the patch carrying capacities and the environmental gradient shifted at a constant rate along the $x$ dimension.
 Generations were non-overlapping and consisted of discrete dispersal and reproduction phases (Fig.
@@
 
-Individuals were characterized by two traits (dispersal and an environmental niche), each defined by a set of $5$ quantitative diploid loci.
+Individuals were characterized by two traits, each defined by a set of $5$ quantitative diploid loci.
 While the number of loci was arbitrary, $5$ was chosen as a compromise between computational restrictions and the likely polygenic nature of such complex traits.
-The dispersal trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
+The first trait defined an individual's expected dispersal distance, assuming an exponential dispersal kernel.
 An individual's realized dispersal distance was then drawn from the dispersal kernel and dispersal direction was random and unbiased.
 Dispersal occurred in continuous space from the center of an individual's current patch and the individual's new patch was then determined by the mapping from continuous space to discrete patches (see Appendix A).
-An individual's environmental niche value determined its fitness according to the local niche optimum.
-The closer an individual's niche value to the local optimum, the higher the individual's realized fitness.
+The second trait allowed for adaptation to the local conditions along the environmental gradient.
+The environmental gradient determined the local phenotypic optimum in this trait and deviation from this optimum decreased individual fitness.
+As this trait determined the environmental conditions in which individuals would experience maximal fitness, we refer to it as the niche trait when differentiating between it and the dispersal trait.
 Reproduction within each patch occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock, melbourne2008extinction}, scaled by the mean fitness of the patch.
-Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with a close match of their niche value to the local optimum produced more offspring on average.
-Thus, the model used a mixture of  hard selection (realized population growth declined with maladaptation relative to the niche optimum) and soft selection (probability of producing offspring depended on fitness relative to other individuals) for the evolutionary dynamics~\citep{wallace1975hard}.
+Parental pairs formed via random sampling of the local population (with replacement) weighted by individual fitness such that individuals with a close match of their niche trait to the local optimum produced more offspring on average.
+Thus, the model used a mixture of  hard selection (expected population growth declined with maladaptation to the environmental gradient) and soft selection (probability of producing offspring depended on fitness relative to other individuals) for the evolutionary dynamics~\citep{wallace1975hard}.
 Allele inheritance was subject to mutation and assumed no linkages among loci.
@@
 
-We varied parameter values to explore the interacting roles of the slope of the niche optimum gradient and range edge starkness (Table A2) in forming spatial population structure at equilibrium and driving the subsequent eco-evolutionary dynamics of range shifts.
-Specifically, we considered a fully factorial combination of: (1) a flat, shallow, and steep niche optimum gradient, (2) shallow, moderate, and stark declines in carrying capacity at the range edge, and (3) slow, moderate, and fast speeds of climate change.
-This yielded a total of $27$ different scenarios, each explored with $200$ simulations.
-Each simulation ran for $2150$ generations with stable climate conditions for the first $2000$ to reach a spatial equilibrium, followed by $100$ generations of climate change and a final $50$ generations of stable conditions.
+We varied parameter values to explore the interacting roles of the slope of the environmental gradient and different types of range edges (Table A2) in forming spatial population structure at equilibrium and driving the subsequent eco-evolutionary dynamics of range shifts.
+Specifically, we considered a fully factorial combination of: (1) a flat, shallow, or steep environmental gradient, (2) gradual, moderate, or stark declines in carrying capacity at the range edge, and (3) slow, moderate, or fast speeds of climate change.
+This yielded a total of $27$ different scenarios, each explored with $200$ simulations, during which we tracked the mean phenotypes of both traits and their genetic variances in each patch through time.
+The first $2000$ generations of each simulation allowed populations to reach a spatial equilibrium, which we confirmed by assessing the spatial distributions of phenotypes and genetic variances in both traits for stability through time.
+On average, simulations in every scenario arrived at stable distributions in all values by generation $1500$.
+Following the initial $2000$ generations, we imposed $100$ generations of climate change by shifting the environmental gradient and patch carrying capacities at a constant rate.
 Figure 1 shows an example of a single population responding to a moderate speed of climate change.
-For each scenario, we evaluated the role of the equilibrium spatial population structure and dispersal evolution in the dynamics of the range shifting populations.
+For each scenario, we evaluated the impact of the environmental gradient and edge type on trait evolution and extinction dynamics during the range shift.
 We primarily discuss simulations assuming a moderate speed of climate change in the main text, but present the results for slow and fast speeds of climate change in Appendix B.
-
-We calculated dispersal evolution in each patch throughout the landscape as the change in mean dispersal phenotype from the beginning of the period of climate change to the end.
-For this analysis, we defined individual patches by their relative location within the range rather than with their fixed spatial coordinates (e.g.
-leading edge vs.
-core populations).
-Due to local extinctions, not all patches were occupied at the end of the period of climate change.
-To quantify dispersal evolution in these patches, we used data from the last generation in which the population had at least $10$ individuals.
-Changes in mean dispersal phenotype were calculated by subtracting the initial mean dispersal phenotype from the value at the end of climate change (or at the last generation of at least $10$ individuals occupying the patch in the case of population extinctions); positive values indicate an increase in the mean dispersal phenotype.
+To assess spatial patterns in trait evolution, we focused primarily on variation along the $x$ dimension of simulated landscapes for two reasons: (1) to capture variation relevant to the changing environmental conditions and (2) because variation in this dimension was much greater than variation within the $y$ dimension in all scenarios.
 All simulations and data processing were performed in R version $3.4.4$~\citep{team2000r} and the code is available at (links are available from the journal office).
@@
 \section*{Results}
-In all scenarios, some populations shifted their ranges in response to climate change.
-However, the proportion of populations that failed to track the changing climate (and subsequently went extinct) depended on the spatial characteristics of the range.
-Populations defined by a steep niche optimum gradient and by stark declines in carrying capacity at the range edge experienced the greatest probability of extinction due to climate change (quantified by the proportion of simulated populations to go extinct through time; Fig.
+\subsection*{Extinction}
+Extinction dynamics during the climate-driven range shifts varied widely among scenarios, ranging from no extinctions in some scenarios to the extinction of every simulated population in others.
+We quantified extinction probability in each scenario as the proportion of simulated populations to decline to extinction through time during the $100$ generations of climate change.
+Comparing extinction probabilities among scenarios revealed four prominent patterns.
+First, the slope of the environmental gradient strongly impacted extinction risk (Fig.
 2).
-While both aspects of a population's range influenced extinction probabilities, the niche optimum gradient drove more dramatic changes to extinction risk, with steeper gradients causing severe increases in the probability of extinction during climate change.
-We varied both parameters widely, doubling the slope of the niche optimum gradient from the shallow to steep scenario and increasing the parameter defining range edge starkness by a factor of $100$ from shallow to stark edges (Table A2).
-This suggests that the niche optimum gradient may be the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios.
-Additionally, as expected, the pace of climate change also influenced extinction probabilities with faster climate change corresponding to greater extinction risk (Fig.
+Scenarios with a flat gradient only resulted in extinctions at a fast speed of climate change coupled with a stark range edge (Fig.
+B2), while scenarios with a steep gradient experienced high extinction probabilities at even a slow speed of climate change (Fig.
+B1).
+Second, edge type also had an important effect on extinction probability, with stark edges increasing extinction risk over gradual and moderate edges.
+Interestingly, when scenarios with moderate and gradual range edges diverged in extinction probabilities, moderate edges consistently yielded lower extinction risk than scenarios with gradual edges (Fig.
+2, B1, \& B2).
+Third, while both the environmental gradient and edge type influenced extinction probabilities, the environmental gradient drove more dramatic changes to extinction risk.
+We varied both parameters widely among scenarios, doubling the slope of the environmental gradient and increasing edge starkness by a factor of $100$ (Table A2).
+Thus, our results suggest the environmental gradient is the stronger driver of extinction risk during climate-induced range shifts across a wide region of parameter space and corresponding biological scenarios.
+Finally, as expected, the speed of climate change also influenced extinction probability with faster climate change corresponding to greater extinction risk (Fig.
 B1 \& B2).
-However, this effect was independent of the niche optimum gradient and range edge starkness in determining extinction probability during range shifts.
 
-In accordance with previous results~\citep{garcia1997genetic}, populations at the range edge had lower fitness than central populations at equilibrium and this effect was amplified with more gradual range edges.
-Counterintuitively, populations that survived climate change tended to be characterized by even greater reductions in fitness at the range edges at equilibrium compared to populations that went extinct (Fig.
-B3-B5).
-While discernible in all simulations with a non-zero slope in the niche optimum gradient, this pattern was most evident in the scenarios defined by a gradual range edge.
-As expected, there was no spatial variation in fitness for populations with no variation in the niche optimum across space.
-Despite the spatial variation in fitness in some scenarios, variance in relative fitness within a patch was relatively low (about $0.3$ across all scenarios).
-This implies that (1) the reduction in fitness at the edge caused by the niche optimum gradient was relatively uniform across all individuals, and (2) as a result, there was relatively low variance in reproductive success in these populations, meaning that evolution at the edge was not driven by only a handful of higher fitness individuals.
+\subsection*{Dispersal evolution} 
+The distribution of dispersal phenotypes at spatial equilibrium varied across the $x$ dimension of simulated landscapes, with more dispersive phenotypes characterizing the range edges.
+While this pattern held generally across all scenarios, the difference between edge and core phenotypes was reduced in the presence of an environmental gradient and starker range edges, leading to lower dispersal phenotypes at the range margins in these populations (Fig.
+3).
+At equilibrium, the distributions of dispersal phenotypes were symmetric, but spatial sorting during climate change not only increased dispersal phenotypes in surviving populations, but led to asymmetric distributions with the highest phenotypes at the leading edge.
+Despite these initial changes, the spatial distributions remained remarkably stable between $50$ and $100$ generations, regardless of the speed of climate change (Fig.
+3, B3, \& B4).
+This stands in contrast to models of dispersal evolution in unbounded range expansions, which predict dispersal to continually increase as the expansion progresses~\citep{shaw2015dispersal, shine2011evolutionary, fronhofer2015eco}.
+In our model, the shifting environmental gradient and distribution of carrying capacities simultaneously exposed new habitat for colonization, thus allowing spatial sorting, and also reduced the fitness of individuals that dispersed too far, resulting in an upper limit on dispersal evolution.
 
-Dispersal evolution is predicted to play a key role in aiding populations as they shift to track a changing climate.
-While some individual simulations confirmed these expectations with average dispersal phenotypes increasing through time (e.g.
-Fig.
-1), examining all simulations from each experimental scenario revealed no differences in the magnitude or direction of dispersal evolution between successful and extinct populations (Fig.
-3a\&b).
-Populations in all parameter combinations experienced both increases and decreases in average dispersal phenotypes, with all distributions of observed changes in dispersal phenotypes centered on $0$ (Fig.
-B3-B5).
-Further, calculating the coefficient of variation (CV) in dispersal genotypes over the $y$ dimension, in which environmental conditions did not vary, revealed that edge populations had between $3$ and $4$ times higher CVs than core populations across all scenarios, indicating genetic diversity at the edge was not a limiting factor in dispersal evolution.
-The similarity in evolved changes in dispersal between surviving and extinct populations suggests that dispersal evolution alone cannot explain which populations successfully tracked moving conditions and which became extinct.
+Both the severity of the environmental gradient and edge starkness also determined the spatial distribution of genetic variance in the dispersal trait.
+Increasing the slope of the environmental gradient reduced the genetic variance in dispersal across the range, but did not alter the qualitative patterns.
+The type of range edge, on the other hand, qualitatively altered the spatial distribution of genetic variance in dispersal (Fig.
+4).
+Scenarios with gradual and moderate edges displayed similar patterns of heightened variance at the range center compared to the margins, but stark edges resulted in the opposite pattern.
+This discrepancy makes sense as scenarios with stark declines in carrying capacity lacked the low abundance, and hence low genetic variance, populations that characterized the edges of scenarios with gradual or moderate declines in carrying capacity (Fig.
+A3).
+In all scenarios, genetic variance in dispersal decreased throughout the range in response to climate change (Fig.
+4), consistent with selection for increased dispersal via spatial sorting.
+This pattern was qualitatively similar for all speeds of climate change (Fig.
+B5 \& B6).
 
-Instead, the distribution of dispersal phenotypes at equilibrium played a key role in determining a population's fate.
-A range of dispersal phenotypes evolved in populations over the $2000$ generations of stable climatic conditions in response to the niche optimum gradient and range edge starkness (Fig.
-B9-B11).
-Populations that survived climate change were composed primarily of individuals with heightened dispersal phenotypes (Fig.
-3c\&d).
-Previous research has demonstrated that defining the range edges via a decline in the intrinsic growth rate (as opposed to carrying capacity as done here) resulted in less dispersive phenotypes at the range edge~\citep{henry2013eco}, meaning extinction risks would be even higher under such a scenario.
-Comparing the full distribution of equilibrium dispersal phenotypes present in a given experimental scenario to the distribution of phenotypes just from surviving populations revealed a threshold value delineating individuals from surviving versus extinct populations.
-Comparison of the different experimental scenarios revealed this threshold to be constant for a given speed of climate change (Fig.
-B9-B11).
-To explain this phenomenon, we used the well-known approximation for the speed of an expanding population, $2\sqrt{rD}$~\citep{fisher1937wave, hastings2005spatial}, in which $r$ is the intrinsic growth rate and $D$ is the diffusion coefficient, to calculate the dispersal phenotype necessary to produce an expansion wave exactly matching the speed of climate change in our simulations (see the model description in Appendix A).
-This estimated dispersal phenotype matched the observed threshold value distinguishing surviving from extinct populations in all experimental scenarios (Figures B9-B11, vertical dashed line).
-Thus, surviving populations in each scenario happened to be the lucky few already composed primarily of individuals with dispersal phenotypes capable of spreading at the pace of climate change, rather than populations in which heightened dispersal evolved over time in response to climate change.
-However, this threshold effect weakened slightly in scenarios with a slow speed of climate change, indicating a potential role for dispersal evolution if the climate were to change at a slow enough rate (Figure B11).
+\subsection*{Adaptation to the environmental gradient} 
+Equilibrium values of the niche trait closely matched the environmental gradient in the range core and diverged at the edges (Fig.
+5), likely due to a combination of gene swamping from abundant central populations~\citep{kirkpatrick1997evolution} and the loss of genetic diversity due to drift in small, peripheral populations~\citep{polechova2018sky, polechova2015limits}.
+In agreement with previous results, the starkness of the range edge influenced this pattern, with increased maladaptation of peripheral populations in scenarios with gradual edges~\citep{garcia1997genetic}.
+Across all scenarios, the clines in the niche trait became much shallower during the range shift (Fig.
+5).
+As a result, the match between the environmental gradient and local trait values substantially deteriorated during climate change, resulting in reduced fitness across most of the range.
+This reduction in fitness, observed at all speeds of climate change (Fig.
+B7 \& B8), could partially explain the increased extinction probability observed for scenarios with a steep environmental gradient (Fig.
+2).
+
+In contrast to the dispersal trait, genetic variance in the niche trait at equilibrium actually increased with the slope of the environmental gradient (Fig.
+6).
+This is likely due to the increased range of environmental conditions encountered by populations arranged along a steeper environmental gradient.
+The spatial distributions of genetic variance in the niche trait were qualitatively similar for scenarios with moderate and stark edges, peaking at the center of the range and declining towards the edges.
+However, scenarios with gradual edges displayed different patterns in the genetic variance of the niche trait, with a peak in variance occurring between the range center and each edge (Fig.
+6).
+This pattern was lost during range shifts, though, leading to qualitatively similar distributions of genetic variance in the niche traits among different edge types (Fig.
+6).
+In contrast to the loss of genetic variance observed in the dispersal trait during climate change, the magnitude of genetic variance in the niche trait during climate change was roughly equivalent to that characterizing equilibrium populations in most scenarios, regardless of the speed of climate change (Fig.
+B9 \& B10).
+Only in scenarios with gradual edges did genetic variance decrease slightly in response to climate change.
 
 \section*{Discussion}
-Range shifts due to climate change represent a global threat to biodiversity and much recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
-We developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and environmental niche traits defined by multiple loci.
+Range shifts due to climate change represent a global threat to biodiversity and recent research has focused on exploring the underlying ecological and evolutionary dynamics of such range shifts to inform conservation efforts.
+Here we showed that extinctions due to climate-induced range shifts are more likely in populations structured by steeper environmental gradients or starker range edges.
+For our approach, we developed an individual-based model to explore the eco-evolutionary dynamics of climate-induced range shifts in sexually reproducing, diploid populations with both dispersal and niche traits defined by multiple loci.
 In contrast, previous models have focused on a subset of these factors: ecological dynamics (e.g.~\citep{brooker2007modelling}), evolution in a single trait only (e.g.~\citep{atkins2010local, henry2013eco}), and relatively simple genetic scenarios (e.g.
 single-locus haploid genetics in asexual populations~\citep{boeye2013more, hargreaves2015fitness}).
-Here, we tested the generality of previous results predicting an important role for dispersal evolution in range shifts~\citep{boeye2013more, henry2013eco} and the interplay of eco-evolutionary dynamics under increased levels of biological complexity.
-Specifically, we demonstrated the role of spatial population structure, driven by a niche optimum gradient and range edge starkness, in determining extinction risk for range shifting populations via impacts on the equilibrium distribution of dispersal phenotypes and environmental niche values.
+Previous studies on eco-evolutionary dynamics of populations responding to climate change have generally predicted that evolution of increased dispersal might provide relief for populations struggling to keep pace with changing environmental conditions~\citep{boeye2013more, henry2013eco, hargreaves2015fitness}.
+Here, we tested the generality of these predictions in populations with varying types of spatial structure induced by the underlying environmental conditions.
 
-Our results suggest that populations characterized by local adaptation to a spatially varying trait optimum and by stark range edges will be less able to track changing climatic conditions (Fig.
+While all surviving populations evolved increased dispersal phenotypes during range shifts, the environmental gradient and edge starkness may have hampered the ability of populations to achieve such increases in dispersal under certain scenarios.
+For example, scenarios characterized by stark edges or steep environmental gradients led to edge populations composed of lower dispersal phenotypes at equilibrium due to the increased cost of dispersal in these scenarios.
+In the presence of a steep environmental gradient, selection favors lower dispersal phenotypes due to the fitness cost of dispersing to a location with a drastically different environmental optimum~\citep{kirkpatrick1997evolution}.
+Similarly, stark range edges can select for lower dispersal due to the increased risk of dispersing into unfavorable habitat~\citep{shaw2014population, shaw2019evolution}.
+Previous research has documented a similar reduction in dispersal phenotypes due to an explicit mortality cost of dispersal~\citep{kubisch2013predicting}, whereas the costs to dispersal in our model result from the environmental gradient and range edge starkness.
+By starting from lower dispersal phenotypes, populations structured by steep environmental gradients or stark range edges needed to achieve greater evolutionary increases in dispersal, since populations in all scenarios evolved to similar levels of dispersal in response to climate change (Fig.
+3).
+Presumably, the increased magnitude of the required evolutionary change in dispersal might have contributed to the increased extinction risk observed in such scenarios.
+
+Additionally, both the environmental gradient and the edge starkness impacted the spatial distribution of genetic variance in dispersal at equilibrium, thus altering the evolutionary potential of populations in different scenarios.
+Equilibrium genetic variance in dispersal decreased with an increasingly steep environmental gradient.
+Coupled with the reduced dispersal phenotypes characteristic of these scenarios at equilibrium, this reduction in genetic diversity could have amplified the difficulty of populations evolving the increased dispersal necessary to track climate change.
+While the effect of edge type on equilibrium patterns of genetic variance in dispersal was less straightforward, resulting in qualitative differences among different levels of edge starkness, it provides a clue to the lower extinction risk faced by scenarios defined by moderate versus gradual edges (Fig.
 2).
-A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
-While it is difficult to exactly map adaptation to the niche optimum gradient to empirical measures of local adaptation, the parameters defining the steepest gradient used here resulted in the intrinsic formation of stable range boundaries, as seen in previous theoretical studies~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}, suggesting they provide reasonable approximations of empirical patterns.
-Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, suggesting many species exhibit similar abundances at the edge and center of their ranges similar to the starkest range edges imposed in our study.
-While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, combined with our results they suggest many species will face elevated extinction risks in climate-induced range shifts due to their spatial population structure.
+Scenarios defined by moderate edges consistently had higher levels of genetic variance in dispersal compared to scenarios with gradual edges, suggesting the evolution of increased dispersal during range shifts may have been easier in such scenarios.
 
-Our results emphasize the importance of the equilibrium distribution of dispersal phenotypes (i.e.
-prior to the onset of rapid climate change) in determining a population's extinction risk during climate change (Fig.
-3c\&d).
-Populations primarily composed of high dispersal phenotypes at equilibrium successfully tracked changing climate conditions, while populations of lower dispersal phenotypes lagged behind the changing conditions to eventually go extinct.
-Importantly, the threshold dividing the equilibrium dispersal phenotypes of successful and extinct populations was constant across all scenarios for a given speed of climate change (Fig.
-B9-11).
-Thus, the difference in survival probability among scenarios was driven by the effects of the niche optimum gradient and range edge starkness on the evolution of dispersal ability throughout the range.
-Scenarios with a steep niche optimum gradient selected for lower dispersal phenotypes at equilibrium due to the potential mismatch of dispersing individuals' niche phenotype and their new location~\citep{kirkpatrick1997evolution}.
-Similarly, a stark range edge selected for lower dispersal due to the risk of dispersing into unsuitable habitat~\citep{shaw2014population, shaw2019evolution}.
-Previous research has documented a similar reduction in dispersal phenotypes due to an explicit mortality cost of dispersal~\citep{kubisch2013predicting}, whereas the costs to dispersal in our model result from the niche optimum gradient and range edge starkness.
-Such a cost to dispersal results in lower dispersal phenotypes at equilibrium, hampering a population's ability to successfully track a changing climate.
-Importantly, dispersal evolution during climate change was unable to counter the influence of equilibrium spatial population structure on extinction dynamics.
+Adaptation to the environmental gradient and patterns in the genetic variance of the niche trait in the different scenarios likely contributed to the extinction risk of range shifting populations as well.
+The process of shifting in response to climate change resulted in dramatically increased mismatches between the environmental gradient and local phenotypes (Fig.
+5), thus reducing fitness throughout the range.
+While this pattern resulted from the shifting locations of the environmental optima, it was likely exacerbated by heightened gene flow throughout the range driven by the increased dispersal phenotypes evolved during range shifts~\citep{lenormand2002gene}.
+This means even the populations successfully tracking climate change experienced substantially reduced fitness relative to equilibrium, which could partially explain the increased extinction risk characterizing scenarios with steep environmental gradients.
+Contrary to the dispersal trait, genetic variance in the niche trait did not decline during the range shift, except in scenarios with gradual edges (Fig.
+6).
+This finding agrees with previous theoretical investigations of stable ranges which found that loss of genetic variance due to genetic drift can constrain range expansion in the presence of an environmental gradient~\citep{polechova2018sky, polechova2015limits}.
+Thus, a reduction in genetic variance of the niche trait, as observed for scenarios with gradual edges, could constrain a population's ability to track changing environmental conditions.
+Along with the patterns in genetic variance of the dispersal trait, this could have contributed to the increased extinction risk faced by populations defined by gradual compared to moderate range edges.
 
-While high dispersal phenotypes at equilibrium increased the probability that populations tracked changing conditions, they had the additional effect of reducing average fitness at the range edges when the niche optimum varied through space (Fig.
-B3-B5).
-In the model, range edge populations tended to have lower abundance than core populations, increasing their susceptibility to gene flow from the core~\citep{kirkpatrick1997evolution, garcia1997genetic}.
-Thus, in populations with high dispersal phenotypes at equilibrium, increased gene flow from the core reduced fitness at the range edge via gene swamping~\citep{lenormand2002gene}.
-As a result, the populations most likely to survive climate change were, counterintuitively, also those characterized by lower fitness at the range edges at equilibrium.
-While not all populations are characterized by small populations at the range edges~\citep{dallas2017species}, our results suggest that high fitness in edge populations may be a warning sign of future difficulty in tracking climate change when the population is structured along a spatial gradient in a trait optimum.
-For the purposes of this investigation, we assumed the niche optima shifted spatially with climate change, as would be expected if the niche optima corresponded to local temperature or precipitation conditions~\citep{davis2001range}.
-However, in systems characterized by a niche optimum gradient defined by other factors (e.g.
-geography or biotic interactions) the gradient might remain stable or even shift in an opposing direction to climate change.
-Future research should investigate the impact of a niche optimum gradient unrelated to climate on extinction risk during range shifts.
+Previous models have predicted that dispersal evolution may be capable of rescuing populations that otherwise would be unable to keep pace with climate change~\citep{boeye2013more, henry2013eco}.
+While our results generally agree with this prediction, comparison with results from previous models suggests an important role for the biological and environmental context of dispersal evolution in range shifts.
+In a model assuming asexual reproduction, adaptation to an environmental gradient increased dispersal evolution during range shifts compared to scenarios with no gradient, contrary to the results presented here~\citep{hargreaves2015fitness}.
+In sexually reproducing populations, however, the evolution of increased dispersal can have an antagonistic interaction with local adaptation to an environmental gradient.
+Increased dispersal, and therefore increased gene flow, can reduce adaptation to the gradient via gene swamping~\citep{lenormand2002gene, kirkpatrick1997evolution}.
+When increased dispersal is necessary to track changing environmental conditions, our results suggest the fitness costs to the population can be severe, contributing to increased extinction risks in the presence of an environmental gradient.
+Thus, while dispersal evolution may be able to improve a population's ability to track changing conditions, it may also incur a severe cost depending on the environmental context of the population.
+For the purposes of this investigation, we assumed the environmental gradient shifted spatially with climate change, as would be expected if it corresponded to a gradient in temperature or precipitation~\citep{davis2001range}.
+However, in systems characterized by a gradient defined by other factors (e.g.
+geography or biotic interactions), the gradient might remain stable or even shift in an opposing direction to climate change.
+Future research should investigate the impact of an environmental gradient unrelated to climate on extinction risk during range shifts.
 
-Previous research has suggested that evolution of increased dispersal ability during climate change may be capable of rescuing populations that would otherwise be unable to keep pace with shifting environmental conditions~\citep{boeye2013more, henry2013eco}.
-Our results suggest this is not always the case.
-In fact, it may only be possible under certain, relatively narrow conditions.
-Previous models showing that dispersal evolution may rescue range shifting populations have typically used relatively simple genetic frameworks to model dispersal, including haploid genetics with a single-locus defining dispersal~\citep{boeye2013more, hargreaves2015fitness}.
-As dispersal evolution during range expansions and shifts occurs via the spatial sorting of dispersal alleles throughout the range~\citep{shine2011evolutionary}, such simplified genetic frameworks may allow more efficient sorting of such alleles compared to situations with more complex genetic structure underlying the dispersal trait.
-Additionally, the interaction of mutation and genetic architecture in different models (e.g.
-few mutations of large effects or many mutations of small effects) undoubtedly plays a role in dispersal evolution during range shifts.
-Increasing mutation rate or effect size might have the equivalent effect of a slower speed of climate change in allowing dispersal evolution to play a greater role in range shift dynamics.
-Further, life history has been shown to impact the maintenance of genetic diversity, and hence evolutionary potential, with stage and age-structured populations shown to harbor greater diversity then populations with non-overlapping generations as modeled here~\citep{ellner1996environmental}.
+In addition to sexual versus asexual reproduction, other factors undoubtedly play important roles in dispersal evolution during range shifts.
+Different models of range shifting populations have made different assumptions regarding life histories, mutation processes, and the genetic architecture of the dispersal trait, all of which have the potential to significantly alter patterns of dispersal evolution.
+For example, increasing the rate or effect size of mutation would likely increase the ability of dispersal evolution to improve a population's ability to track climate change.
+Further, life history has been shown to impact the maintenance of genetic diversity, and hence evolutionary potential, with stage and age-structured populations shown to harbor greater diversity than populations with non-overlapping generations as modeled here~\citep{ellner1996environmental}.
 Populations defined by more complex life histories might, therefore, contain more genetic diversity in dispersal at equilibrium, making evolution of increased dispersal during range shifts more likely.
@@
 
+Previous research has identified multiple factors underlying increased extinction risk during climate change, including niche breadth and range extent~\citep{thuiller2005niche, schwartz2006predicting}.
+While we did not vary niche breadth in our model, both edge starkness and the severity of the environmental gradient resulted in decreased range extents and greater extinction probabilities, agreeing with previous findings~\citep{schwartz2006predicting}.
+Our model framework provides new mechanistic insight, however, indicating that adaptation to an environmental gradient and range edge starkness could be key indicators of increased risk during climate-driven range shifts.
+A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
+While it is difficult to exactly map adaptation to the environmental gradient to empirical measures of local adaptation, the steepest gradient used in our study resulted in the intrinsic formation of stable range boundaries rather than population collapse, which can occur when the environmental gradient is too steep~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}.
+This suggests the parameters used in the scenarios examined here provide reasonable approximations of empirical patterns.
+Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, a pattern matching that imposed by the starkest range edges examined here (Fig.
+A3).
+While some of these patterns could represent a publication bias, for example against negative results in studies of local adaptation, they suggest many species may have spatial population structures similar to those imposed by steep environmental gradients or stark range edges in our model.
+Our results suggest such spatial population structures can dramatically increase extinction risk in the face of climate change, meaning many species may be at greater risk than previously understood.
+
 \section*{Conclusion}
 As climate change continues to threaten populations, communities, and ecosystems~\citep{chen2011rapid, hobbs2009novel, gonzalez2010global}, it is increasingly important to understand population responses to changing environmental conditions.
-In particular, a deeper, process-based understanding of extinction risk in range shifting populations will, in turn, allow more focused conservation interventions.
-Our results suggest that spatial population structure, as determined by the niche optimum gradient and range edge starkness, has the potential to dramatically alter extinction risk for species responding to climate change.
-Further, in contrast to other studies assuming more simplified genetic architecture, we find very little role for the evolution of heightened dispersal in allowing a population to successfully track climate change.
-Future work should continue to examine the circumstances determining the potential for rescue via dispersal evolution in range shifts.
-As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to identify those factors leading to increased extinction risk in range shifting populations, and use that knowledge to develop meaningful conservation strategies to mitigate such risk.
+In particular, a deeper, process-based understanding of extinction risk in range shifting populations will, in turn, allow conservation practitioners to identify those populations most vulnerable to climate-driven range shifts.
+Our results identify certain range characteristics likely to increase extinction risk during range shifts, such as adaptation to a steep environmental gradient or stark range edges, while providing mechanistic insight into the eco-evolutionary dynamics responsible.
+Specifically, and in contrast to previous models assuming asexual reproduction, we find an antagonistic relationship between dispersal evolution and adaptation to an environmental gradient.
+Thus, dispersal evolution in response to climate change can impose significant costs in certain environmental contexts.
+Future work should continue to examine the circumstances determining the potential for population rescue via dispersal evolution in range shifts.
+As climate change continues to accelerate~\citep{chen2017increasing}, it is imperative to identify those factors leading to increased extinction risks in range shifting populations, and use that knowledge to develop meaningful conservation strategies to mitigate such risk.
 
@@
 This model tested an evolving population's ability to track a changing climate under a variety of conditions.
-Specifically, populations were simulated under different combinations of (1) the slope of the niche optimum gradient and (2) the starkness of the range edge.
-In all simulations, an individual's expected dispersal distance and environmental niche were defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
+Specifically, populations were simulated under different combinations of (1) the slope of an environmental gradient and (2) the starkness of the range edge.
+In all simulations, an individual's dispersal and niche trait were defined by an explicit set of quantitative diploid loci subject to mutation, thus allowing both traits to evolve over time.
 All simulations began with stable climate conditions for $2000$ generations to allow the populations to reach a spatial equilibrium before the onset of climate change.
 Climate change was then modeled as a constant, directional shift in environmental conditions (see \textit{Submodels} below).
-Finally, simulations ended with another short period of climate stability to assess a population's ability to persist and recover after shifting its range.
 
 \subsection*{State variables and scales} 
-The model simulated a population of males and females characterized by diploid loci for both their expected dispersal distance and environmental niche.
+The model simulated a population of females and males characterized by diploid loci for both their expected dispersal distance and niche trait.
 Space was modeled as a lattice of discrete patches overlaying a continuous Cartesian coordinate system.
@@
 
-The model implemented climate change by shifting the location of patch carrying capacities and the niche optimum gradient along the $x$ dimension of the landscape.
+The model implemented climate change by shifting the location of patch carrying capacities and the environmental gradient along the $x$ dimension of the landscape.
 Patch carrying capacities were defined by the location of the range center along the $x$ dimension, the starkness of the decline characterizing the range edges, and the width of the range along the $x$ dimension (See Figure A1).
-The niche optimum gradient was linear and shifted at the same speed and in the same direction as carrying capacities during climate change.
+The environmental gradient was linear and shifted at the same speed and in the same direction as carrying capacities during climate change.
 
@@
 After dispersal, reproduction occurred via a stochastic implementation of the classic Ricker model~\citep{ricker1954stock} taking into account the mean fitness of individuals within the patch.
-Reproduction occurred via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their environmental niche and local conditions) were likely to produce multiple offspring while individuals with low relative fitness might not produce any.
+Reproduction occurred via random sampling of the local population (with replacement) weighted by individual relative fitness such that individuals with high relative fitness (as determined by the match between their niche trait and the environmental gradient) were likely to produce multiple offspring while individuals with low relative fitness might not produce any.
 Individuals inherited one allele from each parent at each loci, assuming independent segregation and a mutation process.
@@
 \subsection*{Emergence} 
-Emergent phenomena in this model included the spatial distribution of population abundance, dispersal abilities, and relative fitness throughout the range (i.e.
+Emergent phenomena in this model included the spatial distributions of population abundance, trait phenotypes, and the genetic variance in each trait (i.e.
 the spatial population structure).
@@
 Additionally, the evolutionary trajectories of the two different traits had the potential to interact via the relationship between gene flow (dispersal trait) and local adaptation (niche trait).
-Further, the niche optimum gradient and range edge starkness could interact with trait evolution both during stable climate conditions and during climate change.
+Further, the environmental gradient and range edge starkness could interact with trait evolution both during stable climate conditions and during climate change.
 
@@
 \subsection*{Initialization} 
-The following parameters were set at the beginning of each simulation and formed the initial conditions of the model: the mean and variance for allele values of each trait, initial population size, location of the range center, number of generations for the pre-, post-, and rapid climate change periods of the simulation, and all other necessary parameters for the submodels defined below.
+The following parameters were set at the beginning of each simulation and formed the initial conditions of the model: the mean and variance for allele values of each trait, initial population size, location of the range center, the number of generations of stationary climate conditions (to reach spatial equilibrium), the number of generations of climate change, and all other necessary parameters for the submodels defined below.
 Simulated populations were initialized in the center of the range and allowed to spread and equilibrate throughout the range during the period of stable climate conditions.
@@
 Initial population size was chosen to minimize the risk of stochastic extinction in the early stages of the simulation.
-The time frames defining climate change were designed to give a reasonable period for the population to reach a spatial equilibrium and a long enough period of climate change for extinction dynamics to play out.
+The time period required to reach spatial equilibrium was determined by initial simulations examining the spatial distributions of phenotypes and genetic variances of both traits through time.
+We defined populations as at spatial equilibrium once all of these distributions became stable through time.
+For all parameter combinations, this point was reached by generation $1500$, and thus we chose $2000$ generations to provide enough time for all simulated populations to reach spatial equilibrium prior to imposing climate change.
+The length of climate change was chosen to allow sufficient time for extinction dynamics to play out in simulated populations.
 The number of patches defining the $y$ dimension and the relationship between Cartesian space and discrete patches were chosen to allow a reasonable number of patches to contribute to the eco-evolutionary dynamics of range shifts while not proving computationally restrictive.
@@
     $t_{\Delta}$ & Duration of climate change & $100$ generations \\
-    $t_{max}$ & Total number of generations in the simulation & $2150$ generations \\
+    $t_{max}$ & Total number of generations in the simulation & $2100$ generations \\
     $\eta$ & Width of square habitat patches in Cartesian space & $50$ \\
@@
 \end{equation}
-where $\beta_{t}$ defined the center of the range at time $t$, $\tau$ affected the width of the range, and $\gamma$ affected the slope of the function at the range edges (See Figure A1).
+where $\beta_{t}$ defined the center of the range at time $t$, $\tau$ affected the width of the range, and $\gamma$ affected the function's slope at the range edges (See Figure A1).
 Population dynamics occurred within discrete patches, so to calculate a $K_{x}$ value for a discrete patch from the continuous function $f(x,t)$, we used another parameter defining the spatial scale of each patch ($\eta$).
@@
 \end{equation}
-Thus, ranges defined by different $\gamma$ values could also result in different range-wide carrying capacities, potentially altering both the ecological (e.g.
+As a result, ranges defined by different $\gamma$ values could also yield different range-wide carrying capacities, potentially altering both the ecological (e.g.
 through stochastic extinction events) and evolutionary (e.g.
 through more mutations arising in larger populations) dynamics of the ranges.
-Additionally, different combinations of $\gamma$ and $\tau$ could result in different range widths, which have been shown to impact dispersal evolution within the ranges~\citep{van1997integrodifference}.
+Additionally, different combinations of $\gamma$ and $\tau$ could result in different range widths, which have been shown to impact the dispersal distance necessary for population persistence~\citep{van1997integrodifference}.
 To control for these confounding factors, we fixed the range widths for all scenarios and altered $K_{max}$ to maintain a constant range-wide carrying capacity.
@@
     \hline
-    Range edge starkness & Niche optimum gradient & $\gamma$ & $\tau$ & $\lambda$ & $K_{max}$ \\ \hline \hline
+    Range edge starkness & Environmental gradient & $\gamma$ & $\tau$ & $\lambda$ & $K_{max}$ \\ \hline \hline
      & Flat & $0.0025$ & $-240$ & $0$ & $240$ \\
-    Shallow & Shallow & $0.0025$ & $-240$ & $0.004$ & $240$ \\
+    Gradual & Shallow & $0.0025$ & $-240$ & $0.004$ & $240$ \\
      & Steep & $0.0025$ & $-240$ & $0.008$ & $240$ \\ \hline
@@
 \caption[LoF entry]{Descriptions and parameter values for the $9$ different experimental scenarios.
-As defined in the text, $\gamma$ affects range edge starkness, $\tau$ affects the range width, $\lambda$ is the slope of the niche optimum gradient, and $K_{max}$ is the maximum carrying capacity for patches in the landscape.}
+As defined in the text, $\gamma$ affects range edge starkness, $\tau$ affects the range width, $\lambda$ is the slope of the environmental gradient, and $K_{max}$ is the maximum carrying capacity for patches in the landscape.}
 \label{table:Scenarios}
@@
 Thus, $\gamma$ and $\tau$ were both fixed within a given simulation and $\beta_{t}$ (the location of the range center) was used to simulate climate change.
-During the periods before and after climate change $\beta_{t}$ was constant, but to simulate climate change it varied with time as follows
+During the period before climate change $\beta_{t}$ was constant, but to simulate climate change it varied with time as follows
 \begin{equation}
@@
 
-\paragraph{Niche optimum}
-The niche optimum ($z_{opt,x}$) varied in space according to
+\paragraph{Environmental gradient}
+The local optimum for the niche trait ($z_{opt,x}$) varied in space according to
 \begin{equation}
@@
 where $\omega$ defined the strength of stabilizing selection and $z_{i}$ was an individual's niche phenotype~\citep{lande1976natural}.
-Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the environmental optimum of the patch it occupied ($z_{opt,x}$).
-All loci were assumed to contribute additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values.
+Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the environmental gradient value of the patch it occupied ($z_{opt,x}$).
+All loci contributed additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values.
 As defined above, $z_{opt,x}$ also shifted with climate change (i.e.
@@
 polygynandrous mating assuming a well-mixed population within each patch).
-The sampling was weighted by individual fitness and occurred with replacement so highly fit individuals were likely to have multiple offspring while low fitness individuals might not have had any.
+The sampling was weighted by individual fitness and occurred with replacement so highly fit individuals were likely to have multiple offspring while low fitness individuals might not have any.
 Each offspring inherited one allele per locus from each parent, assuming no linkage among loci.
@@
     $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
-    $\rho$ & Determines the slope of the transition in dispersal phenotypes from $0$ to $D$ & $0.5$ \\
+    $\rho$ & Determined the slope of the transition in dispersal phenotypes from $0$ to $\hat{d}$ & $0.5$ \\
     \hline
@@
     Parameter & Description & Value \\ \hline \hline
-    $\omega$ & Defines the strength of stabilizing selection on fitness traits & $3$ \\
+    $\omega$ & Defined the strength of stabilizing selection on the niche trait & $3$ \\
     $U^{T}$ & Diploid mutation rate for each trait & $0.02$ \\
@@
 
-Since the dispersal phenotype was the expected value of the exponential dispersal kernel, it could be used directly to calculate the two-dimensional diffusion coefficient of population spread ($D$).
-Specifically, since $d_{i}^{2}$ represented the mean squared displacement of an individual with dispersal phenotype $d_{i}$, the two-dimensional diffusion coefficient could be calculated as
-\begin{equation}
-D = \frac{1}{4}d_{i}^{2}
-\end{equation}
-and subsequently used to calculate the approximate speed of an expansion wave defined by that dispersal phenotype.
-
 \newpage{}
@@
 As in the main text, we calculated the cumulative probability of extinction for both slow and fast speeds of climate change.
-The figures in this section use the same layout and line types as Figure 2 in the main text to allow for direct comparisons.
+The figures in this section use the same layout and color scheme as Figure 2 in the main text to allow for direct comparisons.
 
@@
 
-\section*{Equilibrium fitness throughout the landscape}
-To assess trends in realized fitness values throughout the landscape, we calculated patch-level mean individual fitness for each landscape at equilibrium.
-To simplify the figures, we averaged over the $y$ dimension in which the CV in niche genotypes was minimal throughout the range for all scenarios (typically below $0.25$) due to the constant environmental conditions.
-Populations at the range edge were characterized by reduced fitness compared to core populations, as found in previous models~\citep{garcia1997genetic}.
-However, this trend was exacerbated in populations that successfully tracked climate change compared to those that went extinct, especially in the presence of gradual range edges and steep niche optimum gradients.
-As realized fitness values do not vary spatially in simulations with no gradient in the niche optimum, the following figures only show results for scenarios with a shallow or steep gradient.
+\section*{Dispersal evolution}
+Patterns in the spatial distributions of dispersal phenotypes and genetic variance largely followed similar patterns at slow and fast speeds of climate change as shown in the main text for a moderate speed of climate change.
+The main difference was the number of populations able to track climate change when it progressed at different speeds.
+As in the previous section, all figures use the same layout and color scheme as Figures 3 \& 4 in the main text.
 
-[Figures B3-B5 go here.]
+[Figures B3-B6 go here.]
 
-\section*{Dispersal evolution}
-Using the same metric of dispersal evolution from the main text (change in average phenotype for each patch), we display here the observed dispersal evolution over the course of climate change for all experimental scenarios.
-Each histogram in the following figures represents a single experimental scenario as indicated by the figure text.
-The lower left panel and upper right panel from Figure B7 are the same histograms shown in Figure 3a\&b, but are here placed in the context of all other experimental scenarios.
+\section*{Adaptation to the environmental gradient}
+As with dispersal evolution, the spatial patterns in adaptation to the environmental gradient and genetic variance of the niche trait were broadly similar at varying speeds of climate change.
+One important difference, likely related to the reduced extinction risk experienced at low speeds of climate change, is that the mismatch between phenotypes and the environmental gradient imposed by range shifting was reduced at a slow speed of climate change.
+Thus, while fitness was still reduced in these scenarios, it was reduced less than scenarios experiencing faster speeds of climate change.
+Again, all figures use the same layout and color scheme as Figures 5 \& 6 to allow for direct comparisons.
 
-[Figures B6-B8 go here.]
-
-\section*{Equilibrium dispersal phenotypes}
-Here, we present histograms of the equilibrium distribution of dispersal phenotypes to demonstrate the importance of those phenotypes in determining population success or extinction during climate change.
-Dispersal phenotypes are log transformed for easier comparison.
-As with the dispersal evolution section, each histogram represents a single experimental scenario as indicated by the figure text.
-Similarly, the lower left panel and upper right panel from Figure B10 are the same histograms shown in Figure 3c\&d, but are here placed in the context of all other experimental scenarios.
-All histograms additionally have a vertical dashed line indicating the dispersal phenotype necessary to produce an expansion wave traveling at exactly the speed of climate change in each simulation.
-This value serves as a threshold to distinguish individuals from ultimately successful versus extinct populations.
-
-[Figures B9-B11 go here.]
+[Figures B7-B10 go here.]
 
@@
 \includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/SchematicFigures/SimExample"}
-\caption{A single example of a simulation with a steep niche optimum gradient and a moderately stark range edge.
+\caption{A single example of a simulation with a steep environmental gradient and a moderately stark range edge.
 Information on the (a) abundance, (b) dispersal ability, and (c) fitness of individuals in each patch is shown for time periods beginning with the last generation of stable climate conditions ($t = 0$) to $40$ generations after the start of climate change.
 Log transformed mean dispersal phenotypes (b) are shown for each patch.
-Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the niche optimum for each patch.}
+Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the gradient value at each patch.}
 \label{fig:SimExample}
@@
 \caption{The cumulative probability of extinction due to a moderate speed of climate change in different experimental scenarios.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by shallow (solid line), moderate (dashed line), or stark (dotted line) edges.
-In all graphs, a horizontal grey line shows $\%100$ extinction.}
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep environmental gradient, and in ranges characterized by gradual, moderate, or stark edges, indicated by line color as shown in the legend.
+In all graphs, a horizontal grey line shows $100\%$ extinction.}
 \label{fig:ExtProb}
@@
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/DispComposite"}
-%\caption{Patterns in the evolution and the equilibrium distribution of the dispersal trait, highlighting extant simulations from a moderate speed of climate change.
-Evolution in dispersal (a and b) is shown as the change in the mean dispersal phenotype of each patch from the beginning of the period of climate change to the end.
-Positive values indicate an increase in average dispersal ability in the patch.
-Equilibrium distributions of the dispersal trait (c and d) are shown as log transformed dispersal phenotypes of individuals in populations after $2000$ generations of stable climate conditions.
-In all panels, values associated with extant populations are shown in dark blue.
-Results are shown for populations with a flat niche optimum gradient and a gradual range edge (a and c; $n = 155$ extant populations) and for populations with a steep niche optimum gradient and a stark range edge (b and d; $n = 14$ extant populations).
-Full results for all parameter combinations are provided in Appendix B.}
-%\label{fig:Disp}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainMeanDisp"}
+\caption{Spatial distributions of dispersal phenotypes at three different time points during climate change.
+Solid lines indicate the average dispersal phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of dispersal phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
+\label{fig:MeanDisp}
+\end{figure}
 
-%\clearpage
+\clearpage
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainDispGenVar"}
+\caption{Spatial distributions of genetic variance in dispersal at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in dispersal in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
+\label{fig:DispGenVar}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainMeanNiche"}
+\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change.
+Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
+\label{fig:MeanNiche}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainFitGenVar"}
+\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a moderate speed of climate change.}
+\label{fig:FitGenVar}
+\end{figure}
+
+\clearpage
 
@@
 Points represent the carrying capacity of a discrete $\eta$ x $\eta$ patch in the range.
-The vertical dashed lines indicate the $x$ coordinates at which $f(x,t)$ declines below $0.1$ and the $\gamma$ value for each plot is listed above.}
+The vertical dashed lines indicate the $x$ coordinates at which $f(x,t)$ declines below $0.1$ and the $\gamma$ value for each plot is listed above.
+Parameter values correspond to those listed in Table A2.}
 \label{Fig:LifeCycle}
@@
 \caption{The cumulative probability of extinction due to a slow speed of climate change in different experimental scenarios.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) edge.
-In all graphs, a horizontal grey line shows $\%100$ extinction.}
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep environmental gradient, and in ranges characterized by gradual, moderate, or stark edges, indicated by line color as shown in the legend.
+In all graphs, a horizontal grey line shows $100\%$ extinction.}
 \label{Fig:ExtProbSlow}
@@
 \caption{The cumulative probability of extinction due to a fast speed of climate change in different experimental scenarios.
-Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep niche optimum gradient, and in ranges characterized by a shallow (solid line), moderate (dashed line), or stark (dotted line) edge.
-In all graphs, a horizontal grey line shows $\%100$ extinction.}
+Graphs show the proportion of simulated populations that went extinct through time for scenarios with a (a) flat, (b) shallow, and (c) steep environmental gradient, and in ranges characterized by gradual, moderate, or stark edges, indicated by line color as shown in the legend.
+In all graphs, a horizontal grey line shows $100\%$ extinction.}
 \label{Fig:ExtProbFast}
@@
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowInitFitSpace"}
-%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a slow speed of climate change.}
-%\label{Fig:InitFitSlow}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowMeanDisp"}
+\caption{Spatial distributions of dispersal phenotypes at three different time points during climate change.
+Solid lines indicate the average dispersal phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of dispersal phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
+\label{Fig:MeanDispSlow}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainInitFitSpace"}
-%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a moderate speed of climate change.}
-%\label{Fig:InitFit}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastMeanDisp"}
+\caption{Spatial distributions of dispersal phenotypes at three different time points during climate change.
+Solid lines indicate the average dispersal phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of dispersal phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
+\label{Fig:MeanDispFast}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastInitFitSpace"}
-%\caption{Individual fitness along the $x$ dimension of the landscape at equilibrium.
-Points represent the mean across simulations and error bars are interquartile ranges.
-Population status (extinct or successful) was determined for a fast speed of climate change.}
-%\label{Fig:InitFitFast}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowDispGenVar"}
+\caption{Spatial distributions of genetic variance in dispersal at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in dispersal in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
+\label{Fig:DispGenVarSlow}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowDispEvol"}
-%\caption{Observed dispersal evolution in populations responding to a slow speed of climate change.
-Positive values indicate an increase in average dispersal ability during climate change.
-The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-%\label{Fig:DispEvolSlow}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastDispGenVar"}
+\caption{Spatial distributions of genetic variance in dispersal at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in dispersal in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
+\label{Fig:DispGenVarFast}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainDispEvol"}
-%\caption{Observed dispersal evolution in populations responding to a moderate speed of climate change.
-Positive values indicate an increase in average dispersal ability during climate change.
-The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-%\label{Fig:DispEvolMain}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowMeanNiche"}
+\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change.
+Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
+\label{Fig:MeanNicheSlow}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastDispEvol"}
-%\caption{Observed dispersal evolution in populations responding to a fast speed of climate change.
-Positive values indicate an increase in average dispersal ability during climate change.
-The values associated with populations successfully tracking climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-The experimental scenario corresponding to each histogram is indicated on the figure.}
-%\label{Fig:DispEvolFast}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastMeanNiche"}
+\caption{Spatial distributions of niche trait phenotypes at three different time points during climate change.
+Solid lines indicate the average phenotype at each $x$ coordinate and dotted lines are the interquartile ranges of patch means.
+Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
+\label{Fig:MeanNicheFast}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowInitDispVals"}
-%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a slow speed of climate change.}
-%\label{Fig:InitDispSlow}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/SlowFitGenVar"}
+\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a slow speed of climate change.}
+\label{Fig:FitGenVarSlow}
+\end{figure}
 
-%\clearpage
+\clearpage
 
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/MainInitDispVals"}
-%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a moderate speed of climate change.}
-%\label{Fig:InitDispMain}
-%\end{figure}
-
-%\clearpage
-
-%\begin{figure}[h!]
-%\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastInitDispVals"}
-%\caption{Distributions of the dispersal phenotypes observed in equilibrium populations.
-Phenotypes associated with populations that ultimately survived climate change are shown in dark blue and the total number of surviving populations is indicated in the top left corner.
-Vertical dashed lines indicate the dispersal phenotype necessary to produce an expansion wave exactly matching a fast speed of climate change.}
-%\label{Fig:InitDispFast}
-%\end{figure}
+\begin{figure}[h!]
+\includegraphics[width=1\textwidth]{"/Users/Topher/Desktop/PostdocResearch/ShiftingSlopesOther/ResultFigures/FastFitGenVar"}
+\caption{Spatial distributions of genetic variance in the niche trait at three different time points during climate change.
+Solid lines indicate the average genetic variance of patches at each $x$ coordinate and dotted lines are the interquartile ranges.
+Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
+Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Colors indicate different edge types as shown in the figure legend.
+Results are shown for the equilibrium distribution ($t = 0$), halfway through climate change ($t = 50$), and the end of climate change ($t = 100$) for a fast speed of climate change.}
+\label{Fig:FitGenVarFast}
+\end{figure}
 

### 2019-05-01 Equilibrium/EquilibriumAnimations.R

@@ -1,0 +1,223 @@
+# This script will create an animated graph of four metrics: 1) mean dispersal 
+#    phenotype, 2) mean nich phenotype, 3) genetic variance in dispersal, and 
+#    4) genetic variance in the  niche trait. The animation will go
+#    through time during the range shift in increments of 10 generations.
+
+# Set the working directory and create some graphical objects to be used across
+#    all the animations
+setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
+FigWidth <- 1200
+FigHeight <- 900
+FigPointSize <- 18
+GenSteps <- 10
+LengthShift <- 2000
+GenSeq <- seq(1510, LengthShift, by = GenSteps)
+OuterMar <- c(2, 2, 2, 2)
+InnerMar <- c(0,0,0,0)
+LineWidth <- 2.5
+
+# -------------------------------------------------- 1) mean dispersal phenotype
+load("SimData/EquilibriumDispData.rdata")
+# Figure out appropriate values for xlim and ylim
+xLwr <- Inf
+xUpr <- -Inf
+yLwr <- Inf
+yUpr <- -Inf
+for(p in 1:9){
+          CurData <- DispData[[p]]
+          xLwr <- min(xLwr, min(CurData$x))
+          xUpr <- max(xUpr, max(CurData$x))
+          yLwr <- min(yLwr, min(CurData$lwr))
+          yUpr <- max(yUpr, max(CurData$upr))
+}
+
+# Create a temporary directory to hold the individual images comprising the
+#    animation
+TempDirPath <- "ResultFigures/temp"
+dir.create(TempDirPath)
+# Now loop through each time point to create a still image for that generation
+for(gen in GenSeq){
+     if(gen == 0){
+          GenNum <- "000"
+     } else if(gen < 100){
+          GenNum <- paste("0", gen, sep = "")
+     } else{
+          GenNum <- gen
+     }
+     FigName <- paste(TempDirPath, "/gen", GenNum, ".png", sep = "")
+     png(filename = FigName, width = FigWidth, height = FigHeight, pointsize = FigPointSize)
+          par(mfrow = c(3,3), oma = OuterMar, mar = InnerMar)
+          for(i in 1:9){
+               CurData <- subset(DispData[[i]], g == gen)
+               CurData <- CurData[order(CurData$x),]
[…]

### 2019-05-01 Equilibrium/EquilibriumDisp.R

@@ -1,0 +1,82 @@
+# This script will extract the dispersal phenotypes of individuals throughout their
+#    ranges every 10 generations
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:9
+
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 2000
+GenSeq <- seq(10, LengthShift, by = GenSteps)
+
+DispExtract <- function(i){
+     # Set the current parameter combination and isolate the relevant
+     #    simulation IDs.
+     Prefix <- paste("~/ShiftingSlopes/EquilibriumTest/Params", i, sep = "")
+     CurSims <- list.files(Prefix)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste(Prefix, CurSims[1], "parameters.R", sep = "/")
+     source(ParamFile)
+     
+     # Create the data frame for all simulations
+     MeanDisps <- data.frame(x = NA, g = NA, dBar = NA, lwr = NA, upr = NA)
+     
+     # Now loop through each row of the data frame to fill it in
+     for(g in GenSeq){
+          # Create a temporary data frame to hold the results from all simulations
+          #    for each time point, which will then be condensed across simulations
+          #    for the main data frames
+          Temp <- data.frame(x = NA, dBar = NA)
+          for(s in 1:length(CurSims)){
+               InFile <- paste(Prefix, CurSims[s], "SummaryStats.csv", sep = "/")
+               SumStats <- read.csv(InFile)
+               CurSumStats <- subset(SumStats, gen == g)
+               # Isolate the x coordinates from the current data and loop through
+               #    them, adding their data to the temporary data frames
+               xSeq <- unique(CurSumStats$x)
+               for(j in xSeq){
+                    LocalData <- subset(CurSumStats, x == j)
+                    # Add the simulation data to the Temp data frame
+                    Temp <- rbind(c(j, mean(LocalData$muDisp)), Temp)
+               }
+          }
+          # remove the final row of NA's from Temp
[…]

### 2019-05-01 Equilibrium/EquilibriumDisp.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=EquilibriumDisp.R
+LogFile=EquilibriumDisp.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 Equilibrium/EquilibriumDispGenVar.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=EquilibriumDispGenVar.R
+LogFile=EquilibriumDispGenVar.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 Equilibrium/EquilibriumNiche.R

@@ -1,0 +1,86 @@
+# This script will extract the realized fitness of individuals throughout their
+#    ranges every 10 generations
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:9
+
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 2000
+GenSeq <- seq(10, LengthShift, by = GenSteps)
+
+FitExtract <- function(i){
+     # Set the current parameter combination and isolate the relevant
+     #    simulation IDs.
+     Prefix <- paste("~/ShiftingSlopes/EquilibriumTest/Params", i, sep = "")
+     CurSims <- list.files(Prefix)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste(Prefix, CurSims[1], "parameters.R", sep = "/")
+     source(ParamFile)
+     
+     # Create the data frame for all simulations
+     MeanFits <- data.frame(x = NA, g = NA, Mismatch = NA, lwr = NA, upr = NA, Zopt = NA)
+     
+     # Now loop through each row of the data frame to fill it in
+     for(g in GenSeq){
+          # Create a temporary data frame to hold the results from all simulations
+          #    for each time point, which will then be condensed across simulations
+          #    for the main data frames
+          Temp <- data.frame(x = NA, Mismatch = NA, Zopt = NA)
+          for(s in 1:length(CurSims)){
+               InFile <- paste(Prefix, CurSims[s], "SummaryStats.csv", sep = "/")
+               SumStats <- read.csv(InFile)
+               CurSumStats <- subset(SumStats, gen == g)
+               # Isolate the x coordinates from the current data and loop through
+               #    them, adding their data to the temporary data frames
+               xSeq <- unique(CurSumStats$x)
+               for(j in xSeq){
+                    LocalData <- subset(CurSumStats, x == j)
+                    # Calculate the mismatch between optimum and genotype and
+                    #    its direction (i.e. positive or negative)
+                    Zopt <- lambda * (j * eta - LocalData$beta[1])
[…]

### 2019-05-01 Equilibrium/EquilibriumTest.R

@@ -1,0 +1,114 @@
+# This script will run simulations on the MSI compute cluster to characterize
+#    population structure in a stable range for all of the 9 range parameter
+#    combinations to be explored.
+
+# Set the number of processors and number of simulations to be run
+nProc <- 24*18
+NumSims <- 200
+
+# Set the working directory and load necessary data and libraries
+setwd("~/ShiftingSlopes/EquilibriumTest/")
+source("~/ShiftingSlopes/SimFunctions.R")
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+library(parallel)
+library(Rmpi)
+
+# Now set all parameters
+BetaInit <- 0
+omega <- 3
+U <- c(0.02, 0.02)
+Vm <- c(0.0004, 0.0004)
+Lf <- 5
+Ld <- 5
+Rmax <- 2
+width <- 10
+kern <- "exp"
+EnvGradType <- "K"
+monoecious <- FALSE
+BurnIn <- 1500
+BurnOut <- 0
+LengthShift <- 500
+v <- 0
+InitPopSize <- 2500
+FitInit <- 0
+FitDiv <- 0.025
+DispInit <- -1
+DispDiv <- 1
+NumRands <- 1000000
+z <- 0.5
+dmax <- 1000
+rho <- 0.5
+
+AllParams <- vector(mode = "list", length = 9)
+
+for(i in 1:9){
+     gamma <- RangeParams$gamma[i]
+     lambda <- RangeParams$lambda[i]
+     tau <- RangeParams$tau[i]
+     eta <- RangeParams$eta[i]
+     Kmax <- RangeParams$Kmax[i]
+     
+     AllParams[[i]] <- list(BetaInit, gamma, tau, lambda, omega, U, Vm, Lf, Ld, Rmax,
+                            Kmax, width, kern, EnvGradType, monoecious, BurnIn, BurnOut, 
+                            LengthShift, v, InitPopSize, FitInit, FitDiv, DispInit,
+                            DispDiv, eta, NumRands, z, dmax, rho)
+     names(AllParams[[i]]) <- c("BetaInit", "gamma", "tau", "lambda", "omega", "U", "Vm", 
+                                "Lf", "Ld", "Rmax", "Kmax", "width", "kern", "EnvGradType", 
+                                "monoecious", "BurnIn", "BurnOut", "LengthShift", "v", 
+                                "InitPopSize", "FitInit", "FitDiv", "DispInit", "DispDiv", 
[…]

### 2019-05-01 Equilibrium/EquilibriumTest.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=08:00:00,nodes=18:ppn=24,mem=500gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the parameter combination to use and generate names of R scripts and log files
+Rscript=EquilibriumTest.R
+LogFile=EquilibriumTest.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 xVar/xDisp.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=xDisp.R
+LogFile=xDisp.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 xVar/xDispGenVar.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=xDispGenVar.R
+LogFile=xDispGenVar.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 xVar/xFitGenVar.R

@@ -1,0 +1,107 @@
+# This script will extract the genetic variance in fitness throughout the range
+#    every 10 generations
+
+# Set the number of processors available for this script 
+nProc <- 2*24
+
+# Set the working directory
+setwd("~/ShiftingSlopes/")
+source("~/ShiftingSlopes/SimFunctions.R")
+library(parallel)
+library(Rmpi)
+RangeParams <- read.csv("~/ShiftingSlopes/RangeParameters.csv")
+load("ExtSimIDs.rdata")
+
+# Create the vectors to be passed to the function in the data extraction
+SimVec <- 1:27
+SpeedWords <- c("Slow", "MainSim", "Fast")
+SpeedVec <- rep(SpeedWords, 9)
+ParamVec <- rep(1:9, each = 3)
+
+# Create other useful objects to pass to the nodes
+GenSteps <- 10
+LengthShift <- 100
+GenSeq <- seq(0, LengthShift, by = GenSteps)
+
+GenVarExtract <- function(i){
+     # Set the current speed and parameter combination and isolate the relevant
+     #    simulation IDs.
+     SpeedWord <- SpeedVec[i]
+     ParamCombo <- ParamVec[i]
+     CurSims <- subset(SimIDs, Params == ParamCombo)
+     
+     # Source a representative parameter file for useful values
+     ParamFile <- paste("~/ShiftingSlopes/", SpeedWord, "/Params", ParamCombo, 
+                        "/", CurSims$SimID[1], "/parameters.R", sep = "")
+     source(ParamFile)
+     
+     # Create a data frame for all simulations
+     MeanGenVar <- data.frame(x = NA, g = NA, GenVar = NA, lwr = NA, upr = NA)
+     
+     # Now loop through each row of the data frame to fill it in
+     for(g in GenSeq){
+          # Create a temporary data frame to hold the results from all simulations
+          #    for each time point, which will then be condensed across simulations
+          #    for the main data frames
+          Temp <- data.frame(x = NA, GenVar = NA)
+          for(s in 1:nrow(CurSims)){
+               if(g == 0){
+                    InFile <- paste("~/ShiftingSlopes/MainSim/Params", ParamCombo,
+                                    "/", CurSims$SimID[s], "/SummaryStats.csv", sep = "")
+                    SumStats <- read.csv(InFile)
+                    CurSumStats <- subset(SumStats, gen == 2000)
+               } else{
[…]

### 2019-05-01 xVar/xNiche.sh

@@ -1,0 +1,21 @@
+#!/bin/bash -l
+#PBS -l walltime=01:30:00,nodes=2:ppn=24,mem=124gb
+#PBS -m abe
+#PBS -M cweissle@umn.edu
+#PBS -j oe
+
+# Set the names of R scripts and log files
+Rscript=xNiche.R
+LogFile=xNiche.log
+
+# Change to the relevant working directory
+cd ~/ShiftingSlopes/
+
+# Load R and MPI
+module load R/3.4.4
+module load ompi/3.0.0/gnu-7.2.0-centos7
+
+export RMPI_TYPE=OPENMPI
+export OMPI_MCA_mpi_warn_on_fork=0
+
+mpirun -np 1 R CMD BATCH --no-restore --no-save --quiet $Rscript $LogFile

### 2019-05-01 xVar/xVarFigures.R

@@ -1,0 +1,339 @@
+# This script will create figures showing extinction probabilities and the change 
+#    in fitness, dispersal, and genetic variance in both during the range shift.
+
+setwd("~/Desktop/PostdocResearch/ShiftingSlopesOther/")
+library(RColorBrewer)
+
+# First set some common graphical parameters to be used among all the figures
+# Placement
+xLetterAdj <- 1.1
+yLetterAdj <- 0.95
+xLabLine <- 3
+GenTextAdj <- 0.9
+BetaPos <- matrix(c(0,50,100,0,62.5,125,0,75,150), nrow = 3, ncol = 3, byrow = TRUE)
+LegInset <- -0.2
+LegInsetExt <- -0.15
+# Sizing
+FigWidth <- 12
+FigHeight <- 9
+FigPointSize <- 18
+LineWidth <- 2.5
+# Labels and colors
+xLabel <- "Space"
+EdgeWords <- c("Gradual", "Moderate", "Stark")
+LetterSeq <- c("a", "b", "c")
+EdgeCols <- brewer.pal(n = 3, name = "Dark2")
+# Miscelaneous
+SimMat <- matrix(c(7,8,9,4,5,6,1,2,3), nrow = 3, ncol = 3, byrow = TRUE)
+PlotGens <- c(0, 50, 100)
+InnerMar <- c(0.45,0,0.45,0)
+SpeedWords <- c("Slow", "Main", "Fast")
+
+# Now set some graphical parameters specific to the three panel dispersal graphs
+yLabLine <- 5
+TextSize <- 1.75
+AxisSize <- 2.25
+LegSize <- 2.5
+LetterSize <- 2.5
+OuterMar <- c(5, 9, 4, 2)
+
+# -------------------------------------------------- 1) mean dispersal phenotype
+yLabel <- "Within simulation SD in mean dispersal"
+load("SimData/xVarDispData.rdata")
+# Figure out appropriate values for xlim and ylim
+xLwr <- rep(Inf, 3)
+xUpr <- rep(-Inf, 3)
+yLwr <- rep(Inf, 3)
+yUpr <- rep(-Inf, 3)
+for(p in 1:9){
+     for(v in 1:3){
+          CurData <- DispData[[p]][[v]]
+          xLwr[v] <- min(xLwr[v], min(CurData$x, na.rm = TRUE))
+          xUpr[v] <- max(xUpr[v], max(CurData$x, na.rm = TRUE))
[…]

### 2019-06-27 Manuscript/AmNat_ReSubmission.tex
Updated the manuscript with the revisions from the editor's comments

@@
 2, B1, \& B2).
+This is likely due to the different impacts of gradual versus moderate edges on the evolutionary dynamics of both traits as we discuss below.
 Third, while both the environmental gradient and edge type influenced extinction probabilities, the environmental gradient drove more dramatic changes to extinction risk.
@@
 \subsection*{Adaptation to the environmental gradient} 
+In scenarios defined by a flat environmental gradient, patterns in the niche trait displayed no meaningful spatial variation as the trait optimum was constant across the entire range.
+Thus, we primarily present results on evolution of the niche trait in the presence of an environmental gradient.
+
 Equilibrium values of the niche trait closely matched the environmental gradient in the range core and diverged at the edges (Fig.
 5), likely due to a combination of gene swamping from abundant central populations~\citep{kirkpatrick1997evolution} and the loss of genetic diversity due to drift in small, peripheral populations~\citep{polechova2018sky, polechova2015limits}.
-In agreement with previous results, the starkness of the range edge influenced this pattern, with increased maladaptation of peripheral populations in scenarios with gradual edges~\citep{garcia1997genetic}.
+In agreement with previous results, the starkness of the range edge influenced the degree of maladaptation (i.e.
+divergence from the phenotypic optimum) in edge populations~\citep{garcia1997genetic}.
+In scenarios defined by more gradual edges, peripheral populations exhibited increased maladaptation compared to the peripheral populations of scenarios with stark edges.
 Across all scenarios, the clines in the niche trait became much shallower during the range shift (Fig.
@@
 A survey of the scientific literature found evidence for local adaption in approximately $71\%$ of studies, suggesting a high prevalence of local adaptation in natural populations~\citep{hereford2009quantitative}.
-While it is difficult to exactly map adaptation to the environmental gradient to empirical measures of local adaptation, the steepest gradient used in our study resulted in the intrinsic formation of stable range boundaries rather than population collapse, which can occur when the environmental gradient is too steep~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}.
-This suggests the parameters used in the scenarios examined here provide reasonable approximations of empirical patterns.
+Our results suggest such adaptation to an environmental gradient corresponds to a greater extinction risk during climate driven range shifts (Fig.
+2).
 Further, a recent meta-analysis of $1400$ bird, mammal, fish, and tree species found no evidence for consistent declines in abundance towards range edges~\citep{dallas2017species}, a pattern matching that imposed by the starkest range edges examined here (Fig.
@@
 Our results suggest such spatial population structures can dramatically increase extinction risk in the face of climate change, meaning many species may be at greater risk than previously understood.
+
+While our model suggests certain spatial patterns in both the dispersal and niche traits may be responsible for the increased extinction risk faced by populations in certain scenarios, it is not clear how prevalent such patterns may be in natural populations.
+For example, while many studies have found evidence for local adaptation~\citep{hereford2009quantitative}, it is difficult to exactly map adaptation to an environmental gradient to empirical measures of local adaptation.
+However, the steepest gradient used in our study resulted in the intrinsic formation of stable range boundaries rather than population collapse, which can occur when the environmental gradient is too steep~\citep{kirkpatrick1997evolution, alleaume2006geographical, polechova2015limits, polechova2018sky}.
+This suggests the parameters used in the scenarios examined here provide reasonable approximations of empirical patterns.
+Additionally, $64\%$ of studies investigating spatial patterns in genetic diversity found evidence for a decline in genetic diversity from core to edge populations~\citep{eckert2008genetic}, as predicted by our model.
+Matching the model's predictions for the dispersal trait to empirical results, however, returns more mixed results.
+Edge populations of \textit{Abronia umbellata}, a coastal dune plant, exhibit greater dispersal capability than core populations~\citep{darling2008increased}, matching the patterns produced by our model.
+However, examination of edge populations in four cricket species revealed little difference in dispersal capabilities relative to core populations~\citep{simmons2004changes}, contradicting our model.
+Overall, though, empirical studies of spatial variation in dispersal capability within a range are relatively sparse~\citep{hargreaves2014evolution}, making empirical validation of this model prediction difficult.
+However, our results suggest such research could prove highly informative in predicting a species' fate during climate change.
 
@@
 Means and interquartile ranges were calculated across the pooled patch level means along the $y$ dimension of each simulation at each $x$ coordinate.
-Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients.
 Colors indicate different edge types as shown in the figure legend.
@@
 Means and interquartile ranges were calculated across the pooled patch level variances along the $y$ dimension of each simulation at each $x$ coordinate.
-Graphs show the spatial distributions of genetic variance in the niche trait in scenarios with (a) steep, (b) shallow, and (c) flat environmental gradients.
+Graphs show the spatial distributions of niche trait phenotypes in scenarios with (a) steep and (b) shallow environmental gradients.
 Colors indicate different edge types as shown in the figure legend.

### 2019-06-28 Manuscript/AmNat_ReSubmission.tex
Revised the manuscript for resubmission

@@
 While our results generally agree with this prediction, comparison with results from previous models suggests an important role for the biological and environmental context of dispersal evolution in range shifts.
-In a model assuming asexual reproduction, adaptation to an environmental gradient increased dispersal evolution during range shifts compared to scenarios with no gradient, contrary to the results presented here~\citep{hargreaves2015fitness}.
+In a model assuming asexual reproduction, adaptation to an environmental gradient increased dispersal evolution during range shifts compared to scenarios with no gradient~\citep{hargreaves2015fitness}, contrary to the results presented here.
 In sexually reproducing populations, however, the evolution of increased dispersal can have an antagonistic interaction with local adaptation to an environmental gradient.
@@
 where $\omega$ defined the strength of stabilizing selection and $z_{i}$ was an individual's niche phenotype~\citep{lande1976natural}.
-Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the environmental gradient value of the patch it occupied ($z_{opt,x}$).
+Thus, an individual's realized fitness was higher the closer its niche phenotype ($z_{i}$) was to the trait optimum at the patch it occupied ($z_{opt,x}$).
 All loci contributed additively to an individual's niche value with no dominance or epistasis, meaning an individual's phenotype was simply the sum of the individual's allele values.
@@
 \end{equation}
-where $F_{t,x}$ was the number of females in patch $x$ at time $t$, $R$ was the intrinsic growth rate for the population and remained constant in both time and space, $\psi$ was the expected sex ratio of the population, $N_{t,x}$ was the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ was the local carrying capacity based on the environmental conditions.
+where $F_{t,x}$ was the number of females in patch $x$ at time $t$, $R$ was the intrinsic growth rate for the population and remained constant in both time and space, $\psi$ was the expected proportion of females in the population, $N_{t,x}$ was the number of individuals (males and females) in patch $x$ at time $t$, and $K_{x}$ was the local carrying capacity based on the environmental conditions.
 To incorporate demographic stochasticity, the realized number of offspring for each patch was then drawn from a Poisson distribution.
@@
     $R$ & Intrinsic growth rate of the population & $2$ \\
-    $\psi$ & Expected sex ratio (proportion of females) in the population & $0.5$ \\
+    $\psi$ & Expected proportion of females in the population & $0.5$ \\
     $\hat{d}$ & Maximum achievable dispersal phenotype & $1000$ \\
@@
     $L^{T}$ & Number of diploid loci defining each trait & $5$ loci \\
-    $\mu_{1}^{f}$ & Initial mean allele value for the niche trait & $0$ \\
+    $\mu_{1}^{n}$ & Initial mean allele value for the niche trait & $0$ \\
     $\mu_{1}^{d}$ & Initial mean allele value for the dispersal trait & $-1$ \\
-    $\sigma_{1}^{f}$ & Initial standard deviation of allele values for the niche trait & $0.025$ \\
+    $\sigma_{1}^{n}$ & Initial standard deviation of allele values for the niche trait & $0.025$ \\
     $\sigma_{1}^{d}$ & Initial standard deviation of allele values for the dispersal trait & $1$ \\
@@
 Log transformed mean dispersal phenotypes (b) are shown for each patch.
-Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the gradient value at each patch.}
+Average patch fitness (c) was calculated based on the mean niche trait of local individuals and the trait optimum at each patch.}
 \label{fig:SimExample}
